---
title: "Diversity Metrics at Every Stage of a Hybrid Maize Pipeline"
subtitle: "Formulas, verified identities, R code, and the computational traps — heterotic groups, DH induction, line per se, hybrid mining"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

# Scope and how to read this

This document takes the metric recommendation established in the companion
report (`metric_selection.Rmd`) and applies it stage by stage to a hybrid maize
programme: **S1** conservation of the heterotic groups, **S2** choice of parents
and crosses for doubled-haploid (DH) induction, **S3** selection of lines per
se, **S4** hybrid mining. For each stage it gives the metric, the formula, the
computational cost, and the R route.

Two rules govern everything below.

**Every formula here was verified numerically before it was written.** Where a
result is an exact algebraic identity, the maximum deviation from simulation is
reported next to it. Where it is a Monte Carlo agreement, the residual is
reported. Nothing appears in this document that was not checked against a
simulated oracle in `R/00_sim_ref.R`.

**The metric set is the one established previously**: molecular (IBS) coancestry
as the kernel, gene diversity `GD = 1 - theta` as the diversity scale, status
number `Ns = 1/(2 theta)` as the interpretable effective size, allelic richness
as the one genuinely non-redundant addition, and proportional loss measured
against a **same-size resampling baseline** rather than against the full pool.

Code: `R/pipeline_metrics.R` (this document) and `R/fast_metrics.R` (the
inner-loop routes from the companion report).

---

# The kernel: one matrix, all four stages

Molecular coancestry `f_ij` is the probability that an allele drawn at random
from individual `i` is identical in state to one drawn from `j`
(Nejati-Javaremi et al. 1997; Caballero & Toro 2002). For inbred lines coded
`{0,1}` at `m` biallelic markers:

$$f_{ij} \;=\; 1 - \frac{1}{m}\sum_{k=1}^{m} (x_{ik} - x_{jk})^2$$

Group coancestry of a set is the mean of the full submatrix **including the
diagonal** (Cockerham 1967; Caballero & Toro 2000):

$$\theta = \frac{1}{n^2}\sum_{i}\sum_{j} f_{ij}, \qquad GD = 1 - \theta,
\qquad N_s = \frac{1}{2\theta}$$

The diagonal is not a technicality. It is why a small set always looks highly
coancestral, and it is the reason the naive greedy constraint in §S3 fails.

## Trap 1 — modified Rogers distance is the same statistic

For homozygous lines,

$$f_{ij} = 1 - MRD_{ij}^2$$

**exactly** — maximum deviation `1.1e-16` over 120 lines and 6000 markers, i.e.
floating-point noise. Modified Rogers distance is the measure dominant in the
maize literature; molecular coancestry is the measure recommended here. They are
one measurement in two parameterisations. Reporting both as separate evidence of
diversity double-counts. Use either, convert freely:

```{r}
f <- coancestry_ibs(GL)          # from genotypes
D <- mrd_matrix(GL)              # = sqrt(1 - f)
all.equal(f, coancestry_from_mrd(D))   # TRUE
```

---

# S1 — Conserving the heterotic groups

## What has to be measured

Diversity in a hybrid programme is **two quantities moving in opposite
directions**: within-pool diversity erodes under selection, while between-pool
divergence inflates. A single pool-wide statistic cannot express both, and a
programme that monitors only total diversity can look stable while both pools
collapse internally and drift apart.

The exact partition, for any grouping with weights `w_p`:

$$\theta_T = \sum_p w_p^2\,\theta_{pp} \;+\; \sum_{p \ne q} w_p w_q\,\theta_{pq}$$

Verified to **0.00e+00** deviation against the direct mean of the full matrix.

## Trap 2 — two F_ST definitions differ by up to 2x

Both of these are in routine use, both are computed from the same coancestry
matrix, and they are **not interchangeable**:

$$F_{ST}^{(W)} = \frac{\theta_w - \theta_B}{1 - \theta_B}
\qquad\qquad
G_{ST} = \frac{\theta_w - \theta_T}{1 - \theta_T}$$

The first references *between-pool* coancestry; the second references *total*
(pooled-frequency) coancestry and equals Nei's `G_ST = (H_T - H_S)/H_T` — an
identity verified here to `3.2e-16` against the classical frequency-based
computation. Because `theta_T > theta_B` whenever the pools differ at all,
`G_ST < F_ST` **always**. Across simulated divergence from 0.02 to 0.40 the
ratio ran from **1.96x to 1.59x** — they never converge (panel B of the pitfalls
figure).

Consequence for practice: state which definition you used, and never compare
your value against a published threshold computed the other way.

## Trap 3 — F_ST is the wrong target to maximise

`F_ST` keeps rising as pools drift apart at loci that are *already* divergent,
which buys no additional heterosis. Under a dominance model the quantity that
generates heterosis is the fraction of loci fixed for **opposite** alleles in
the two pools. In the simulation these two decouple: at target `F_ST` = 0.15,
`mean|Δp|` = 0.19 but the fraction of oppositely-fixed loci was 0.00025;
reaching 2.3% of loci oppositely fixed required `F_ST` = 0.40, by which point
within-pool `GD` had fallen from 0.347 to 0.216.

Hold `F_ST` in a **band**, not at a maximum, and monitor complementarity
directly.

## The S1 panel

```{r}
s1_monitor(GL, pool, cycle = 12)
#  GD_within, GD_total, Fst_Wright, Gst_Nei, Ns_total,
#  GD_SS, GD_NSS, frac_opposite_fixed, mean_abs_delta_p
```

Log this every cycle and plot as a time series. Defend `GD_within` per pool;
watch `frac_opposite_fixed` for heterosis potential; treat `F_ST` as a
descriptor, not an objective.

Empirical anchor: Kadoumi et al. (2025, *TAG*) tracked diversity erosion
and heterotic-group divergence over decades of a commercial European dent maize
programme — the reference case for what these curves look like in a real
programme. For grouping methodology, Oyetunde et al. (2020, *Agronomy*) compare
four heterotic-grouping methods including SNP-based distance, and White et al.
(2020, *Crop Science*) characterise 329 ex-PVP inbreds for diversity structure
and heterotic pattern.

---

# S2 — Choosing crosses for DH induction

This is the stage where the largest share of pipeline diversity is lost, and it
is also the stage where the cheapest exact algebra is available.

## Three verified identities

For a DH family from cross `(i,j)` of two inbreds, each locus is inherited from
`i` or `j` with probability 1/2 and then doubled:

**(a)** coancestry between two DH sibs of the same family:
$$\mathbb{E}[f] = \frac{1 + f_{ij}}{2}$$

**(b)** coancestry between a DH of family `(i,j)` and one of family `(k,l)`:
$$\mathbb{E}[f] = \frac{f_{ik} + f_{il} + f_{jk} + f_{jl}}{4}$$

**(c)** group coancestry of a DH pool built from crosses `c = 1..C`, each
contributing `n_c` lines, `D = Σ n_c` in total:
$$\theta_{pool} = \mathbf{w}^{\top} \mathbf{F}_L \mathbf{w}
\;+\; \frac{1}{D}\left(1 - \overline{\left(\tfrac{1+f_c}{2}\right)}\right)$$

where `w` is the **parent-usage weight vector** over the candidate lines.
Verified against 800 simulated DH: residual `2.6e-04` with the self-coancestry
correction, `3.6e-04` without — the correction is `O(1/D)` and worth keeping.

**This is the key structural result of S2.** The diversity of the DH pool is a
quadratic form in *which parents you use and how often*. The DH lines never need
to be genotyped — or to exist — for their pool diversity to be optimised. You
can evaluate a crossing plan before making a single cross:

```{r}
dh_pool_theta(crosses, n_per = 40, fL)   # crosses: C x 2 matrix of line indices
```

## Progeny variance, and the trap of using distance for it

Under an additive model only loci segregating in the cross contribute:

$$\sigma^2_{DH(i,j)} = \frac{1}{4}\sum_{k \,:\, x_{ik} \ne x_{jk}} \beta_k^2$$

Verified against 3000 simulated DH per family (observed/predicted ratios
0.96–1.01). Combined with the mean it gives the usefulness criterion
(Schnell & Utz 1975):

$$UC = \mu_{ij} + i\,h\,\sigma_{DH(i,j)}$$

with `i = φ(z_q)/q` — `sel_intensity(0.05)` returns 2.0627 against the textbook
2.063.

## Trap 4 — genetic distance is not a proxy for progeny variance

It is common practice to rank crosses by marker distance on the argument that
more distant parents segregate for more loci. Once the heterotic pattern is
fixed, this fails:

| stratum | cor(σ_DH, MRD) | R² |
|---|---|---|
| within-pool (A × A) | +0.22 | 0.048 |
| inter-pool (A × B)  | +0.10 | 0.011 |
| **pooled**          | **+0.58** | **0.332** |

The pooled correlation is a **stratification artifact**: it separates inter-
from within-pool crosses and carries almost no information within either. A raw
count of segregating loci explains the same 0.331 — distance is adding nothing
beyond "are these parents in different pools", which you already know.

Rank crosses on `sigma_dh_crosses()` computed from marker effects, not on
distance. Panel A of the pitfalls figure shows the decoupling directly.

The applied literature here is genomic-prediction-based rather than
distance-based, consistent with this result: Beyene et al. (2021,
*Frontiers in Plant Science*) apply genomic selection at the early DH stage of a
tropical maize pipeline, and Cheng et al. (2025, *IJMS*) predict
inter-heterotic-group line crosses genomically for DH screening.

---

# S3 — Selecting lines per se

The unit is now the line and the metric set is the benchmarked one. Two points
of practice.

**Apply the constraint within each pool separately.** A pooled constraint can be
satisfied by keeping the pools apart while both erode internally — exactly the
failure mode S1 exists to prevent.

**Measure loss against a same-size resampling baseline.** Comparing a selected
set of `n` against the pool of `L` confounds selection with sample size:

```{r}
loss_vs_resampling(idx, fL, B = 200)
#  GD_selected, GD_random_mean, loss_pct, percentile
```

## Trap 5 — the incremental greedy constraint silently returns nothing

The natural implementation — "add the next-best line while `theta` stays below
the ceiling" — **cannot work**, and fails silently by returning an empty set.
Group coancestry includes the diagonal, so a single line has `theta = 1` and any
pair has `theta = (1 + f_ij)/2`. A ceiling calibrated on a large pool is
unreachable at small set sizes; the greedy never gets past the first candidate.
`theta` falls monotonically as the set grows, so **the constraint is only
meaningful at the target size**.

The working route is truncate-then-repair: take the top `n_sel` by merit, then
swap the most-related member for the least-related non-member until the ceiling
is met, breaking ties on merit.

```{r}
tmax <- theta_ceiling(fL, n_sel = 40, alpha = 0.005)   # 0.5% loss vs resampling
sel  <- select_lines_constrained(merit, fL, n_sel = 40, theta_max = tmax)
```

`theta_ceiling()` expresses the ceiling as a permitted **proportional loss**
against the same-size baseline, so the constraint means the same thing at every
set size.

---

# S4 — Hybrid mining

For F1 hybrids of homozygous lines the hybrid-by-hybrid coancestry matrix is
never needed. With `u` the line-usage count vector over the `n` selected hybrids
(`u_k` = number of selected hybrids using line `k`, `Σu = 2n`):

$$\theta_{hyb} = \frac{\mathbf{u}^{\top}\mathbf{F}_L\,\mathbf{u}}{(2n)^2}$$

Verified against the direct hybrid matrix: deviation **exactly 0**. This is the
collapse derived in the companion report — cost drops from `O(n²)` over an
`N × N` hybrid matrix to `O(L²)` with no hybrid matrix formed at all, and it is
what makes a differential-evolution inner loop over tens of thousands of
candidate hybrids affordable. The companion report benchmarks the cost and shows
that encoding the search over **line weights** rather than hybrid keys is what
determines whether the optimiser finds a good solution under many restrictions.

Expected heterozygosity of the hybrid set needs no separate computation: for
hybrids of fixed lines it is exactly `1 - theta_hyb`.

```{r}
u <- line_usage(idx, parents, L)
c(theta = hybrid_theta_from_usage(u, fL),
  He    = hybrid_He(u, fL),
  Ne    = effective_number(u),
  bal   = pool_balance(idx, parents, pool))
```

---

# Where the diversity actually goes

The four stages were simulated end to end over **8 independent founder pools**
(120 lines, 6000 markers, target `F_ST` = 0.15; 20 crosses → 40 DH each → 80
lines selected → 200 hybrids mined), under two cross-selection strategies:
merit only (UC), and UC plus a diversity term on parental coancestry.

Loss at each stage is the proportional loss of gene diversity against that
stage's **own same-size resampling baseline**, so the stages are on one scale.

| stage | merit only | merit + diversity |
|---|---|---|
| S2 DH induction  | **9.40%** (77% of total) | 3.43% (36%) |
| S3 line per se   | 0.32% (3%) | 2.98% (29%) |
| S4 hybrid mining | 2.49% (20%) | 3.47% (35%) |

**DH induction dominates.** Under merit-only cross selection it accounts for
**77% of all diversity lost in the pipeline** — more than three times the loss
at hybrid mining, which is where optimisation effort conventionally goes. The
reason is structural: 20 crosses out of 1770 possible within-pool combinations
is a far harsher bottleneck than 200 hybrids out of 1600, and the S2 collapse
(§S2c) makes it the cheapest stage to constrain.

## What the diversity term buys, and what it does not

| outcome | merit only | merit + div | paired t (n=8) |
|---|---|---|---|
| hybrid merit | 33.40 | 29.71 | **p = 0.0002** |
| effective no. parent lines | 24.20 | 25.33 | p = 0.24 |
| marker alleles retained | 11274 | 11314 | p = 0.52 |

The merit cost is **11%, consistent in all 8 pools**. Neither end-of-pipeline
diversity outcome improves significantly. The diversity term redistributes loss
across stages — it flattens the S2 spike — but with downstream selection still
running on merit alone, the pipeline reconverges. Diversity spent at S2 is
partly reclaimed at S3 and S4.

**The practical reading**: a diversity constraint applied at one stage buys
stage-level diversity, not end-of-pipeline richness. If allelic richness in the
commercial hybrid set is the target, the constraint has to be carried through
all four stages, and the loss ledger is the instrument for allocating a budget
across them.

---

# Cost summary

| stage | metric | cost | notes |
|---|---|---|---|
| S1 | `theta_decompose` | `O(L²)` once per cycle | monitoring, not inner loop |
| S2 | `dh_pool_theta` | `O(L²)` per plan | no DH genotypes needed |
| S2 | `sigma_dh_crosses` | `O(C·m)` | the real cost at this stage |
| S3 | `select_lines_constrained` | `O(swaps · n)` | swaps typically < 20 |
| S4 | `hybrid_theta_from_usage` | `O(L²)` per evaluation | no `N × N` matrix |

The `O(L²)` quadratic form is the recurring shape. With `L` in the hundreds it
is microseconds, which is why the whole pipeline can be optimised inside a
metaheuristic. The dominant cost anywhere in this document is
`sigma_dh_crosses` at S2, because it touches the marker matrix; cache it per
candidate cross rather than recomputing inside the optimiser loop.

---

# Checklist

1. Compute the coancestry matrix once per cycle; never report modified Rogers distance alongside it as separate evidence.
2. Monitor within-pool `GD` per pool, not pooled `GD`; state which `F_ST` definition you use; hold it in a band.
3. Rank crosses on `sigma_DH` from marker effects, never on parental distance.
4. Evaluate the crossing plan's `theta_pool` before making the crosses — it is a quadratic form in parent usage.
5. Calibrate every diversity ceiling against a same-size resampling baseline, and apply it at the target set size.
6. Constrain within pool, at every stage — a single-stage constraint does not survive to the hybrid set.
7. At S4, use the line-usage collapse; the hybrid coancestry matrix is never needed.
8. Keep the per-stage loss ledger. It is the only way to see that DH induction, not hybrid mining, is where the diversity goes.

---

# References

Verified from the screened corpus (`report/bibliography_plant.csv`); classical
results cited from the source literature.

**Classical.** Cockerham (1967) *Crop Science* 7:167–169 — group coancestry.
Wright (1978) *Evolution and the Genetics of Populations* Vol. 4 — modified
Rogers distance. Nei (1973) *PNAS* 70:3321–3323 — `G_ST`. Schnell & Utz (1975)
— usefulness criterion. Nejati-Javaremi, Smith & Gibson (1997) *J. Anim. Sci.*
75:1738–1745 — molecular/IBS coancestry. Caballero & Toro (2000) *Genet. Res.*
75:331–343; (2002) *Conserv. Genet.* 3:289–299 — coancestry, partition, and
effective size.

**Post-2018, screened.**

- Kadoumi R. et al. (2025) Evolution of population structure in a commercial European hybrid dent maize breeding program and consequences on genetic diversity. *Theoretical and Applied Genetics*. doi:10.1007/s00122-025-05008-5
- White M.R., Mikel M.A., de Leon N., Kaeppler S.M. (2020) Diversity and heterotic patterns in North American proprietary dent maize germplasm. *Crop Science*. doi:10.1002/csc2.20050
- Oyetunde O.A., Badu-Apraku B., Ariyo O.J., Alake C.O. (2020) Efficiencies of heterotic grouping methods for classifying early maturing maize inbred lines. *Agronomy* 10:1198. doi:10.3390/agronomy10081198
- Xue Y. et al. (2025) Insights into the genomic divergence of maize heterotic groups in China. *Journal of Integrative Plant Biology*. doi:10.1111/jipb.13884
- Beyene Y. et al. (2021) Application of genomic selection at the early stage of breeding pipeline in tropical maize. *Frontiers in Plant Science*. doi:10.3389/fpls.2021.685488
- Cheng D. et al. (2025) Genomic prediction for germplasm improvement through inter-heterotic-group line crossing in maize. *International Journal of Molecular Sciences* 26:2662. doi:10.3390/ijms26062662
- Hölker A.C. et al. (2022) Theoretical and experimental assessment of genome-based prediction in landraces of allogamous crops. *PNAS*. doi:10.1073/pnas.2121797119
- Polzer C. et al. (2025) Rapid cycling genomic selection in maize landraces. *Theoretical and Applied Genetics*. doi:10.1007/s00122-025-04855-6
- Arca M. et al. (2023) Genotyping of DNA pools identifies untapped landraces and genomic regions to develop next-generation varieties. *Plant Biotechnology Journal*. doi:10.1111/pbi.14022
- Mayer M. et al. (2020) Discovery of beneficial haplotypes for complex traits in maize landraces. *Nature Communications*. doi:10.1038/s41467-020-18683-3
- Beugnot A. et al. (2023) Identifying QTLs involved in hybrid performance and heterotic group complementarity. *Theoretical and Applied Genetics*. doi:10.1007/s00122-023-04431-w
- González-Diéguez D.O. et al. (2026) Sparse testcrossing for early-stage genomic prediction of general combining ability. *Theoretical and Applied Genetics*. doi:10.1007/s00122-026-05169-x
- Godoy dos Santos J.C. et al. (2026) Dissecting genetic variance structure and evaluating genomic prediction models for single-cross hybrids from Stiff Stalk and Non-Stiff Stalk heterotic groups. *G3*. doi:10.1093/g3journal/jkag163
- Mowers R.P., Nyhus K.A., Tragesser S.L., Foster D.J. (2018) Doubled haploid testcross design for predicting corn hybrid performance using molecular markers. *Crop Science*. doi:10.2135/cropsci2017.02.0086
- Bernardo R. (2021) Upgrading a maize breeding program via two-cycle genomewide selection. *Crop Science*. doi:10.1002/csc2.20516
