---
title: "From the Marker Matrix to Diversity Metrics"
subtitle: "Maize hybrid selection under a genetic diversity constraint"
author: "Technical document"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
    df_print: kable
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, comment = "#>")
options(digits = 6)
root <- if (basename(getwd()) == "doc") ".." else "."
for (f in list.files(file.path(root, "R"), full.names = TRUE)) source(f)
```

# Scope

This document describes, with explicit algebra and the code that implements
it, the path from the lines' marker matrix to the metrics used to constrain
diversity loss when selecting hybrids.

The central thesis is that **two different matrices are needed, and swapping
them is the easiest mistake to make**: the genomic matrix of @vanraden2008 to
*predict* performance, and the molecular coancestry matrix of @caballero2002
to *measure* diversity. Section \@ref(why-not-g) shows algebraically and
numerically why the first one cannot serve the second.

As a running example, a small simulated set:

```{r data}
cfg <- modifyList(sim_config, list(n_pool_A = 25, n_pool_B = 25, m = 1500))
ctx <- simulate_data(cfg)
n_sel <- 40
c(hybrids = ctx$N, markers = ctx$m, lines = ctx$n_lines)
```

# From Line to Hybrid

## Coding

Let $\mathbf{M}^{L}$ be the $L \times m$ matrix of line genotypes, with
$M^{L}_{ik} \in \{0, 2\}$ = count of the reference allele at marker $k$.
Inbred lines are homozygous, so the value 1 never occurs — residual
heterozygotes must be imputed or discarded beforehand.

## Hybrid Genotype

The $F_1$ hybrid between lines $a$ and $b$ receives one gamete from each
parent. Since both parents are homozygous, the gamete from $a$ carries
$M^{L}_{ak}/2$ copies of the reference allele, and therefore

$$
M^{H}_{(ab)k} \;=\; \frac{M^{L}_{ak} + M^{L}_{bk}}{2} \;\in\; \{0, 1, 2\}.
$$

The value 1 appears exactly at markers where the two lines differ — these are
the hybrid's heterozygous loci, and they are the ones carrying heterosis.
Predicting the hybrid's genotype is **deterministic**, not statistical: it is
the one piece of information that requires no assumption at all
[@technow2014].

```{r hybrids, eval=FALSE}
# R/00_data.R
ped <- expand.grid(a = a_ids, b = b_ids)                    # all A x B combinations
M   <- (L[ped$a, , drop = FALSE] + L[ped$b, , drop = FALSE]) / 2
```

From here on we work with the **within-individual reference-allele
frequency**, which is the natural quantity for coancestry:

$$
x_{ik} \;=\; \frac{M^{H}_{ik}}{2} \;\in\; \{0,\; 0.5,\; 1\}.
$$

```{r x}
str(ctx$X[1:3, 1:6])
```

# The VanRaden Genomic Matrix {#g-vanraden}

## Definition

With $p_k$ = reference-allele frequency in the hybrid population,
$\mathbf{Z} = \mathbf{M}^{H} - \mathbf{1}\,2\mathbf{p}'$ is the centered matrix
and

$$
\mathbf{G} \;=\; \frac{\mathbf{Z}\mathbf{Z}'}{2\sum_{k} p_k (1-p_k)}.
$$

This is @vanraden2008's first parametrization. It estimates realized additive
relationship and is what feeds GBLUP to predict traits from already-estimated
marker effects.

```{r g, eval=FALSE}
vanraden_G <- function(M) {
  p <- colMeans(M) / 2
  Z <- sweep(M, 2, 2 * p, "-")
  list(G = tcrossprod(Z) / (2 * sum(p * (1 - p))), Z = Z, p = p)
}
```

## Why G Cannot Measure Diversity {#why-not-g}

Centering on $2\mathbf{p}$ uses frequencies **from the population itself**.
So every column of $\mathbf{Z}$ sums to zero:

$$
\mathbf{Z}'\mathbf{1} = \mathbf{0}
\quad\Longrightarrow\quad
\mathbf{1}'\mathbf{G}\mathbf{1}
\;=\; \frac{\mathbf{1}'\mathbf{Z}\mathbf{Z}'\mathbf{1}}{2\sum_k p_k q_k}
\;=\; \frac{\lVert \mathbf{Z}'\mathbf{1}\rVert^2}{2\sum_k p_k q_k}
\;=\; 0 .
$$

The sum of **all** elements of $\mathbf{G}$ is identically zero. Since group
coancestry is proportional to that sum, it equals zero for the whole
population, and the status number $N_s = 1/(2\theta)$ diverges:

```{r g-zero}
c(sum_G       = sum(ctx$G),
  theta_on_G  = mean(ctx$G) / 2,
  Ns_on_G     = 1 / (2 * mean(ctx$G) / 2))
```

In subsets the sum isn't zero, but it is a **centered** quantity that
oscillates around zero and takes negative values. It is not a probability and
admits no identity-by-descent interpretation.

> The root of the problem is the implicit base population. $\mathbf{G}$
> measures relatedness *relative* to the current population; diversity
> requires an *absolute* scale. See @gomezromano2016 and @villanueva2020 for
> a discussion of which genomic coancestry matrix to use in which context.

# The Molecular Coancestry Matrix {#f-molecular}

## Definition

Malécot's coancestry [@malecot1948] between $i$ and $j$ is the probability
that two randomly sampled alleles, one from each individual, at the same
locus, are identical. Replacing identity by descent with **identity by
state** — what markers actually observe — yields molecular coancestry
[@caballero2002; @caballero2000]:

$$
f_{ij} \;=\; \frac{1}{m}\sum_{k=1}^{m}
\Big[\, x_{ik}\,x_{jk} \;+\; (1-x_{ik})(1-x_{jk}) \,\Big].
$$

The first term is the probability that both sampled alleles are the
reference allele; the second, that both are the alternative allele.
Properties:

- $f_{ij} \in [0, 1]$ always. **A negative value is a symptom of the wrong
  matrix being used.**
- Self-coancestry is $f_{ii} = (1 + F_i)/2$, with $F_i$ the individual's
  molecular homozygosity [@wright1922]. For a fully heterozygous $F_1$,
  $f_{ii} = 0.5$.
- No centering, no scaling, no arbitrary base population.

## Computable Form

Computing the double sum naively is $O(n^2 m)$. Expanding:

$$
\sum_k \big[x_{ik}x_{jk} + (1-x_{ik})(1-x_{jk})\big]
= 2\sum_k x_{ik}x_{jk} \;-\; \sum_k x_{ik} \;-\; \sum_k x_{jk} \;+\; m,
$$

which gives, with $\mathbf{s} = \mathbf{X}\mathbf{1}$ (row sums),

$$
\boxed{\;\mathbf{f} \;=\; \frac{1}{m}\Big( 2\,\mathbf{X}\mathbf{X}'
\;-\; \mathbf{s}\mathbf{1}' \;-\; \mathbf{1}\mathbf{s}' \;+\; m\,\mathbf{J} \Big)\;}
$$

with $\mathbf{J}$ the all-ones matrix. **A single matrix product**, the same
cost as a $\mathbf{G}$ matrix.

```{r f-code, eval=FALSE}
molecular_coancestry <- function(X) {
  m <- ncol(X)
  s <- rowSums(X)
  (2 * tcrossprod(X) - outer(s, rep(1, nrow(X))) -
     outer(rep(1, nrow(X)), s) + m) / m
}
```

```{r f-check}
c(minimum = min(ctx$f), maximum = max(ctx$f),
  diagonal_mean = mean(diag(ctx$f)), off_diagonal_mean = mean(ctx$f[upper.tri(ctx$f)]))
```

The value 0.5 would be the floor of the diagonal, reached by a hybrid
heterozygous at *every* marker. The excess above 0.5 measures exactly the
fraction of markers at which the two parent lines carry the same allele —
loci where the $F_1$ is homozygous and, by definition, contribute nothing to
heterosis:

```{r f-diagonal}
fd <- mean(diag(ctx$f))
c(diagonal_mean = fd,
  molecular_F = 2 * fd - 1,             # f_ii = (1 + F_i)/2
  heterozygous_fraction = 2 * (1 - fd))
```

This fraction depends directly on the divergence between heterotic pools
(`fst` in `sim_config`): more divergent pools produce more heterozygous $F_1$
hybrids and a diagonal closer to 0.5.

# Metrics Derived from f

## Group Coancestry and the Off-Diagonal Metric Problem

The group coancestry [@cockerham1967] of a set $S$ of $n$ individuals is the
mean of **all** $n^2$ elements, diagonal included:

$$
\theta_S \;=\; \frac{1}{n^2}\sum_{i \in S}\sum_{j \in S} f_{ij}
\;=\; \frac{1}{n^2}\,\mathbf{1}'\mathbf{f}_S\mathbf{1}.
$$

It is the probability that two alleles sampled at random *from the whole
set* are identical — including the possibility of both coming from the same
individual. That is why the diagonal enters: it represents the chance of
sampling the same individual twice, which is real when the set is used as a
crossing pool.

Splitting diagonal ($\bar f_d$) from off-diagonal ($\bar f_o$):

$$
\theta_S \;=\; \frac{\bar f_d}{n} \;+\; \frac{n-1}{n}\,\bar f_o .
\tag{1}
$$

The off-diagonal mean $\bar f_o$ is therefore just one of the two terms. When
$\bar f_d$ varies little across individuals — the case of an all-$F_1$
population — the two metrics rank almost identically, and the off-diagonal
metric works fine as a *ranking*. What it does not provide is the
interpretable scale and the correct baseline (Sections \@ref(ns) and
\@ref(baseline)).

```{r theta}
idx <- sample.int(ctx$N, n_sel)
theta_group(idx, ctx)
```

## Gene Diversity: θ and He Are the Same Quantity {#identity}

This is the result that simplifies everything else. Starting from the
definition and swapping the order of summation:

$$
\theta_S \;=\; \frac{1}{m}\sum_k \left[
\frac{1}{n^2}\sum_i\sum_j x_{ik}x_{jk}
+ \frac{1}{n^2}\sum_i\sum_j (1-x_{ik})(1-x_{jk}) \right]
$$

The double sum factors: $\frac{1}{n^2}\sum_i\sum_j x_{ik}x_{jk} =
\big(\frac{1}{n}\sum_i x_{ik}\big)^2 = \bar p_k^2$, where $\bar p_k$ is the
**group's** allele frequency. Hence

$$
\theta_S \;=\; \frac{1}{m}\sum_k \big(\bar p_k^2 + \bar q_k^2\big),
\qquad\text{and therefore}\qquad
\boxed{\;GD_S \;=\; 1 - \theta_S \;=\; \frac{1}{m}\sum_k 2\,\bar p_k \bar q_k
\;\equiv\; H_e\;}
$$

which is exactly Nei's gene diversity [@nei1973]. **Group coancestry and
expected heterozygosity are not independent checks: they are the same
quantity written two ways.**

Three practical consequences:

1. One fewer metric to justify to the technical committee.
2. "Losing 5% of diversity" gains an exact meaning: 5% of the set's expected
   heterozygosity.
3. Two computation routes with the same result and very different costs —
   the matrix route is $O(n^2)$, the frequency route is $O(nm)$. For
   $n=200$ and $m=25000$ that is 40 thousand versus 5 million operations.
   **The matrix route is ~100x cheaper, and it's the one that belongs in the
   optimizer's inner loop.**

```{r identity}
c(via_submatrix  = theta_group(idx, ctx),
  via_frequency   = theta_from_freq(idx, ctx),
  difference      = theta_group(idx, ctx) - theta_from_freq(idx, ctx))

c(GD = gene_diversity(idx, ctx), He_Nei = he_nei(idx, ctx))
```

Equality to machine precision is the implementation's strongest sanity check:
if it fails, matrix `f` is centered, rescaled, or has the wrong coding.

## Status Number Ns {#ns}

@lindgren1997 and @lindgren1998 propose expressing group coancestry on the
scale of individual counts:

$$
N_s \;=\; \frac{1}{2\,\theta_S}.
$$

The reading is direct: $N_s$ is the number of **unrelated, non-inbred**
individuals that would have the same gene diversity as the observed set.
Edge cases:

```{r ns-limits}
fake <- ctx
fake$f <- matrix(0.5, 5, 5)      # unrelated, p = 0.5
c(theta = theta_group(1:5, fake), Ns = status_number(1:5, fake))

fake$f <- matrix(1, 5, 5)        # all identical and homozygous
c(theta = theta_group(1:5, fake), Ns = status_number(1:5, fake))
```

### An Important Caveat About Ne {-}

@toro2020 show that the **rate-based** effective size, $N_e = 1/(2\Delta F)$,
loses meaning under molecular diversity management: optimization increases
diversity in early generations, producing a negative $\Delta F$ and therefore
a negative $N_e$, and its long-term behavior fails to converge to an
asymptotic value.

This **does not invalidate** $N_s$. The distinction is that $N_s$ is a
monotonic reparametrization of $\theta$ — a *static descriptor* of the set —
while $N_e$ is a *rate* between generations. Use $N_s$ to compare sets; don't
extrapolate it to project future drift.

For the same reason, do not use the linkage-disequilibrium-based $N_e$ here
($N_e = 1/(3(\bar r^2 - 1/S))$): it estimates population history, not the
composition of a selected sample.

## Effective Number of Parents

A metric independent of $\mathbf{f}$, computed purely from the hybrids'
pedigree. Letting $p_i$ be the frequency with which line $i$ appears as a
parent among the selected hybrids,

$$
N_{e(\text{par})} \;=\; \frac{1}{\sum_i p_i^2},
$$

the inverse Simpson index. It captures precisely what coancestry blurs: 200
hybrids can have acceptable $\theta$ while still drawing on only 12 lines, if
those 12 happen to be genetically distant from one another.

```{r ne-par}
c(random          = ne_parents(sel_random(ctx, n_sel), ctx),
  truncation      = ne_parents(sel_truncation(ctx, n_sel), ctx),
  theoretical_max = ctx$n_lines)
```

Its cost is negligible (a `tabulate` count), which makes it a natural
candidate to **enter the objective function alongside $\theta$**.

## Allelic Richness

$\theta$ and $H_e$ are averages across markers and are dominated by
intermediate-frequency alleles. A lost rare allele barely moves them, yet the
loss is irreversible. We count markers that were polymorphic in the
population and became rare or fixed in the selection:

$$
A_{\text{lost}} \;=\; \#\{k : \text{MAF}_k^{(S)} < 0.05 \;\wedge\;
\text{MAF}_k^{(\text{pop})} \ge 0.05 \}.
$$

```{r richness}
c(random      = alleles_lost(sel_random(ctx, n_sel), ctx),
  truncation  = alleles_lost(sel_truncation(ctx, n_sel), ctx))
```

It is the only metric in the set **genuinely non-redundant** with $\theta$.

## Distances: E-NE and A-NE

With distance $d_{ij} = 1 - f_{ij}$, @debeukelaer2018 define in Core Hunter 3
two measures that capture opposite aspects that $\theta$ alone cannot
separate:

$$
\text{E-NE} = \frac{1}{n}\sum_{i \in S} \min_{j \in S,\, j \ne i} d_{ij},
\qquad
\text{A-NE} = \frac{1}{N}\sum_{i=1}^{N} \min_{j \in S} d_{ij}.
$$

- **E-NE** (*entry-to-nearest-entry*) measures **non-redundancy within the
  selection**. It tells apart "200 individuals spread out" from "100
  near-identical pairs spread out", configurations that can share the same
  $\theta$.
- **A-NE** (*accession-to-nearest-entry*) measures **representativeness of
  the population**. It detects a selection that optimizes $\theta$ by
  concentrating in one region of genetic space while ignoring the rest.

```{r ene-ane}
sapply(list(random = sel_random(ctx, n_sel),
            truncation = sel_truncation(ctx, n_sel),
            greedy_div = sel_greedy(ctx, n_sel, w = 0)),
       function(s) c(ENE = ene(s, ctx), ANE = ane(s, ctx)))
```

## Effective Dimensionality

Over the eigenvalues $\lambda_1 \ge \dots \ge \lambda_n$ of the submatrix
$\mathbf{f}_S$, the participation ratio

$$
D_{\text{eff}} \;=\; \frac{\left(\sum_i \lambda_i\right)^2}{\sum_i \lambda_i^2}
$$

counts how many independent directions of genetic variation remain. It
equals 1 if all variation lies on a single axis and $n$ if it is spread
evenly.

```{r effdim}
c(random     = eff_dim(sel_random(ctx, n_sel), ctx),
  truncation = eff_dim(sel_truncation(ctx, n_sel), ctx))
```

## Decomposition by Heterotic Group

In hybrid maize the diversity **between pools** is the engine of heterosis
and cannot be spent; the diversity **within each pool** is the fuel for
future progress. Following the logic of subdivided populations
[@caballero2002; @fernandez2019], we compute molecular coancestry between
lines $\mathbf{f}^{L}$ and weight it by each line's usage among the selected
hybrids:

$$
\theta_A = \mathbf{w}_A' \mathbf{f}^{L} \mathbf{w}_A, \qquad
\theta_B = \mathbf{w}_B' \mathbf{f}^{L} \mathbf{w}_B, \qquad
\theta_{AB} = \mathbf{w}_A' \mathbf{f}^{L} \mathbf{w}_B,
$$

with $\mathbf{w}_A$ the vector of usage frequencies of pool-A lines.

```{r pools}
rbind(random     = theta_pools(sel_random(ctx, n_sel), ctx),
      truncation = theta_pools(sel_truncation(ctx, n_sel), ctx))
```

A blind optimizer can keep global $\theta$ acceptable while eroding one of
the pools. Without the decomposition this goes unnoticed, and the
consequence only shows up the following cycle.

# The Correct Baseline {#baseline}

## Why Not Compare Against the 5000 × 5000 Matrix

The natural comparison — coancestry of the selected set against that of the
whole population — is **biased by construction**. Applying equation (1) to
both sets:

$$
\theta_S - \theta_N
= \left(\frac{1}{n} - \frac{1}{N}\right)\left(\bar f_d - \bar f_o\right)
\;>\; 0,
$$

since $\bar f_d > \bar f_o$ whenever individuals are not clones. A
**random** sample of 200 already looks less diverse than the population of
5000, without any selection having occurred. The bias is purely from
sampling and grows the smaller the sample.

Checking the formula numerically:

```{r bias}
fd <- mean(diag(ctx$f)); fo <- mean(ctx$f[upper.tri(ctx$f)])
predicted_bias <- (1/n_sel - 1/ctx$N) * (fd - fo)

set.seed(11)
observed_bias <- mean(replicate(400, theta_group(sample.int(ctx$N, n_sel), ctx))) -
                 theta_group(seq_len(ctx$N), ctx)

c(predicted = predicted_bias, observed = observed_bias)
```

Note also that the **off-diagonal** mean of a random subset is an *unbiased*
estimator of the population's off-diagonal mean. In other words, the metric
you already use is free of this bias — but it is also not comparable across
sample sizes on the $N_s$ scale.

## Operational Definition of Loss

The "0% loss" reference is the **distribution of random subsets of the same
size**, obtained by resampling:

$$
\alpha \;=\; \frac{GD_{\text{random}(n)} - GD_S}{GD_{\text{random}(n)}}.
$$

So $\alpha = 0$ means "as diverse as a random sample of the same size", and
$\alpha < 0$ means more diverse than chance. Resampling also throws in the
standard deviation, and hence the z-score of any strategy.

```{r null}
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
c(GD_population = gene_diversity(seq_len(ctx$N), ctx),
  GD_reference = null$gd_ref, sd = null$gd_sd)

sapply(list(random = sel_random(ctx, n_sel),
            truncation = sel_truncation(ctx, n_sel),
            trunc_cap = sel_truncation_cap(ctx, n_sel)),
       alpha_loss, ctx = ctx, gd_ref = null$gd_ref)
```

## How to Set the α Limit

The acceptable limit **should not be chosen a priori**. Pure truncation —
the greediest possible selection — defines the maximum attainable $\alpha$.
If that ceiling is lower than the intended limit, the constraint restricts
nothing.

```{r alpha-max}
attainable_alpha_max <- alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref)
c(attainable_alpha_max = attainable_alpha_max,
  intended_limit = 0.05,
  active_constraint = attainable_alpha_max > 0.05)
```

On the full dataset (2500 hybrids, 5000 markers, selecting 100) that ceiling
came out at **3.63%** — a 5% limit would be inoperative. The correct
procedure is to trace the $\alpha \times$ gain frontier and choose the knee
of the curve.

# Use Inside the Optimizer

## Constraint Formulation, Not a Weighted Sum

With $\mathbf{u}$ the multi-trait index and $S$ the set of $n$ selected
individuals:

$$
\max_{S,\; |S| = n} \; \frac{1}{n}\sum_{i \in S} u_i
\qquad \text{subject to} \qquad
\alpha(S) \;\le\; \alpha_{\max}.
$$

The common alternative, $\max\; \bar u - \lambda\,\theta_S$, requires
calibrating $\lambda$, which has no direct interpretation. The constraint
form preserves the operational meaning: "I accept losing at most
$\alpha_{\max}$ of diversity". It is the same principle behind optimal
contribution selection [@meuwissen1997], differing in that there
contributions $\mathbf{c}$ are continuous while here the problem is
combinatorial (0/1 with equal weight), which motivates using differential
evolution [@storn1997; @mullen2011].

```{r fitness, eval=FALSE}
make_fitness <- function(ctx, n_sel, alpha_max, gd_ref, penalty = 1000) {
  function(par) {
    idx  <- decode(par, ctx$N, n_sel)
    gd   <- 1 - mean(ctx$f[idx, idx])          # matrix route: 40k operations
    viol <- max(0, (gd_ref - gd) / gd_ref - alpha_max)
    -(mean(ctx$index[idx]) - penalty * viol)   # DEoptim minimizes
  }
}
```

## Which Metric Belongs in the Inner Loop

The optimizer performs on the order of $10^5$ to $10^6$ evaluations. The cost
per call decides what goes into the objective function and what is left for
post-hoc validation:

```{r cost}
cost_per_eval(ctx, n_sel, reps = 20)
```

Practical conclusion: $\theta$ (matrix route) and $N_{e(\text{par})}$ go into
the objective; allelic richness, A-NE, and $G_{st}$ are computed **once** on
the final solution.

## Convergence Diagnostic

The problem's continuous relaxation — allowing $0 \le c_i \le 1/n$ with
$\sum_i c_i = 1$ instead of $c_i \in \{0, 1/n\}$ — provides a valid upper
bound, since every discrete solution is feasible under the relaxed problem. A
large gap between the differential evolution result and that bound indicates
**the optimizer hasn't converged**, not that the problem is intrinsically
hard.

In practice, seeding the initial population with heuristic solutions (greedy
on $\theta$, truncation with a per-parent usage cap) is decisive: without it
the optimizer ends up *below* a plain greedy.

# Operational Summary

| Step | Object | Formula |
|---|---|---|
| 1 | Hybrid genotype | $M^{H}_{(ab)k} = (M^{L}_{ak} + M^{L}_{bk})/2$ |
| 2 | Trait prediction | VanRaden $\mathbf{G}$, $\mathbf{Z}\mathbf{Z}'/2\sum p_kq_k$ |
| 3 | **Diversity** | $\mathbf{f} = (2\mathbf{X}\mathbf{X}' - \mathbf{s}\mathbf{1}' - \mathbf{1}\mathbf{s}' + m\mathbf{J})/m$ |
| 4 | Group coancestry | $\theta_S = \mathbf{1}'\mathbf{f}_S\mathbf{1}/n^2$ |
| 5 | Gene diversity | $GD = 1 - \theta_S \equiv H_e$ |
| 6 | Interpretable scale | $N_s = 1/(2\theta_S)$ |
| 7 | Baseline | resampling of $n$ individuals, not the full population |
| 8 | Loss | $\alpha = (GD_{\text{random}} - GD_S)/GD_{\text{random}}$ |

Mistakes to avoid, in order of severity:

1. Computing coancestry over centered $\mathbf{G}$ — a meaningless result.
2. Comparing the $n \times n$ submatrix against the $N \times N$ matrix —
   sampling bias read as diversity loss.
3. Fixing $\alpha_{\max}$ before knowing the attainable ceiling.
4. Omitting the diagonal from $\theta$ — small impact in an $F_1$ population,
   but required for the $N_s$ scale to be meaningful.
5. Running the optimizer without a warm start and concluding the problem is
   hard.

# References
