# Coancestry, and the two matrices {#sec-coancestry}

```{r}
#| include: false
source("_common.R")
```

Two matrices are built from the same marker data and they are not
interchangeable. One predicts traits. The other measures diversity. Using the
first for the second is the most consequential mistake available in this
subject, and it fails silently -- it returns a number, and the number is
meaningless.

This chapter builds both, shows the failure, and derives the matrix used for
the rest of the book.

## From line to hybrid

Let $\mathbf{M}^{L}$ be the $L \times m$ matrix of line genotypes, with
$M^{L}_{ik} \in \{0, 2\}$ the count of the reference allele at marker $k$.
Inbred lines are homozygous, so the value 1 never occurs -- residual
heterozygotes must be imputed or discarded beforehand.

The F1 hybrid between lines $a$ and $b$ receives one gamete from each parent.
Since both parents are homozygous, the gamete from $a$ carries $M^{L}_{ak}/2$
copies of the reference allele, so

$$
M^{H}_{(ab)k} \;=\; \frac{M^{L}_{ak} + M^{L}_{bk}}{2} \;\in\; \{0, 1, 2\}.
$$

The value 1 appears exactly at markers where the two lines differ. Those are
the hybrid's heterozygous loci, and they are the ones carrying heterosis.

Predicting the hybrid's *genotype* is deterministic, not statistical. It is the
one piece of information in the whole pipeline that requires no assumption at
all -- unlike its phenotype, which requires a great many.

From here on we work with the **within-individual reference-allele frequency**,

$$
x_{ik} \;=\; \frac{M^{H}_{ik}}{2} \;\in\; \{0,\; 0.5,\; 1\},
$$

which is the natural quantity for coancestry: it is the probability that an
allele drawn at random from individual $i$ at marker $k$ is the reference
allele.

```{r}
ctx <- simulate_data(book_cfg)
ctx$X[1:4, 1:8]
```

Rows are hybrids, columns markers. A 0.5 is a heterozygous locus.

## The VanRaden matrix, and why it cannot measure diversity {#sec-why-not-g}

With $p_k$ the reference-allele frequency in the hybrid population and
$\mathbf{Z} = \mathbf{M}^{H} - \mathbf{1}\,2\mathbf{p}'$ the centered matrix,

$$
\mathbf{G} \;=\; \frac{\mathbf{Z}\mathbf{Z}'}{2\sum_{k} p_k (1-p_k)}.
$$

This is VanRaden's first parametrisation. It estimates realised additive
relationship and it is the right matrix for GBLUP. It is implemented as
`vanraden_G()`:

```{r}
#| file: !expr file.path("..", "R", "00_data.R")
#| eval: false
#| echo: false
```

```{r}
args(vanraden_G)
```

Now the failure. Centering on $2\mathbf{p}$ uses frequencies from the
population itself, so every column of $\mathbf{Z}$ sums to zero:

$$
\mathbf{Z}'\mathbf{1} = \mathbf{0}
\quad\Longrightarrow\quad
\mathbf{1}'\mathbf{G}\mathbf{1}
\;=\; \frac{\lVert \mathbf{Z}'\mathbf{1}\rVert^2}{2\sum_k p_k q_k}
\;=\; 0 .
$$

The sum of **all** elements of $\mathbf{G}$ is identically zero. Group
coancestry is proportional to that sum, so it is zero for the whole population,
and the status number $N_s = 1/(2\theta)$ diverges:

```{r}
c(sum_G      = sum(ctx$G),
  theta_on_G = mean(ctx$G) / 2,
  Ns_on_G    = 1 / (2 * mean(ctx$G) / 2))
```

That is not a rounding artefact to be worked around. It is exact, and it is
what centering means. In a *subset* the sum is not zero, but the quantity is
still centered: it oscillates around zero, takes negative values, is not a
probability, and admits no identity-by-descent interpretation.

::: {.callout-important}
## The root cause is the implicit base population

$\mathbf{G}$ measures relatedness *relative to the current population*.
Diversity requires an *absolute* scale. Any matrix whose centering constant is
estimated from the same data whose diversity you are measuring will have this
problem, whatever it is called.
:::

## Molecular coancestry {#sec-f-molecular}

Malécot's coancestry between $i$ and $j$ is the probability that two randomly
sampled alleles, one from each individual at the same locus, are identical.
Replacing identity *by descent* with identity *by state* -- which is what
markers actually observe -- yields molecular coancestry
[@caballero2002; @caballero2000]:

$$
f_{ij} \;=\; \frac{1}{m}\sum_{k=1}^{m}
\Big[\, x_{ik}\,x_{jk} \;+\; (1-x_{ik})(1-x_{jk}) \,\Big].
$$

The first term is the probability that both sampled alleles are the reference
allele; the second, that both are the alternative. Three properties follow
immediately:

- $f_{ij} \in [0, 1]$ always. **A negative value is a symptom of the wrong
  matrix.**
- Self-coancestry is $f_{ii} = (1 + F_i)/2$, with $F_i$ the individual's
  molecular homozygosity. For a fully heterozygous F1, $f_{ii} = 0.5$.
- No centering, no scaling, no arbitrary base population.

### Computable form

The double sum is $O(n^2 m)$ if computed naively. Expanding,

$$
\sum_k \big[x_{ik}x_{jk} + (1-x_{ik})(1-x_{jk})\big]
= 2\sum_k x_{ik}x_{jk} \;-\; \sum_k x_{ik} \;-\; \sum_k x_{jk} \;+\; m,
$$

so with $\mathbf{s} = \mathbf{X}\mathbf{1}$ the row sums,

$$
\mathbf{f} \;=\; \frac{1}{m}\Big( 2\,\mathbf{X}\mathbf{X}'
\;-\; \mathbf{s}\mathbf{1}' \;-\; \mathbf{1}\mathbf{s}' \;+\; m\,\mathbf{J} \Big),
$$

a single matrix product -- the same cost as building $\mathbf{G}$.

```{r}
molecular_coancestry
```

```{r}
fmt(c(minimum       = min(ctx$f),
      maximum       = max(ctx$f),
      diagonal_mean = mean(diag(ctx$f)),
      offdiag_mean  = mean(ctx$f[upper.tri(ctx$f)])))
```

Everything is inside $[0, 1]$, as it must be.

### Reading the diagonal

The value 0.5 is the floor of the diagonal, reached by a hybrid heterozygous at
*every* marker. The excess above 0.5 measures exactly the fraction of markers
at which the two parent lines carry the same allele -- loci where the F1 is
homozygous and therefore contributes nothing to heterosis.

```{r}
fd <- mean(diag(ctx$f))
fmt(c(diagonal_mean         = fd,
      molecular_F           = 2 * fd - 1,     # f_ii = (1 + F_i) / 2
      heterozygous_fraction = 2 * (1 - fd)))
```

This fraction depends directly on the divergence between the heterotic pools
(`fst` in the configuration): more divergent pools give more heterozygous F1
hybrids and a diagonal closer to 0.5. That relationship is what
@sec-simulating-crosses puts under experimental control.

## `f` is not an ad hoc substitute

Two independent lines of work converge on this matrix, which is worth stating
because it is otherwise easy to read $\mathbf{f}$ as "the thing we used because
$\mathbf{G}$ broke".

Nejati-Javaremi et al. [-@nejatijavaremi1997] proposed the *allelic-similarity*
method, $f_{ij} = \tfrac{1}{4}\sum_{k=1}^{2}\sum_{l=1}^{2}\delta_{kl}$ with
$\delta_{kl} = 1$ when the $k$-th allele of $i$ equals the $l$-th of $j$. That
is exactly the identity-by-state average above, generalised from single loci to
a genome-wide mean. @gomezromano2016 contrast it directly against
crossproduct-based matrices: the allelic-similarity matrix is guaranteed to stay
in $[0, 1]$ by construction, while crossproduct matrices normalised by *current*
allele frequencies are not -- precisely the failure demonstrated above.

The same matrix arrives from a second direction. @toro2014 and @meuwissen2020
show that fixing the VanRaden centering frequency at $p_k = 0.5$ for every
marker, instead of the sample frequency, produces a matrix -- written
$\mathbf{G}_{0.5}$ -- that is proportional to homozygosity-based inbreeding and
molecular coancestry. Structurally that is $\mathbf{f}$: replacing $2p_k$ with
$1$ removes the population-dependent term that forced
$\mathbf{1}'\mathbf{G}\mathbf{1} = 0$.

So $\mathbf{f}$ is the member of the VanRaden family that has an absolute scale,
reached independently from two unrelated derivations. That it also happens to be
the one that does not break is not a coincidence.

## The division of labour

```{r}
#| echo: false
knitr::kable(
  data.frame(
    Matrix = c("`G` (VanRaden)", "`f` (molecular coancestry)"),
    Centering = c("sample allele frequencies", "none"),
    Range = c("centered, signed", "[0, 1]"),
    `Sum of all elements` = c("exactly 0", "positive, informative"),
    Use = c("trait prediction (GBLUP)", "every diversity quantity"),
    check.names = FALSE
  ),
  caption = "Two matrices, two jobs. They are built from the same markers and must not be substituted for one another."
)
```

This split is enforced throughout `hybdiv`: every metric takes `ctx$f`, and
`ctx$G` appears only where traits are predicted. The test suite asserts
`sum(G) == 0` precisely so that a future change which silently starts feeding
`G` to a diversity metric fails loudly.

The next chapter takes $\mathbf{f}$ and asks what happens when the population it
describes is not one population, but several.
