# Diversity in a subdivided population {#sec-subdivided}

```{r}
#| include: false
source("_common.R")
```

A breeding programme is never one population. It is heterotic groups, or
accessions, or landraces, or cycles -- and the diversity of the whole is not the
average of the diversity of the parts. @caballero2002 give the partition that
makes this precise, and it is the analytical backbone of everything that
follows: the constraint in @sec-ocs is a special case of it, and the
within/between asymmetry that makes plant hybrid breeding different from animal
breeding is a statement about which term of it you are allowed to minimise.

The paper's central claim is uncomfortable and worth stating first: **a method
that ranks accessions by distance alone can invert the correct conclusion.** An
accession can be the most distant *because* it is internally fixed, and removing
it can *raise* the diversity of the pool. We reproduce that result numerically
below.

## Notation

- A metapopulation of $n$ subpopulations; subpopulation $i$ has $N_i$
  individuals, $N_T = \sum_i N_i$.
- A tilde, $\tilde{x}$, is a weighted mean **within** subpopulations:
  $\tilde{f} = \sum_i f_{ii} N_i / N_T$.
- A bar, $\bar{x}$, is a mean over **all pairs** in the metapopulation,
  between-subpopulation pairs included:
  $\bar{f} = \sum_{i,j} f_{ij} N_i N_j / N_T^2$.
- Always $\bar{f} \le \tilde{f} \le \tilde{s}$. The differences between those
  three numbers *are* the partition of diversity.
- Everything is computed per locus and then averaged over loci. Never average
  frequencies first.

::: {.callout-warning}
## The subtle point, and the one most often got wrong

Every coancestry here includes the pairs of an individual with **itself** --
group coancestry in the sense of Cockerham. That is what makes $1 - f_{ii}$
exactly Nei's expected heterozygosity, and it is what makes the algebra close.
Drop the diagonal and none of the identities below hold.
:::

## The three coancestry blocks

### $f_{ij}$, between subpopulations

The probability that two alleles, one sampled at random in subpopulation $i$ and
one in $j$, are identical. From markers, $f_{ij} = \sum_k \tilde p_{k,i}
\tilde p_{k,j}$.

Read it as **redundancy** between two materials. If you cross plants of $i$ with
plants of $j$, the progeny has expected inbreeding $F = f_{ij}$. It answers:
what do I gain by recombining these two accessions?

In maize, $f_{ij}$ between a Stiff Stalk pool and a Non-Stiff Stalk pool is
typically low -- and that low between-group coancestry is exactly what sustains
the heterotic pattern. Between two elite lines of the same family it can exceed
0.6, and the cross generates nothing new.

### $f_{ii}$, within a subpopulation

The case $i = j$. From markers $f_{ii} = \sum_k \tilde p_{k,i}^2$, the expected
homozygosity, so

$$1 - f_{ii} \;=\; H_{e,i} \;=\; \text{Nei's gene diversity of accession } i.$$

A landrace accession regenerated twenty times with forty ears per cycle
accumulates coancestry and watches $f_{ii}$ rise silently. It is the cheapest
alarm for genetic erosion *inside* a genebank -- and it is precisely what a
distance tree cannot see.

### $s_i$ and $F_i$, self-coancestry and molecular inbreeding

$$s_i = \sum_k \sum_l \frac{p_{k,l,i}^2}{N_i}, \qquad
  F_i = 2 s_i - 1, \qquad 1 - s_i = H_{o,i}$$

with $p_{k,l,i}$ the dosage of allele $k$ in individual $l$.

::: {.callout-important}
## $F_i$ here is molecular, and is not $F_{IS}$

It is measured against a hypothetical base in which all alleles are distinct. It
is **not** comparable across studies using different marker panels. For the
question "is this population mating at random?", use $\alpha_i$ below.
:::

In soybean, rice and wheat, pure lines have $s_i \to 1$, so $F_i \to 1$ and
$H_o \to 0$. That is expected, not a symptom. It changes only *where* the
diversity resides.

## Two per-subpopulation metrics

$$\alpha_i = \frac{F_i - f_{ii}}{1 - f_{ii}}, \qquad
  G_i = \frac{s_i - f_{ii}}{1 - f_{ii}} = \frac{1 + \alpha_i}{2}$$

$\alpha_i$ is the subpopulation's analogue of $F_{IS}$: positive means an excess
of homozygotes (selfing, mating among relatives, a Wahlund effect from internal
substructure, null alleles); negative means an excess of heterozygotes
(dioecy, self-incompatibility, selection, or a freshly formed F1 or synthetic).

$G_i$ is the proportion of the accession's diversity that sits **between**
individuals rather than inside them, and it is probably the most underused
quantity in the paper. It is a mating-system meter:

```{r}
#| echo: false
knitr::kable(data.frame(
  `G_i` = c("~ 0.5", "< 0.5", "-> 1.0"),
  Situation = c("perfect panmixia", "excess of heterozygotes", "all homozygous"),
  `Typical material` = c("maize, sunflower, open populations",
                         "F1 synthetics, heterozygous clones, self-incompatibles",
                         "inbred lines, doubled haploids, selfing species"),
  check.names = FALSE),
  caption = "G_i as a mating-system meter.")
```

Operationally, $1 - G_i$ is the fraction of an accession's diversity stored
*inside* each plant, and it decides the sampling design. With $G_i \to 1$
(selfers), one individual captures almost nothing and you need many plants per
accession. With $G_i \approx 0.5$ (maize), a few plants already capture half.

## Distances, and what they actually are

Nei's minimum distance between subpopulations is

$$\mathbf{D}_{ij} = \frac{f_{ii} + f_{jj}}{2} - f_{ij}
                  = \sum_k \tfrac{1}{2}(\tilde p_{k,i} - \tilde p_{k,j})^2 .$$

Note what that is: **half the sum of squared allele-frequency differences** -- a
variance component, not an evolutionary distance calibrated in time. It is the
piece that enters $F_{ST}$, and it is legitimate. The error in a
distance-only method is not using $\mathbf{D}_{ij}$; it is using *only*
$\mathbf{D}_{ij}$.

::: {.callout-tip}
## A free $O(N^2)$ saving

The paper's Eq. 13 computes $D_{ij}$ by a double sum over all pairs of
individuals, at cost $O(N_i N_j m)$. That is unnecessary: the identity
$D_{ij} = (s_i + s_j)/2 - f_{ij}$ is **exact** and reduces the cost to $O(Nm)$.
For a 50k-SNP panel and 5000 accessions this is the difference between hours and
seconds.
:::

## The equation that matters most in practice

$$\bar{f} = \sum_{i} \frac{N_i}{N_T}
  \Big[\, \underbrace{f_{ii}}_{\text{internal coancestry}}
        - \underbrace{\textstyle\sum_j \mathbf{D}_{ij} N_j / N_T}_{\text{mean distance to the others}} \Big]$$

Each accession's contribution to global coancestry -- that is, to the *loss* of
diversity in the pool -- has **two terms of opposite sign**:

- **Term 1 (+).** The more internally fixed an accession is, the more it
  *raises* global coancestry: it dilutes the pool with copies of the same
  allele.
- **Term 2 (−).** The more distant it is from the others, the more it *reduces*
  global coancestry.

A distance-only method sees term 2 alone. That is the whole mechanism of the
inverted conclusions.

## F statistics and the partition

$$F_{IS} = \frac{\tilde F - \tilde f}{1 - \tilde f}, \qquad
  F_{ST} = \frac{\tilde f - \bar f}{1 - \bar f} = \frac{\bar D}{1 - \bar f}, \qquad
  F_{IT} = \frac{\tilde F - \bar f}{1 - \bar f}$$

with $(1 - F_{IT}) = (1 - F_{IS})(1 - F_{ST})$.

The partition itself is the heart of the paper:

$$(1 - \bar f) = \underbrace{(1 - \tilde s)}_{GD_{WI}}
              + \underbrace{(\tilde s - \tilde f)}_{GD_{BI}}
              + \underbrace{(\tilde f - \bar f)}_{GD_{BS}}$$

```{r}
#| echo: false
knitr::kable(data.frame(
  Component = c("GD_T = 1 - f_bar", "GD_WI = 1 - s_til", "GD_BI = s_til - f_til",
                "GD_WS = 1 - f_til", "GD_BS = f_til - f_bar"),
  Definition = c("total diversity", "within individuals",
                 "between individuals, within accession",
                 "within subpopulations", "between subpopulations"),
  Biologically = c("He of the whole pool if everything were recombined",
                   "heterozygosity carried inside each plant",
                   "segregating variation available to within-population selection",
                   "GD_WI + GD_BI",
                   "divergence; the only component a distance method sees"),
  check.names = FALSE),
  caption = "The partition of gene diversity (Caballero & Toro 2002, Eq. 7).")
```

with the clean identities

$$\frac{GD_{WI}}{GD_T} = (1-G)(1-F_{ST}), \quad
  \frac{GD_{BI}}{GD_T} = G(1-F_{ST}), \quad
  \frac{GD_{WS}}{GD_T} = 1-F_{ST}, \quad
  \frac{GD_{BS}}{GD_T} = F_{ST}.$$

### Why the partition changes the strategy by species

- **Outcrossers** (maize, sunflower, forages, forest trees): $GD_{WI}$ is large.
  The diversity is literally *inside* the plants. Conserving means maintaining
  heterozygosity -- regenerate with many individuals and balanced contributions.
- **Selfers** (soybean, wheat, rice, beans) **and DH or line collections**:
  $GD_{WI} \approx 0$. All diversity is $GD_{BI} + GD_{BS}$. Conserving means
  maintaining the *number of distinct lines*; heterozygosity is not a target.
  Regenerating with few plants per line is acceptable; losing lines is not.
- **Clonal crops** (potato, sugarcane, fruit trees): $GD_{WI}$ is high and
  **frozen** -- it does not decay by drift while the clone exists. The $N_e$
  algebra does not apply without adaptation.

This is the reason a hybrid maize programme cannot manage diversity with a
single pool-wide number: it needs $GD_{WS}$ defended and $GD_{BS}$ held, and
those two pull in opposite directions under selection.

## The worked example, reproduced

The paper's metapopulation: four subpopulations of $N$ = 12, 4, 8, 6, three loci
with 5, 3 and 3 alleles. Its published coancestry matrix:

```{r}
f <- matrix(c(0.3391, 0.3021, 0.2535, 0.3229,
              0.3021, 0.8750, 0.3073, 0.3889,
              0.2535, 0.3073, 0.4766, 0.2483,
              0.3229, 0.3889, 0.2483, 0.3704), 4, 4, byrow = TRUE,
            dimnames = list(paste0("Sub", 1:4), paste0("Sub", 1:4)))
N <- c(12, 4, 8, 6)
w <- N / sum(N)
```

Note subpopulation 2: $f_{22} = 0.875$, so $H_e = 0.125$. It is nearly fixed
internally. Now the global quantities, with the paper's published values beside
them:

```{r}
f_bar <- drop(w %*% f %*% w)
f_til <- sum(w * diag(f))
s_til <- 0.7056                       # published; needs individual genotypes

data.frame(
  quantity = c("f_bar", "f_til", "GD_T", "GD_WI", "GD_BI", "GD_BS", "F_ST"),
  computed = round(c(f_bar, f_til, 1 - f_bar, 1 - s_til, s_til - f_til,
                     f_til - f_bar, (f_til - f_bar) / (1 - f_bar)), 4),
  published = c(0.3256, 0.4535, 0.6744, 0.2944, 0.2521, 0.1279, 0.1897)
)
```

Every value reproduces exactly. Note that $GD_{BS}$ is only 19% of the total:
the between-subpopulation divergence -- the only thing a distance method
measures -- accounts for less than a fifth of the diversity in this
metapopulation.

### Removing an accession can *increase* diversity

```{r}
res <- list(f = f, by_subpop = data.frame(N = N))
cbind(ct_loss_gain(res),
      delta_pct_equal = round(ct_loss_gain(res, "equal")$delta_pct, 2))
```

The number to look at is subpopulation 2. A Weitzman-style analysis calls it the
most valuable in the set -- losing it would cost 65.5% of the diversity, because
it is the most divergent. The coancestry analysis says removing it **raises**
pool diversity by 1.8% (census weights) or 6.5% (equal weights).

It is distant *because* it is fixed. Alleles fixed in a sample of four
individuals are not a genetic resource; they are the result of drift. The same
pattern appears in the paper's reanalysis of eleven European pig breeds: the
breed a distance method nominates as most valuable turns out to raise diversity
by 0.67% when removed.

::: {.callout-caution}
## What this does and does not license

For a genebank: a rare accession with few plants and high homozygosity -- the
archetypal "unique" accession on a distance tree -- may be contributing
*negatively* to the pool. That does **not** mean discard it. It means the
correct priority may be to **regenerate it at larger $N$**, not to replicate it
into a core collection. Storing seed is cheap; extinction is irreversible.
:::

### Optimal contributions

Maximising $GD_T = 1 - \sum_{i,j} f_{ij} c_i c_j$ subject to $c_i \ge 0$ and
$\sum_i c_i = 1$ is a quadratic programme:

```{r}
oc <- optimal_contributions(f)
oc$c_i
c(GD_before = 1 - f_bar, GD_after = oc$GD_max,
  gain_pct = 100 * (oc$GD_max - (1 - f_bar)) / (1 - f_bar))
```

Subpopulation 2 receives weight **zero**, and total diversity rises from 0.6744
to 0.6874. This reproduces the paper's published solution
$\mathbf{c} = (0.522,\, 0,\, 0.270,\, 0.208)$ exactly.

Three direct uses in plant breeding:

- **Base population for recurrent selection.** $c_i$ gives the seed proportion
  from each source that maximises long-term additive variance.
- **Core collection.** Instead of clustering heuristics plus proportional
  sampling, solve the quadratic programme directly with operational constraints
  ($c_i \le$ multiplication capacity, $c_i \ge$ a policy minimum for accessions
  of cultural importance).
- **Optimal contribution selection.** Replace the objective with
  $\max \mathbf{c}'\mathbf{g} - \lambda\,\mathbf{c}'\mathbf{F}\mathbf{c}$ and you
  have Meuwissen's framework, now standard in genomic programmes. This chapter is
  the case $\lambda \to \infty$, pure diversity. A breeder operates in the middle
  of the curve -- which is @sec-ocs.

## Minimising coancestry is maximising $N_e$

Since $N_s = 1/(2\bar f)$, minimising group coancestry and maximising effective
size are the same operation. The paper's Eq. 14 makes the operational
consequence explicit. With **equal** contributions ($S_W^2 = S_B^2 = 0$),
$N_e \approx 2Nn/(1 - F_{IT})$ -- roughly **double** the $N_e$ obtained under
random (Poisson) contributions.

> Harvesting the same number of seeds from every plant, and the same number of
> plants from every family, doubles the effective size. No genotyping, no
> technology, no cost.

It is probably the best cost-benefit intervention available in germplasm
management. The opposite -- bulk harvest from a plot, which in practice samples
the most productive plants disproportionately -- inflates $S_W^2$ and collapses
$N_e$.

## The honest caveats

The paper makes these, and a breeder should take them seriously.

Criteria based on within-population variation systematically **favour larger
populations**. A $\bar f$ criterion can push a genebank toward abundant elite
material -- the opposite of the conservation objective. Beyond that:

- Neutral markers do not measure adaptive value, disease resistance, stress
  adaptation or cultural value. A fixed accession may carry the only resistance
  allele to an emerging rust.
- **Allelic richness** is more sensitive to bottlenecks than heterozygosity and
  is often the more relevant criterion for pre-breeding. The argument that
  minimising $\bar f$ maximises allelic richness in the long run (via the
  effective number of alleles being the inverse of mean coancestry) is valid
  asymptotically but does **not** protect rare alleles in the short term. This
  is why allelic richness survives the redundancy analysis in @sec-choosing as
  the one genuinely independent axis.
- Weighting between-group variation more heavily is arbitrary. A more defensible
  fix is to optimise $\mathbf{c}$ with per-accession minimum constraints.

**Rule of thumb:** use $\bar f$ to decide *proportions and regeneration effort*;
never use $\bar f$ alone to decide *discard*.

## Two errata, found while verifying

Worth recording, since the numbers above were checked against the published
tables.

- Table 2b lists $F_2 = 0.8383$, but $F_i = 2s_i - 1 = 2(0.9167) - 1 = 0.8333$.
  A typographical error. Every other value in Tables 2 and 3 reproduces exactly.
- Conceptually, the Discussion states that equal contributions
  ($S_B^2 = S_W^2 = 0$) are required "to minimise effective size" and "maximise
  mean coancestry". That is inverted: Eq. 15 and the entire argument of the
  paper show that equal contributions **maximise** $N_e$ and **minimise**
  $\bar f$.

## What manages this quantity, and where

This chapter is the book's keystone, and it is easy to read it as a
self-contained treatment of Caballero & Toro. The partition it derives is not
background: it is the object every later chapter operates on. Stated once, so
the rest of the book can refer back rather than re-introduce it:

| Term | What it is | Where it is managed |
|---|---|---|
| $GD_{WI}$ | diversity within individuals | exactly zero with inbred parents (@sec-catalogue); the hybrid-level $GD_{WI\text{-}hyb}$ is what carries signal |
| $GD_{BI}$ | between individuals, within pool | the quantity $\alpha$ budgets (@sec-ocs), and the one recurrent selection erodes (@sec-cycles) |
| $GD_{BS}$ | between pools | the heterosis engine, to be **preserved** not minimised -- derived as an exact identity in @sec-heterosis |

Three consequences follow, and each is a chapter.

**A single pool-wide statistic cannot express "conserve within, hold between".**
That is why @sec-catalogue reports the partition rather than only $GD_T$, and why
@sec-stages makes S1 a separate monitoring stage instead of folding pool
conservation into the hybrid-selection constraint.

**The between term is not a bonus to be maximised.** @sec-heterosis-flat shows
that under neutral drift every unit of $GD_{BS}$ gained is paid for out of the
shared term, so pushing the pools apart buys nothing. The partition tells you
where diversity *is*, not which direction to move it.

**The within term is the one with a future.** $GD_{BI}$ is the raw material of
next cycle's response, which @sec-response makes precise: it is the same sum over
the same loci as genic variance, weighted equally instead of by squared effect.
That is the entire reason a diversity constraint is worth writing down, and
@sec-limits is where the equivalence is pushed until it breaks.
