# Response to selection: the other half of the objective {#sec-response}

```{r}
#| include: false
source("_common.R")
```

Every chapter so far measures what selection *spends*. None of them writes down
what it *buys*.

That asymmetry is not cosmetic. @sec-cycles will show a constrained programme
overtaking an unconstrained one after eight cycles, and the book can report that
number because it simulated it -- not because anything in the preceding chapters
predicts it. A reader who wants to know whether the crossover arrives at cycle 8
in *their* programme has no way to find out except to re-run the simulation with
their own parameters.

The quantitative genetics that answers such questions is older than any of the
genomic machinery here, and this chapter installs it: the variance selection
acts on, the equation that turns that variance into gain, and the classical
index theory that turns out to be the direct ancestor of this book's $\alpha$
constraint.

```{r}
sim <- simulate_pools(n_A = book_cfg$n_pool_A, n_B = book_cfg$n_pool_B,
                      m = book_cfg$m, fst = 0.15, n_qtl = 300, h2 = 0.5, seed = 1)
X  <- sim$X                       # N x m hybrids, 0/0.5/1
gv <- drop(X %*% sim$beta)        # true genetic value, known here by construction
I  <- sim$index                   # what the breeder actually sees
c(hybrids = nrow(X), markers = ncol(X), qtl = sum(sim$beta != 0))
```

## The breeder's equation, and how well it does here

Response to one round of truncation selection is

$$R \;=\; i \, r \, \sigma_A$$

with $i$ the selection intensity (a property of the selected fraction alone),
$r$ the accuracy of the criterion as a predictor of genetic value, and
$\sigma_A$ the additive standard deviation. `sel_intensity()` has been in this
package since @sec-stages; it has never once been used to compute a response.

```{r}
n_sel <- 60
N     <- nrow(X)
i_norm <- sel_intensity(n_sel / N)
r      <- cor(I, gv)
sel    <- order(I, decreasing = TRUE)[seq_len(n_sel)]

fmt(data.frame(i = i_norm, r = r, sigma_A = sd(gv),
               predicted = i_norm * r * sd(gv),
               realised  = mean(gv[sel]) - mean(gv)), 3)
```

The equation overpredicts, by about `r round(100 * (1 - (mean(gv[sel]) - mean(gv)) / (i_norm * r * sd(gv))))` percent. That is worth taking apart rather than
rounding away, because both halves of the discrepancy are instructive.

```{r}
i_real <- (mean(I[sel]) - mean(I)) / sd(I)
b      <- cov(gv, I) / var(I)                 # regression of value on criterion
fmt(data.frame(
  i_normal_theory = i_norm,
  i_realised      = i_real,
  R_with_realised_i = i_real * r * sd(gv),
  R_from_regression = b * (mean(I[sel]) - mean(I)),
  R_actual          = mean(gv[sel]) - mean(gv),
  index_skew        = mean(scale(I)^3)), 3)
```

Two separate failures, and they are different in kind.

**The realised intensity is below normal theory.** $i = \varphi(z)/q$ is exact
for a normal criterion. This index is slightly left-skewed, so its upper tail is
thinner than a normal's and the top `r n_sel` individuals sit closer to the mean
than the formula assumes. This is a property of *this* population, not an error.

**The regression is exact, and still wrong.** `R_from_regression` and
`R_with_realised_i` agree to machine precision -- they are the same quantity
written twice, since $b\,S_I = i\,r\,\sigma_A$ identically. Both still exceed the
realised response. What fails is not the algebra but the assumption underneath
it: $R = b\,S_I$ requires $E[g \mid I]$ to be linear in $I$ *in the selected
tail*, which is a bivariate-normality claim, and bivariate normality is exactly
what a two-pool hybrid population does not have.

::: {.callout-important}
## Read the breeder's equation as an upper bound here

Every response number in this book's cycle simulations is a *realised* response,
measured from the simulation. Where a predicted response is quoted, it will run
above the realised one for the reason above. The gap is not noise and does not
shrink with more markers; it shrinks when the criterion becomes more normal and
the population less structured.
:::

## Genic variance, and what it has to do with gene diversity

The additive variance in the equation above is not a fixed property of the
population. Split it in two.

**Genic variance** is what the loci would contribute if they were independent:

$$V_{\text{genic}} \;=\; \sum_k \beta_k^2 \operatorname{Var}(x_k)$$

**Realised variance** is what the population actually has, $\operatorname{Var}
(\mathbf{X}\boldsymbol\beta)$, which is the sum of the *whole* covariance matrix
rather than only its diagonal. The difference is exactly the gametic-phase
disequilibrium:

$$\operatorname{Var}(\mathbf{X}\boldsymbol\beta) - V_{\text{genic}}
\;=\; 2\sum_{k<l}\beta_k\beta_l \operatorname{Cov}(x_k, x_l)$$

```{r}
c(realised = var(gv),
  genic    = genic_var(X, sim$beta),
  gap      = var(gv) - genic_var(X, sim$beta),
  oracle_deviation = abs(genic_var(X, sim$beta) - ref_genic_var(X, sim$beta)))
```

Now the sentence that ties this chapter to every one that follows.

$V_{\text{genic}}$ is $\sum_k \beta_k^2\,p_k q_k$ up to a coding constant, and
gene diversity is $\frac{1}{m}\sum_k 2 p_k q_k$. **They are the same sum over the
same loci, weighted differently** -- one by squared effect, one by nothing at
all. That is the precise sense in which managing gene diversity manages the raw
material of response, and it is why a diversity constraint is a defensible proxy
for a variance constraint in the first place.

It is also the precise sense in which the proxy is incomplete. Gene diversity
weights every locus equally, so it cannot distinguish a marker carrying half the
genetic variance from one carrying none. And -- more seriously --

::: {.callout-warning}
## No metric in this book can see the disequilibrium term

$\theta$, $GD$, $N_s$, allelic richness, `F_hom` and `F_drift` are all functions
of allele frequencies **one locus at a time**. The covariance term above does not
move a single allele frequency. A flat gene diversity trajectory therefore does
not imply flat additive variance, and @sec-limits measures how far apart the two
can travel.
:::

## The decomposition this book already had

Additive variance is the variance of *breeding values*, not of genotypic values,
and in a two-pool hybrid population the average effect of an allele substitution
depends on which pool it is substituted into. That sounds like it needs new
machinery. It does not: @sec-heterosis-gca-sca already built the exact
orthogonal split, and general combining ability *is* the additive part.

```{r}
set.seed(7)
d  <- rnorm(ncol(X), 0.3, 0.1)                  # dominance deviations
U  <- sim$GL[sim$pool == "A", ]
V  <- sim$GL[sim$pool == "B", ]
gs <- heterosis_gca_sca(U, V, d)
H  <- heterosis_value(X, d)

nA <- nrow(U); nB <- nrow(V)
gA <- rep(gs$gca_A, times = nB)                 # ped varies pool A fastest
gB <- rep(gs$gca_B, each  = nA)
sc <- as.vector(gs$sca)

c(reconstruction_dev = max(abs(H - (gs$mu + gA + gB + sc))),
  var_H              = var(H),
  sum_of_components  = var(gA) + var(gB) + var(sc),
  identity_dev       = abs(var(H) - (var(gA) + var(gB) + var(sc))))
```

The three components are orthogonal, so their variances add exactly. Read the
shares:

```{r}
p <- c(gca_A = var(gA), gca_B = var(gB), sca = var(sc))
fmt(data.frame(component = names(p), variance = p,
               share = p / sum(p)), 3)
```

`gca_A` and `gca_B` are the part of heterosis an additive model can express and
that therefore behaves like additive variance under selection. `sca` is the
part it cannot, and it does not accumulate across cycles the way the general
components do. A programme selecting on an additive index is selecting on the
first two and getting the third by luck.

## What `weights = NULL` is assuming

`stage2_select()` defaults to equal weight on every trait, and every chapter of
this book including @sec-case-study takes that default. It is a choice, not a
neutral one.

The classical answer is the Smith-Hazel index (Smith 1936; Hazel 1943): the
weights maximising the correlation between the index and the aggregate genotype
$H = \mathbf{a}'\mathbf{g}$ are $\mathbf{b} = \mathbf{P}^{-1}\mathbf{G}\mathbf{a}$.

```{r}
P <- matrix(c(4.0, 1.2, 1.2, 1.0), 2)      # phenotypic covariance
G <- matrix(c(2.0, 0.9, 0.9, 0.3), 2)      # genetic covariance
a <- c(1, 1)                               # equal economic weight

fmt(data.frame(trait = c("yield", "moisture"),
               economic_weight = a,
               smith_hazel     = index_smith_hazel(P, G, a)), 3)
```

Equal economic weights do **not** give equal index weights. They coincide only
when $\mathbf{P}^{-1}\mathbf{G}$ is proportional to the identity, which requires
the traits to be equally variable, equally heritable and mutually uncorrelated.
Nothing guarantees that, and where it fails the index being maximised is not the
index anyone intended.

## The $\alpha$ constraint is a restricted index

@sec-ocs will argue that a diversity *budget* is approvable where a weighted sum
is not: "I accept losing five percent" is a sentence a breeding committee can
sign, and "diversity is worth $\lambda$ index units" is not.

That argument is usually presented as a point about human decision-making. It is
older and harder than that. Kempthorne & Nordskog (1959) posed exactly the same
problem -- maximise aggregate gain subject to **zero expected change** in a
specified direction -- and solved it:

$$\mathbf{b} \;=\; \mathbf{P}^{-1}\left[\mathbf{G}\mathbf{a} - \mathbf{C}
(\mathbf{C}'\mathbf{P}^{-1}\mathbf{C})^{-1}\mathbf{C}'\mathbf{P}^{-1}
\mathbf{G}\mathbf{a}\right], \qquad \mathbf{C} = \mathbf{G}\mathbf{R}$$

```{r}
R  <- cbind(c(0, 1))                       # hold moisture at zero gain
b_r <- index_restricted(P, G, a, R)
b_u <- index_smith_hazel(P, G, a)

fmt(data.frame(
  index = c("Smith-Hazel", "restricted"),
  w_yield    = c(b_u[1], b_r[1]),
  w_moisture = c(b_u[2], b_r[2]),
  gain_in_restricted_trait = c(drop(crossprod(R, G %*% b_u)),
                               drop(crossprod(R, G %*% b_r)))), 4)
```

The restricted index moves the restricted trait by exactly zero -- not
approximately, not on average. That is the property distinguishing a restriction
from a penalty, and it is the property $\alpha$ inherits.

::: {.callout-note}
## What is genuinely new in this book, then

Not the idea of constraining rather than pricing: that is 1959. What is new is
the restricted direction. Kempthorne and Nordskog restricted a *trait*. This
book restricts the loss of molecular gene diversity, which is not a trait, has no
economic weight, and is measured on a matrix rather than a phenotype. The algebra
of restriction carries over; the object being restricted does not.

Stating the lineage makes @sec-ocs stronger, not weaker. A sixty-year-old
theorem stands behind the move.
:::

## Where this connects

@sec-genomic-era and @sec-catalogue measure the frequencies whose weighted sum is
the genic variance defined here, which is why a diversity constraint is a
variance constraint in disguise. @sec-genomic-prediction measures the $r$ this
chapter took as given, and Finding 5 of @sec-choosing shows what happens when it
is assumed rather than measured. @sec-ocs is where the restricted index becomes
the $\alpha$ constraint, and @sec-limits is where the disequilibrium term this
chapter isolated turns out to matter more than the equation suggests.
