# Simulating lines, pools and crosses {#sec-simulating-crosses}

```{r}
#| include: false
source("_common.R")
```

Everything downstream of this chapter consumes three objects and knows nothing
about where they came from. That is deliberate: it is the seam at which a
simulation is swapped for real data, and it is the only place that has to change
when it is.

This chapter builds those objects. It also builds the candidate pool the rest of
the book selects from, so the choices made here -- how divergent the pools are,
how the index is constructed -- determine what the later chapters have to work
with.

## What has to be produced

```{r}
#| echo: false
knitr::kable(data.frame(
  Object = c("`X`", "`f`", "`hybrids`", "`fL`"),
  Shape = c("N x m", "N x N", "N x (3 + n_traits)", "L x L, optional"),
  Content = c("hybrid marker matrix, values in 0/0.5/1",
              "molecular coancestry of the hybrids",
              "hybrid id, the two parent lines, predicted traits",
              "coancestry BETWEEN LINES"),
  `Needed for` = c("every frequency-based metric",
                   "every coancestry-based metric",
                   "the index, and the line-usage vector",
                   "only the heterotic-group decomposition"),
  check.names = FALSE),
  caption = "The stage-1 contract. Stage 2 runs without `fL`; it loses only the pool decomposition.")
```

## Simulating the lines

Two heterotic pools are generated with a Balding-Nichols model. Ancestral allele
frequencies are drawn uniform on $[0.05, 0.95]$, then each pool's frequency at
marker $k$ is drawn as

$$p_{A,k} \sim \text{Beta}\!\left(p_k \frac{1 - F_{ST}}{F_{ST}},\;
                                (1 - p_k) \frac{1 - F_{ST}}{F_{ST}}\right),$$

and independently for pool B. The Beta has mean $p_k$ and variance
$F_{ST}\, p_k (1 - p_k)$, so the parameter `fst` in the configuration *is* the
expected between-pool divergence -- it is not a proxy for it.

Genotypes are then drawn as $2 \times \text{Bernoulli}(p)$, which makes every
line fully homozygous, as an inbred line should be.

```{r}
simulate_lines
```

```{r}
L <- simulate_lines(book_cfg)
dim(L)
table(as.vector(L[1:5, 1:200]))
```

Only 0 and 2 appear: no residual heterozygotes. With real data those must be
imputed or dropped before this point, because the hybrid-genotype identity in
@sec-coancestry assumes homozygous parents.

### The divergence knob

`fst` is the single most consequential parameter in the simulation, because it
controls how much heterosis there is to find and therefore how much a diversity
constraint can cost. It is worth seeing directly:

```{r}
#| fig-cap: "Between-pool divergence sets the hybrid heterozygous fraction, and with it the floor of the coancestry diagonal."
fst_grid <- c(0.02, 0.05, 0.10, 0.15, 0.25, 0.40)
div <- vapply(fst_grid, function(fst) {
  cfg <- modifyList(book_cfg, list(fst = fst, m = 1500))
  Lx  <- simulate_lines(cfg)
  fL  <- molecular_coancestry(Lx / 2)
  A   <- seq_len(cfg$n_pool_A); B <- cfg$n_pool_A + seq_len(cfg$n_pool_B)
  # a hybrid A x B is heterozygous wherever its parents differ
  c(between_pool_f = mean(fL[A, B]),
    within_pool_f  = mean(c(fL[A, A], fL[B, B])),
    het_fraction   = 1 - mean(fL[A, B]))
}, numeric(3))

d <- data.frame(fst = fst_grid, t(div))
plot(d$fst, d$het_fraction, type = "b", pch = 19, ylim = c(0, 1),
     xlab = expression(F[ST]~"between pools"),
     ylab = "expected heterozygous fraction of the F1")
grid()
```

```{r}
fmt(d)
```

More divergent pools give more heterozygous F1 hybrids. That is the heterosis
engine, and @sec-subdivided is the reason it must be *held*, not maximised: the
same divergence that raises `GD_BS` is being paid for out of within-pool
diversity.

## From lines to hybrids

Every A x B combination is formed. With 25 lines per pool that is 625
candidates; at the production scale of 71 per pool it is 5041.

```{r}
cfg <- book_cfg
a_ids <- seq_len(cfg$n_pool_A)
b_ids <- cfg$n_pool_A + seq_len(cfg$n_pool_B)
ped <- expand.grid(a = a_ids, b = b_ids)
M   <- (L[ped$a, , drop = FALSE] + L[ped$b, , drop = FALSE]) / 2   # 0/1/2
dim(M)
head(ped, 3)
```

The factorial structure matters more than it looks. Because every line appears
in many candidate hybrids, selecting hybrids is really selecting *line usage* --
and that observation, made precise in @sec-collapse, is what makes the whole
problem tractable at production scale.

## Predicted traits

Marker effects are drawn correlated across traits, through a Cholesky factor of
a fixed correlation matrix, and the predicted values are the centered marker
matrix times those effects:

```{r}
set.seed(cfg$seed + 99)
R <- matrix(c(1, 0.3, -0.3,
              0.3, 1,  0.1,
             -0.3, 0.1, 1), 3, 3)
vr <- vanraden_G(M)
B <- matrix(rnorm(cfg$m * cfg$n_traits), cfg$m) %*% chol(R)
traits <- scale(vr$Z %*% B)
round(cor(traits), 3)
```

The realised trait correlations recover the target `R`, which is the point of
the Cholesky step: without it, a multi-trait index would be a weighted sum of
independent noise, and the trade-off it induces against diversity would be
unrealistically easy.

::: {.callout-note}
## What this simulation deliberately does not model

These predicted values are constructed from *known* marker effects, so they
carry no prediction error. Real GEBVs do. The book therefore treats `ctx$index`
as given throughout, and the reader should read every gain figure as an upper
bound on what a real programme would realise. The additive model also carries no
dominance, so it cannot express specific combining ability -- see the caveats in
@sec-case-study.
:::

## The two matrices, both built

```{r}
X <- M / 2
ctx <- build_ctx(X = X, f = molecular_coancestry(X), G = vr$G, traits = traits,
                 ped = ped, n_lines = nrow(L),
                 pool = rep(c("A", "B"), c(cfg$n_pool_A, cfg$n_pool_B)),
                 fL = molecular_coancestry(L / 2), cfg = cfg)

c(N = ctx$N, m = ctx$m, n_lines = ctx$n_lines,
  sum_G = sum(ctx$G), min_f = min(ctx$f), max_f = max(ctx$f))
```

`sum(G)` is zero to machine precision and `f` is inside $[0, 1]$: the two
properties from @sec-coancestry, on this dataset.

The context also freezes two things that must never be recomputed per scenario:

```{r}
str(ctx[c("p0", "maf_pop", "index")], max.level = 1)
```

`p0` is the candidate-pool allele frequency vector. It is the **base** against
which `F_hom` and `F_drift` are measured in @sec-genomic-era, and recomputing it
per scenario would silently zero out the historical memory of erosion -- the
single most common error in applying these metrics.

## The stage-1 contract in one call

All of the above is `stage1_simulate()`:

```{r}
stage1_build
```

```{r}
st1 <- stage1_simulate(book_cfg)
str(st1, max.level = 1)
head(st1$hybrids, 3)
```

For real data the same function is called with your own matrices:

```{r}
#| eval: false
st1 <- stage1_build(
  X      = hybrid_markers,   # N x m, coded 0/1/2 or 0/0.5/1 (auto-detected)
  ped    = pedigree,         # data.frame with columns a, b
  traits = predicted_traits, # N x n_traits
  fL     = line_coancestry   # optional
)
```

`stage1_build()` detects the marker coding and rescales, so both conventions
give an identical result:

```{r}
tr <- st1$hybrids[, grep("^trait", names(st1$hybrids))]
pd <- data.frame(a = st1$hybrids$line_A, b = st1$hybrids$line_B)
identical(stage1_build(st1$X, pd, tr)$f,
          stage1_build(st1$X * 2, pd, tr)$f)
```

## What the candidate pool looks like

```{r}
#| fig-cap: "The candidate pool. Left: the index the selection maximises. Right: gene diversity of random subsets of 60, the baseline every later chapter measures against."
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.5, 1))
hist(ctx$index, breaks = 40, col = "grey85", border = "white",
     main = "multi-trait index", xlab = "index (s.d. units)")
gd_random <- replicate(400, gene_diversity(sample.int(ctx$N, 60), ctx))
hist(gd_random, breaks = 30, col = "grey85", border = "white",
     main = "GD of random subsets of 60", xlab = "gene diversity")
abline(v = gene_diversity(seq_len(ctx$N), ctx), col = "#b2182b", lwd = 2)
par(op)
```

The red line is the gene diversity of the *whole* pool. It does not sit at the
centre of the distribution of random subsets of 60 -- it sits above it. That gap
is pure sampling, it has nothing to do with selection, and mistaking it for a
loss is the subject of @sec-baseline.

```{r}
c(GD_population = gene_diversity(seq_len(ctx$N), ctx),
  GD_random_60  = mean(gd_random),
  apparent_loss_pct = 100 * (gene_diversity(seq_len(ctx$N), ctx) - mean(gd_random)) /
                            gene_diversity(seq_len(ctx$N), ctx))
```
