# Sizing a segregating population, F1 to F4 {#sec-sizing}

```{r}
#| include: false
source("_common.R")
f2 <- read.csv(book_data("f2_results.csv"))
ml <- read.csv(book_data("f2_multilocus.csv"))
```

The previous chapters took the candidate pool as given. This one asks where it
comes from: a wide cross between two divergent parents, carried F1 -> F2 -> F3
-> F4, and the question every programme has to answer before it plants anything
-- **how many plants?**

The answer turns out to depend entirely on what you are sizing *for*. There is
no single optimal $N$, and the most common failure is to size for the objective
that saturates earliest and then be surprised when a different objective was
never met.

## The simulated space is ancestry, not alleles

Each genome position is coded 0 (from parent P1) or 1 (from parent P2). For a
wide biparental cross between two divergent heterotic groups this is the correct
abstraction: essentially every locus polymorphic between the parents is
biallelic, so ancestry is a sufficient statistic. Throughout this chapter,
"allele frequency" means "frequency of P1 ancestry".

## The assumptions, one at a time

These are the assumptions the numbers rest on. Read them before trusting any of
the numbers.

**P1. Ancestry, not alleles.** As above.

**P2. Parents fully homozygous and fixed for alternative alleles.** Consequently
every F1 plant is genetically identical and the genetic $N_e$ of the F1 is 1.
Sizing the F1 is a **logistical** decision, not a genetic one -- unless the
parents carry residual heterozygosity, which is handled separately below.

**P3. The genetic map.** Ten chromosomes, lengths from a classical composite
maize map, about 1670 cM (~17 Morgans). Do **not** use IBM-type map lengths from
intermated RILs: they are inflated 5-10x by construction and will make every
answer wrong.

```{r}
map <- build_map(maize_chr_len, spacing_cM = 2)
c(chromosomes = length(map$chr_len), markers = map$m,
  total_cM = map$total_cM, morgans = map$total_cM / 100)
```

**P4. Haldane recombination, no interference,** identical in male and female.
The probability of a haplotype switch between adjacent markers $d$ cM apart is
$r = 0.5(1 - e^{-2d/100})$; the first marker of each chromosome gets $r = 0.5$,
which produces independent segregation between chromosomes automatically.

*Limitation:* without interference the variance in crossover number is inflated
(Poisson). Maize has strong positive interference, so the simulated block-length
distribution has a slightly longer tail than reality. Mean block lengths and
allele frequencies are unaffected.

**P5. Segregation distortion by rejection sampling,** therefore exact rather
than approximate. Two mechanisms: *gametophytic* (ga1/Ga1-s type, pollen
incompatibility) acting only on the male gamete, and *zygotic* (hybrid lethality
or subvitality) as genotype fitnesses. Because rejection acts on the **whole
gamete**, the region linked to the distorter is dragged along -- which is
exactly the phenomenon of interest.

```{r}
f2_distorters
```

::: {.callout-warning}
## The distorters are plausible, not measured

This is the parameter that moves the results most. Replace `f2_distorters` with
your own F2 genotyping data before using any number here to decide field area.
:::

**P6. Five schemes,** all ending with $H = 0.125$ so they are directly
comparable: `ssd` (one seed per plant per generation -- the neutral reference,
no selection, no drift beyond the meiotic), `ssd_attr` (SSD with 15% random loss
per generation: vigour, partial sterility, flowering asynchrony -- common in wide
crosses with exotic material), `bulk` (differential seed contribution, a
Wright-Fisher model with weights $w_i = e^{\beta(z_i - \bar z)}$ where $z_i$ is
the share of genome from the adapted parent -- a *phenomenological* model of
polygenic natural selection, not a model of adaptation loci), and `syn1`/`syn2`
(one or two random intercrossing cycles in the F2 before the two selfings).

**P7. The F4 panel is not a RIL panel.** Residual heterozygosity is
$H = 0.125$, and that changes every theoretical expectation:

```{r}
#| echo: false
H <- 0.125
knitr::kable(data.frame(
  Quantity = c("Var(dosage per line)", "SE(p_hat)", "Diversity retained D",
               "P(P2 homozygote per locus)", "Junctions per haplotype"),
  Formula = c("(1 - H) / 4", "0.5 sqrt((1 - H) / N)", "1 - (1 - H) / N",
              "(1 - H) / 2", "A_2 = L; A_(t+1) = A_t + L H_t"),
  `Value in F4` = c(sprintf("%.5f", (1 - H) / 4),
                    sprintf("%.3f / sqrt(N)", 0.5 * sqrt(1 - H)),
                    sprintf("1 - %.3f / N", 1 - H),
                    sprintf("%.4f", (1 - H) / 2),
                    "1.75 L"),
  check.names = FALSE),
  caption = "Theoretical identities for an F4 panel. The commonly quoted 0.5/sqrt(N) holds only at F-infinity.")
```

**P8. The standard error is estimated between replicates,** not between markers.
This is essential: linked markers are strongly autocorrelated, so the spread
across markers within one replicate is not an estimator of sampling error.

**P9. What is not modelled,** and therefore cannot be answered here: deliberate
artificial selection, QTL architecture, phenotypic inbreeding depression,
mutation (irrelevant on this timescale), and structural variation. If you know
there is a large inversion between your parents, zero the recombination rate
over that interval in `build_map()`.

*Known bias:* at 2 cM spacing, double crossovers within an interval are not
counted. The bias is $-2.0\%$ in junction number, always downward. Use
`spacing_cM = 1` if you need a faithful count.

## Verifying the identities live

The theory above is not decoration -- it is checkable, and the simulator is only
trustworthy if it reproduces it.

```{r}
map10 <- build_map(maize_chr_len, spacing_cM = 10)
neutral <- make_distortion(map10, NULL, active = FALSE)

set.seed(4)
F1 <- make_F1(map10, 400)
F2 <- cross_indices(F1, 1:400, 1:400, map10, neutral)
F3 <- cross_indices(F2, 1:400, 1:400, map10, neutral)
F4 <- cross_indices(F3, 1:400, 1:400, map10, neutral)

data.frame(
  generation = c("F1", "F2", "F3", "F4"),
  H_observed = round(c(panel_het(F1), panel_het(F2), panel_het(F3), panel_het(F4)), 4),
  H_expected = c(1, 0.5, 0.25, 0.125)
)
```

Heterozygosity halves at every selfing, as it must. And the diversity identity:

```{r}
set.seed(5)
check <- do.call(rbind, lapply(c(20, 50, 100, 200), function(N) {
  pop <- run_scheme("ssd", N, map10, neutral, f2_opt)
  H <- panel_het(pop)
  data.frame(N = N,
             D_observed = diversity_retained(panel_freq(pop)),
             D_expected = 1 - (1 - H) / N)
}))
fmt(check, 4)
```

Map expansion is an algebraic prediction, and the payoff is a result many
programmes get wrong:

```{r}
Lm <- map$total_cM / 100
rbind(SSD  = expected_junctions("ssd",  map),
      Syn1 = expected_junctions("syn1", map),
      Syn2 = expected_junctions("syn2", map))
```

Each Syn cycle adds only $0.5L$ junctions, not $1.0L$. The extra meiosis creates
a junction only where the parent is heterozygous, and $H = 0.5$ at the F2/Syn
stage. Budgeting an intercrossing cycle as "one more map length of
recombination" overstates what it buys by a factor of two.

## The results

The full grid -- five schemes, `r length(unique(f2$N))` population sizes,
adaptive replication -- takes several minutes and is shipped precomputed.

```{r}
#| echo: false
ssd <- f2[f2$scheme == "ssd", ]
sel <- ssd[ssd$N %in% c(10, 30, 50, 70, 90, 110, 150, 190, 290, 390), ]
knitr::kable(data.frame(
  N = sel$N,
  D = round(sel$div, 3),
  `SE(p_hat)` = round(sel$se_p, 3),
  `skewed genome` = sprintf("%.1f%%", 100 * sel$skew),
  `min(p, 1-p)` = round(sel$pmin, 3),
  `N equivalent` = round(sel$n_eq),
  check.names = FALSE),
  caption = "SSD, the neutral reference. Everything below comes from this table.")
```

### The diversity curve saturates at 0.988, not at 1

```{r}
#| fig-cap: "Diversity retained against population size, five schemes. The theoretical curve is 1 - (1 - H)/N under no distortion."
schemes <- unique(f2$scheme)
cols <- c(ssd = "#1b6ca8", ssd_attr = "#7fb3d5", bulk = "#c0392b",
          syn1 = "#27ae60", syn2 = "#16a085")
plot(NA, xlim = range(f2$N), ylim = c(0.88, 1.0),
     xlab = expression(N[F2]~"plants"), ylab = "diversity retained, D")
grid()
for (s in schemes) {
  d <- f2[f2$scheme == s, ]
  lines(d$N, d$div, type = "b", pch = 19, cex = 0.6, col = cols[s], lwd = 1.8)
}
curve(1 - 0.875 / x, add = TRUE, lty = 2, col = "grey30", lwd = 1.5)
legend("bottomright", c(schemes, "theory, no distortion"),
       col = c(cols[schemes], "grey30"), lwd = c(rep(1.8, length(schemes)), 1.5),
       lty = c(rep(1, length(schemes)), 2), pch = c(rep(19, length(schemes)), NA),
       bty = "n", cex = 0.8)
```

The gap between the observed curve and the theoretical one is **segregation
distortion** -- a deterministic bias that does not shrink with more plants.
Adding plants buys you drift protection; it does not buy you distortion
protection.

### The bulk is the worst scheme across the whole range

```{r}
#| echo: false
b400 <- f2$div[f2$scheme == "bulk"  & f2$N == max(f2$N)]
s40  <- f2$div[f2$scheme == "ssd"   & f2$N == 30]
data.frame(comparison = c("bulk, largest N", "SSD, N = 30"),
           N = c(max(f2$N), 30),
           D = round(c(b400, s40), 3))
```

Bulk at the largest population size retains **less** diversity than SSD at 30
plants. With a less adapted donor parent, natural selection in the bulk removes
precisely the germplasm the wide cross was made to introduce. If the objective is
introgression, the bulk is not a cheaper SSD -- it is a different experiment.

### Block size does not depend on N

```{r}
#| fig-cap: "Ancestry frequency along the ten chromosomes, at N = 20 and N = 300. At N = 20 the whole curve is noisy: that is drift, and more plants remove it. At N = 300 the noise is gone and three peaks remain, exactly at the distorters. Those are the deterministic bias, and no number of plants removes them."
map4 <- build_map(maize_chr_len, spacing_cM = 4)
dist <- make_distortion(map4, f2_distorters)
set.seed(6)
small <- panel_freq(run_scheme("ssd", 20,  map4, dist, f2_opt))
large <- panel_freq(run_scheme("ssd", 300, map4, dist, f2_opt))

plot(small, type = "l", col = "#e08214", ylim = c(0, 1),
     xlab = "genome (chromosomes 1-10)", ylab = "P1 ancestry frequency", xaxt = "n")
lines(large, col = "#1b6ca8", lwd = 1.6)
abline(h = 0.5, lty = 2, col = "grey50")
# chromosome boundaries and distorter positions
bnd <- which(map4$first)
axis(1, at = (c(bnd, map4$m + 1)[-1] + bnd) / 2, labels = seq_along(map4$chr_len))
abline(v = bnd[-1] - 0.5, col = "grey85")
for (i in seq_len(nrow(f2_distorters)))
  abline(v = marker_index(map4, f2_distorters$chr[i], f2_distorters$pos_cM[i]),
         col = "#b2182b", lty = 3)
legend("topleft", c("N = 20", "N = 300", "distorter"), col = c("#e08214", "#1b6ca8", "#b2182b"),
       lty = c(1, 1, 3), lwd = c(1, 1.6, 1), bty = "n", cex = 0.8, horiz = TRUE)
```

Block length depends only on the number of meioses. Against linkage drag the
instrument is an **intercrossing cycle, not field area** -- which is what the
Syn curves in the theory table above quantify.

### The metrics saturate at different N

This is the central practical result, and the reason the question "what is the
optimal $N$?" has no single answer.

```{r}
#| echo: false
knitr::kable(data.frame(
  Objective = c("Diversity retained (D)",
                "Worst region of the genome, min(p, 1-p)",
                "Precision of the panel's p_hat",
                "Recovery of 5 loci",
                "Recovery of 8-10 loci"),
  `Saturates near` = c("100-120", "60-80",
                       "**never** -- falls as 1/sqrt(N)",
                       "200-260",
                       "**not saturated** at the largest N tested"),
  check.names = FALSE),
  caption = "Different objectives, different answers. Size for the one that binds.")
```

```{r}
#| fig-cap: "Multilocus recovery: P(at least one line homozygous for P2 at k loci). This is the objective that keeps paying for larger N."
kk <- sort(unique(ml$k))
plot(NA, xlim = range(ml$N), ylim = c(0, 1), xlab = expression(N[F2]),
     ylab = "P(at least one line)")
grid()
for (i in seq_along(kk)) {
  d <- ml[ml$k == kk[i], ]
  lines(d$N, d$P, type = "b", pch = 19, cex = 0.6, col = i + 1, lwd = 1.8)
}
legend("topright", paste("k =", kk), col = seq_along(kk) + 1, lwd = 1.8,
       pch = 19, bty = "n")
```

The marginal gain in $D$ tells the same story from the other side:

```{r}
#| echo: false
d <- f2[f2$scheme == "ssd", ]
d <- d[order(d$N), ]
data.frame(
  step = paste0(utils::head(d$N, -1), " -> ", d$N[-1]),
  gain_in_D = round(diff(d$div), 4)
)[1:6, ]
```

The knee is early. Past it, twenty more plants buy thousandths.

## Sizing the F1

With fully homozygous parents the F1 is genetically uniform and the decision is
logistical. With residual heterozygosity -- $F \approx 0.95$ is common in
programme lines -- each parent still segregates at a fraction of loci, and each
F1 plant receives one random gamete from each parent. At a segregating locus the
chance of losing one of the two parental alleles with $n$ F1 plants is
$2(1/2)^n$.

```{r}
s <- f1_sizing(n_F1 = 1:20, h_res = 0.05, n_loci_eff = 5000)
s[s$n_F1 %in% c(4, 8, 10, 12, 15, 20), ]
```

Twelve selfed F1 plants already put the expected number of lost loci below one.
That is the whole F1 sizing argument.

## Working numbers

```{r}
#| echo: false
knitr::kable(data.frame(
  Generation = c("F1", "F2", "", "F3", "F4"),
  Plants = c("8-15 selfed", "**120** to retain diversity",
             "**250-400** for multilocus selection or panel frequencies",
             "same N (SSD, one seed per plant)", "lines x 3-8 plants"),
  Rationale = c(
    "Ne = 1, a logistical call. With 5% residual heterozygosity, 12 plants put expected lost loci below 1",
    "Above this, +280 plants buy +0.005 of D",
    "These are the objectives that keep gaining with N",
    "Does not create diversity, only avoids losing it",
    "Seed multiplication. Plants of the same line are not independent replicates: they share 87.5% of the genome"),
  check.names = FALSE),
  caption = "Working numbers for a wide cross. Size for the objective that binds, not for D.")
```

::: {.callout-caution}
## On the "equivalent N"

$N_{eq} = (1 - H)/(1 - D)$ answers "how many F4 lines without distortion would
give the diversity actually observed". It is a **communication device, not a
Wright effective size**: it converts a deterministic bias into a drift
equivalent. Use it to compare scenarios; never substitute it into an $N_e$
formula.
:::

## Where this connects

The panel this chapter sizes is the raw material for everything else, and the
claim that its diversity is a *ceiling* should be measured rather than asserted.
The rest of the book works on one kernel -- molecular coancestry $f$ and
$GD = 1 - \theta$ (@sec-coancestry). The F4 panel can be pushed straight into it:

```{r}
set.seed(3)
map4  <- build_map(spacing_cM = 10)
dist4 <- make_distortion(map4, active = TRUE)

bridge <- do.call(rbind, lapply(c(30, 120, 480), function(n) {
  pop <- run_scheme("SSD", n, map4, dist4, opt = f2_opt)
  X   <- t((pop$H1 + pop$H2) / 2)          # lines x markers, 0/0.5/1
  th  <- mean(molecular_coancestry(X))
  data.frame(n_F2 = n, D = diversity_retained(panel_freq(pop)),
             theta = th, GD = 1 - th)
}))
fmt(bridge, 4)
```

$D$, the chapter's own diversity-retained statistic, and $GD$, the quantity every
later chapter constrains, move together: they are the same loss in two coordinate
systems. The sizing decision made here therefore sets the $GD$ that
@sec-choosing, @sec-ocs and @sec-combinatorial spend the rest of the book
budgeting, and it sets it *once*. An optimiser cannot select back up to variation
the panel never sampled; the best it can do is spend what is left more slowly.

::: {.callout-note}
## $N_s$ is not informative inside a biparental cross

Computing the status number on these panels returns approximately 1 whatever the
sizing, and that is correct rather than broken. Every line descends from two
parents, so the panel has the genetic base of a single diploid individual, which
is exactly what $N_s = 1/(2\theta)$ reports. @sec-ns warns that $N_s$ describes a
base rather than projecting drift; here the base is genuinely one cross.

Use $D$ and $GD$ to compare sizing scenarios within a cross, and $N_s$ only once
material from several crosses has been pooled.
:::

@sec-stages puts a number on how the loss is distributed across the whole
pipeline, and this stage is where the cheapest insurance is available -- more
plants, once, at the start. @sec-limits is where the same argument is made about
the *variance* rather than the frequencies, and reaches a different conclusion
about what is recoverable.
