# Simulating a breeding programme: AlphaSimR from the ground up {#sec-alphasimr}

```{r}
#| include: false
source("_common.R")
library(AlphaSimR)
rrs_cfg <- utils::modifyList(rrs_config,
  list(n_pool_A = 12, n_pool_B = 12, n_dh = 3, n_sel = 6, n_testers = 2,
       n_cycles = 4, seg_sites = 150, n_chr = 5, n_traits = 2, seed = 1))
```

@sec-simulating-crosses built the `{X, f, hybrids, fL}` contract with a
deliberately fast simulator: Balding-Nichols allele frequencies, independent
markers, no recombination, no dominance. That is enough to exercise the
diversity and selection machinery, but it is not how a breeding programme is
actually run, and it is not how anyone would build one from scratch.

This chapter does both. Steps 1 to 7 are a **step-by-step construction of a
simulation** with AlphaSimR (@gaynor2021), written out in the primitive calls --
`runMacs()`, `SimParam`, `addTraitA()`, `makeCross()`, `makeDH()`, `setPheno()`,
`selectInd()` -- with a plant breeder's reading of each argument. Step 8 then
assembles those primitives into **reciprocal recurrent selection** (RRS;
@comstock1949), the standard commercial scheme for two heterotic pools, and
shows that the same stage-1 contract comes out the other end, unchanged, so
`stage2_select()` runs on it exactly as it does on the book's own simulated data.

The order matters. Everything in Step 8 is a wrapper in
`R/13_alphasimr_bridge.R` -- `rrs_config`, `alphasimr_founder_pools()`,
`alphasimr_rrs_cycle()`, `alphasimr_make_hybrids()`, `alphasimr_to_stage1()`,
`alphasimr_pipeline()` -- and a wrapper is only readable once you know what it
is wrapping. AlphaSimR is a `Suggests` dependency, guarded exactly like
`quadprog` and `highs` elsewhere in the package; this chapter, and
@sec-programme, are the only ones that need it.

@sec-programme takes the same primitives in the other direction: a complete
*inbred-line* programme with staged yield trials, rather than a hybrid one.

## Step 1 -- Founder genomes, and which engine to use

Every AlphaSimR simulation starts with founder haplotypes. There are two
engines, and the choice is not cosmetic.

`runMacs()` runs a backward-in-time coalescent (@chen2009) under a
species-specific demography -- `species = "MAIZE"`, `"WHEAT"`, `"CATTLE"` --
so the founders carry a realistic allele-frequency spectrum and realistic
linkage disequilibrium. `quickHaplo()` samples 0s and 1s directly: instant, and
with no history behind it.

`inbred = TRUE` is the argument a plant breeder almost always wants. It makes
every founder fully homozygous, which is what a panel of inbred lines or
released varieties actually is. Without it the founders are outbred individuals,
which is the right model for cattle and the wrong one for wheat.

```{r}
tm <- function(expr) round(as.numeric(system.time(expr)["elapsed"]), 2)

set.seed(7); t_q <- tm(qa <- quickHaplo(nInd = 20, nChr = 10, segSites = 50,
                                        inbred = TRUE))
set.seed(7);        qb <- quickHaplo(nInd = 20, nChr = 10, segSites = 50,
                                     inbred = TRUE)

set.seed(7); t_m <- tm(ma <- runMacs(nInd = 20, nChr = 10, segSites = 50,
                                     inbred = TRUE, species = "MAIZE"))
set.seed(7);        mb <- runMacs(nInd = 20, nChr = 10, segSites = 50,
                                  inbred = TRUE, species = "MAIZE")

fmt(data.frame(engine     = c("quickHaplo", "runMacs"),
               seconds    = c(t_q, t_m),
               same_twice = c(identical(qa@geno, qb@geno),
                              identical(ma@geno, mb@geno))))
```

Read the third column again. **`runMacs()` is not reproducible from
`set.seed()`.** It hands the coalescent off to the MaCS binary, which seeds
itself outside R's random number stream, so two calls with an identical seed
return different founders. `quickHaplo()` is pure R sampling and is exactly
reproducible.

(The `seconds` column is wall clock and will differ on your machine and between
renders; the ratio between the two engines is the part that transfers.)

::: {.callout-warning}
## The founders are not seeded, and everything downstream is

This is the single most expensive trap in AlphaSimR, because it fails silently:
the code looks reproducible, runs without complaint, and returns different
numbers. It is already documented in `alphasimr_founder_pools()`'s help page --
the note there says `runMacs2()`, and the table above shows it is not
`runMacs2`-specific but true of MaCS generally.

Two consequences:

1. **Found once, then branch.** Any experiment comparing two schemes must draw
   the founders a single time and start both arms from that one object. Draw
   them twice and the comparison measures the founders as much as the schemes.
   @sec-limits' linked arm is built this way for exactly this reason.
2. **Save what you cannot reseed.** `saveRDS(founderGenomes, ...)` is the only
   way to get the same coalescent history back in a later session.

Everything after the founders -- the crosses, the doubled haploids, the
phenotyping error, the recombination -- *is* controlled by `set.seed()`.
:::

So what does `quickHaplo()` give up? Linkage disequilibrium among the founders.
The honest way to measure that is against the sampling floor: with `n`
individuals, two completely independent loci still show `r^2` of about
`1 / (n - 1)` by chance alone.

```{r}
ldstat <- function(fg) {
  SPx <- SimParam$new(fg)
  g <- pullSegSiteGeno(newPop(fg, simParam = SPx), simParam = SPx)
  g <- g[, apply(g, 2, stats::sd) > 0, drop = FALSE]
  r <- stats::cor(g)
  c(adjacent = mean(diag(r[-nrow(r), -1])^2),   # neighbouring sites
    overall  = mean(r[upper.tri(r)]^2),         # all pairs, mostly distant
    noise    = 1 / (nrow(g) - 1))               # what pure chance gives
}
set.seed(3); lq <- ldstat(quickHaplo(nInd = 100, nChr = 5, segSites = 200,
                                     inbred = TRUE))
set.seed(3); lm_ <- ldstat(runMacs(nInd = 100, nChr = 5, segSites = 200,
                                   inbred = TRUE, species = "MAIZE"))
fmt(rbind(quickHaplo = lq, runMacs = lm_))
```

`quickHaplo()`'s adjacent-site `r^2` sits exactly on the noise floor: there is
no founder LD at all, as designed. `runMacs()`'s is an order of magnitude above
it, and its *overall* `r^2` is barely above the floor -- so the LD it creates is
local, decaying with map distance, which is what real LD does.

The `runMacs` row of that table is itself not reproducible, for the reason the
row above it establishes; the `quickHaplo` row is, to the last digit.

That contamination spreads further than it looks. `runMacs()` consumes an
unpredictable amount of R's own random stream on its way to and from the MaCS
binary, so *everything executed after it in the same session* inherits an
unpredictable seed -- an earlier draft of this chapter had reproducible-looking
Steps 2 to 7 that quietly moved whenever Step 1 was re-executed. The remedy is
the rule stated above, applied to the chapter itself: each of the steps below
calls `set.seed()` immediately before drawing its own founders, rather than
trusting a seed set once at the top. With that in place, two independent
executions agree on every number in Steps 1 to 7 except the wall-clock seconds
and the `runMacs` row itself.

That measurement decides the policy this chapter follows:

- **`quickHaplo(inbred = TRUE)` for Steps 2--7 and all of @sec-programme.** They
  teach polygenic response arithmetic, where founder LD is not the point, and
  in exchange every number printed is reproducible from a clean clone.
  AlphaSimR still applies a real genetic map and real meiosis to every
  descendant, so linkage and the Bulmer effect of @sec-limits still operate --
  only the founders' *history* is stylised.
- **`runMacs2(split = )` in Step 8.** There the divergence between two heterotic
  pools *is* the object of study, and no amount of speed compensates for not
  having it.

Founder LD does matter where marker-QTL association is the mechanism, which is
genomic prediction -- see @sec-genomic-prediction, and note that the accuracies
there are computed on a different simulator for this reason.

## Step 2 -- `SimParam`, and defining a trait

`SimParam` holds everything global: the genetic map, which sites are QTL, the
trait effects, the error variances. It is an R6 object, so its methods are
called with `$`, and a variable named `SP` is picked up automatically by every
AlphaSimR function that needs it.

```{r}
# Seeded here, not once at the top of the chapter: the runMacs() calls in Step 1
# consume an unpredictable amount of R's random stream, so anything downstream of
# them is only reproducible by accident unless the seed is reset. This is the
# "found once, then branch" rule of Step 1 applied to the chapter itself.
set.seed(101)
founders <- quickHaplo(nInd = 60, nChr = 10, segSites = 100, inbred = TRUE)
SP <- SimParam$new(founders)
SP$addTraitA(nQtlPerChr = 100, mean = 10, var = 1)
SP$setVarE(h2 = 0.4)
```

There are four trait constructors, and choosing between them is choosing a model
of gene action:

| Method | Adds | When a plant programme needs it |
|---|---|---|
| `SP$addTraitA()` | additive only | line per se performance; the default teaching case |
| `SP$addTraitAD()` | + dominance | testcross and hybrid value. Without dominance a line's cross performance is just its own breeding value and there is nothing for RRS to exploit (@sec-heterosis). This is what Step 8 uses |
| `SP$addTraitAG()` | + genotype-by-environment | multi-location yield trials, where the same genotype is scored in several places. @sec-programme uses it |
| `SP$addTraitADE()` | + epistasis | available; used nowhere in this book |

`nQtlPerChr` sets how polygenic the trait is. Setting it equal to `segSites`
makes *every* segregating site a QTL, which is what @sec-alphasimr's own
configuration does -- convenient, because then the marker matrix and the causal
loci are the same set and there is no marker-versus-QTL gap to reason about.
Real programmes have that gap; @sec-genomic-era is where the book takes it
seriously.

`mean` and `var` fix the trait's mean and *genetic* variance in the founder
population by rescaling the sampled effects. `corA` (Step 7) sets the genetic
correlation between traits.

The environmental variance has two routes, and plant breeders use the second:

- `SP$setVarE(h2 = 0.4)` fixes an error variance once, from the founders' genetic
  variance, and every later `setPheno()` reuses it.
- `setPheno(varE = , reps = )` sets it per call. **`reps` is the plant idiom for
  locations**: a genotype scored in `reps` environments has its error variance
  divided by `reps`. That is how a staged programme raises selection accuracy
  from one stage to the next without ever changing the trait's heritability --
  the mechanism @sec-programme is built on.

## Step 3 -- The base population, and where the variance components live

`newPop()` turns founder haplotypes into individuals and, because a trait is
now defined, computes their genetic values immediately.

```{r}
pop <- setPheno(newPop(founders, simParam = SP), simParam = SP)
c(n = pop@nInd, meanG = as.numeric(meanG(pop)), varG = as.numeric(varG(pop)),
  varP = as.numeric(varP(pop)),
  h2_realised = as.numeric(varG(pop) / varP(pop))) |> fmt(3)
```

The realised heritability is near but not equal to the requested 0.4:
`setVarE()` fixes a variance, and the ratio it produces in any finite sample is
itself a random variable. This is a general and useful habit -- **check what the
simulation realised, do not assume what you requested.**

`genParam()` returns the full decomposition, and two of its fields are the ones
this book cares about most:

```{r}
mixed <- pop
for (k in 1:3) mixed <- randCross(mixed, nCrosses = 1000, simParam = SP)

vrow <- function(p, lab) {
  g <- genParam(p, simParam = SP)
  data.frame(population = lab,
             varA      = as.numeric(g$varA),
             genicVarA = as.numeric(g$genicVarA),
             ratio     = as.numeric(g$varA / g$genicVarA))
}
fmt(rbind(vrow(pop,   "founders (fully inbred)"),
          vrow(mixed, "after 3 random matings")))
```

`varA` is the realised additive variance; `genicVarA` is what it would be if the
loci were independent and in Hardy-Weinberg proportions. Their ratio is the
whole subject of @sec-limits, and the package's `genic_var()` computes the same
quantity from a marker matrix directly.

The founders' ratio above 1 is not disequilibrium -- it is **inbreeding**. A
fully homozygous population has per-locus genotypic variance `4pq` where a
Hardy-Weinberg one has `2pq`, so `varA` runs up to twice the genic value by
construction. Three rounds of random mating restore Hardy-Weinberg and pull the
ratio back towards 1; what remains is the disequilibrium among only 60 founders'
worth of gametes. A ratio away from 1 in an *outbred* population is the
gametic-phase disequilibrium @sec-response defines and @sec-limits measures; a
ratio away from 1 in an inbred one is mostly just the homozygosity.

## Step 4 -- The DNA lottery, and why an F1 is not a family

The course this chapter's structure follows demonstrates Mendelian sampling by
crossing outbred founders and showing that full sibs vary. With **inbred**
parents the result is different, and it is the fact plant breeding is organised
around.

```{r}
lottery <- do.call(rbind, lapply(list(c(1, 2), c(1, 3), c(2, 3)), function(ij) {
  plan <- matrix(ij, ncol = 2)
  f1 <- makeCross(pop, plan, nProgeny = 200, simParam = SP)
  dh <- makeDH(makeCross(pop, plan, nProgeny = 1, simParam = SP),
               nDH = 200, simParam = SP)
  better <- max(gv(pop)[ij])
  data.frame(cross         = paste(ij, collapse = " x "),
             parent_mean   = mean(gv(pop)[ij]),
             better_parent = better,
             f1_sd         = as.numeric(sqrt(varG(f1))),
             dh_sd         = as.numeric(sqrt(varG(dh))),
             dh_best       = max(gv(dh)),
             transgressive = mean(gv(dh) > better))
}))
fmt(lottery, 3)
```

`f1_sd` is **exactly zero**. Two fully inbred parents are homozygous everywhere,
so every gamete each one makes is identical, and every F1 is the same
heterozygote. Two hundred F1 plants from a cross of two inbred lines are two
hundred copies of one genotype. There is no lottery yet.

The lottery starts one meiosis later. `dh_sd` is the standard deviation among
doubled haploids drawn from that F1, and it is substantial -- this is the
segregation that F2, DH and RIL populations exist to sample. It is also why
@sec-sizing is a chapter: the number of DH lines you pull from a cross is the
number of tickets you buy, and no downstream selection stage can recover a
combination that was never sampled.

`transgressive` is the fraction of DH lines exceeding the *better* parent. It is
not small. Crossing two mediocre parents whose favourable alleles sit at
different loci produces progeny beyond either of them -- transgressive
segregation, which @mackay2020 argues is the same phenomenon as heterosis and
inbreeding depression viewed from a third angle. @sec-heterosis is where this
book takes up that argument.

## Step 5 -- Selfing, doubled haploids, and what the shortcut buys

A plant breeder has to get from a heterozygous F1 to a homozygous line. There
are two routes, and AlphaSimR implements both.

```{r}
homozygosity <- function(p)
  mean(pullSegSiteGeno(p, simParam = SP) %in% c(0, 2))

f1 <- makeCross(pop, matrix(c(1, 2), ncol = 2), nProgeny = 200, simParam = SP)

gen <- f1
traj <- data.frame(generation = "F1", homozygosity = homozygosity(f1))
for (k in 2:7) {
  gen <- self(gen, nProgeny = 1, simParam = SP)
  traj <- rbind(traj, data.frame(generation = paste0("F", k),
                                 homozygosity = homozygosity(gen)))
}
traj <- rbind(traj, data.frame(generation = "DH",
                               homozygosity = homozygosity(makeDH(f1, nDH = 1,
                                                                  simParam = SP))))
fmt(traj)
```

`self()` halves the remaining heterozygosity every generation -- the familiar
`1 - (1/2)^(t-1)` approach to fixation, which never quite arrives. `makeDH()`
gets there in one step, at 1.0 exactly, by doubling a haploid gamete.

The saving is seasons, and seasons are the currency of a breeding programme:
six or seven selfing generations against one DH cycle. @sec-programme's scheme
uses `makeDH()` for precisely this reason, and the cost that never shows up in
the simulation is that DH production is a laboratory operation with its own
failure rate and its own genotype-dependent response.

The other operations a scheme needs are all one call each: `randCross()` for
random mating within a pool, `makeCross()` for a specified crossing plan,
`hybridCross()` for a full factorial between two groups (Step 8), and
`mergePops()` or `c()` to combine populations.

## Step 6 -- Phenotyping, selection, and checking the breeder's equation

Selection needs a criterion. `setPheno()` supplies the noisy one; `selectInd()`
acts on it.

```{r}
mixed <- setPheno(mixed, simParam = SP)

sweep_q <- do.call(rbind, lapply(c(0.05, 0.10, 0.20), function(q) {
  n_sel <- round(q * mixed@nInd)
  sel   <- selectInd(mixed, nInd = n_sel, use = "pheno", simParam = SP)
  prog  <- randCross(sel, nCrosses = 1000, simParam = SP)

  S     <- as.numeric(meanP(sel) - meanP(mixed))
  vA    <- as.numeric(varA(mixed, simParam = SP))
  h2r   <- vA / as.numeric(varP(mixed))
  i_emp <- S / sqrt(as.numeric(varP(mixed)))
  r_emp <- stats::cor(gv(mixed)[, 1], pheno(mixed)[, 1])

  data.frame(q = q,
             i_empirical = i_emp, i_normal = sel_intensity(q),
             R_S_h2      = S * h2r,
             R_i_r_sdA   = i_emp * r_emp * sqrt(vA),
             R_observed  = as.numeric(meanG(prog) - meanG(mixed)))
}))
fmt(sweep_q)
```

Both forms of the breeder's equation are here. `R = S * h2` uses the selection
differential directly; `R = i * r * sigma_A` decomposes it into intensity,
accuracy and available variation -- the form @sec-response builds the book's
objective side on. `R_observed` is the realised change in *genetic* mean between
the selected parents' progeny and the starting population, which is what the
prediction is a prediction of.

They agree to a few percent, and the agreement degrades as `q` rises, which is
the expected direction: the normal-theory intensity assumes an infinite
population and a normal criterion, and a 20% selected fraction of a finite
population departs from both.

The book exports its own selection intensity, and AlphaSimR exports one too.
They are independent implementations of `phi(z) / q`:

```{r}
q <- 0.05
c(hybdiv = sel_intensity(q), AlphaSimR = selInt(q),
  difference = abs(sel_intensity(q) - selInt(q))) |> fmt(10)
```

`selectInd()` takes `use = "pheno"`, `"gv"` or `"ebv"`. Selecting on `"gv"` is
selecting with perfect accuracy -- impossible in a real programme, and the
correct way to compute the ceiling a scheme is being measured against.
`use = "ebv"` selects on whatever has been written into the population's `ebv`
slot, which is how genomic selection enters a simulation (@sec-programme).
`selectTop = FALSE` selects downwards, for traits where less is better.

`selectWithinFam()` is the plant-specific one. A programme that truncates
globally on a set of families will take every entry from the two best crosses
and nothing from the rest; capping the number retained per family holds the
programme's genetic base open at some cost in immediate gain. It is the same
trade-off as the `max_use` parental cap of @sec-combinatorial, applied one stage
earlier.

## Step 7 -- Two traits, and what equal weighting assumes

`corA` makes the traits genetically correlated, which is where selection stops
being one-dimensional.

```{r}
set.seed(202)
f2 <- quickHaplo(nInd = 80, nChr = 10, segSites = 100, inbred = TRUE)
SP2 <- SimParam$new(f2)
SP2$addTraitA(nQtlPerChr = 100, mean = c(10, 10), var = c(1, 4),
              corA = matrix(c(1, -0.4, -0.4, 1), 2, 2))
SP2$setVarE(h2 = c(0.6, 0.1))

mt <- newPop(f2, simParam = SP2)
for (k in 1:3) mt <- randCross(mt, nCrosses = 1000, simParam = SP2)
mt <- setPheno(mt, simParam = SP2)

G <- varG(mt); P <- varP(mt); a <- c(1, 1)
c(cor_genetic = stats::cov2cor(G)[2, 1],
  cor_pheno   = stats::cov2cor(P)[2, 1],
  h2_trait1   = as.numeric(diag(G)[1] / diag(P)[1]),
  h2_trait2   = as.numeric(diag(G)[2] / diag(P)[2])) |> fmt(3)
```

The genetic correlation lands near the requested `-0.4`; the phenotypic one is
much weaker, because the environmental components were left uncorrelated
(`SP$setCorE()` is where that would change). The two traits differ in variance
and, sharply, in heritability -- the ordinary situation.

The book's pipeline defaults to `weights = NULL`, meaning equal weight on every
trait. Step 7 is where that default gets examined, using the index functions
@sec-response introduced:

```{r}
cor_with_H <- function(w)
  drop(crossprod(w, G %*% a)) /
    sqrt(drop(crossprod(w, P %*% w)) * drop(crossprod(a, G %*% a)))

B <- rbind(equal       = a,
           smith_hazel = index_smith_hazel(P, G, a),
           restricted  = index_restricted(P, G, a, R = cbind(c(0, 1))))
dG <- t(apply(B, 1, function(w) drop(G %*% w)))
colnames(dG) <- c("gain_trait1", "gain_trait2")
fmt(cbind(B, dG, cor_with_H = apply(B, 1, cor_with_H)))
```

Three readings, in the order they matter:

1. **Equal weights are not the Smith-Hazel solution.** Smith-Hazel puts 0.46
   units of weight on trait 2 for every unit on trait 1, not 1.00. Equal
   weighting is optimal only when `P^-1 G` is proportional to the identity,
   which needs the traits equally variable, equally heritable and uncorrelated.
   None of that holds here, and it rarely holds anywhere.
2. **The cost of getting it wrong is modest here, and the direction of the
   error is not.** Equal weighting loses about 2% of the attainable correlation
   with the aggregate genotype, but it drives a large unintended gain in trait 2
   simply because trait 2 has four times the variance. An index that is nearly
   as good on the objective can still be badly wrong about where the gain lands.
3. **The restricted index holds trait 2 at exactly zero.** That is
   Kempthorne and Nordskog's construction (@kempthorne1959) and, as
   @sec-response argues, the classical ancestor of this book's `alpha`
   constraint -- restrict a direction rather than price it.

Smith-Hazel is defined as the weight vector maximising that correlation, so the
claim is checkable against a numerical optimum rather than taken on faith:

```{r}
opt <- stats::optim(c(1, 1), function(w) -cor_with_H(w), method = "BFGS")
c(numerical = -opt$value, smith_hazel = cor_with_H(B["smith_hazel", ]),
  difference = abs(-opt$value - cor_with_H(B["smith_hazel", ]))) |> fmt(10)
```

Multi-trait genomic selection in wheat, balancing yield against protein content
and quality, is the standing plant example -- @michel2019.

## Step 8 -- The parts assembled: reciprocal recurrent selection

Everything above was one operation at a time. A breeding scheme is those
operations in a loop, and the loop for two heterotic pools is reciprocal
recurrent selection: each pool is improved using the *other* pool as tester.
Doubled-haploid lines are drawn from pool A, testcrossed to pool B (and vice
versa), the testcross hybrids are phenotyped, and the lines with the best mean
testcross performance -- a proxy for general combining ability, GCA -- are
recombined to found the next cycle's pool. Elite lines from both pools, at the
end, are crossed directly to produce the commercial hybrids. It is exactly the
asymmetry @sec-cycles argues for: selection happens on the hybrid, recycling
happens within pool.

### Founding two pools with a real coalescent history

This is the one place the chapter pays for `runMacs2()` rather than
`quickHaplo()`. `AlphaSimR::runMacs2(..., split = )` draws founder haplotypes
from a population-split coalescent history: one ancestral population, then two
demes diverging `split` generations ago. That is the realistic analogue of
`simulate_lines()`'s Balding-Nichols draw in @sec-simulating-crosses, and
between-pool divergence is the entire subject of @sec-subdivided, so a stylised
substitute would not do.

```{r}
fp <- alphasimr_founder_pools(rrs_cfg)
c(pool_A = fp$pop_A@nInd, pool_B = fp$pop_B@nInd,
  markers = rrs_cfg$n_chr * rrs_cfg$seg_sites)
alphasimr_gd(fp$pop_A, fp$SP)
alphasimr_gd(fp$pop_B, fp$SP)
```

`alphasimr_gd()` is `1 - mean(f)`, the exact definition `gene_diversity()` uses,
computed directly from AlphaSimR's genotype pull -- so the two pools are
already comparable to every diversity number in the rest of the book on the
same scale.

Reading `alphasimr_founder_pools()` against Steps 1 and 2, it is
`runMacs2(nInd = , nChr = , segSites = , split = )` followed by
`SP$addTraitAD(nQtlPerChr = seg_sites, meanDD = `r rrs_cfg$mean_dd`,
varDD = `r rrs_cfg$var_dd`)` and `SP$setVarE(h2 = )`. The requested `nInd`
individuals are sampled across both demes -- the first `n_pool_A` become pool A,
the rest pool B. There is no separate marker panel: every one of the
`r rrs_cfg$seg_sites` segregating sites per chromosome is a QTL, so the sites
that carry the trait are the ones `pullSegSiteGeno()` reads for `X`, `f` and
`fL` throughout. And per Step 2's table, the trait is `addTraitAD` because
without dominance there is nothing for testcross selection to exploit that
within-pool selection could not already get.

Because these founders come from MaCS, the numbers in the rest of this section
change whenever the chapter is re-executed -- see Step 1's warning. The numbers
in Steps 1 to 7, and in @sec-programme, do not.

### One cycle

```{r}
cy1 <- alphasimr_rrs_cycle(fp$pop_A, fp$pop_B, fp$SP, rrs_cfg)
fmt(cy1$summary, 4)
```

`alphasimr_rrs_cycle()` runs the same six steps for each pool, symmetrically
(pool A tested on B, pool B tested on A), and every one is a primitive from
above:

1. **Double haploids** (Step 5). `makeDH()` draws `r rrs_cfg$n_dh` fully
   homozygous lines from *each* individual in the current pool -- at book scale,
   `r rrs_cfg$n_pool_A` parents x `r rrs_cfg$n_dh` DH each =
   `r rrs_cfg$n_pool_A * rrs_cfg$n_dh` DH lines per pool, per cycle.
2. **Pick reciprocal testers.** The first `r rrs_cfg$n_testers` individuals of
   the *opposite* pool are used as testers -- a fixed slice, not a random or
   merit-based draw. This is the RRS-defining move: a line's value is judged
   by its cross onto the other heterotic pool, not by itself.
3. **Testcross and phenotype** (Step 5, Step 6). `hybridCross()` forms the full
   factorial between every DH line and every tester --
   `r rrs_cfg$n_pool_A * rrs_cfg$n_dh` DH x `r rrs_cfg$n_testers` testers =
   `r rrs_cfg$n_pool_A * rrs_cfg$n_dh * rrs_cfg$n_testers` testcross hybrids
   per pool -- and `setPheno()` adds phenotyping error sized to hit the
   heritability set by `SP$setVarE(h2 = )`.
4. **GCA proxy.** Testcross phenotypes are grouped by DH-line parent
   (`tapply(pheno[, 1], mother, mean)`) and averaged across the
   `r rrs_cfg$n_testers` testers each line was crossed to -- a general
   combining ability estimate, following @comstock1949. Only trait 1 is
   used for this ranking even though this chapter's config simulates
   `r rrs_cfg$n_traits` traits per line; the second trait rides along in
   `X`/`hybrids` but plays no role in selection here. Step 7 is what it would
   take to use both.
5. **Keep the elites** (Step 6). The `r rrs_cfg$n_sel` DH lines per pool with
   the highest mean testcross phenotype are kept as `elite_A`/`elite_B`.
6. **Recombine** (Step 5). `randCross()` random-mates the elites back up to
   `r rrs_cfg$n_pool_A` individuals, founding the next cycle's pool A
   (symmetrically for B).

`elite_A`/`elite_B` are what step 6 mates -- and, in the *last* cycle, what
the final commercial cross uses instead of being recombined again. The cycle's
`summary` mixes two different populations on purpose: `mean_tc_A`/`mean_tc_B`
average over *all* DH lines tested that cycle (the population's testcross
performance before selection), while `gd_A`/`gd_B` are gene diversity of only
the `r rrs_cfg$n_sel` surviving elites (the population after selection). That
asymmetry is exactly what the next section plots -- response measured
pre-selection, diversity measured post-selection.

### The trajectory across cycles

`alphasimr_pipeline()` just runs `alphasimr_rrs_cycle()` in a loop for
`r rrs_cfg$n_cycles` cycles, carrying `pop_A`/`pop_B` forward from one
cycle's output to the next's input -- there is no reseeding inside the loop,
only the founder draw at the start is seeded, so the whole trajectory is one
continuous random process. Only the *last* cycle's `elite_A`/`elite_B` --
not an intermediate cycle's -- go on to found the commercial hybrid pool
below; every earlier cycle's elites are spent entirely on `randCross()`ing
the next cycle's pool.

```{r}
res <- alphasimr_pipeline(rrs_cfg)
fmt(res$cycles, 4)
```

```{r}
#| label: fig-alphasimr-cycles
#| echo: false
#| fig-width: 9
#| fig-height: 4.2
#| fig-cap: "Reciprocal recurrent selection over four cycles, book scale. Left: mean testcross phenotype (selection response) per pool. Right: gene diversity of the elite lines retained each cycle."
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.3, 2.2, 0.8))
cols <- c(A = "#1b6ca8", B = "#c0392b")
cy <- res$cycles
plot(cy$cycle, cy$mean_tc_A, type = "b", pch = 19, col = cols["A"],
     ylim = range(cy$mean_tc_A, cy$mean_tc_B), xlab = "Cycle",
     ylab = "Mean testcross phenotype", main = "Selection response")
lines(cy$cycle, cy$mean_tc_B, type = "b", pch = 19, col = cols["B"])
legend("bottomright", c("pool A tested on B", "pool B tested on A"), col = cols,
       lwd = 2, pch = 19, bty = "n", cex = 0.8)
plot(cy$cycle, cy$gd_A, type = "b", pch = 19, col = cols["A"],
     ylim = c(0, max(cy$gd_A, cy$gd_B)), xlab = "Cycle",
     ylab = "Gene diversity of the elite lines", main = "Diversity erosion")
lines(cy$cycle, cy$gd_B, type = "b", pch = 19, col = cols["B"])
par(op)
```

Both pools respond -- mean testcross phenotype rises -- and both pools lose
gene diversity in their retained elite lines, cycle over cycle. Nobody pays this
cost for its own sake: @sec-cycles makes the same point with the package's own
faster simulator, over a longer trajectory and a grid of selection intensities,
and @sec-programme makes it a third time in an inbred-line programme. This
section shows it holds with actual meiosis and dominance behind it too.

### The final hybrid pool through `stage2_select()`, unchanged

`alphasimr_make_hybrids()` testcrosses the final cycle's elites directly
against each other -- a full factorial diallel between the
`r rrs_cfg$n_sel` pool-A elites and `r rrs_cfg$n_sel` pool-B elites, so
`r rrs_cfg$n_sel * rrs_cfg$n_sel` commercial hybrids -- and phenotypes them
the same way every testcross in this chapter was phenotyped. `alphasimr_to_stage1()`
then does the bookkeeping that makes this look like any other stage-1 output:
it remaps AlphaSimR's internal line ids into the `1..n_A`, `n_A+1..n_A+n_B`
convention `hybrids$line_A`/`line_B` and `fL` use elsewhere in the book, and
builds `fL` by stacking the elite lines from *both* pools' genotypes into one
matrix and computing a single joint coancestry matrix over all of them --
not two separate per-pool matrices -- which is what lets `theta_pools()` and
the rest of the pipeline index into pool A and pool B lines with one
consistent numbering.

```{r}
st1 <- res$stage1
dim(st1$X)
sel <- stage2_select(st1$X, st1$f, st1$hybrids, n_sel = 8, fL = st1$fL)
fmt(sel$metrics[, c("scenario", "alpha", "GD", "GD_BS", "F_ST", "index")], 4)
```

Nothing here is AlphaSimR-specific: `stage2_select()` never sees where `X`,
`f`, `hybrids` and `fL` came from, only that they satisfy the contract stage 1
promises. `GD_BS`, the between-pool divergence that carries the heterosis
signal (@sec-choosing), is computed from `fL` exactly as it is everywhere else
in the book.

## Assumptions, and what this adds over @sec-simulating-crosses

**What is more realistic here.** Actual recombination and linkage (a genetic
map per chromosome, not independent markers), a dominance-driven trait so
testcross selection has something real to select on, and multi-generation
recurrent selection rather than one static pool.

**What is still simplified.** Testers are a small fixed subset of the opposite
pool (`r rrs_cfg$n_testers` individuals) taken by index, not chosen at random
or by merit, and not the single unrelated inbred tester a real programme
would typically use -- a fixed pool slice makes the same testers reused every
cycle, which a production programme would treat as a confound; there is one
phenotyping environment (`setPheno()`'s error variance, no
genotype-by-environment term -- @sec-programme adds one); the population-split
coalescent history (`runMacs2(split = )`) is a stylised divergence process, not
calibrated to a named heterotic-pool pair; doubled-haploid production is
instantaneous, skipping the field seasons Step 5 counted; and selection acts on
trait 1 only -- the `r rrs_cfg$n_traits` traits this chapter's config
simulates are all carried through `X`/`hybrids`, but only the first ever
enters the GCA ranking, so the rest are along for the ride, not selected on.

**Scale.** This chapter runs live at a book scale (`r rrs_cfg$n_pool_A` +
`r rrs_cfg$n_pool_B` founders, `r rrs_cfg$n_chr * rrs_cfg$seg_sites` markers,
`r rrs_cfg$n_cycles` cycles) -- AlphaSimR is fast enough at this size that no
caching is needed, unlike the MABC scans in @sec-backcross. A heavier,
production-scale run is `inst/scripts/run_alphasimr.R`:

```sh
Rscript inst/scripts/run_alphasimr.R --test          # smoke run, ~10 s
Rscript inst/scripts/run_alphasimr.R --out=DIR        # full run, ~2 min
```

It runs `alphasimr_pipeline()` at a larger configuration, feeds the resulting
hybrid pool through `stage2_select()`, and writes the cycle trajectory and
selection metrics as CSVs.

## Where this connects

@sec-simulating-crosses and Step 8 build the same three-object contract
from two different generative models -- the fast one and the realistic one --
which is the point of drawing that seam where the book does: everything past
it, `build_ctx()` through `stage2_select()`, cannot tell which one it is
looking at. @sec-heterosis explains why the dominance term here is what makes
testcross selection informative at all. @sec-response supplies the breeder's
equation Step 6 checks and the index theory Step 7 uses; @sec-limits is where
Step 3's `varA`-against-`genicVarA` gap becomes the subject rather than a
diagnostic. @sec-cycles runs the same diversity-erosion argument at a longer
horizon with the package's own faster simulator; this chapter is its cross-check
against an engine with real meiosis behind it.

@sec-programme is the other half of the method: the same primitives assembled
into an inbred-line programme with staged yield trials, where the question is
not which hybrids to sell but how fast a variety pipeline can run and what it
spends to run faster.
