# Simulating a hybrid breeding programme: AlphaSimR from the ground up {#sec-alphasimr}

```{r}
#| include: false
source("_common.R")
library(AlphaSimR)
rrs_cfg <- utils::modifyList(rrs_config,
  list(n_pool_A = 12, n_pool_B = 12, n_dh = 3, n_sel = 6, n_testers = 2,
       n_cycles = 4, seg_sites = 150, n_chr = 5, n_traits = 2, seed = 1))
```

@sec-simulating-crosses built the `{X, f, hybrids, fL}` contract with a
deliberately fast simulator: Balding-Nichols allele frequencies, independent
markers, no recombination, no dominance. That is enough to exercise the
diversity and selection machinery, but it is not how a hybrid breeding
programme is actually run, and it is not how anyone would build one from
scratch.

This chapter builds one. The crop in mind is maize, the canonical allogamous
hybrid crop: two heterotic pools, inbred (doubled-haploid) lines drawn from
each, lines judged by their testcrosses onto the other pool rather than by
themselves, pools improved by reciprocal recurrent selection (RRS;
@comstock1949), and a commercial product that is a single cross between one
line from each pool. Steps 1 to 7 are a **step-by-step construction** with
AlphaSimR (@gaynor2021), written out in the primitive calls -- `runMacs()`,
`SimParam`, `addTraitAD()`, `makeCross()`, `hybridCross()`, `makeDH()`,
`setPheno()`, `selectInd()` -- with a plant breeder's reading of each argument
and a measurement at each step. Step 8 assembles those primitives into RRS,
replicates it against a paired no-selection control, and shows that the same
stage-1 contract comes out the other end, so `stage2_select()` runs on it
exactly as it does on the book's own simulated data.

**Why this machinery, and why it has no counterpart in a self-pollinated
programme.** In an autogamous (self-pollinated) crop -- wheat, rice, soybean,
common bean -- recombination happens once, in the F1, and the programme then
narrows a segregating population toward homozygosity by selfing or
single-seed descent, selecting throughout on a line's own (per se) phenotype,
because a self-pollinated genotype breeds true and its own performance *is*
what a farmer will grow: the released variety is very often the same
homozygous line the breeder selected. @sec-programme builds exactly that
programme with the same AlphaSimR primitives -- `makeDH()`, `setPheno()`,
`selectInd()` -- and is the chapter to read this one against, step by step.
None of Steps 2, 4 and
6 below would have a job there: there is no dominance for per se selection to
be blind to (an inbred line's value is its additive merit alone -- Step 3
makes this an identity, `varD = 0` exactly on a homozygous panel), no tester
to cross a candidate onto, no general or specific combining ability to
estimate (@sprague1942), and no second, deliberately divergent population to
keep apart, because there is only one gene pool, not two heterotic ones held
apart on purpose. What follows instead is specific to an allogamous,
hybrid-product crop: the inbred line is not the product, the F1 is (Step 5);
a line's breeding value depends on which pool it will be crossed to, not on
itself alone (Step 6's tester dependence); and the trait improvement a
programme has made is realised only in a cross remade every season, never in
either parental line by itself.

The pedagogical source is the AlphaSimR course of Gorjanc and colleagues
[@gorjanc2022]: its vignettes on simulating DNA, simulating traits, the DNA
lottery, single- and multi-trait selection, and a plant breeding programme.
The course has no hybrid programme of its own; the wheat programme it does build is the
subject of @sec-programme. This chapter takes the course's primitives and
assembles them into the hybrid case.

The order matters. Everything in Step 8 is a wrapper in
`R/13_alphasimr_bridge.R` -- `rrs_config`, `alphasimr_founder_pools()`,
`alphasimr_rrs_cycle()`, `alphasimr_make_hybrids()`, `alphasimr_to_stage1()`,
`alphasimr_pipeline()` -- and a wrapper is only readable once you know what it
is wrapping. AlphaSimR is a `Suggests` dependency, guarded exactly like
`quadprog` and `highs` elsewhere in the package; this chapter and
@sec-programme are the only ones that need it.

## Step 1 -- Founder genomes: what a founder is, and which engine to use

Every AlphaSimR simulation starts with founder haplotypes. Before choosing an
engine, it is worth seeing what one produces, at a size small enough to read.
`inbred = TRUE` makes every founder fully homozygous, which is what a panel of
inbred lines is.

```{r}
set.seed(1)
tiny <- quickHaplo(nInd = 3, nChr = 2, segSites = 4, inbred = TRUE)
SPt  <- SimParam$new(tiny)
tiny_pop <- newPop(tiny, simParam = SPt)
pullSegSiteHaplo(tiny_pop, simParam = SPt)
```

One row per haplotype, two per diploid individual (`1_1` and `1_2` are the two
chromosome copies of individual 1), one column per segregating site across both
chromosomes (`1_3` is chromosome 1, site 3). Alleles are coded 0 and 1 --
ancestral and derived under a coalescent engine, arbitrary labels under
`quickHaplo()`. Because the founders are inbred, the two rows of each
individual are identical. `pullSegSiteGeno()` sums the two
rows into an allele dosage in `{0, 1, 2}`:

```{r}
tiny_geno <- pullSegSiteGeno(tiny_pop, simParam = SPt)
tiny_geno
stopifnot(all(tiny_geno %in% c(0, 2)))
stopifnot(isTRUE(all.equal(diag(molecular_coancestry(tiny_geno / 2)),
                           rep(1, 3))))
```

The two assertions fix the coding for the rest of the book. AlphaSimR's dosage
is 0/1/2; the contract of @sec-simulating-crosses codes the same information as
0/0.5/1, the within-individual frequency of the reference allele (AlphaSimR's
derived allele plays that role), and every bridge function divides by two on
the way across (`stage1_build()` also detects
and rescales 0/1/2 on its own). On an inbred individual the molecular
self-coancestry of @sec-coancestry is exactly 1, which is the second assertion.

There are two engines, and the choice is not cosmetic. `runMacs()` runs a
backward-in-time coalescent (@chen2009) under a species-specific demography --
`species = "MAIZE"`, `"WHEAT"`, `"CATTLE"` -- so the founders carry a realistic
allele-frequency spectrum and realistic linkage disequilibrium. `quickHaplo()`
samples 0s and 1s directly: instant, and with no history behind it.

```{r}
tm <- function(expr) as.numeric(system.time(expr)["elapsed"])

# reproducibility: two calls from the same seed, and R's own stream after each
set.seed(7); qa <- quickHaplo(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE)
rs_qa <- .Random.seed
set.seed(7); qb <- quickHaplo(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE)
rs_qb <- .Random.seed
set.seed(7); ma <- runMacs(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE,
                           species = "MAIZE")
rs_ma <- .Random.seed
set.seed(7); mb <- runMacs(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE,
                           species = "MAIZE")
rs_mb <- .Random.seed
stopifnot(identical(rs_qa, rs_qb), identical(rs_ma, rs_mb))

# timing: one runMacs() call; quickHaplo() over n_rep_q calls, because a
# single call is below the clock's resolution
n_rep_q <- 100
t_m <- tm(runMacs(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE,
                  species = "MAIZE"))
t_q <- tm(for (k in seq_len(n_rep_q))
            quickHaplo(nInd = 20, nChr = 10, segSites = 50, inbred = TRUE)) / n_rep_q

fmt(data.frame(engine              = c("quickHaplo", "runMacs"),
               seconds_per_call    = c(t_q, t_m),
               same_twice          = c(identical(qa@geno, qb@geno),
                                       identical(ma@geno, mb@geno)),
               R_stream_same_after = c(identical(rs_qa, rs_qb),
                                       identical(rs_ma, rs_mb))), 4)
```

Read the last two columns together. **`runMacs()` is not reproducible from
`set.seed()`, and R's own random stream is untouched by that.** `runMacs()`
draws a seed for MaCS from R's stream (one `sample.int()` per chromosome) and
hands it to the binary. The fourth column, asserted in the chunk, says that
after two calls from the same seed R's `.Random.seed` is identical: the
amount of R's stream the call consumes is fixed, and every draw R makes
afterwards is the same in both runs. Yet the third column says the haplotypes
differ. What was measured, then, is that MaCS's output is not a deterministic
function of the seed it is handed. The only thing that moves between two runs
is the founder haplotypes themselves -- and everything computed *from* those
founders. `quickHaplo()` is pure R sampling and is exactly reproducible.

(The `seconds_per_call` column is wall clock and will differ on your machine
and between renders. `runMacs()` takes seconds; `quickHaplo()`, timed over
`r n_rep_q` calls because one call is below the clock's resolution, takes
about `r fmt(1000 * t_q, 2)` ms per call.)

::: {.callout-warning}
## The founders are not seeded, and everything downstream is

This is the single most expensive trap in AlphaSimR, because it fails silently:
the code looks reproducible, runs without complaint, and returns different
numbers. The non-reproducibility is documented in `alphasimr_founder_pools()`'s
help page for `runMacs2()`; the table above shows it is not `runMacs2`-specific
but true of MaCS generally.

Two consequences, both because the founders move and not because of anything
R's stream does:

1. **Found once, then branch.** Any experiment comparing two schemes must draw
   the founders a single time and start both arms from that one object. Draw
   them twice and the two arms start from different founders, and the
   comparison measures the founders as much as the schemes. @sec-limits'
   linked arm is built this way, and so is Step 8's paired control below.
2. **Save what you cannot reseed.** `saveRDS(founderGenomes, ...)` is the only
   way to get the same coalescent history back in a later session.

Everything after the founders -- the crosses, the doubled haploids, the
phenotyping error, the recombination -- *is* controlled by `set.seed()`, given
the founders. Reseeding downstream does not recover the founders and is not
needed for anything else.
:::

So what does `quickHaplo()` give up? Linkage disequilibrium among the founders.
The honest way to measure that is against the sampling floor: with `n`
individuals, two completely independent loci still show `r^2` of about
`1 / (n - 1)` by chance alone.

```{r}
ldstat <- function(fg) {
  SPx <- SimParam$new(fg)
  g <- pullSegSiteGeno(newPop(fg, simParam = SPx), simParam = SPx)
  g <- g[, apply(g, 2, stats::sd) > 0, drop = FALSE]
  chr <- sub("_.*$", "", colnames(g))           # "chr_site" column names
  r <- stats::cor(g)
  adj <- diag(r[-nrow(r), -1])                  # consecutive columns ...
  adj <- adj[chr[-length(chr)] == chr[-1]]      # ... on the same chromosome
  c(adjacent = mean(adj^2),                     # neighbouring sites, same chromosome
    overall  = mean(r[upper.tri(r)]^2),         # all pairs, mostly distant
    noise    = 1 / (nrow(g) - 1))               # what pure chance gives
}
set.seed(3); lq <- ldstat(quickHaplo(nInd = 100, nChr = 5, segSites = 200,
                                     inbred = TRUE))
set.seed(3); lm_ <- ldstat(runMacs(nInd = 100, nChr = 5, segSites = 200,
                                   inbred = TRUE, species = "MAIZE"))
fmt(rbind(quickHaplo = lq, runMacs = lm_))
```

`adjacent` is the mean `r^2` between neighbouring sites on the same
chromosome (pairs that straddle a chromosome boundary are excluded);
`overall` is over all pairs, most of them on different chromosomes.
`quickHaplo()`'s adjacent-site `r^2` sits on the noise floor: there is no
founder LD at all, as designed. `runMacs()`'s is
`r fmt(lm_[["adjacent"]] / lm_[["noise"]], 0)` times the floor in this render,
and its *overall* `r^2` is barely above it -- so the LD it creates is local,
decaying with map distance, which is what real LD does.

The `runMacs` row of that table is itself not reproducible, for the reason the
row above it establishes; the `quickHaplo` row is, to the last digit.

What that does and does not imply for the rest of the chapter. R's stream is
unaffected by MaCS, so any step that draws its own `quickHaplo()` founders
from a fresh `set.seed()` is exactly reproducible, and so would be a step that
did not reseed: the `runMacs()` calls above leave the stream exactly where a
fixed number of `sample.int()` calls leaves it. What is not reproducible is
anything computed *from* MaCS founders -- the `runMacs` rows of both tables
here, and all of Step 8. The chapter still calls `set.seed()` immediately
before each step's own founder draw rather than once at the top, so that each
step is reproducible on its own and not only as part of this chapter's
execution order. With that in place, two independent executions agree on every
number in Steps 1 to 7 except the wall-clock seconds and the `runMacs` rows.

That measurement decides the policy this chapter follows:

- **`quickHaplo(inbred = TRUE)` for Steps 2--7 and all of @sec-programme.** They
  teach the arithmetic of dominance, heterosis and response, where founder LD
  is not the point, and in exchange every number printed is reproducible from
  a clean clone. AlphaSimR still applies a real genetic map and real meiosis to
  every descendant, so linkage and the Bulmer effect of @sec-limits still
  operate -- only the founders' *history* is stylised.
- **`runMacs2(split = )` in Step 8.** There the divergence between two heterotic
  pools *is* the object of study, and no amount of speed compensates for not
  having it. Every Step 8 result is therefore replicated over founder draws.

Founder LD does matter where marker-QTL association is the mechanism, which is
genomic prediction -- see @sec-genomic-prediction, and note that the accuracies
there are computed on a different simulator for this reason.

## Step 2 -- `SimParam`, and defining a trait with dominance

`SimParam` holds everything global: the genetic map, which sites are QTL, the
trait effects, the error variances. It is an R6 object, so its methods are
called with `$`, and a variable named `SP` is picked up automatically by every
AlphaSimR function that needs it.

A hybrid programme exists because of dominance: without it a line's cross
performance is its own breeding value and there is nothing a testcross can see
that the line per se could not (@sec-heterosis). So the trait defined here, and
used through Step 6, carries a dominance term from the start. This is exactly
the reason an autogamous programme has no equivalent of this step: there,
dominance is invisible by construction (a released, self-pollinated variety is
homozygous, so no heterozygote is ever evaluated or sold) and `SP$addTraitA()`
-- additive effects only -- is the whole model, the case @sec-programme runs
throughout.

```{r}
# Seeded here, not once at the top of the chapter: Step 1 draws from R's random
# stream (each runMacs() call consumes a fixed amount of it), so everything
# downstream would move with any edit or re-execution of Step 1 unless the seed
# is reset immediately before the draw that matters.
set.seed(101)
founders <- quickHaplo(nInd = 60, nChr = 10, segSites = 100, inbred = TRUE)
SP <- SimParam$new(founders)
SP$addTraitAD(nQtlPerChr = 100, mean = 10, var = 1,
              meanDD = rrs_cfg[["mean_dd"]], varDD = rrs_cfg[["var_dd"]])
SP$setVarE(h2 = 0.4)
```

There are four trait constructors, and choosing between them is choosing a model
of gene action:

| Method | Adds | When a plant programme needs it |
|---|---|---|
| `SP$addTraitA()` | additive only | line per se performance; the course's teaching case [@gorjanc2022], and Step 7 here |
| `SP$addTraitAD()` | + dominance | testcross and hybrid value. This is what Steps 2--6 and Step 8 use |
| `SP$addTraitAG()` | + genotype-by-environment | multi-location yield trials, where the same genotype is scored in several places. @sec-programme uses it |
| `SP$addTraitADE()` | + epistasis | available; used nowhere in this book |

### The dominance model, as AlphaSimR writes it

For a diploid with QTL $k = 1, \dots, m$, let $x_{ik} \in \{0, 1, 2\}$ be the
dosage of the derived allele of individual $i$ at QTL $k$ (the column of
`pullQtlGeno()`). AlphaSimR's `TraitAD` genetic value is

$$G_i \;=\; \mu_0 + \sum_{k=1}^{m} a_k\,(x_{ik} - 1) + \sum_{k=1}^{m} d_k\,\mathbb{1}[x_{ik} = 1]$$

with $\mu_0$ an intercept, $a_k \in \mathbb{R}$ the additive effect (half the
difference between the two homozygotes, on the classical scale of
@falconer1996, Ch. 7, which puts the origin at the midpoint of the two
homozygotes and assigns them $+a$ and $-a$ and the heterozygote $d$; no
equation number is attributed here, the scale is stated in full) and $d_k \in
\mathbb{R}$ the dominance deviation, the departure of the heterozygote from that
midpoint. It is the model of @sec-heterosis with AlphaSimR's coding, and the
chunk checks it rather than trusts it:

```{r}
tr <- SP$traits[[1]]
a  <- tr@addEff
d  <- tr@domEff
pop0 <- newPop(founders, simParam = SP)
x0   <- pullQtlGeno(pop0, simParam = SP)
gv_by_hand <- as.numeric(tr@intercept + (x0 - 1) %*% a + (x0 == 1) %*% d)
stopifnot(isTRUE(all.equal(gv_by_hand, as.numeric(gv(pop0)))))
```

`meanDD` and `varDD` parameterise the **dominance degree**
$\delta_k = d_k / |a_k|$: AlphaSimR draws $\delta_k \sim N(\texttt{meanDD},
\texttt{varDD})$ and sets $d_k = \delta_k\,|a_k|$, so $\delta_k = 0$ is
additivity, $\delta_k = 1$ complete dominance, $\delta_k > 1$ overdominance,
and a *positive mean* is directional dominance: on average the heterozygote
sits above the homozygote midpoint, whichever allele is favourable. Directional
dominance is what makes heterosis positive and inbreeding depression a loss on
the trait scale, and it is the standing hypothesis for yield in maize.

```{r}
c(n_qtl = length(a), meanDD_requested = rrs_cfg[["mean_dd"]],
  meanDD_realised = mean(d / abs(a)), varDD_requested = rrs_cfg[["var_dd"]],
  varDD_realised = stats::var(d / abs(a))) |> fmt(3)
```

`nQtlPerChr` sets how polygenic the trait is. Setting it equal to `segSites`
makes *every* segregating site a QTL, which is what this chapter does
throughout -- convenient, because then the marker matrix and the causal loci
are the same set and there is no marker-versus-QTL gap to reason about. Real
programmes have that gap; @sec-genomic-era is where the book takes it
seriously.

`mean` fixes the mean genetic value of the *founders*. `var` fixes their
**additive** variance (`useVarA = TRUE`, the default), not their total genetic
variance -- with inbred founders the two coincide, for a reason Step 3 shows.
`corA` (Step 7) sets the genetic correlation between traits.

The environmental variance has two routes, and plant breeders use the second:

- `SP$setVarE(h2 = 0.4)` fixes an error variance once, from the founders'
  genetic variance (here `r fmt(SP$varE, 2)`), and every later `setPheno()`
  reuses it.
- `setPheno(varE = , reps = )` sets it per call. **`reps` is the plant idiom for
  locations**: `varE` is the plot-level error, and a genotype scored in `reps`
  plots has its error variance divided by `reps`, so the phenotype is an
  entry mean. That is how a staged programme raises selection accuracy from
  one stage to the next without ever changing the trait's heritability -- the
  mechanism @sec-programme is built on, and the one Step 6 uses for testers.

## Step 3 -- The base population, and where the variance components live

`newPop()` turns founder haplotypes into individuals and, because a trait is
now defined, computes their genetic values immediately.

```{r}
pop <- setPheno(newPop(founders, simParam = SP), simParam = SP)
c(n = pop@nInd, meanG = as.numeric(meanG(pop)), varG = as.numeric(varG(pop)),
  varP = as.numeric(varP(pop)),
  h2_realised = as.numeric(varG(pop) / varP(pop))) |> fmt(3)
```

The realised heritability is near but not equal to the requested 0.4:
`setVarE()` fixes a variance, and the ratio it produces in any finite sample is
itself a random variable. This is a general and useful habit -- **check what the
simulation realised, do not assume what you requested.**

`genParam()` returns the full decomposition. Four of its fields are the ones a
hybrid programme cares about, and they read differently on a panel of inbred
lines and on a random-mating population made from them:

```{r}
mixed <- pop
for (k in 1:3) mixed <- randCross(mixed, nCrosses = 1000, simParam = SP)

vrow <- function(p, lab) {
  g <- genParam(p, simParam = SP)
  data.frame(population = lab,
             meanG     = as.numeric(g[["mu"]]),
             varA      = as.numeric(g[["varA"]]),
             genicVarA = as.numeric(g[["genicVarA"]]),
             ratio     = as.numeric(g[["varA"]] / g[["genicVarA"]]),
             varD      = as.numeric(g[["varD"]]),
             genicVarD = as.numeric(g[["genicVarD"]]))
}
vtab <- rbind(vrow(pop,   "founders (fully inbred)"),
              vrow(mixed, "after 3 random matings"))
fmt(vtab)
stopifnot(vtab[["varD"]][1] < 1e-10)

# Among fully homozygous lines the locus value is a (x - 1), so at linkage
# equilibrium varA = sum 4 p q a^2, while genicVarA = sum 2 p q alpha^2 with
# alpha = a + d (q - p): under dominance the LE ratio is close to, not
# exactly, 2. What varA carries above varA_LE is disequilibrium.
p_f      <- colMeans(x0) / 2                        # founder allele frequencies
varA_f   <- vtab[["varA"]][1]
genic_f  <- vtab[["genicVarA"]][1]
varA_LE  <- sum(4 * p_f * (1 - p_f) * a^2)
ratio_LE <- varA_LE / genic_f
varA_LD  <- varA_f - varA_LE
ratio_LD <- varA_LD / genic_f
stopifnot(isTRUE(all.equal(as.numeric(bv(pop, simParam = SP)[, 1]),
                           as.numeric(sweep(x0, 2, 2 * p_f) %*% a))))
c(varA = varA_f, varA_LE = varA_LE, ratio_LE = ratio_LE,
  varA_LD = varA_LD, ratio_LD = ratio_LD) |> fmt(4)
```

`varA` is the realised additive variance; `genicVarA` is what it would be if the
loci were independent and in Hardy-Weinberg proportions. Their ratio is the
whole subject of @sec-limits, and the package's `genic_var()` computes the same
quantity from a marker matrix directly.

The founders' ratio of `r fmt(vtab[["ratio"]][1], 2)` has two parts, and the
chunk separates them. Among fully homozygous lines the value at QTL $k$ is
$a_k (x_{ik} - 1)$ with $x_{ik} \in \{0, 2\}$, and AlphaSimR's `bv()` on these
founders is exactly $\sum_k a_k (x_{ik} - 2 p_k)$ (asserted in the chunk). At
linkage equilibrium its variance is `varA_LE`
$= \sum_k 4 p_k q_k a_k^2 =$ `r fmt(varA_LE, 4)`, against a genic value of
`r fmt(genic_f, 4)`: a ratio of `r fmt(ratio_LE, 2)`, exact given the effects
and the frequencies. That part is **inbreeding** -- a homozygous population
has per-locus variance `4pq` where a Hardy-Weinberg one has `2pq`. It is close
to 2 and not exactly 2 because `genicVarA` is written with the Hardy-Weinberg
average effect $\alpha_k = a_k + d_k (q_k - p_k)$ defined below, whereas the
homozygous lines carry $a_k$; twice the genic value equals `varA_LE` only
where $d_k (q_k - p_k) = 0$ at every locus. The remainder,
`varA_LD = varA - varA_LE` $=$ `r fmt(varA_LD, 4)`, or `r fmt(ratio_LD, 2)`
in ratio units, is sampling (gametic-phase) disequilibrium among only 60
founders' worth of gametes. Three rounds of random mating restore
Hardy-Weinberg and pull the ratio to `r fmt(vtab[["ratio"]][2], 2)`; what
remains is that same disequilibrium. A ratio away from 1 in an *outbred*
population is the gametic-phase disequilibrium @sec-response defines and
@sec-limits measures; a ratio away from 1 in an inbred one is mostly the
homozygosity.

The dominance columns carry the hybrid lesson. **`varD` among the inbred
founders is exactly zero** (the assertion above): a homozygous line has no
heterozygous locus, so its dominance contribution is zero whatever $d$ is,
and no amount of per se evaluation of inbred lines can see dominance. This
single identity is the operational reason an autogamous programme stops at
per se selection and never needs a testcross: its released lines are exactly
this "founders" row, `varD = 0` for the whole life of the programme, so per
se phenotype already is breeding value up to the additive/genic gap discussed
below. After random mating `varD` appears. Two identities make that quantitative. With
$p_k$ the founders' frequency of the derived allele at QTL $k$ and
$q_k = 1 - p_k$, the population mean under Hardy-Weinberg proportions is
[@falconer1996, Ch. 7, Eq. 7.2]

$$M_{HW} \;=\; \mu_0 + \sum_k \big[a_k\,(p_k - q_k) + 2\,p_k q_k\,d_k\big]$$

while the inbred panel at the same frequencies has only the first term. The
difference, $2\sum_k d_k p_k q_k$, is the inbreeding depression of
@sec-heterosis at $F = 1$, which the package computes as
`inbreeding_depression(p, d, F_coef = 1)`. And the genic additive variance is
$\sum_k 2 p_k q_k \alpha_k^2$ with $\alpha_k = a_k + d_k\,(q_k - p_k)$ the
average effect of a gene substitution. @falconer1996 define $\alpha$ in
Ch. 7 and the single-locus additive variance $2pq\alpha^2$ in Ch. 8; the
equation numbers are not attributed here, so both identities are stated in
this chapter's notation: $\alpha_k = a_k + d_k\,(q_k - p_k)$ and, summed over
independent loci, $\sum_k 2 p_k q_k \alpha_k^2$, which is the genic variance:

```{r}
gp  <- genParam(pop, simParam = SP)
alpha_f <- a + d * (1 - 2 * p_f)
stopifnot(isTRUE(all.equal(as.numeric(gp[["mu_HW"]] - gp[["mu"]]),
                           inbreeding_depression(p_f, d, F_coef = 1))))
stopifnot(isTRUE(all.equal(as.numeric(gp[["genicVarA"]]),
                           sum(2 * p_f * (1 - p_f) * alpha_f^2))))
c(mean_inbred = as.numeric(gp[["mu"]]), mean_HW = as.numeric(gp[["mu_HW"]]),
  inbreeding_depression = inbreeding_depression(p_f, d, F_coef = 1)) |> fmt(3)
```

Both hold to machine precision against AlphaSimR's own `genParam()`. The gap
between the two means is the amount a random-mating population built from these
lines would sit *above* the lines themselves -- and, read in the other
direction, the amount an open-pollinated population loses on being driven to
homozygosity. It is the same $d$ that Step 4 turns into heterosis.

## Step 4 -- The DNA lottery, heterosis, and why an F1 is not a family

The course [@gorjanc2022] demonstrates Mendelian sampling by crossing outbred
founders and showing that full sibs vary. With **inbred** parents the result is different,
and it is the fact hybrid breeding is organised around.

```{r}
crosses <- list(c(1, 2), c(1, 3), c(2, 3))
lottery <- do.call(rbind, lapply(crosses, function(ij) {
  plan <- matrix(ij, ncol = 2)
  f1 <- makeCross(pop, plan, nProgeny = 200, simParam = SP)
  dh <- makeDH(makeCross(pop, plan, nProgeny = 1, simParam = SP),
               nDH = 200, simParam = SP)
  better <- max(gv(pop)[ij])
  data.frame(cross         = paste(ij, collapse = " x "),
             parent_mean   = mean(gv(pop)[ij]),
             better_parent = better,
             f1_mean       = as.numeric(meanG(f1)),
             f1_sd         = as.numeric(sqrt(varG(f1))),
             dh_sd         = as.numeric(sqrt(varG(dh))),
             dh_best       = max(gv(dh)),
             transgressive = mean(gv(dh) > better),
             # how far the better parent sits above the DH mean, in DH s.d.
             gap_in_dh_sd  = (better - mean(gv(pop)[ij])) / as.numeric(sqrt(varG(dh))))
}))
fmt(lottery, 3)
stopifnot(all(lottery[["f1_sd"]] < 1e-8))
stopifnot(which.max(lottery[["transgressive"]]) == which.min(lottery[["gap_in_dh_sd"]]))
```

`f1_sd` is **zero**, asserted to machine precision. Two fully inbred parents
are homozygous everywhere, so every gamete each one makes is identical, and
every F1 is the same heterozygote. Two hundred F1 plants from a cross of two inbred lines are
two hundred copies of one genotype. There is no lottery yet -- and this
uniformity is the first reason a single-cross hybrid is a saleable variety.

`f1_mean` sits well above `better_parent`. That is heterosis, and under the
model of Step 2 it has an exact value: for inbred parents $i$ and $j$ with
allele indicators $u_{ik}, u_{jk} \in \{0, 1\}$ (their dosage halved), the F1
is heterozygous exactly where they differ, so

$$H_{ij} \;=\; G_{ij} - \tfrac{1}{2}(G_i + G_j) \;=\; \sum_k d_k\,\mathbb{1}[u_{ik} \neq u_{jk}]$$

The additive terms cancel from the mid-parent difference because an inbred
parent's additive contribution is exactly half the F1's. `heterosis_value(X, d)`
is this sum, with `X` the F1's 0/0.5/1 genotype:

```{r}
H <- vapply(seq_along(crosses), function(s) {
  ij <- crosses[[s]]
  f1 <- makeCross(pop, matrix(ij, ncol = 2), nProgeny = 1, simParam = SP)
  hv <- as.numeric(heterosis_value(pullQtlGeno(f1, simParam = SP) / 2, d))
  stopifnot(isTRUE(all.equal(as.numeric(gv(f1)) - mean(gv(pop)[ij]), hv)))
  hv
}, numeric(1))
names(H) <- lottery[["cross"]]
fmt(H, 3)
```

Over a *set* of crosses the same identity has a closed form in the allele
frequencies. Take the 60 founders as two arbitrary halves, "pool A" (lines
1--30) and "pool B" (lines 31--60) -- arbitrary because `quickHaplo()` gave
them a common history and no divergence, which Step 8 will add -- and cross
every A line to every B line. `hybridCross(crossPlan = "testcross")` is the
full factorial. With $p_{Ak}$ and $p_{Bk}$ the derived-allele frequencies of
the two pools at QTL $k$, the mean of $H_{ij}$ over the factorial is
`heterosis_expected(pA, pB, d)`, the identity of @sec-heterosis:

$$\bar H \;=\; \sum_k d_k\,\big(p_{Ak} + p_{Bk} - 2\,p_{Ak}\,p_{Bk}\big)$$

```{r}
A <- pop[1:30]
B <- pop[31:60]
fact <- hybridCross(A, B, crossPlan = "testcross", simParam = SP)
U  <- pullQtlGeno(A, simParam = SP) / 2
V  <- pullQtlGeno(B, simParam = SP) / 2
pA <- colMeans(U)
pB <- colMeans(V)
H_bar_obs <- as.numeric(meanG(fact)) - (as.numeric(meanG(A)) + as.numeric(meanG(B))) / 2
H_bar_exp <- heterosis_expected(pA, pB, d)
stopifnot(isTRUE(all.equal(H_bar_obs, H_bar_exp)))
c(n_hybrids = fact@nInd, mean_A = as.numeric(meanG(A)), mean_B = as.numeric(meanG(B)),
  mean_hybrids = as.numeric(meanG(fact)), heterosis_observed = H_bar_obs,
  heterosis_expected = H_bar_exp,
  pct_of_midparent = 100 * H_bar_obs / ((as.numeric(meanG(A)) + as.numeric(meanG(B))) / 2)) |>
  fmt(3)
```

Exact, on the full factorial, because every A x B pairing is present once.
The size -- `r fmt(100 * H_bar_obs / ((as.numeric(meanG(A)) + as.numeric(meanG(B))) / 2), 1)`
percent of the mid-parent -- follows from `meanDD` and from `quickHaplo()`'s
frequencies near one half, where $p_A + p_B - 2 p_A p_B$ is near its
maximum; it is not calibrated to a crop. @falconer1996, Ch. 14, derive the
heterosis of the F1 between two *random-mating* populations as
$H_{F1} = \sum_k d_k y_k^2$ with $y_k = p_{Ak} - p_{Bk}$ (stated here in this
chapter's notation rather than by their equation number; it is twice the
`divergence` term of `heterosis_partition()`). The two do not conflict; they
use different mid-parents. Theirs carries each population's own
random-mating heterozygosity; an inbred line carries none, so the whole F1
heterozygosity is heterosis and the expression above keeps the `shared` term
as well -- @sec-heterosis works the reconciliation through.

The lottery starts one meiosis later. `dh_sd` is the standard deviation among
doubled haploids drawn from that F1, and it is substantial -- this is the
segregation that F2, DH and RIL populations exist to sample. It is also why
@sec-sizing is a chapter: the number of DH lines you pull from a cross is the
number of tickets you buy, and no downstream selection stage can recover a
combination that was never sampled.

`transgressive` is the fraction of DH lines exceeding the *better* parent per
se: `r fmt(lottery[["transgressive"]][1], 3)`,
`r fmt(lottery[["transgressive"]][2], 3)` and
`r fmt(lottery[["transgressive"]][3], 3)` for the three crosses. It depends
strongly on the cross, and `gap_in_dh_sd` says why: the DH lines are centred
on the mid-parent, so what matters is how far the better parent sits above it
in units of the DH standard deviation -- `r fmt(lottery[["gap_in_dh_sd"]][1], 2)`,
`r fmt(lottery[["gap_in_dh_sd"]][2], 2)` and
`r fmt(lottery[["gap_in_dh_sd"]][3], 2)`. The cross with the smallest gap is
the one with the most transgressive lines (asserted), about a third of them;
the other two, whose better parent is two DH standard deviations above the
mid-parent, give a few percent. Crossing two parents whose favourable alleles
sit at different loci produces progeny beyond either of them -- transgressive
segregation, which @mackay2020 argue is the same phenomenon as heterosis and
inbreeding depression viewed from a third angle -- and how often depends on
the gap between the parents relative to the segregation variance, not on a
constant.

## Step 5 -- Selfing, doubled haploids, and why hybrid seed is remade every year

A plant breeder has to get from a heterozygous F1 to a homozygous line. There
are two routes, and AlphaSimR implements both. Following the same cross through
selfing generations also answers a question every hybrid grower has asked:
what happens if the F1 seed is saved and sown again.

This is the one step where the allogamous, hybrid case and the autogamous
case use literally the same primitive -- `makeDH()` is exactly what
@sec-programme calls to take its wheat F1s straight to a homozygous line, one
generation instead of the six or seven selfing generations traced below --
and where they diverge again in what the resulting inbred line is *for*. In
an autogamous programme that homozygous line is commonly the release: it
breeds true, so what leaves the DH pipeline is what a farmer eventually
grows. In this chapter's programme the homozygous DH line is a parent, held
inside a heterotic pool, testcrossed and recycled (Step 6, Step 8), and never
sold on its own; the object a grower plants is the F1 these lines make
together, which is the reason it cannot be saved.

Under selfing without selection, the fraction of the F1's heterozygous loci
still heterozygous in generation $F_t$ is $(1/2)^{t-1}$ in expectation, so the
expected mean of the selfed generation is

$$\mathbb{E}[M_{F_t}] \;=\; \tfrac{1}{2}(G_i + G_j) + H_{ij}\,(1/2)^{t-1}$$

with $H_{ij}$ the F1 heterosis of Step 4. At $t = 2$ this is the classical
result that the F2 retains half the F1 heterosis, $H_{F2} = \tfrac{1}{2} H_{F1}$,
which @falconer1996 derive in Ch. 14 for the F2 of a cross between two
populations (stated here in this chapter's notation rather than by their
equation number); the form above extends it to later selfed generations by
the same argument, since the expected heterozygosity at every locus is halved
each generation.

```{r}
homozygosity <- function(p)
  mean(pullSegSiteGeno(p, simParam = SP) %in% c(0, 2))

# Seeded here so that the |z| < 4 check below does not inherit the RNG state
# of Step 4: this z-statistic is not exactly Gaussian at this sample size, so
# an unrelated upstream edit that reseeded Step 4 could otherwise flip the
# check and break the render even though it passes for the seed used here.
set.seed(55)
ij <- c(1, 2)
midparent <- mean(gv(pop)[ij])
f1  <- makeCross(pop, matrix(ij, ncol = 2), nProgeny = 400, simParam = SP)
gen <- f1
traj <- data.frame(generation = "F1", homozygosity = homozygosity(f1),
                   meanG = as.numeric(meanG(f1)), sdG = as.numeric(sqrt(varG(f1))),
                   expected = midparent + H[["1 x 2"]])
for (t in 2:7) {
  gen  <- self(gen, nProgeny = 1, simParam = SP)
  traj <- rbind(traj, data.frame(generation = paste0("F", t),
                                 homozygosity = homozygosity(gen),
                                 meanG = as.numeric(meanG(gen)),
                                 sdG = as.numeric(sqrt(varG(gen))),
                                 expected = midparent + H[["1 x 2"]] * (1 / 2)^(t - 1)))
}
dh1 <- makeDH(f1, nDH = 1, simParam = SP)
traj <- rbind(traj, data.frame(generation = "DH", homozygosity = homozygosity(dh1),
                               meanG = as.numeric(meanG(dh1)),
                               sdG = as.numeric(sqrt(varG(dh1))),
                               expected = midparent))
traj[["z"]] <- (traj[["meanG"]] - traj[["expected"]]) / (traj[["sdG"]] / sqrt(400))
traj[["z"]][1] <- 0   # F1: no variance, and the identity of Step 4 is exact
fmt(traj, 3)
stopifnot(all(abs(traj[["z"]]) < 4))
```

Three columns, three lessons.

`homozygosity`: `self()` halves the remaining heterozygosity every generation
-- the familiar `1 - (1/2)^(t-1)` approach to fixation, which never quite
arrives. `makeDH()` gets there in one step, at 1.0 exactly, by doubling a
haploid gamete.

`meanG` against `expected`: the F2 mean falls by half the heterosis, the F3 by
three quarters, and the DH lines land on the mid-parent. Each row is 400
plants, one per F1 plant: `self(nProgeny = 1)` carries 400 parallel,
independent selfing lineages forward, so within a row the 400 values are
independent and `sdG / sqrt(400)` is the standard error of the row's mean.
The check is statistical: `z` is the observed minus expected mean in units of
that standard error, and the chunk asserts every `|z| < 4`. Across rows the
same 400 lineages are followed -- each $F_t$ plant is the selfed progeny of
one $F_{t-1}$ plant -- so successive rows are correlated, and the table is
400 lineages checked at seven generations, not seven independent samples.
Nothing about the parents changed; the loss is the heterozygosity
being halved, times $d$. This is why hybrid seed is remade from the two inbred
parents every season and why a maize F2 crop yields visibly less than the F1
it was saved from.

`sdG`: the F1 has none and the F2 has plenty. The F2 is not a variety; it is a
segregating population, and a uniform, high-yielding one cannot be recovered
from it by anything but selection and re-fixation -- which is the line
development a breeding programme does, not something a grower can do.

The saving DH buys is seasons, and seasons are the currency of a breeding
programme: six or seven selfing generations against one DH cycle. Step 8's
scheme uses `makeDH()` for precisely this reason, and the cost that never
shows up in the simulation is that DH production is a laboratory operation
with its own failure rate and its own genotype-dependent response.

The other operations a scheme needs are all one call each: `randCross()` for
random mating within a pool, `makeCross()` for a specified crossing plan,
`hybridCross()` for a full factorial or a testcross between two groups, and
`mergePops()` or `c()` to combine populations.

## Step 6 -- Phenotyping and selection: per se, then on the testcross

Selection needs a criterion. `setPheno()` supplies the noisy one; `selectInd()`
acts on it. This step does it twice: first on a line's own phenotype, as the
course does [@gorjanc2022], to check the breeder's equation; then on the testcross, which is
how a hybrid programme actually ranks lines.

### Per se selection and the breeder's equation

```{r}
be_draw <- function(q, r) {
  set.seed(500 + r)
  mx    <- setPheno(mixed, simParam = SP)
  sel   <- selectInd(mx, nInd = round(q * mx@nInd), use = "pheno", simParam = SP)
  prog  <- randCross(sel, nCrosses = 1000, simParam = SP)
  S     <- as.numeric(meanP(sel) - meanP(mx))
  vA    <- as.numeric(varA(mx, simParam = SP))
  i_emp <- S / sqrt(as.numeric(varP(mx)))
  r_emp <- stats::cor(bv(mx, simParam = SP)[, 1], pheno(mx)[, 1])
  c(i_empirical = i_emp,
    R_S_h2      = S * vA / as.numeric(varP(mx)),
    R_i_r_sdA   = i_emp * r_emp * sqrt(vA),
    R_observed  = as.numeric(meanG(prog) - meanG(mixed)))
}
n_be <- 20
sweep_q <- do.call(rbind, lapply(c(0.05, 0.10, 0.20), function(q) {
  draws <- t(sapply(seq_len(n_be), function(r) be_draw(q, r)))
  gap   <- draws[, "R_observed"] - draws[, "R_S_h2"]
  data.frame(q = q, i_normal = sel_intensity(q),
             t(colMeans(draws)),
             R_observed_sd = stats::sd(draws[, "R_observed"]),
             gap_mean = mean(gap), gap_sd = stats::sd(gap))
}))
fmt(sweep_q, 3)
```

Both forms of the breeder's equation are here, each averaged over `r n_be`
independent draws of the phenotypes, the selection and the progeny (the
population selected from is the same `mixed` every time, so `varA` is fixed
and only `S`, the accuracy and the Mendelian sampling vary). `R = S * h2` uses
the selection differential directly, with `h2` the *narrow-sense* heritability
`varA / varP` of the population selected from; `R = i * r * sigma_A`
decomposes it into intensity, accuracy and available variation -- the form
@sec-response builds the book's objective side on. Because the trait has
dominance, the accuracy `r` is the correlation of the phenotype with the
**breeding value** (`bv()`), not with the total genetic value: only the
additive part is transmitted. `R_observed` is the realised change in
*genetic* mean between the selected parents' random-mated progeny and the
starting population, which is what the prediction is a prediction of.

`gap_mean` is the paired difference between the observed response and
`R_S_h2`, with `gap_sd` its standard deviation over the draws. The mean gap is
`r fmt(100 * sweep_q[["gap_mean"]][1] / sweep_q[["R_S_h2"]][1], 1)` percent of
the prediction at `q = 0.05` and
`r fmt(100 * sweep_q[["gap_mean"]][3] / sweep_q[["R_S_h2"]][3], 1)` percent at
`q = 0.20`; its standard error over the `r n_be` draws is
`gap_sd / sqrt(`r n_be`)`, and the largest gap in units of that standard
error is `r fmt(max(abs(sweep_q[["gap_mean"]] / (sweep_q[["gap_sd"]] / sqrt(n_be)))), 1)`.
A single draw is another matter: `R_observed_sd` says that
one realisation of the response scatters by around
`r fmt(100 * sweep_q[["R_observed_sd"]][1] / sweep_q[["R_S_h2"]][1], 0)` percent
of the prediction at `q = 0.05`, which is why the course's one-off comparison
[@gorjanc2022] "did not match perfectly" and why this one is replicated. The equation
predicts the additive response to first order in the allele-frequency change;
what it leaves out is second order in that change and the dominance
contribution of the progeny's departure from Hardy-Weinberg proportions, both
small with `r 1000` progeny from `r round(0.05 * mixed@nInd)` or more parents.
The empirical intensity agrees with the normal-theory value to within
`r fmt(100 * max(abs(sweep_q[["i_empirical"]] / sweep_q[["i_normal"]] - 1)), 1)`
percent: with `r mixed@nInd` candidates and a criterion that is a sum over
`r length(a)` loci plus normal error, the two assumptions of `sel_intensity()`
-- an infinite population and a normal criterion -- are nearly met.

The book exports its own selection intensity, and AlphaSimR exports one too.
They are independent implementations of `phi(z) / q`:

```{r}
q <- 0.05
c(hybdiv = sel_intensity(q), AlphaSimR = selInt(q),
  difference = abs(sel_intensity(q) - selInt(q))) |> fmt(10)
```

`selectInd()` takes `use = "pheno"`, `"gv"` or `"ebv"`. Selecting on `"gv"` is
selecting with perfect accuracy -- impossible in a real programme, and the
correct way to compute the ceiling a scheme is being measured against.
`use = "ebv"` selects on whatever has been written into the population's `ebv`
slot, which is how genomic selection enters a simulation (@sec-programme).
`selectTop = FALSE` selects downwards, for traits where less is better.

`selectWithinFam()` is the plant-specific one. A programme that truncates
globally on a set of families will take every entry from the two best crosses
and nothing from the rest; capping the number retained per family holds the
programme's genetic base open at some cost in immediate gain. It is the same
trade-off as the `max_use` parental cap of @sec-combinatorial, applied one stage
earlier.

### Selection on the testcross: general combining ability

A maize breeder does not rank inbred lines on their own yield. A line is
crossed to a **tester** from the opposite pool, the testcross hybrids are grown
in yield trials, and the line is ranked on the mean of its testcrosses. This
subsection is the one with no autogamous counterpart at all: a self-pollinated
variety is judged, released and grown as itself, so selection there stops at
the per se phenotype of Step 6's first half. A hybrid parent is judged by a
cross it will never be sold as, which is the entire reason general and
specific combining ability exist as concepts. The
quantity being estimated is the line's **general combining ability** (GCA), the
average performance of its crosses expressed as a deviation from the mean of
all crosses (@sprague1942); what a particular cross does beyond the sum of its
two parents' GCAs is the specific combining ability (SCA). Comstock, Robinson
and Harvey's reciprocal recurrent selection (@comstock1949) is built on using
the opposite pool as the tester so that selection captures both.

On the full factorial of Step 4 the GCA of every A line against pool B can be
computed exactly, and under the model of Step 2 it has a closed form. With
$u_{ik}$ the line's allele indicator and $\bar G$ the factorial mean,

$$\mathrm{GCA}_i \;=\; \frac{1}{n_B}\sum_{j} G_{ij} - \bar G \;=\; \sum_k \alpha^{(B)}_k\,(u_{ik} - p_{Ak}),
\qquad \alpha^{(B)}_k = a_k + d_k\,(1 - 2\,p_{Bk})$$

$\alpha^{(B)}_k$ is the average effect of substituting the line's allele into a
*pool-B* background -- the same $\alpha = a + d(q - p)$ of Step 3, with the
frequency of the tester pool, not the line's own pool. That is the algebraic
content of the sentence "combining ability is a property of a line against a
tester": change the tester pool and the weights change. Its dominance part,
$\sum_k d_k (1 - 2 p_{Bk})(u_{ik} - p_{Ak})$, is the `gca_A` of
`heterosis_gca_sca()` (@sec-heterosis-gca-sca), and the interaction residual
of the factorial is that function's `sca`:

```{r}
gca_true <- tapply(gv(fact)[, 1], fact@mother, mean)[A@id] - as.numeric(meanG(fact))
gca_formula <- drop(sweep(U, 2, pA) %*% (a + d * (1 - 2 * pB)))
stopifnot(isTRUE(all.equal(as.numeric(gca_true), as.numeric(gca_formula))))

hg <- heterosis_gca_sca(U, V, d)
stopifnot(isTRUE(all.equal(as.numeric(gca_true) - as.numeric(sweep(U, 2, pA) %*% a),
                           as.numeric(hg[["gca_A"]]))))

G_ij <- tapply(gv(fact)[, 1], list(fact@mother, fact@father), mean)[A@id, B@id]
sca  <- sweep(sweep(G_ij, 1, rowMeans(G_ij)), 2, colMeans(G_ij)) + mean(G_ij)
stopifnot(isTRUE(all.equal(unname(sca), unname(hg[["sca"]]))))

var_gca <- mean(gca_true^2)
var_sca <- mean(sca^2)
c(var_GCA_A = var_gca, var_SCA = var_sca,
  cor_perse_gv_vs_GCA = stats::cor(gv(A)[, 1], gca_true)) |> fmt(4)
```

The last number is the one to hold on to. With pools that have no divergence,
$p_{Bk}$ sits near one half at most loci, $(1 - 2 p_{Bk})$ is near zero, and
the dominance contribution to GCA nearly vanishes: a line's own genetic value
ranks it for combining ability almost perfectly. Step 8 measures the same
correlation on pools with a real split, and it is lower there. Divergence is
what makes the testcross informative beyond the line itself, and it is the
reason heterotic pools are kept apart.

### How many testers, how many locations

A programme cannot cross every candidate to every line of the other pool. It
uses $t$ testers and $r$ locations, and the estimate of $\mathrm{GCA}_i$ is the
mean phenotype of the line's $t$ testcrosses, each an entry mean over $r$
plots. Writing the testcross phenotype as $\bar G + \mathrm{GCA}_i +
\mathrm{GCA}_j + \mathrm{SCA}_{ij} + e_{ij}$ with $e_{ij}$ the entry-mean error
of variance $\sigma^2_e / r$, and averaging over the $t$ testers (whose own
GCAs are common to every candidate and drop out of the ranking), the estimate
is $\widehat{\mathrm{GCA}}_i = \mathrm{GCA}_i + \bar s_i + \bar e_i$, with
$\bar s_i$ the mean of the line's SCA with its $t$ testers and $\bar e_i$ the
mean error. The statistic the chunk computes is a *sample* correlation
between $\widehat{\mathrm{GCA}}_i$ and $\mathrm{GCA}_i$ over the
$n_A =$ `r A@nInd` fixed candidate lines, with the $t$ testers drawn without
replacement from the $n_B =$ `r B@nInd` lines of pool B. To first order its
expectation is

$$\rho(t, r) \;=\; \sqrt{\frac{v_g}{v_g + v_s + v_e}}, \qquad
v_g = \frac{1}{n_A - 1}\sum_{i=1}^{n_A} \mathrm{GCA}_i^2, \quad
v_s = \frac{\sigma^2_{SCA}}{t}\cdot\frac{n_A}{n_A - 1}\cdot\frac{n_B - t}{n_B - 1}, \quad
v_e = \frac{\sigma^2_e}{t\,r}$$

where $v_g$ is the sample variance ($n_A - 1$ divisor) of the candidates'
true GCA, because the correlation is taken over exactly those $n_A$ points;
$\sigma^2_{SCA} = \frac{1}{n_A n_B}\sum_{ij} \mathrm{SCA}_{ij}^2$ is the SCA
variance on the factorial (computed above), $(n_B - t)/(n_B - 1)$ is the
finite-population correction for $t$ testers drawn without replacement from
$n_B$, $n_A/(n_A - 1)$ turns the mean of the per-candidate variances of
$\bar s_i$ into the expected sample variance, and $\sigma^2_e$ is `SP$varE`,
the plot error of Step 2. The textbook form of this prediction replaces $v_g$
by the population variance of GCA and $v_s$ by $\sigma^2_{SCA}/t$: it assumes
an infinite tester pool and population variances. It is kept in the table as
`predicted_infinite`, so the two can be compared. Neither corrects the
second-order term of a sample correlation (the Jensen term); it is measured
directly, not assumed away, in the Monte Carlo chunk below the accuracy grid,
and left uncorrected because it is small at this $n_A$. The simulation
measures the thing itself: repeated draws of the tester set and of the field
error, at every combination.

```{r}
grid  <- expand.grid(testers = c(1, 2, 4, 8), reps = c(1, 4))
n_acc <- 20
acc   <- sapply(seq_len(n_acc), function(rep) {
  set.seed(300 + rep)
  testers <- B[sample(B@nInd, max(grid[["testers"]]))]
  tc <- hybridCross(A, testers, crossPlan = "testcross", simParam = SP)
  apply(grid, 1, function(g) {
    ph  <- setPheno(tc, reps = g[["reps"]], simParam = SP)@pheno[, 1]
    use <- tc@father %in% testers@id[seq_len(g[["testers"]])]
    est <- tapply(ph[use], tc@mother[use], mean)[A@id]
    stats::cor(est, gca_true)
  })
})
# The prediction, for the statistic actually computed above: a sample
# correlation over the n_A FIXED candidates, with t testers drawn WITHOUT
# replacement from the n_B lines of pool B.
#   est_i = GCA_i + sbar_i + ebar_i
#   v_g : the candidates' sample variance of GCA (n_A - 1 divisor), because
#         the correlation is taken over exactly these n_A points;
#   v_s : sbar_i is the mean of t values from row i of sca, whose n_B entries
#         average zero, so Var(sbar_i) = (s2_i / t) (n_B - t) / (n_B - 1) with
#         s2_i = mean_j sca_ij^2; the column means of sca are zero too, so the
#         sample mean of sbar over candidates is exactly zero and the expected
#         sample variance is mean_i Var(sbar_i) n_A / (n_A - 1)
#         = var_sca n_A / (n_A - 1) (n_B - t) / (n_B - 1) / t;
#   v_e : varE / (t r), independent of everything else.
# To first order E[cor] = sqrt(v_g / (v_g + v_s + v_e)). The second-order
# (Jensen) term of the sample correlation is not corrected here; it is
# measured by Gaussian Monte Carlo in the chunk below the accuracy grid and
# reported there (`jensen_max`) rather than asserted in advance.
# `predicted_infinite` is the textbook form -- population variances, infinite
# tester pool -- kept so the two can be compared at this replication.
n_A <- A@nInd; n_B <- B@nInd
var_gca_n1 <- stats::var(gca_true)
fpc <- (n_B - grid[["testers"]]) / (n_B - 1)
v_s <- var_sca * (n_A / (n_A - 1)) * fpc / grid[["testers"]]
v_e <- SP$varE / (grid[["testers"]] * grid[["reps"]])
grid[["predicted"]] <- sqrt(var_gca_n1 / (var_gca_n1 + v_s + v_e))
grid[["predicted_infinite"]] <- sqrt(var_gca / (var_gca + var_sca / grid[["testers"]] + v_e))
grid[["observed_mean"]] <- rowMeans(acc)
grid[["observed_sd"]]   <- apply(acc, 1, stats::sd)
grid[["z"]] <- (grid[["observed_mean"]] - grid[["predicted"]]) /
  (grid[["observed_sd"]] / sqrt(n_acc))
fmt(grid, 3)
acc_max_z <- max(abs(grid[["z"]]))
acc_max_z_infinite <- max(abs(grid[["observed_mean"]] - grid[["predicted_infinite"]]) /
                            (grid[["observed_sd"]] / sqrt(n_acc)))
acc_max_gap <- max(abs(grid[["predicted"]] - grid[["predicted_infinite"]]))
c(max_abs_z = acc_max_z, max_abs_z_infinite = acc_max_z_infinite,
  max_gap_between_predictions = acc_max_gap) |> fmt(3)
```

```{r}
# The second-order term the first-order prediction leaves out: the bias of a
# sample correlation over n_A points, by Gaussian Monte Carlo per cell,
# y = x + N(0, v_s + v_e), E[cor(x, y)] minus the first-order value. In the
# table above the candidates are the SAME 30 lines in every draw, so `jensen`
# holds x fixed at gca_true and draws only the noise -- the design actually
# measured. `jensen_random_x` redraws x ~ N(0, v_g) as well, the textbook
# small-sample bias of a correlation between two random variables, which is
# a different (larger) quantity and not the one the table is exposed to.
set.seed(600)
n_mc <- 4000
cor_cols <- function(x, y) {
  xc <- sweep(x, 2, colMeans(x)); yc <- sweep(y, 2, colMeans(y))
  colSums(xc * yc) / sqrt(colSums(xc^2) * colSums(yc^2))
}
jensen_mc <- function(x_of) vapply(seq_len(nrow(grid)), function(i) {
  x <- x_of()
  y <- x + matrix(stats::rnorm(n_A * n_mc, 0, sqrt(v_s[i] + v_e[i])), n_A)
  mean(cor_cols(x, y)) - grid[["predicted"]][i]
}, numeric(1))
jensen <- jensen_mc(function() matrix(gca_true, n_A, n_mc))
jensen_random_x <- jensen_mc(function()
  matrix(stats::rnorm(n_A * n_mc, 0, sqrt(var_gca_n1)), n_A))
names(jensen) <- names(jensen_random_x) <- paste0("t", grid[["testers"]], "_r", grid[["reps"]])
jensen_max <- max(abs(jensen))
fmt(rbind(fixed_candidates = c(jensen, max_abs = jensen_max),
          random_x = c(jensen_random_x, max_abs = max(abs(jensen_random_x)))), 4)
```

The Jensen term for the design actually run (`fixed_candidates`, `gca_true`
held fixed at the `r n_A` candidates) is at most `r fmt(jensen_max, 3)` in
absolute value across the eight cells, small next to the accuracies themselves
(`r fmt(min(grid[["predicted"]]), 2)` to `r fmt(max(grid[["predicted"]]), 2)`),
which is what "left uncorrected" means quantitatively; `random_x` is the
larger, textbook small-sample bias of a correlation between two random
variables, shown only for contrast and not the quantity this chapter's
statistic carries.

Accuracy rises with both, and the two arguments buy different things. More
locations shrink only the error term; more testers shrink the error term *and*
the SCA term, because a line's specific interactions with its testers average
out. The table is `r n_acc` draws of the tester set and of the field error
per cell; the corrected and the textbook predictions differ by at most
`r fmt(acc_max_gap, 3)` here. `z` is the observed mean minus the corrected
prediction in units of the standard error of that mean over the `r n_acc`
draws; the largest `|z|` over the eight cells is `r fmt(acc_max_z, 2)` (it
would be `r fmt(acc_max_z_infinite, 2)` against the textbook form), so at
this replication the observed accuracies agree with the prediction. Each
`observed_sd` is across tester sets and error draws, on a fixed set of
candidates: what a single testcross trial's ranking is worth depends on which
testers were drawn, not only on how many; with one tester at one location the
spread across tester sets is
`r fmt(grid[["observed_sd"]][1] / grid[["observed_mean"]][1], 2)` of the
accuracy itself.

In the scheme of Step 8 the testers are `r rrs_cfg[["n_testers"]]` individuals
of the opposite pool and the trials have one location, which puts it in the
low-accuracy corner of this table. A real maize programme would use elite
inbred testers and several locations; @hallauer2010 is the reference on the
choice of tester.

## Step 7 -- Two traits, and what equal weighting assumes

`corA` makes the traits genetically correlated, which is where selection stops
being one-dimensional. This step uses an additive trait pair on its own founder
draw: index theory is written on breeding values, and the point being made
does not need dominance.

```{r}
set.seed(202)
f2 <- quickHaplo(nInd = 80, nChr = 10, segSites = 100, inbred = TRUE)
SP2 <- SimParam$new(f2)
SP2$addTraitA(nQtlPerChr = 100, mean = c(10, 10), var = c(1, 4),
              corA = matrix(c(1, -0.4, -0.4, 1), 2, 2))
SP2$setVarE(h2 = c(0.6, 0.1))

mt <- newPop(f2, simParam = SP2)
for (k in 1:3) mt <- randCross(mt, nCrosses = 1000, simParam = SP2)
mt <- setPheno(mt, simParam = SP2)

G <- varG(mt); P <- varP(mt); a2 <- c(1, 1)
c(cor_genetic = stats::cov2cor(G)[2, 1],
  cor_pheno   = stats::cov2cor(P)[2, 1],
  h2_trait1   = as.numeric(diag(G)[1] / diag(P)[1]),
  h2_trait2   = as.numeric(diag(G)[2] / diag(P)[2])) |> fmt(3)
```

The genetic correlation lands near the requested `-0.4`; the phenotypic one is
much weaker, because the environmental components were left uncorrelated
(`SP$setCorE()` is where that would change). The two traits differ in variance
and, sharply, in heritability -- the ordinary situation.

The book's pipeline defaults to `weights = NULL`, meaning equal weight on every
trait. Step 7 is where that default gets examined, using the index functions
@sec-response introduced:

```{r}
cor_with_H <- function(w)
  drop(crossprod(w, G %*% a2)) /
    sqrt(drop(crossprod(w, P %*% w)) * drop(crossprod(a2, G %*% a2)))

Bw <- rbind(equal       = a2,
            smith_hazel = index_smith_hazel(P, G, a2),
            restricted  = index_restricted(P, G, a2, R = cbind(c(0, 1))))
# expected response per unit selection intensity on the index I = b'x:
# the vector of phenotypic differentials is S = i P b / sqrt(b'Pb), and the
# multi-trait breeder's equation R = G P^-1 S gives R = i G b / sqrt(b'Pb)
dG <- t(apply(Bw, 1, function(w) drop(G %*% w) / sqrt(drop(crossprod(w, P %*% w)))))
colnames(dG) <- c("gain_trait1", "gain_trait2")
idx_tab <- cbind(Bw, dG, aggregate = rowSums(dG),
                 cor_with_H = apply(Bw, 1, cor_with_H))
fmt(idx_tab)
stopifnot(isTRUE(all.equal(idx_tab[, "aggregate"],
                           idx_tab[, "cor_with_H"] * sqrt(drop(crossprod(a2, G %*% a2))),
                           check.attributes = FALSE)))
```

`gain_trait1` and `gain_trait2` are the expected genetic change in each trait
per unit of selection intensity on the index, `G b / sqrt(b' P b)`, and
`aggregate` is their `a`-weighted sum -- which equals `cor_with_H` times the
standard deviation of the aggregate genotype, asserted. Three readings, in the
order they matter:

1. **Equal weights are not the Smith-Hazel solution.** Smith-Hazel puts
   `r fmt(idx_tab["smith_hazel", 2] / idx_tab["smith_hazel", 1], 2)` units of
   weight on trait 2 for every unit on trait 1, not 1.00. The index is
   `b = P^-1 G a` (@smith1936; @hazel1943), with no further structure, and
   `b` is proportional to `a` if and only if `a` is an eigenvector of
   `P^-1 G`. A sufficient condition is `G = c P` -- every trait with the same
   heritability `c` and the genetic correlations equal to the phenotypic ones
   -- which says nothing about the traits being equally variable or
   uncorrelated: `index_smith_hazel()`'s help page gives a correlated-trait
   pair (`G = I`, `P = [[2, 0.5], [0.5, 2]]`) for which equal weights are
   exact. Here the heritabilities are
   `r fmt(diag(G)[1] / diag(P)[1], 2)` and `r fmt(diag(G)[2] / diag(P)[2], 2)`
   and the genetic and phenotypic correlations
   `r fmt(stats::cov2cor(G)[2, 1], 2)` and `r fmt(stats::cov2cor(P)[2, 1], 2)`,
   so `G` is nowhere near a multiple of `P`, and the table shows that `(1, 1)`
   is not an eigenvector of `P^-1 G`.
2. **The cost on the objective is modest here; the composition of the gain is
   not.** Equal weighting loses
   `r fmt(100 * (1 - idx_tab["equal", "cor_with_H"] / idx_tab["smith_hazel", "cor_with_H"]), 1)`
   percent of the attainable correlation with the aggregate genotype. But it
   moves trait 1 by `r fmt(idx_tab["equal", "gain_trait1"], 3)` per unit
   intensity where Smith-Hazel moves it by
   `r fmt(idx_tab["smith_hazel", "gain_trait1"], 3)`,
   `r fmt(idx_tab["smith_hazel", "gain_trait1"] / idx_tab["equal", "gain_trait1"], 1)`
   times as much; both indices put most of the aggregate into trait 2, whose
   variance is four times larger, and the weights decide how much trait 1
   moves at all. An index that is nearly as good on the objective can still
   land the gain somewhere else.
3. **The restricted index holds trait 2 at exactly zero**, at a cost of
   `r fmt(100 * (1 - idx_tab["restricted", "aggregate"] / idx_tab["smith_hazel", "aggregate"]), 0)`
   percent of the aggregate. That is Kempthorne and Nordskog's construction
   (@kempthorne1959) and, as @sec-response argues, the classical ancestor of
   this book's `alpha` constraint -- restrict a direction rather than price it.

Smith-Hazel is defined as the weight vector maximising that correlation, so the
claim is checkable against a numerical optimum rather than taken on faith:

```{r}
opt <- stats::optim(c(1, 1), function(w) -cor_with_H(w), method = "BFGS")
c(numerical = -opt$value, smith_hazel = cor_with_H(Bw["smith_hazel", ]),
  difference = abs(-opt$value - cor_with_H(Bw["smith_hazel", ]))) |> fmt(10)
```

Multi-trait genomic selection in wheat, balancing yield against protein content
and quality, is the standing plant example -- @michel2019. In the hybrid scheme
below only trait 1 enters the ranking; the second trait rides along in the
contract so that `stage2_select()` has an index to build.

## Step 8 -- The parts assembled: reciprocal recurrent selection

Everything above was one operation at a time. A breeding scheme is those
operations in a loop, and the loop for two heterotic pools is reciprocal
recurrent selection: each pool is improved using the *other* pool as tester.
Doubled-haploid lines are drawn from pool A, testcrossed to pool B (and vice
versa), the testcross hybrids are phenotyped, and the lines with the best mean
testcross performance -- the GCA estimate of Step 6 -- are recombined to found
the next cycle's pool. Elite lines from both pools, at the end, are crossed
directly to produce the commercial hybrids. It is exactly the asymmetry
@sec-cycles argues for: selection happens on the hybrid, recycling happens
within pool.

### Founding two pools with a real coalescent history

This is the one place the chapter pays for `runMacs2()` rather than
`quickHaplo()`. `AlphaSimR::runMacs2(..., split = )` draws founder haplotypes
from a population-split coalescent history: one ancestral population, then two
demes of equal size that stopped exchanging migrants `split` generations ago,
with the first half of the sampled individuals from one deme and the second
half from the other. That is the realistic analogue of `simulate_lines()`'s
Balding-Nichols draw in @sec-simulating-crosses, and between-pool divergence
is the entire subject of @sec-subdivided, so a stylised substitute would not do.

```{r}
n_split_draws <- 20   # founder draws per split time, below
n_dh_gca      <- 10   # DH lines per founder for the GCA measurement, below
fp <- alphasimr_founder_pools(rrs_cfg)
founder_geno <- pullSegSiteGeno(c(fp[["pop_A"]], fp[["pop_B"]]), simParam = fp[["SP"]])
c(pool_A = fp[["pop_A"]]@nInd, pool_B = fp[["pop_B"]]@nInd,
  markers = ncol(founder_geno),
  heterozygous_fraction = mean(founder_geno == 1),
  gd_A = alphasimr_gd(fp[["pop_A"]], fp[["SP"]]),
  gd_B = alphasimr_gd(fp[["pop_B"]], fp[["SP"]])) |> fmt(3)
```

Reading `alphasimr_founder_pools()` against Steps 1 and 2, it is
`runMacs2(nInd = , nChr = , segSites = , split = )` followed by
`SP$addTraitAD(nQtlPerChr = seg_sites, meanDD = `r rrs_cfg[["mean_dd"]]`,
varDD = `r rrs_cfg[["var_dd"]]`)` -- the same dominance model as Step 2 -- and
`SP$setVarE(h2 = `r rrs_cfg[["h2"]]`)`. Two things about it must be said
before any number is read. **The founders are outbred**: `runMacs2()` is called
without `inbred = TRUE`, so each pool is a sample of a random-mating
population, heterozygous at the fraction of sites printed above, as an
open-pollinated maize population would be; the inbred lines of the scheme are
made from them by `makeDH()` inside each cycle. And there is no separate
marker panel: every one of the `r rrs_cfg[["seg_sites"]]` segregating sites per
chromosome is a QTL, so the sites that carry the trait are the ones
`pullSegSiteGeno()` reads for `X`, `f` and `fL` throughout. `alphasimr_gd()`
is `1 - mean(f)`, the exact definition `gene_diversity()` uses, computed from
that pull, so the two pools are already on the same scale as every diversity
number in the rest of the book.

Because these founders come from MaCS, no number in the rest of this section
is reproducible from its seed -- see Step 1: the founders move between runs,
and everything below is computed from them. The chapter's response is to
replicate every Step 8 measurement over independent founder draws and report
the spread.

**What `split` does.** The bridge's default is `split_gen = `r rrs_cfg[["split_gen"]]``.
The sweep below draws `r n_split_draws` independent founder sets at each of four split
times and measures three things on each: the Caballero & Toro partition of
@sec-subdivided over the two pools (`ct_partition()`, the general form of
`gd_partition()`: the $F$-statistics of their Eq. 6 and the partition
$GD_T = GD_{WI} + GD_{BI} + GD_{BS}$ of their Eq. 7, from which
$F_{ST} = GD_{BS} / GD_T$ follows as an identity derived in @sec-subdivided,
not stated in the paper), the share of expected heterosis that the divergence
term carries (`heterosis_partition()`, @sec-heterosis), and -- the Step 6
question with divergence -- how much of a DH line's exact GCA against the
other pool's DH lines its own genetic value can see. For that last one,
`r n_dh_gca` DH lines are drawn from every founder, the full factorial between
the two DH sets is made, and the GCA of each pool-A line is split into its
additive and dominance parts with `heterosis_gca_sca()`; the identity behind
that split was asserted in Step 6. The chunk also records, per draw, an
ascertainment bookkeeping that the reading of `GD_T` will need: the fraction
of sites monomorphic within each pool, the fraction fixed for opposite
alleles in the two pools, and the part of `GD_T` those sites carry.

```{r}
split_draw <- function(split, seed) {
  cfg <- utils::modifyList(rrs_cfg, list(split_gen = split, seed = seed))
  fp_s <- alphasimr_founder_pools(cfg)
  SP_s <- fp_s[["SP"]]
  pool <- rep(c("A", "B"), c(cfg[["n_pool_A"]], cfg[["n_pool_B"]]))
  geno <- pullSegSiteGeno(c(fp_s[["pop_A"]], fp_s[["pop_B"]]), simParam = SP_s)
  ct   <- ct_partition(geno, pool)
  pA_s <- colMeans(geno[pool == "A", ]) / 2
  pB_s <- colMeans(geno[pool == "B", ]) / 2
  d_s  <- SP_s$traits[[1]]@domEff
  a_s  <- SP_s$traits[[1]]@addEff
  hp   <- heterosis_partition(pA_s, pB_s, d_s)
  # Ascertainment: runMacs2() returns sites that segregate in the POOLED
  # sample, so a site fixed for different alleles in the two pools is kept
  # and, with equal pool sizes, sits at pooled frequency 1/2 -- the maximum
  # of 2 p q. GD_T = mean 2 pbar qbar exactly (census weights, equal sizes),
  # so it splits into the part those fixed differences carry and the rest.
  mono_A <- pA_s %in% c(0, 1)
  mono_B <- pB_s %in% c(0, 1)
  fixed  <- mono_A & mono_B & (pA_s != pB_s)
  pbar_s <- (pA_s + pB_s) / 2
  h_s    <- 2 * pbar_s * (1 - pbar_s)
  stopifnot(isTRUE(all.equal(ct[["diversity"]][["GD_T"]], mean(h_s))))
  set.seed(seed)
  dh_A <- makeDH(fp_s[["pop_A"]], nDH = n_dh_gca, simParam = SP_s)
  dh_B <- makeDH(fp_s[["pop_B"]], nDH = n_dh_gca, simParam = SP_s)
  hyb  <- hybridCross(dh_A, dh_B, crossPlan = "testcross", simParam = SP_s)
  gca  <- tapply(gv(hyb)[, 1], hyb@mother, mean)[dh_A@id] - as.numeric(meanG(hyb))[1]
  U_s  <- pullQtlGeno(dh_A, simParam = SP_s) / 2
  hg   <- heterosis_gca_sca(U_s, pullQtlGeno(dh_B, simParam = SP_s) / 2, d_s)
  # GCA = additive part + dominance part (the Step 6 identity), so its
  # variance is var(add) + var(dom) + 2 cov(add, dom): three terms, not two.
  gca_add <- as.numeric(sweep(U_s, 2, colMeans(U_s)) %*% a_s)
  gca_dom <- as.numeric(hg[["gca_A"]])
  stopifnot(isTRUE(all.equal(as.numeric(gca), gca_add + gca_dom)))
  data.frame(split = split, seed = seed,
             F_ST  = ct[["global"]][["F_ST"]],
             GD_T  = ct[["diversity"]][["GD_T"]],
             GD_BS = ct[["diversity"]][["GD_BS"]],
             mono_A = mean(mono_A), mono_B = mean(mono_B),
             fixed_diff = mean(fixed),
             GD_T_fixed = sum(h_s[fixed]) / length(h_s),
             GD_T_other = sum(h_s[!fixed]) / length(h_s),
             maf_pooled = mean(pmin(pbar_s, 1 - pbar_s)),
             heterosis_divergence_share = hp[["divergence"]] / hp[["total"]],
             GCA_var = stats::var(gca),
             GCA_var_add = stats::var(gca_add),
             GCA_var_dom = stats::var(gca_dom),
             GCA_cov2 = 2 * stats::cov(gca_add, gca_dom),
             GCA_dominance_share = stats::var(gca_dom) / stats::var(gca),
             cor_perse_GCA = stats::cor(gv(dh_A)[, 1], gca))
}
splits <- c(0, 10, 100, 1000)
sw <- do.call(rbind, lapply(splits, function(s)
  do.call(rbind, lapply(seq_len(n_split_draws), function(r)
    split_draw(s, seed = 100 * s + r)))))
sw_mean <- aggregate(sw[, -(1:2)], by = list(split = sw[["split"]]), FUN = mean)
sw_sd   <- aggregate(sw[, -(1:2)], by = list(split = sw[["split"]]), FUN = stats::sd)
rownames(sw_mean) <- rownames(sw_sd) <- paste0("split_", sw_mean[["split"]])
cols_div <- c("split", "F_ST", "GD_T", "GD_BS", "heterosis_divergence_share")
cols_asc <- c("split", "GD_T", "mono_A", "mono_B", "fixed_diff", "GD_T_fixed",
              "GD_T_other", "maf_pooled")
cols_gca <- c("split", "GCA_var", "GCA_var_add", "GCA_var_dom", "GCA_cov2",
              "GCA_dominance_share", "cor_perse_GCA")
fmt(sw_mean[, cols_div], 3)
fmt(sw_sd[, cols_div], 3)
fmt(sw_mean[, cols_asc], 3)
fmt(sw_mean[, cols_gca], 3)
fmt(sw_sd[, cols_gca], 3)
```

```{r}
# the contrast between the two extreme split times, for the two GCA columns
contrast <- function(m) {
  x <- sw[[m]][sw[["split"]] == max(splits)]
  y <- sw[[m]][sw[["split"]] == min(splits)]
  se <- sqrt(stats::var(x) / length(x) + stats::var(y) / length(y))
  c(at_split_max = mean(x), at_split_0 = mean(y), difference = mean(x) - mean(y),
    se = se, t = (mean(x) - mean(y)) / se)
}
ctr <- rbind(GCA_dominance_share = contrast("GCA_dominance_share"),
             cor_perse_GCA       = contrast("cor_perse_GCA"))
fmt(ctr, 3)
```

```{r}
#| echo: false
# prose helpers: the ratio var(dom)/var(GCA) with only one part moved from
# its split-0 to its split-1000 value (covariance held at split 0), and the
# covariance term in standard errors of zero
gca_only <- function(move) {
  g0 <- sw_mean["split_0", ]; g1 <- sw_mean["split_1000", ]
  add <- if (move == "add") g1[["GCA_var_add"]] else g0[["GCA_var_add"]]
  dom <- if (move == "dom") g1[["GCA_var_dom"]] else g0[["GCA_var_dom"]]
  dom / (add + dom + g0[["GCA_cov2"]])
}
cov_frac <- abs(sw_mean[["GCA_cov2"]]) / sw_mean[["GCA_var"]]      # |2 cov| as a share of var(GCA)
cov_se   <- abs(sw_mean[["GCA_cov2"]]) / (sw_sd[["GCA_cov2"]] / sqrt(n_split_draws))
names(cov_frac) <- names(cov_se) <- rownames(sw_mean)
cov_frac_max <- max(cov_frac)
cov_se_max   <- max(cov_se)
dom_change   <- sw_mean["split_1000", "GCA_var_dom"] - sw_mean["split_0", "GCA_var_dom"]
```

Read the first two tables together: the first is the mean over the
`r n_split_draws` draws, the second their standard deviation. What was
measured: `F_ST` rises from `r fmt(sw_mean["split_0", "F_ST"], 3)` at
`split = 0` to `r fmt(sw_mean["split_1000", "F_ST"], 3)` at `split = 1000`,
and `GD_BS` from `r fmt(sw_mean["split_0", "GD_BS"], 3)` to
`r fmt(sw_mean["split_1000", "GD_BS"], 3)`, each step many standard
deviations wide; the divergence share of expected heterosis rises from
`r fmt(sw_mean["split_0", "heterosis_divergence_share"], 3)` to
`r fmt(sw_mean["split_1000", "heterosis_divergence_share"], 3)`. `GD_T` does
not stay flat: it is `r fmt(sw_mean["split_0", "GD_T"], 3)` at `split = 0`,
`r fmt(sw_mean["split_100", "GD_T"], 3)` at `split = 100` and
`r fmt(sw_mean["split_1000", "GD_T"], 3)` at `split = 1000`, and the rise
from `split = 100` to `split = 1000`,
`r fmt(sw_mean["split_1000", "GD_T"] - sw_mean["split_100", "GD_T"], 3)`, is
`r fmt((sw_mean["split_1000", "GD_T"] - sw_mean["split_100", "GD_T"]) / sqrt((sw_sd["split_1000", "GD_T"]^2 + sw_sd["split_100", "GD_T"]^2) / n_split_draws), 0)`
standard errors at this replication. The third table is the bookkeeping
behind that rise. Across the same range the fraction of sites monomorphic
within pool A goes from `r fmt(sw_mean["split_0", "mono_A"], 3)` to
`r fmt(sw_mean["split_1000", "mono_A"], 3)` (pool B:
`r fmt(sw_mean["split_0", "mono_B"], 3)` to
`r fmt(sw_mean["split_1000", "mono_B"], 3)`), the fraction of sites fixed for
opposite alleles in the two pools from
`r fmt(sw_mean["split_0", "fixed_diff"], 3)` to
`r fmt(sw_mean["split_1000", "fixed_diff"], 3)`, the part of `GD_T` those
sites carry (`GD_T_fixed`) from `r fmt(sw_mean["split_0", "GD_T_fixed"], 3)`
to `r fmt(sw_mean["split_1000", "GD_T_fixed"], 3)`, and the part carried by
every other site (`GD_T_other`) falls from
`r fmt(sw_mean["split_0", "GD_T_other"], 3)` to
`r fmt(sw_mean["split_1000", "GD_T_other"], 3)`.

What is inferred from those three tables: `runMacs2()` returns the sites that
segregate in the *pooled* sample of the two demes, so a site fixed for
opposite alleles in the two pools is kept and, with equal pool sizes, sits at
pooled frequency one half -- the maximum of $2\bar p \bar q$, which is exactly
what `GD_T` averages over sites (asserted in the chunk). The rise in `GD_T`
therefore comes entirely from sites fixed for opposite alleles, an
ascertainment effect of how the founder sites are sampled, while diversity at
every other site falls with divergence. At those other sites the reading is
the one @sec-heterosis gives for expected heterosis: drift moves diversity
from within pools to between them and does not create it. How much of `GD_T`
a marker panel shows under divergence therefore depends on how its sites were
ascertained, which is a property of the panel and not of the pools.

The fourth and fifth tables are the combining-ability side, mean and standard
deviation, and the sixth is the contrast between the two extreme split times,
unpaired because the draws are independent. GCA is the sum of its additive and
dominance parts (asserted in the chunk), so its variance is
$\mathrm{var}(\mathrm{add}) + \mathrm{var}(\mathrm{dom}) + 2\,\mathrm{cov}(\mathrm{add}, \mathrm{dom})$,
three terms. What was measured, from `split = 0` to `split = 1000`: `GCA_var`
goes from `r fmt(sw_mean["split_0", "GCA_var"], 3)` to
`r fmt(sw_mean["split_1000", "GCA_var"], 3)`, `GCA_var_add` from
`r fmt(sw_mean["split_0", "GCA_var_add"], 3)` to
`r fmt(sw_mean["split_1000", "GCA_var_add"], 3)`, `GCA_var_dom` from
`r fmt(sw_mean["split_0", "GCA_var_dom"], 3)` to
`r fmt(sw_mean["split_1000", "GCA_var_dom"], 3)`, and `GCA_cov2`
($2\,\mathrm{cov}$) from `r fmt(sw_mean["split_0", "GCA_cov2"], 3)` to
`r fmt(sw_mean["split_1000", "GCA_cov2"], 3)`, the last with between-draw
standard deviations of `r fmt(sw_sd["split_0", "GCA_cov2"], 3)` and
`r fmt(sw_sd["split_1000", "GCA_cov2"], 3)`, so at every split it is within
`r fmt(cov_se_max, 1)` standard errors of zero. The ratio
`GCA_dominance_share` $= \mathrm{var}(\mathrm{dom}) / \mathrm{var}(\mathrm{GCA})$
rises from `r fmt(ctr["GCA_dominance_share", "at_split_0"], 3)` to
`r fmt(ctr["GCA_dominance_share", "at_split_max"], 3)`
(`t = `r fmt(ctr["GCA_dominance_share", "t"], 1)``). Which part drives it:
holding the dominance part at its `split = 0` value and moving only the
additive part to its `split = 1000` value gives a ratio of
`r fmt(gca_only("add"), 3)`; holding the additive part and moving only the
dominance part gives `r fmt(gca_only("dom"), 3)`. And the per
se-versus-GCA correlation falls from `r fmt(ctr["cor_perse_GCA", "at_split_0"], 3)`
to `r fmt(ctr["cor_perse_GCA", "at_split_max"], 3)`
(`t = `r fmt(ctr["cor_perse_GCA", "t"], 1)``). Both correlations are below the
`r fmt(stats::cor(gv(A)[, 1], gca_true), 3)` of Step 6's undifferentiated
halves, but that comparison crosses a change of founders, trait draw and
line number, so the within-sweep contrast is the one that isolates `split`.

What is inferred: in this render the
`r if (gca_only("add") >= gca_only("dom")) "fall of the *additive* part of the GCA variance accounts for more of the rise in the ratio than the growth of the dominance part" else "growth of the dominance part of the GCA variance accounts for more of the rise in the ratio than the fall of the additive part"`;
the additive part falls with divergence because the sites that fix within a
pool stop contributing to it, so the ratio can rise without the dominance
part growing at all. The ratio is a partition of $\mathrm{var}(\mathrm{GCA})$
only insofar as the covariance term is negligible; here that term is within
`r fmt(cov_se_max, 1)` standard errors of zero at every split, which is the
only sense in which "share" is used. As the pools separate, the dominance
weights $(1 - 2 p_{Bk})$ of Step 6 move away from zero and a growing part of
a line's combining ability is invisible to its own genetic value; the evidence
for that at this scale is the `t` of the contrast, and nothing here separates
how much of the between-draw spread is the trait drawn with each founder set
and how much is the DH sampling. The limit of the inference is the scale:
`r rrs_cfg[["n_pool_A"]]` individuals per pool,
`r rrs_cfg[["n_chr"]] * rrs_cfg[["seg_sites"]]` sites, `r n_split_draws`
draws per split, and a `split` in coalescent generations that is not
calibrated to any named heterotic-pool pair. The direction transfers; the
numbers do not.

### One cycle

```{r}
set.seed(11)
cy1 <- alphasimr_rrs_cycle(fp[["pop_A"]], fp[["pop_B"]], fp[["SP"]], rrs_cfg)
fmt(cy1[["summary"]], 4)
```

`alphasimr_rrs_cycle()` runs the same six steps for each pool, symmetrically
(pool A tested on B, pool B tested on A), and every one is a primitive from
above:

1. **Double haploids** (Step 5). `makeDH()` draws `r rrs_cfg[["n_dh"]]` fully
   homozygous lines from *each* individual in the current pool -- at book scale,
   `r rrs_cfg[["n_pool_A"]]` parents x `r rrs_cfg[["n_dh"]]` DH each =
   `r rrs_cfg[["n_pool_A"]] * rrs_cfg[["n_dh"]]` DH lines per pool, per cycle.
   Because the pool is outbred, the DH lines from one parent differ from each
   other: this is the lottery of Step 4, drawn one generation earlier than in
   a line programme.
2. **Pick reciprocal testers.** The first `r rrs_cfg[["n_testers"]]` individuals
   of the *current* opposite pool are used as testers -- an arbitrary slice by
   index, not a random or merit-based draw, outbred individuals rather than
   inbred lines, and a different set each cycle because step 6 recreates the
   pool. This is the RRS-defining move: a line's value is judged by its cross onto the other
   heterotic pool, not by itself.
3. **Testcross and phenotype** (Step 6). `hybridCross()` forms the full
   factorial between every DH line and every tester --
   `r rrs_cfg[["n_pool_A"]] * rrs_cfg[["n_dh"]]` DH x `r rrs_cfg[["n_testers"]]`
   testers = `r rrs_cfg[["n_pool_A"]] * rrs_cfg[["n_dh"]] * rrs_cfg[["n_testers"]]`
   testcross hybrids per pool -- and `setPheno()` adds phenotyping error sized
   to the heritability set by `SP$setVarE(h2 = )`, at one location.
4. **GCA estimate** (Step 6). Testcross phenotypes are grouped by DH-line
   parent (`tapply(pheno[, 1], mother, mean)`) and averaged across the
   `r rrs_cfg[["n_testers"]]` testers each line was crossed to. Only trait 1 is
   used for this ranking even though this chapter's config simulates
   `r rrs_cfg[["n_traits"]]` traits per line; the second trait rides along in
   `X`/`hybrids` but plays no role in selection here. Step 7 is what it would
   take to use both.
5. **Keep the elites** (Step 6). The `r rrs_cfg[["n_sel"]]` DH lines per pool
   with the highest mean testcross phenotype are kept as `elite_A`/`elite_B`.
6. **Recombine** (Step 5). `randCross()` random-mates the elites back up to
   `r rrs_cfg[["n_pool_A"]]` individuals, founding the next cycle's pool A
   (symmetrically for B).

`elite_A`/`elite_B` are what step 6 mates -- and, in the *last* cycle, what
the final commercial cross uses instead of being recombined again. The cycle's
`summary` mixes two different populations on purpose: `mean_tc_A`/`mean_tc_B`
average over *all* DH lines tested that cycle (the population's testcross
performance before selection), while `gd_A`/`gd_B` are gene diversity of only
the `r rrs_cfg[["n_sel"]]` surviving elites (the population after selection).
That asymmetry is exactly what the next section plots -- response measured
pre-selection, diversity measured post-selection.

### The trajectory across cycles, replicated, against a paired control

`alphasimr_pipeline()` runs `alphasimr_rrs_cycle()` in a loop for
`r rrs_cfg[["n_cycles"]]` cycles, carrying `pop_A`/`pop_B` forward from one
cycle's output to the next's input. Only the *last* cycle's
`elite_A`/`elite_B` go on to found the commercial hybrid pool; every earlier
cycle's elites are spent entirely on `randCross()`ing the next cycle's pool.

One run of that loop is an anecdote, and it would be an anecdote about one
founder draw. The loop is written out here instead, so that three things can
be added to it: a **paired no-selection control** (`random_elite = TRUE`,
which draws the elites at random and leaves every other operation -- the DH
lines, the testcrosses, the recombination -- identical, so the difference
between the two arms is selection and nothing else), the between-pool
partition of the *recombined* pools each cycle, and the genetic value of the
actual product, the elite factorial.

```{r}
run_rrs <- function(fp, cfg, random_elite = FALSE, seed = 1) {
  set.seed(seed)
  SP_r  <- fp[["SP"]]
  pop_A <- fp[["pop_A"]]; pop_B <- fp[["pop_B"]]
  pool  <- rep(c("A", "B"), c(cfg[["n_pool_A"]], cfg[["n_pool_B"]]))
  pools_row <- function(pop_A, pop_B) {
    ct <- ct_partition(pullSegSiteGeno(c(pop_A, pop_B), simParam = SP_r), pool)
    data.frame(gd_pool_A = alphasimr_gd(pop_A, SP_r),
               gd_pool_B = alphasimr_gd(pop_B, SP_r),
               GD_BS = ct[["diversity"]][["GD_BS"]],
               F_ST  = ct[["global"]][["F_ST"]])
  }
  rows <- list(cbind(cycle = 0, mean_tc_A = NA, mean_tc_B = NA, gd_A = NA, gd_B = NA,
                     hybrid_meanG = NA, pools_row(pop_A, pop_B)))
  for (i in seq_len(cfg[["n_cycles"]])) {
    cy <- alphasimr_rrs_cycle(pop_A, pop_B, SP_r, cfg, random_elite = random_elite)
    pop_A <- cy[["pop_A"]]; pop_B <- cy[["pop_B"]]
    prod  <- hybridCross(cy[["elite_A"]], cy[["elite_B"]], crossPlan = "testcross",
                         simParam = SP_r)
    rows[[i + 1]] <- cbind(cycle = i, cy[["summary"]],
                           hybrid_meanG = as.numeric(meanG(prod))[1],   # trait 1
                           pools_row(pop_A, pop_B))
  }
  do.call(rbind, rows)
}

n_rep <- 48
arms <- lapply(seq_len(n_rep), function(s) {
  fp_s <- alphasimr_founder_pools(utils::modifyList(rrs_cfg, list(seed = s)))
  list(sel = run_rrs(fp_s, rrs_cfg, random_elite = FALSE, seed = s),
       ctl = run_rrs(fp_s, rrs_cfg, random_elite = TRUE,  seed = s))
})
# pairing check: cycle 1 is identical between arms up to the elite pick
stopifnot(all(vapply(arms, function(a)
  isTRUE(all.equal(a[["sel"]][c("mean_tc_A", "mean_tc_B")][2, ],
                   a[["ctl"]][c("mean_tc_A", "mean_tc_B")][2, ])), logical(1))))
```

`r n_rep` founder draws; each drawn once and both arms branched off it, with
the same seed set before each arm so that cycle 1's DH lines and testcross
phenotypes are identical between arms (asserted) and the two only diverge at
the selection step. The first replicate's selected arm, cycle by cycle:

```{r}
fmt(arms[[1]][["sel"]], 3)
```

`gd_pool_A`/`gd_pool_B` are the gene diversity of the recombined pools that
enter the next cycle, on the same `1 - mean(f)` scale as the elites' `gd_A`/
`gd_B`; `GD_BS` and `F_ST` are the between-pool components of those two pools;
`hybrid_meanG` is the mean genetic value of the `r rrs_cfg[["n_sel"]]` x
`r rrs_cfg[["n_sel"]]` elite factorial, the product the cycle would sell. All
of them, over all replicates, at the last cycle, selected minus control:

```{r}
last <- rrs_cfg[["n_cycles"]] + 1
# The two pools are exchangeable by design, so their mean is a legitimate
# pooled statistic; how much noise it removes is measured below (pool_tab).
arms <- lapply(arms, function(a) lapply(a, function(x) {
  x[["gd_elite_mean"]] <- (x[["gd_A"]] + x[["gd_B"]]) / 2
  x[["gd_pool_mean"]]  <- (x[["gd_pool_A"]] + x[["gd_pool_B"]]) / 2
  x
}))
metrics <- c("mean_tc_A", "mean_tc_B", "hybrid_meanG", "gd_A", "gd_B",
             "gd_elite_mean", "gd_pool_A", "gd_pool_B", "gd_pool_mean",
             "GD_BS", "F_ST")
arm_last <- function(arm, m) vapply(arms, function(a) a[[arm]][[m]][last], numeric(1))
diffs    <- function(m) arm_last("sel", m) - arm_last("ctl", m)   # one per founder draw
paired <- function(m) {
  dd <- diffs(m)
  c(selected = mean(arm_last("sel", m)), control = mean(arm_last("ctl", m)),
    diff_mean = mean(dd), diff_sd = stats::sd(dd),
    t = mean(dd) / (stats::sd(dd) / sqrt(length(dd))))
}
rep_tab  <- t(sapply(metrics, paired))
t_paired <- rep_tab[, "t"]                       # named: t_paired[["gd_elite_mean"]]
t_crit   <- stats::qt(0.975, df = n_rep - 1)
sig_paired <- abs(t_paired) > t_crit             # named logical, same names
fmt(rep_tab, 3)
# Pooling the two pools halves the variance only if their paired differences
# are uncorrelated across founder draws: measured, not assumed.
pool_pair <- function(mA, mB) {
  dA <- diffs(mA); dB <- diffs(mB)
  c(cor_AB = stats::cor(dA, dB), sd_A = stats::sd(dA), sd_B = stats::sd(dB),
    sd_pooled = stats::sd((dA + dB) / 2),
    sd_if_independent = sqrt((stats::var(dA) + stats::var(dB)) / 4))
}
pool_tab <- rbind(gd_elite = pool_pair("gd_A", "gd_B"),
                  gd_pool  = pool_pair("gd_pool_A", "gd_pool_B"),
                  mean_tc  = pool_pair("mean_tc_A", "mean_tc_B"))
fmt(pool_tab, 3)
stopifnot(isTRUE(all.equal(pool_tab["gd_elite", "sd_pooled"],
                           rep_tab["gd_elite_mean", "diff_sd"])))
```

```{r}
loss <- function(arm, m) vapply(arms, function(a)
  (a[[arm]][[m]][1] - a[[arm]][[m]][last]) / a[[arm]][[m]][1], numeric(1))
loss_tab <- rbind(
  `pool A, selected` = c(mean = mean(loss("sel", "gd_pool_A")), sd = stats::sd(loss("sel", "gd_pool_A"))),
  `pool A, control`  = c(mean = mean(loss("ctl", "gd_pool_A")), sd = stats::sd(loss("ctl", "gd_pool_A"))),
  `pool B, selected` = c(mean = mean(loss("sel", "gd_pool_B")), sd = stats::sd(loss("sel", "gd_pool_B"))),
  `pool B, control`  = c(mean = mean(loss("ctl", "gd_pool_B")), sd = stats::sd(loss("ctl", "gd_pool_B"))))
fmt(loss_tab, 3)
fig_cap <- paste0("Reciprocal recurrent selection over ", rrs_cfg[["n_cycles"]],
  " cycles at book scale, ", n_rep, " founder draws, selection (solid) against ",
  "the paired no-selection control (dashed); points are means over draws and ",
  "bars one standard deviation. Left: mean testcross phenotype per pool, ",
  "measured on all DH lines tested that cycle. Right: gene diversity of the ",
  "elite lines retained each cycle.")
```

```{r}
#| label: fig-alphasimr-cycles
#| echo: false
#| fig-width: 9
#| fig-height: 4.2
#| fig-cap: !expr fig_cap
arm_stat <- function(arm, m, f) sapply(seq_len(rrs_cfg[["n_cycles"]]), function(i)
  f(vapply(arms, function(a) a[[arm]][[m]][i + 1], numeric(1))))
cyc  <- seq_len(rrs_cfg[["n_cycles"]])
cols <- c(A = "#1b6ca8", B = "#c0392b")
panel <- function(mA, mB, ylab, main, legend_pos) {
  st <- lapply(c(sel = "sel", ctl = "ctl"), function(arm)
    list(A = arm_stat(arm, mA, mean), B = arm_stat(arm, mB, mean),
         sA = arm_stat(arm, mA, stats::sd), sB = arm_stat(arm, mB, stats::sd)))
  ylim <- range(unlist(lapply(st, function(s)
    c(s[["A"]] - s[["sA"]], s[["A"]] + s[["sA"]], s[["B"]] - s[["sB"]], s[["B"]] + s[["sB"]]))))
  plot(cyc, st[["sel"]][["A"]], type = "n", ylim = ylim, xlab = "Cycle", ylab = ylab,
       main = main, xaxt = "n")
  axis(1, at = cyc)
  for (arm in c("sel", "ctl")) {
    lty <- if (arm == "sel") 1 else 2
    off <- if (arm == "sel") -0.04 else 0.04
    for (p in c("A", "B")) {
      x <- cyc + off; y <- st[[arm]][[p]]; s <- st[[arm]][[paste0("s", p)]]
      lines(x, y, type = "b", pch = 19, lty = lty, col = cols[p])
      arrows(x, y - s, x, y + s, angle = 90, code = 3, length = 0.03, col = cols[p])
    }
  }
  legend(legend_pos, c("pool A, selected", "pool B, selected",
                       "pool A, control", "pool B, control"),
         col = cols[c("A", "B", "A", "B")], lty = c(1, 1, 2, 2), pch = 19,
         bty = "n", cex = 0.75)
}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.3, 2.2, 0.8))
panel("mean_tc_A", "mean_tc_B", "Mean testcross phenotype", "Selection response", "topleft")
panel("gd_A", "gd_B", "Gene diversity of the elite lines", "Diversity erosion", "topright")
par(op)
```

```{r}
#| echo: false
# prose helpers: every verdict below is decided by the render's own t
verdict <- function(m, yes = "is", no = "is not") if (sig_paired[[m]]) yes else no
sgn <- function(x, d = 3) sprintf(paste0("%+.", d, "f"), x)
```

What was measured (@fig-alphasimr-cycles and the tables): over `r n_rep`
paired replicates, the selected arm's mean testcross phenotype at cycle
`r rrs_cfg[["n_cycles"]]` differs from the control's by
`r sgn(rep_tab["mean_tc_A", "diff_mean"], 2)` (pool A) and
`r sgn(rep_tab["mean_tc_B", "diff_mean"], 2)` (pool B) trait units, with paired
standard deviations of `r fmt(rep_tab["mean_tc_A", "diff_sd"], 2)` and
`r fmt(rep_tab["mean_tc_B", "diff_sd"], 2)`; the elite factorial's genetic
value -- the product -- differs by `r sgn(rep_tab["hybrid_meanG", "diff_mean"], 2)`
(`t = `r fmt(t_paired[["hybrid_meanG"]], 1)``). Gene diversity of the elite
lines, averaged over the two pools, differs by
`r sgn(rep_tab["gd_elite_mean", "diff_mean"])` (paired s.d.
`r fmt(rep_tab["gd_elite_mean", "diff_sd"], 3)`, `t = `r fmt(t_paired[["gd_elite_mean"]], 1)``),
and that of the recombined pools by
`r sgn(rep_tab["gd_pool_mean", "diff_mean"])` (`t = `r fmt(t_paired[["gd_pool_mean"]], 1)``).
Between the pools, `GD_BS` differs by `r sgn(rep_tab["GD_BS", "diff_mean"])`
(`t = `r fmt(t_paired[["GD_BS"]], 1)``) and `F_ST` by
`r sgn(rep_tab["F_ST", "diff_mean"])` (`t = `r fmt(t_paired[["F_ST"]], 1)``).
The two-sided 5% point of `t` with `r n_rep - 1` degrees of freedom is
`r fmt(t_crit, 2)`. Against it, in this render, the response -- both pools
and the product --
`r if (all(sig_paired[c("mean_tc_A", "mean_tc_B", "hybrid_meanG")])) "is beyond it" else "is not beyond it"`;
the pooled elite-line gene-diversity difference
`r verdict("gd_elite_mean", "is beyond it", "is not beyond it")`; the pooled
recombined-pool difference
`r verdict("gd_pool_mean", "is beyond it", "is not beyond it")`; the `GD_BS`
difference `r verdict("GD_BS", "is beyond it", "is not beyond it")`; the
`F_ST` difference `r verdict("F_ST", "is beyond it", "is not beyond it")`.

Pooling the two pools' differences is not free noise reduction, and
`pool_tab` measures what it buys. Averaging halves the variance only if the
two pools' paired differences are independent across founder draws. For the
elite-line gene diversity they are correlated at
`r fmt(pool_tab["gd_elite", "cor_AB"], 2)`, and the pooled standard deviation,
`r fmt(pool_tab["gd_elite", "sd_pooled"], 3)`, is close to the
`r fmt(pool_tab["gd_elite", "sd_if_independent"], 3)` it would be if they were
independent: pooling does what it should there. For the testcross mean they
are correlated at `r fmt(pool_tab["mean_tc", "cor_AB"], 2)`, and the pooled
standard deviation is `r fmt(pool_tab["mean_tc", "sd_pooled"], 2)` against
`r fmt(pool_tab["mean_tc", "sd_if_independent"], 2)` if independent: pooling
does not reduce that noise, because the founder draw moves both pools'
response together. The per-pool rows are there to show the asymmetry between
two exchangeable pools that any one set of `r n_rep` draws produces. The loss
table says how much of the pools' loss is selection: the control loses gene
diversity too -- `r rrs_cfg[["n_sel"]]` elites recombined into
`r rrs_cfg[["n_pool_A"]]` individuals is a bottleneck whether or not the
elites were chosen -- and the selected arm's mean loss
`r if (rep_tab["gd_pool_mean", "diff_mean"] < 0) "exceeds" else "does not exceed"`
the control's, with the pooled `t` just stated as the evidence.

What is inferred: with real meiosis and a dominance trait behind it,
reciprocal recurrent selection responds in both pools, and at this
replication the response
`r if (all(sig_paired[c("mean_tc_A", "mean_tc_B", "hybrid_meanG")])) "is beyond" else "is not beyond"`
`t_crit`. Whether it spends within-pool diversity faster than the same
bottleneck under drift alone is what the pooled gene-diversity `t` decides,
and in this render it `r verdict("gd_elite_mean", "does", "does not")` for
the elite lines and `r verdict("gd_pool_mean", "does", "does not")` for the
recombined pools: the selection component of the loss is of the same order as
its paired standard deviation at this scale, so the conclusion, not only the
number, is what the `t` decides. Where the sign holds, the control isolates
that component from the drift part, which is the distinction @sec-cycles
draws with the package's own faster simulator over a longer trajectory and a
grid of selection intensities, and @sec-programme draws a third time in an
inbred-line programme. The between-pool rows: the `GD_BS` difference
`r verdict("GD_BS")` beyond `t_crit`; where it is, and positive, RRS is
moving the pools apart faster than drift does, which is what a scheme
selecting on the cross is expected to do. `F_ST` rises partly because `GD_T`
falls, so `GD_BS` is the row to read. What limits the inference is the scale
-- `r rrs_cfg[["n_pool_A"]]` founders per pool, `r rrs_cfg[["n_cycles"]]`
cycles, `r rrs_cfg[["n_testers"]]` testers at one location, `r n_rep` draws
-- and the fact that the founders come from MaCS, so both the numbers in this
paragraph and any verdict that rests on a `t` near `t_crit` may differ
between renders.

### The final hybrid pool through `stage2_select()`, unchanged

`alphasimr_pipeline()` is the loop above without the instrumentation: founders,
`r rrs_cfg[["n_cycles"]]` cycles, then `alphasimr_make_hybrids()` testcrosses
the final cycle's elites directly against each other -- a full factorial
between the `r rrs_cfg[["n_sel"]]` pool-A elites and `r rrs_cfg[["n_sel"]]`
pool-B elites, so `r rrs_cfg[["n_sel"]] * rrs_cfg[["n_sel"]]` commercial
hybrids -- and phenotypes them the same way every testcross in this chapter
was phenotyped. `alphasimr_to_stage1()` then does the bookkeeping that makes
this look like any other stage-1 output: it remaps AlphaSimR's internal line
ids into the `1..n_A`, `n_A+1..n_A+n_B` convention `hybrids$line_A`/`line_B`
and `fL` use elsewhere in the book, and builds `fL` by stacking the elite lines
from *both* pools' genotypes into one matrix and computing a single joint
coancestry matrix over all of them -- not two separate per-pool matrices --
which is what lets `theta_pools()` and the rest of the pipeline index into pool
A and pool B lines with one consistent numbering.

```{r}
res <- alphasimr_pipeline(rrs_cfg)
fmt(res[["cycles"]], 4)
st1 <- res[["stage1"]]
dim(st1[["X"]])
sel <- stage2_select(st1[["X"]], st1[["f"]], st1[["hybrids"]], n_sel = 8, fL = st1[["fL"]],
                     verbose = FALSE)
fmt(sel[["metrics"]][, c("scenario", "alpha", "GD", "GD_BS", "F_ST", "index")], 4)
```

This is one more founder draw, so its trajectory is one more replicate of the
figure above, not the same run. Depending on the draw, `stage2_select()` may
warn that the attainable `alpha` ceiling is at or below zero: with
`r rrs_cfg[["n_sel"]] * rrs_cfg[["n_sel"]]` hybrids over
`r 2 * rrs_cfg[["n_sel"]]` lines, truncation on the index can pick a set no
less diverse than a random one of the same size, and then the constraint has
nothing to restrict and every `DE_alpha` row coincides with `ref_truncation`.
That is the same ceiling @sec-combinatorial computes on the book's own data,
where the pool is large enough for it to be positive. Nothing here is
AlphaSimR-specific:
`stage2_select()` never sees where `X`, `f`, `hybrids` and `fL` came from, only
that they satisfy the contract stage 1 promises. `GD_BS`, the between-pool
divergence that carries the heterosis signal (@sec-choosing), is computed from
`fL` exactly as it is everywhere else in the book.

## Assumptions, and what this adds over @sec-simulating-crosses

**What is more realistic here.** Actual recombination and linkage (a genetic
map per chromosome, not independent markers), a trait with directional
dominance so that testcross selection has something real to select on and
heterosis and inbreeding depression have a value, outbred founder pools with a
coalescent split behind them, and multi-generation recurrent selection with a
paired control rather than one static pool.

**What is declared, and where.** Steps 2--6 use inbred founders from
`quickHaplo()` with every site a QTL and dominance degrees drawn at
`meanDD = `r rrs_cfg[["mean_dd"]]``, `varDD = `r rrs_cfg[["var_dd"]]``; the
"pools" of Steps 4 and 6 are two undifferentiated halves of one founder set.
Step 7 uses a separate additive founder draw. Step 8 uses outbred founders
from `runMacs2(split = )`, again with every site a QTL; its DH lines are fully
inbred, its testers are outbred individuals, and every loss of diversity is
measured against the same run's founders (cycle 0) and against the paired
no-selection control -- not against a random-subset null of the same size,
which is the reference @sec-combinatorial's `alpha` uses.

**What is still simplified.** Testers are the first `r rrs_cfg[["n_testers"]]`
individuals of the *current* opposite pool, an arbitrary slice by index -- not
chosen at random or by merit, and not the elite inbred tester a real programme
would use; because `randCross()` recreates each pool every cycle, the testers
change from cycle to cycle, so successive cycles' testcross means are not on
a common tester and are comparable only through the pools' genetic level;
there is one phenotyping
environment (`setPheno()`'s error variance, no genotype-by-environment term --
@sec-programme adds one); the population-split coalescent history is a
stylised divergence process, not calibrated to a named heterotic-pool pair;
doubled-haploid production is instantaneous, skipping the field seasons Step 5
counted; and selection acts on trait 1 only -- the `r rrs_cfg[["n_traits"]]`
traits this chapter's config simulates are all carried through `X`/`hybrids`,
but only the first ever enters the GCA ranking. The accuracy prediction of
Step 6 is first order in the sample correlation -- its finite-tester-pool and
sample-variance corrections are applied, its second-order term is not -- and
the selfing trajectory of Step 5 follows 400 parallel lineages from one
cross, checked at seven generations.

**Scale.** This chapter runs live at a book scale (`r rrs_cfg[["n_pool_A"]]` +
`r rrs_cfg[["n_pool_B"]]` founders, `r rrs_cfg[["n_chr"]] * rrs_cfg[["seg_sites"]]`
markers, `r rrs_cfg[["n_cycles"]]` cycles, `r n_rep` paired founder draws) --
AlphaSimR is fast enough at this size that no caching is needed, unlike the
MABC scans in @sec-backcross. Step 8's numbers may change on re-execution
and do change between sessions, because its founders come from MaCS (Step 1);
when they move, they move by about the amounts the standard deviations above
say, and so may a conclusion that rests on a `t` near `t_crit`: between
sessions, the paired `t` for the selection component of gene-diversity
erosion has been seen both above and below `t_crit` at 24 pairs, which is why
the chapter runs `r n_rep` pairs and writes each such inference conditionally
on the render's own `t`. A heavier, production-scale run is
`inst/scripts/run_alphasimr.R`:

```sh
Rscript inst/scripts/run_alphasimr.R --test          # smoke run, ~10 s
Rscript inst/scripts/run_alphasimr.R --out=DIR        # full run, ~2 min
```

It runs `alphasimr_pipeline()` at a larger configuration, feeds the resulting
hybrid pool through `stage2_select()`, and writes the cycle trajectory and
selection metrics as CSVs.

## Where this connects

@sec-simulating-crosses and Step 8 build the same three-object contract
from two different generative models -- the fast one and the realistic one --
which is the point of drawing that seam where the book does: everything past
it, `build_ctx()` through `stage2_select()`, cannot tell which one it is
looking at. @sec-heterosis is the algebra Steps 3, 4 and 6 verify against
AlphaSimR: inbreeding depression against `genParam()`, expected heterosis
against a full factorial, and the GCA/SCA decomposition against the exact
combining abilities. @sec-response supplies the breeder's equation Step 6
checks and the index theory Step 7 uses; @sec-limits is where Step 3's
`varA`-against-`genicVarA` gap becomes the subject rather than a diagnostic.
@sec-subdivided supplies the partition Step 8 measures across `split` and
across cycles. @sec-cycles runs the same diversity-erosion argument at a
longer horizon with the package's own faster simulator; this chapter is its
cross-check against an engine with real meiosis behind it, and @sec-limits
uses this chapter's linked arm for the same reason.

@sec-programme is the other half of the method: the same primitives assembled
into an inbred-line programme with staged yield trials, where the question is
not which hybrids to sell but how fast a variety pipeline can run and what it
spends to run faster.
