# A complete plant breeding programme, and what it costs in diversity {#sec-programme}

```{r}
#| include: false
source("_common.R")
library(AlphaSimR)
```

@sec-alphasimr assembled AlphaSimR's primitives into a *hybrid* scheme: two
heterotic pools, reciprocal testers, a commercial cross at the end. This chapter
assembles the same primitives into the other kind of programme -- an
**inbred-line** pipeline, where the product is a variety rather than a hybrid,
and where the organising problem is not which cross to sell but how many
genotypes to carry through how many stages of field testing.

The scheme is the wheat programme of @gaynor2017, reduced in scale. It is worth
building because it is the shape most public and many commercial line-breeding
programmes actually have, and because it exposes something the hybrid chapter
cannot: **the trade between the number of genotypes tested and the precision
with which each is measured.** A breeder has a fixed number of plots. Spending
them on more entries at one location or fewer entries at eight locations is the
same budget spent two ways, and the staged trial is the standard compromise.

Then the chapter does what the course this scheme comes from does not: it puts
the book's diversity machinery over the result, and asks what the faster
programme spends to be faster.

## The scheme

```{r}
#| echo: false
#| fig-width: 9.5
#| fig-height: 3.6
#| fig-cap: "The staged inbred-line programme. Entries fall by roughly an order of magnitude at each stage while replication -- the number of locations each entry is scored in -- rises. Parents for the next cycle are recycled from the late trials (solid) or, under genomic selection, from the preliminary trial as well (dashed)."
op <- par(mar = c(0, 0, 0, 0)); plot.new(); plot.window(c(0, 100), c(0, 40))
box_ <- function(x, y, w, h, lab, fill, cex = 0.66) {
  rect(x, y, x + w, y + h, col = fill, border = "grey40")
  text(x + w / 2, y + h / 2, lab, cex = cex)
}
arr <- function(x0, y0, x1, y1, lty = 1, col = "grey30")
  arrows(x0, y0, x1, y1, length = 0.06, col = col, lty = lty, lwd = 1.5)

stages <- list(c(1, "Parents", "#d1e5f0"), c(14.5, "F1\n20 crosses", "#d1e5f0"),
               c(28, "DH\n400 lines", "#f7f7f7"), c(41.5, "Headrows\n400, 1/4 rep", "#f7f7f7"),
               c(55, "PYT\n60, 1 loc", "#fddbc7"), c(68.5, "AYT\n12, 4 loc", "#fddbc7"),
               c(82, "EYT\n4, 8 loc", "#fddbc7"))
for (s in stages) box_(as.numeric(s[1]), 16, 12, 11, s[2], s[3])
box_(93, 16, 6, 11, "Variety", "#b8e0b8")
for (x in c(13, 26.5, 40, 53.5, 67, 80.5, 92.5)) arr(x, 21.5, x + 1.4, 21.5)

arr(88, 16, 88, 6, lty = 1, col = "#b2182b"); arr(88, 6, 8, 6, lty = 1, col = "#b2182b")
arr(7, 6, 7, 16, lty = 1, col = "#b2182b")
arr(61, 16, 61, 10.5, lty = 2, col = "#1b6ca8")
arr(61, 10.5, 9, 10.5, lty = 2, col = "#1b6ca8")
text(46, 4, "recycle parents from EYT + AYT (phenotypic selection)",
     cex = 0.6, col = "#b2182b")
text(36, 12.4, "genomic selection also recycles from PYT -- one year sooner",
     cex = 0.6, col = "#1b6ca8")
text(50, 31, "entries fall, replication rises", cex = 0.72, col = "grey25")
arr(30, 29, 78, 29, col = "grey55")
par(op)
```

Seven stages, and the logic connecting them is a single idea: **at every step,
entries fall and replication rises.** Four hundred doubled-haploid lines are
scored once, cheaply, in unreplicated headrows; sixty survive into a preliminary
yield trial at one location; twelve into an advanced trial at four; four into an
elite trial at eight. Selection accuracy at the last stage is high because the
error variance has been divided by eight locations -- not because the trait
became more heritable.

That is what `reps` does in `setPheno()`, and it is the argument @sec-alphasimr's
Step 2 flagged as the plant idiom. The configuration:

```{r}
sch <- list(
  n_founders = 40, n_chr = 10, n_qtl = 60,   # founder genomes
  n_crosses  = 20, n_dh = 20, fam_max = 5,   # crossing and DH production
  n_pyt = 60, n_ayt = 12, n_eyt = 4,         # entries per yield trial
  rep_hdrw = 1/4, rep_pyt = 1, rep_ayt = 4, rep_eyt = 8,   # locations per stage
  var_e = 0.4, gs_boost = 3, n_burn = 7, n_years = 15
)
```

`rep_hdrw = 1/4` is the course's device for a stage *less* precise than a single
plot: headrows are scored visually on a short row, which carries less
information than one replicated plot, so the error variance is multiplied by
four. `fam_max` caps how many DH lines any one cross may contribute to the
preliminary trial -- `selectWithinFam()` from @sec-alphasimr's Step 6, holding
the genetic base open against a truncation that would otherwise take the whole
trial from the two best crosses.

## Why a breeding programme needs a burn-in

The loop below advances every stage by one year, simultaneously. That only means
something if all seven stages are already *occupied*, by cohorts of different
ages: this year's elite trial entered the programme six years ago, this year's
headrows entered last year. Starting the loop from an empty programme would
compare a year-1 elite trial that does not exist against a year-1 headrow that
does.

The fix is the staggered fill in the course's vignette, and it is the most
confusing seven lines in it:

```r
for (year in 1:7) {
  F1 <- randCross(Parents, n_crosses)
  if (year < 7) DH   <- makeDH(F1, n_dh)
  if (year < 6) HDRW <- setPheno(DH, reps = rep_hdrw)
  if (year < 5) PYT  <- ...
  if (year < 4) AYT  <- ...
  if (year < 3) EYT  <- ...
}
```

Read it as a triangle rather than a loop. On the first pass every stage runs, so
every stage holds a cohort. On the second pass the elite trial is skipped, so
the cohort sitting in it is a year older than the one behind it. By the seventh
pass only the crossing block still runs. What comes out is a programme in which
each stage holds a cohort of the right age relative to its neighbours -- a
*steady state*, which is the only state in which a per-year genetic gain is a
meaningful number.

## One programme, fifteen years

The whole scheme is one function. It is written out rather than packaged
because the loop is the thing being taught, and because a reader adapting it to
their own programme will edit the loop, not call it.

```{r}
gd_of <- function(pop, SP)
  1 - mean(molecular_coancestry(pullSegSiteGeno(pop, simParam = SP) / 2))

run_programme <- function(cfg, seed, genomic = FALSE, gs_from = 1) {
  set.seed(seed)
  fg <- quickHaplo(nInd = cfg$n_founders, nChr = cfg$n_chr,
                   segSites = cfg$n_qtl, inbred = TRUE)
  SP <- SimParam$new(fg)
  SP$addTraitAG(nQtlPerChr = cfg$n_qtl, mean = 4, var = 0.1, varGxE = 0.2)
  ph <- function(p, r) setPheno(p, varE = cfg$var_e, reps = r, simParam = SP)

  # --- burn-in: fill every stage with a cohort of the right age ---
  Parents <- newPop(fg, simParam = SP)
  for (year in 1:cfg$n_burn) {
    F1 <- randCross(Parents, cfg$n_crosses, simParam = SP)
    if (year < 7) DH   <- makeDH(F1, cfg$n_dh, simParam = SP)
    if (year < 6) HDRW <- ph(DH, cfg$rep_hdrw)
    if (year < 5) {
      PYT <- selectWithinFam(HDRW, nInd = cfg$fam_max, use = "pheno", simParam = SP)
      PYT <- ph(selectInd(PYT, nInd = cfg$n_pyt, use = "pheno", simParam = SP),
                cfg$rep_pyt)
    }
    if (year < 4)
      AYT <- ph(selectInd(PYT, nInd = cfg$n_ayt, use = "pheno", simParam = SP),
                cfg$rep_ayt)
    if (year < 3)
      EYT <- ph(selectInd(AYT, nInd = cfg$n_eyt, use = "pheno", simParam = SP),
                cfg$rep_eyt)
  }

  # --- advancing years, last stage first so each reads the previous cohort ---
  out <- data.frame(year = 1:cfg$n_years, meanG = NA_real_, varG = NA_real_,
                    acc_pyt = NA_real_, gd_pyt = NA_real_, gd_parents = NA_real_)
  for (year in 1:cfg$n_years) {
    gs <- genomic && year >= gs_from

    Variety <- selectInd(EYT, nInd = 1, use = "pheno", simParam = SP)
    EYT <- ph(selectInd(AYT, nInd = cfg$n_eyt, use = "pheno", simParam = SP),
              cfg$rep_eyt)

    if (gs) {            # lever 1: a more accurate criterion at the PYT stage
      PYT@ebv <- setPheno(PYT, varE = cfg$var_e,
                          reps = cfg$rep_pyt * cfg$gs_boost,
                          onlyPheno = TRUE, simParam = SP)
      AYT <- ph(selectInd(PYT, nInd = cfg$n_ayt, use = "ebv", simParam = SP),
                cfg$rep_ayt)
    } else {
      AYT <- ph(selectInd(PYT, nInd = cfg$n_ayt, use = "pheno", simParam = SP),
                cfg$rep_ayt)
    }

    PYT <- selectWithinFam(HDRW, nInd = cfg$fam_max, use = "pheno", simParam = SP)
    PYT <- ph(selectInd(PYT, nInd = cfg$n_pyt, use = "pheno", simParam = SP),
              cfg$rep_pyt)
    HDRW <- ph(DH, cfg$rep_hdrw)
    DH   <- makeDH(F1, cfg$n_dh, simParam = SP)

    if (gs) {            # lever 2: recycle parents a year earlier, from the PYT
      PYT@ebv <- setPheno(PYT, varE = cfg$var_e,
                          reps = cfg$rep_pyt * cfg$gs_boost,
                          onlyPheno = TRUE, simParam = SP)
      EYT@ebv <- pheno(EYT); AYT@ebv <- pheno(AYT)
      Parents <- selectInd(c(EYT, AYT, PYT), nInd = cfg$n_eyt + cfg$n_ayt,
                           use = "ebv", simParam = SP)
    } else {
      Parents <- c(EYT, AYT)
    }
    F1 <- randCross(Parents, cfg$n_crosses, simParam = SP)

    out$meanG[year]      <- meanG(PYT)
    out$varG[year]       <- varG(PYT)
    out$acc_pyt[year]    <- stats::cor(PYT@gv[, 1], PYT@pheno[, 1])
    out$gd_pyt[year]     <- gd_of(PYT, SP)
    out$gd_parents[year] <- gd_of(Parents, SP)
  }
  out
}
```

Two details are worth pausing on.

**The advancing loop runs backwards.** The elite trial is filled *before* the
advanced trial is refilled, which is before the preliminary trial is refilled.
Written in the other order, `EYT` would be selected from an `AYT` that had
already been overwritten with this year's cohort, and the programme would
advance two stages per year without anyone noticing. Ordering is the whole
correctness argument of a staged simulation.

**Progress is reported at the PYT.** It is the first stage with a yield
phenotype and the last with enough entries for a variance to mean anything --
`meanG` on four elite lines is noise. This is the course's choice and it is the
right one.

```{r}
ps <- run_programme(sch, seed = 1, genomic = FALSE)
fmt(ps[c(1, 5, 10, 15), ])
```

Genetic mean rises, genetic variance falls, and gene diversity of the
preliminary trial falls with it. Nothing here is surprising; what matters is
that all three are now being measured on the same run.

::: {.callout-note}
## These numbers are reproducible

The founders come from `quickHaplo()`, which is seeded by `set.seed()`, so
`run_programme(sch, seed = 1)` returns the same data frame every time -- unlike
@sec-alphasimr's Step 8, whose `runMacs2()` founders are not reproducible.
That is what makes the paired comparison below legitimate: the phenotypic and
genomic arms are run from the *same* seed, so they share founders, and every
difference between them is attributable to the scheme.
:::

## Genomic selection: two levers, not one

Genomic selection is usually described as buying accuracy. It buys two things,
and the second is normally the larger (@gaynor2017; @meuwissen2001):

1. **Accuracy earlier in the cycle.** A prediction is available on a line that
   has been genotyped but barely phenotyped, so the advance from preliminary to
   advanced trial can be made on better information than one location's yield.
2. **A shorter generation interval.** Parents can be recycled from the
   preliminary trial instead of waiting for the elite trial -- and the breeder's
   equation of @sec-response is a rate *per unit time*, so cutting years off the
   cycle raises gain per year even at unchanged accuracy per selection.

This chapter mimics the prediction rather than fitting one: `ebv` is filled with
a phenotype simulated at `gs_boost` times the replication, which is a stand-in
for a genomic prediction of the corresponding accuracy. @sec-genomic-prediction
is where the book actually fits prediction models and measures what accuracy is
attainable; the honest statement here is that this chapter *assumes* an accuracy
and studies its consequences, rather than earning it.

```{r}
gs <- run_programme(sch, seed = 1, genomic = TRUE, gs_from = 6)
fmt(gs[c(1, 5, 10, 15), ])
```

Years 1 to 5 are identical to the phenotypic arm, digit for digit, because
genomic selection only switches on in year 6. That is a matched control in the
sense @sec-limits uses the term: the two arms share a history, so the comparison
is paired rather than between two independent runs.

## The comparison, replicated

One run of one seed is an anecdote. Both arms are cheap enough to replicate.

```{r}
seeds <- 1:24
reps <- do.call(rbind, lapply(seeds, function(s) {
  a <- run_programme(sch, s, genomic = FALSE)
  b <- run_programme(sch, s, genomic = TRUE, gs_from = 6)
  rate <- function(d) stats::coef(stats::lm(meanG ~ year,
                                            data = d[d$year >= 6, ]))[2]
  data.frame(seed = s,
             gain_ps = rate(a),      gain_gs = rate(b),
             gd_ps   = a$gd_pyt[nrow(a)], gd_gs = b$gd_pyt[nrow(b)],
             end_ps  = a$meanG[nrow(a)],  end_gs = b$meanG[nrow(b)])
}))

paired <- function(x, y) {
  d <- y - x
  c(mean_difference = mean(d), t = mean(d) / (stats::sd(d) / sqrt(length(d))))
}
fmt(rbind(`gain per year`      = paired(reps$gain_ps, reps$gain_gs),
          `gene diversity, yr 15` = paired(reps$gd_ps, reps$gd_gs),
          `genetic mean, yr 15`   = paired(reps$end_ps, reps$end_gs)))
```

```{r}
#| label: fig-programme-arms
#| echo: false
#| fig-width: 9
#| fig-height: 4.2
#| fig-cap: "Twenty-four paired runs, genomic selection switched on in year 6. Left: genetic mean of the preliminary yield trial. Right: gene diversity of the same population. The faster programme is also the one spending diversity faster."
traj <- lapply(seeds, function(s)
  list(ps = run_programme(sch, s, FALSE), gs = run_programme(sch, s, TRUE, gs_from = 6)))
avg <- function(arm, col) rowMeans(sapply(traj, function(t) t[[arm]][[col]]))
yr <- seq_len(sch$n_years)
cols <- c(ps = "#1b1b1b", gs = "#7b3294")
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.3, 2.2, 0.8))
for (v in c("meanG", "gd_pyt")) {
  a <- avg("ps", v); b <- avg("gs", v)
  plot(yr, a, type = "l", lwd = 2.4, col = cols["ps"], ylim = range(a, b),
       xlab = "Year", ylab = if (v == "meanG") "Genetic mean (PYT)" else "Gene diversity (PYT)",
       main = if (v == "meanG") "Response" else "Diversity")
  lines(yr, b, lwd = 2.4, lty = 2, col = cols["gs"])
  abline(v = 6, col = "grey70", lty = 3)
  legend(if (v == "meanG") "topleft" else "topright",
         c("phenotypic", "genomic (from yr 6)"), col = cols, lwd = 2.4,
         lty = c(1, 2), bty = "n", cex = 0.78)
}
par(op)
```

Genomic selection raises the rate of gain by roughly half, and the paired
`t`-statistic leaves no doubt about the sign. That is the result @gaynor2017
reports and the reason two-part strategies were adopted.

The second row is the one this book exists for. **Gene diversity in the
preliminary trial is significantly lower under genomic selection at year 15**,
and it is lower for a reason that is structural rather than incidental:
recycling parents from the preliminary trial means recycling them a year before
the trial data could have distinguished full sibs, so the parents chosen are
more closely related than the ones the phenotypic arm would have chosen. A
shorter generation interval is a higher rate of coancestry accumulation per unit
time, not merely per cycle.

## The diversity ledger

The measurement above used gene diversity of the preliminary trial. The parent
set is the population that actually transmits, and it is the one an optimal
contribution scheme would act on:

```{r}
ledger <- data.frame(
  year       = yr,
  gd_pyt_ps  = avg("ps", "gd_pyt"),  gd_pyt_gs  = avg("gs", "gd_pyt"),
  gd_par_ps  = avg("ps", "gd_parents"), gd_par_gs = avg("gs", "gd_parents"))
fmt(ledger[c(1, 5, 10, 15), ])
```

```{r}
loss <- function(v) (ledger[[v]][1] - ledger[[v]][nrow(ledger)]) / ledger[[v]][1]
fmt(c(`PYT, phenotypic`   = loss("gd_pyt_ps"), `PYT, genomic`   = loss("gd_pyt_gs"),
      `parents, phenotypic` = loss("gd_par_ps"), `parents, genomic` = loss("gd_par_gs")))
```

Those four numbers are exactly the quantity this book calls `alpha`: the
proportional loss of gene diversity, here measured against the programme's own
starting point rather than against a random-selection baseline of the same size
(@sec-baseline is where that distinction is made, and it matters -- a
same-size random baseline would attribute part of this loss to drift rather
than to selection).

::: {.callout-note}
## Status number, and why it is not the headline here

`status_number()` on a set of fully inbred lines returns a value near or below
1, because a homozygous line has self-coancestry `f_ii = 1` and the diagonal
dominates the mean. That is the same arithmetic @sec-sizing explains for a
biparental cross, and it is correct rather than broken -- but it makes `N_s` an
unhelpful headline for a line-breeding programme. Gene diversity is the metric
that stays interpretable across both inbred and outbred populations, which is
why the ledger is written in it.
:::

What would an `alpha` budget mean here? Not a constraint on which hybrids to
sell, as in @sec-combinatorial, but a constraint on **which parents to recycle**:
the same optimal-contribution problem of @sec-ocs, applied to the
`c(EYT, AYT, PYT)` candidate set that the genomic arm selects from. The
candidate set is small, the coancestry matrix is cheap, and the choice being
constrained is annual. @sec-cycles is the argument for why that constraint pays
back, and @sec-limits is where the payback is predicted rather than observed.

## Assumptions

**What this models.** Staged testing with replication rising across stages,
genotype-by-environment interaction as a variance component
(`SP$addTraitAG(varGxE = )`), doubled-haploid line development, within-family
caps, a steady-state programme reached through a burn-in, and two distinct
genomic-selection levers separated so their effects can be told apart.

**What it does not.** Genomic prediction is *mimicked* by adding replication,
not fitted -- so the accuracy is assumed rather than earned, and the training
population, its size, and its relationship to the selection candidates play no
role (@sec-genomic-prediction is where they do). There is one trait, so the
index questions of @sec-alphasimr's Step 7 do not arise. Locations differ only
through a variance component: there is no correlated environment structure, no
year effect, and no location-by-year term, so a "location" here is an
independent draw rather than a place. Doubled-haploid production is
instantaneous and never fails. Founders come from `quickHaplo()`, so they carry
no linkage disequilibrium -- @sec-alphasimr's Step 1 measures what that costs
and argues it is the right trade for a chapter about response rather than about
marker-QTL association.

**Scale.** `r sch$n_founders` founders, `r sch$n_chr * sch$n_qtl` loci,
`r sch$n_years` years, `r length(seeds)` paired seeds, all live at render.
@gaynor2017's programme is an order of magnitude larger in every dimension; the
qualitative results here are the ones that survive the reduction, and the
absolute rates are not comparable to a real wheat programme.

## Where this connects

@sec-alphasimr supplies every primitive this chapter uses and builds them into
the hybrid scheme instead; the two chapters are the same toolkit pointed at the
two kinds of programme. @sec-response is where the breeder's equation says why a
shorter generation interval raises gain per year. @sec-genomic-prediction is
where the accuracy this chapter assumes is actually measured, and
@sec-genomic-era is where marker-versus-QTL, the reason that accuracy is finite,
is taken seriously.

The diversity result belongs to the book's spine. @sec-cycles shows the same
tension -- faster response, faster erosion -- with the package's own simulator
over a longer horizon; @sec-limits explains why the erosion is what bounds the
programme and when it can be predicted; @sec-ocs is the machinery for spending
the budget deliberately rather than discovering afterwards what was spent.
