# Diversity metrics in the genomic era {#sec-genomic-era}

```{r}
#| include: false
source("_common.R")
ctx  <- simulate_data(book_cfg)
L    <- simulate_lines(book_cfg)
pool <- rep(c("A", "B"), c(book_cfg$n_pool_A, book_cfg$n_pool_B))
P0   <- pool_freqs(L / 2, pool)
```

In classical theory, inbreeding measured as **loss of heterozygosity** and
inbreeding measured as **drift of allele frequencies** are the same thing.
@meuwissen2020 show that inside genomic optimal-contribution schemes they stop
being the same thing, and can diverge badly, whenever diversity is controlled
through **identity by state** on a marker panel. Only when control is through
**identity by descent** is the classical equivalence restored.

The practical consequence is uncomfortable:

> You hit exactly the target you wrote into the optimiser's constraint, and you
> pay an unmonitored price in the metric you did not write.

If you constrain $\tfrac{1}{2}\mathbf{c}'\mathbf{G}\mathbf{c}$ using a VanRaden
$\mathbf{G}$, you are constraining **drift**. The loss of heterozygosity in your
germplasm may be running at a different rate, in either direction, from the one
you believe you are controlling.

This chapter is written for one kind of programme, stated in full in
@sec-era-programme: an allogamous, maize-type hybrid programme running
reciprocal recurrent selection over two heterotic pools. The first half defines
the metrics; the second half is the sequence of steps that puts them into a
cycle report, with the check and the failure mode of each step.

## The programme this chapter is written for {#sec-era-programme}

Every number below is computed under these assumptions. They are declared here
because several of the results are false outside them.

**A1. The species is allogamous and the product is an F1 hybrid.** Maize-type:
two heterotic pools, A and B, crossed to produce single-cross hybrids. The
commercial product carries the heterozygosity; the parents do not. Nothing here
is addressed to a self-pollinated crop whose product is a line, and nothing here
is addressed to a programme whose commercial unit is a synthetic or a population
cross -- both change which quantity is the product and therefore which loss
matters.

**A2. Parent lines are fully inbred, by DH induction or by selfing to fixation.**
This is a property of the *programme*, not of the mating system: an allogamous
crop bred for hybrids still delivers inbred parents. In this simulation the
lines are homozygous at every locus, which the coancestry kernel reports exactly:

```{r}
#| label: era-inbred-check
c(het_fraction_lines   = mean(L / 2 == 0.5),
  self_coancestry_min  = min(diag(molecular_coancestry(L / 2))),
  het_fraction_F1      = mean(ctx$X == 0.5),
  twice_GD_WI_hyb      = 2 * gd_partition(seq_len(ctx$N), ctx)[["GD_WI_hyb"]])
```

Line heterozygosity is zero and every line's self-coancestry is 1, so
individual-level inbreeding coefficients are 1 by construction and carry no
management information. The F1s carry
`r fmt(mean(ctx$X == 0.5), 3)` of their loci heterozygous, and that fraction is
exactly twice `GD_WI_hyb` -- the identity @sec-heterosis derives. **The
manageable quantity is therefore an allele frequency in a pool, never a
homozygosity within an individual.** Everything below is computed on
frequencies.

**A3. Germplasm is recycled within pool; only information crosses the pool
boundary.** Lines are chosen on hybrid performance against the opposite pool
(@comstock1949) and recycled inside their own pool. A DH taken from an
$A \times B$ F1 would be a blend of both pools and would destroy the heterotic
pattern, so it never happens. `run_cycles()` in @sec-cycles implements exactly
this loop.

**A4. Cycles are discrete.** Pool sizes are constant across cycles and no
generation overlap is modelled. Real programmes overlap cycles and reuse the
same line for years, which biases any $\Delta F$ read off a cycle index; see
@sec-limits.

**A5. Markers are neutral, biallelic, coded $0/0.5/1$** as the frequency of the
reference allele within the individual, and are in linkage equilibrium within
pool because each line's genotype is drawn independently per locus
(`simulate_lines()`). Linkage equilibrium makes the audit step of
@sec-era-step5 a best case, and that is said again where the number is reported.

**A6. The frequency base $p_0$ is frozen at the founding cycle** and never
recomputed. This is `ctx$p0`, set once in `build_ctx()`. @sec-era-step1 is that
step and measures what happens without it.

**A7. No dominance enters any metric in this chapter.** $\theta$, $GD$ and
$\alpha$ are functions of allele frequencies alone. The dominance side of the
same germplasm -- what the divergence between pools is worth in heterosis -- is
@sec-heterosis, and the two must not be mixed inside one constraint.

**A8. The scale is the book's teaching scale**: `r book_cfg$n_pool_A` +
`r book_cfg$n_pool_B` lines, `r book_cfg$m` markers, `r ctx$N` hybrids in the
full factorial. The production scale is 71 + 71 lines and 25000 markers. Where a
result is scale-dependent, it says so.

## Three lenses, not three synonyms

```{r}
#| label: era-lenses
#| echo: false
knitr::kable(data.frame(
  Lens = c("Drift", "Homozygosity", "IBD", "Hybrid"),
  `What the metric actually sees` = c(
    "squared allele-frequency change relative to a base",
    "proportion of heterozygosity remaining",
    "segments inherited by descent from a defined base",
    "haplotype homozygosity used as a proxy for IBD"),
  `Inbreeding metric` = c("F_drift", "F_hom", "F_IBD", "F_ROH"),
  `Relationship matrix` = c("G_VR2, G_VR1, G_i(p)", "G_0.5 (molecular coancestry)",
                            "A, G_LA", "G_ROH"),
  check.names = FALSE),
  caption = "The three lenses of the section title, plus the hybrid of two of them: G_ROH reads homozygosity and is interpreted as IBD, which is why it gets a row of its own and not a lens of its own. G_0.5 is the matrix this book calls f -- see @sec-coancestry.")
```

::: {.callout-important}
## Every inbreeding metric is a ratio to a reference population

Without explicitly declaring your base -- cycle 0 of the recurrent programme? a
landrace panel? the founding heterotic pool? -- none of the numbers below mean
anything. This is the commonest error in plant applications: using the *current*
generation's frequencies as the base, which zeroes the historical memory of
erosion. @sec-era-step1 measures the size of that mistake: it is not an
attenuation, it is an exact zero.
:::

## The metrics, one at a time

### Rate of inbreeding and effective size

What you manage is not $F$ but its rate of accumulation. $F$ is a stock;
$\Delta F$ is a flow.

$$\Delta F = \frac{F_t - F_{t-1}}{1 - F_{t-1}}, \qquad N_e = \frac{1}{2\Delta F}$$

with $F_t \in (-\infty, 1]$ the inbreeding coefficient of cycle $t$ against the
frozen base (negative values are possible for $F_{hom}$, see below),
$\Delta F$ the per-cycle rate and $N_e$ the effective size it implies
[@falconer1996, Ch. 3-4].

Since $1 - F_t = (1 - \Delta F)^t$, a regression of $\log(1 - F_t)$ on $t$ has
slope $\log(1 - \Delta F)$, so $\Delta F = 1 - e^{\text{slope}}$ and
$-\text{slope}$ is only its first-order approximation. **The linearity of that
regression is itself a diagnostic:** in @meuwissen2020, $\log(1 - F_{drift})$
was approximately linear under every scheme, while $\log(1 - F_{hom})$ showed
marked curvature under some, notably the ROH-based one. Curvature means a
non-constant rate -- your $N_e$ target is not being met steadily.

A tropical maize recurrent programme targeting $N_e = 50$ per cycle needs
$\Delta F = 0.01$. Measuring $F$ only at cycle 8 and dividing by 8 hides the
possibility that cycles 1-3 were conservative and cycles 6-8 collapsed the base.
@sec-era-step4 fits the regression on a case where the answer is known in
advance, and then on one where the diagnostic has to fire.

### `F_hom`, homozygosity-based {#sec-era-fhom}

Wright defined an inbreeding coefficient to run from 0 to 1 as heterozygosity
under random mating falls from its initial state to zero. So the metric is the
fraction of heterozygosity already lost relative to the base:

$$F_{hom} = 1 - \frac{1}{N_{SNP}}\sum_k \frac{2p_{t,k}(1 - p_{t,k})}{2p_{0,k}(1 - p_{0,k})}
          = 1 - \frac{1}{N_{SNP}}\sum_k \frac{H_{t,k}}{H_{0,k}}$$

with $p_{t,k} \in [0,1]$ the frequency of the reference allele at marker $k$ in
the population being scored, $p_{0,k} \in (0,1)$ the same frequency in the frozen
base, $H = 2p(1-p)$ the expected heterozygosity under random mating, and
$N_{SNP}$ the number of markers [@meuwissen2020, their Eq. 1A defines it as the
total marker count]. This book restricts the sum to markers polymorphic in the
base, because $H_{0,k} = 0$ makes the ratio undefined otherwise --
`freq_metrics()` does that with `keep <- ctx$p0 > eps & ctx$p0 < 1 - eps` -- and
the restriction changes nothing against their result, since their own panels
were random samples of loci already segregating in the base population.

**It can be negative.** If management pushes frequencies toward 0.5,
heterozygosity rises above the base and the "inbreeding coefficient" goes below
zero. That is not a numerical error: it is the expected behaviour of any scheme
that maximises heterozygosity on the monitored panel.

It protects against inbreeding depression and the expression of deleterious
recessives in homozygosis. It is the right lens when the dominant risk is
*vigour* -- in this programme, the vigour of the segregating crossing block and
not of the fixed lines, which is a narrower claim than it looks and is qualified
at the end of this section.

#### What this lens is computed on, and what it is not

$F_{hom}$ as `freq_metrics()` computes it takes $p_t$ from
`colMeans(X[idx, ])`: the reference-allele frequency of the **merged** hybrid
set. With inbred parents that merged frequency is
$\bar p = (\tilde p_A + \tilde p_B)/2$, with $\tilde p_A$ and $\tilde p_B$ the
two pools' frequencies weighted by the line usage the selection induces. So the
$He$ column of the same call is $\overline{2\bar p(1 - \bar p)}$, which is
$GD_T = 1 - \bar f$ of the Caballero & Toro partition [@caballero2002, Eq. 7]
*identically*, not approximately: $\bar f = (\theta_A + 2\theta_{AB} +
\theta_B)/4 = \overline{\bar p^2 + (1 - \bar p)^2}$ under A2.

**That is gene diversity of the notional merged pool, and it is not the
heterozygosity of the product.** The realised F1 heterozygosity is
$2\,GD_{WI\text{-}hyb}$ -- exactly, by the self-coancestry identity of
assumption A2 -- and on the full factorial it equals $GD_T + GD_{BS}$, the
between-pool component counted twice (@sec-heterosis-identity). The difference
between the two quantities is therefore $GD_{BS}$, which the $F_{hom}$ lens does
not carry, and which moves with the selection:

```{r}
#| label: era-hom-two-levels
set.seed(31)
sels <- list(random     = sample.int(ctx$N, 60),
             max_div    = sel_greedy(ctx, 60, w = 0),
             balanced   = sel_greedy(ctx, 60, w = 1),
             truncation = sel_truncation(ctx, 60))
two_levels <- t(sapply(sels, function(i) {
  gp <- gd_partition(i, ctx)
  fm <- freq_metrics(i, ctx)
  c(He_merged = fm[["He"]], GD_T = gp[["GD_T"]], F_hom = fm[["F_hom"]],
    GD_BS = gp[["GD_BS"]], F1_het = mean(ctx$X[i, ] == 0.5),
    two_GD_WI_hyb = 2 * gp[["GD_WI_hyb"]],
    GD_T_plus_BS = gp[["GD_T"]] + gp[["GD_BS"]])
}))
stopifnot(isTRUE(all.equal(two_levels[, "He_merged"], two_levels[, "GD_T"])),
          isTRUE(all.equal(two_levels[, "F1_het"], two_levels[, "two_GD_WI_hyb"])))
fmt(as.data.frame(two_levels), 4)
```

Two identities hold to machine precision in that table and the `stopifnot()`
enforces both on every render: $He = GD_T$, and the realised F1 heterozygosity
equals $2\,GD_{WI\text{-}hyb}$. The third comparison is *not* an identity on a
selected subset: $GD_T + GD_{BS}$ departs from the realised heterozygosity by at
most `r sprintf("%.4f", max(abs(two_levels[, "GD_T_plus_BS"] - two_levels[, "F1_het"])))`
across these four rows, because $GD_T$ and $GD_{BS}$ are computed from line-usage
weights -- what a random re-pairing of the selected parents would give -- while
$GD_{WI\text{-}hyb}$ scores the pairs actually chosen (@sec-heterosis states the
same limit from the other side).

Now read the ranking. Truncation has the **highest** $F_{hom}$,
`r fmt(two_levels["truncation", "F_hom"], 4)` against
`r fmt(two_levels["random", "F_hom"], 4)` for the random draw, which on the usual
reading is "most heterozygosity lost". Yet its hybrids are *more* heterozygous
than the random draw's:
`r fmt(two_levels["truncation", "F1_het"], 4)` against
`r fmt(two_levels["random", "F1_het"], 4)`. The two readings are not in conflict
because they are two quantities: truncation concentrated usage inside the pools
(@sec-era-step6 measures that) and so lost merged-pool gene diversity, while the
$A \times B$ pairs it kept happen to be more divergent --
$GD_{BS}$ = `r fmt(two_levels["truncation", "GD_BS"], 4)` against
`r fmt(two_levels["random", "GD_BS"], 4)`. Reported alone, $F_{hom}$ would have
called that plan the worst of the four for the very quantity -- product
heterozygosity -- on which it beats the random draw.

Within a heterotic pool under reciprocal recurrent selection, $F_{hom}$ on the
*pool's* own frequencies is still the right lens for one thing, and that one
thing is narrower than it is usually stated to be. It is the **segregating
crossing block**, which is not inbred and in which inbreeding depression is
therefore expressed:
$ID = 2F\sum_k d_k p_k q_k$ [@falconer1996, Ch. 14], with $d_k$ the dominance
deviation at locus $k$, $p_k$ and $q_k = 1 - p_k$ the pool's current
frequencies, and $F \in [0,1]$ the block's inbreeding coefficient relative to
random mating at those same *current* frequencies -- which is **not** the $F_t$
of this chapter's two lenses, measured against the frozen base $p_0$, and the
two must not be substituted for one another. @sec-heterosis-id verifies that
quantity against `heterosis_partition()`'s shared term. It is built from the
same per-locus product $p_k q_k$ that $F_{hom}$ aggregates -- but weighted by
$d_k$, which $F_{hom}$ does not carry, so $F_{hom}$ locates the erosion and
does not price it.

It does **not** anticipate the per-se value of the inbred lines themselves.
Under A2 a line has no heterozygous locus, so its dominance contribution is
exactly zero whatever the pool's frequencies are -- @sec-heterosis-id measures
that exact zero -- and a fully homozygous genotype's value is therefore the
additive sum $\sum_k a_k (2x_{ik} - 1)$ alone, the genotypic-value model of
@sec-heterosis with $x_{ik} \in \{0, 1\}$ at every locus and $a_k$ the per-locus
additive effect. Its pool mean moves with the frequencies in whichever direction
they move, so there is no directional decline there for $F_{hom}$ to anticipate. It also says nothing about between-pool
divergence (@sec-heterosis). Computed
instead on the homozygosity of individual inbred parents it is 1 by construction
-- assumption A2 -- and is **useless for management**. The mating system of the
species does not rescue that: an inbred maize line is as fixed as an inbred wheat
line. What distinguishes the hybrid programme is that the pool of lines can be
intermated, so the pool has a frequency vector worth managing even though none of
its members does.

### `F_drift`, drift-based

$$F_{drift} = \frac{1}{N_{SNP}}\sum_k \frac{(p_{t,k} - p_{0,k})^2}{p_{0,k}(1 - p_{0,k})}$$

with the same symbols [@meuwissen2020].

**Never negative.** It has the algebraic form of the standardised-variance
definition of $F_{ST}$ -- a variance of allele frequencies divided by
$p_0(1-p_0)$, the form @wright1951 introduced and @holsinger2009 separate from
the several other quantities that carry the same name -- but applied to *one
population through time* rather than to a set of populations in space. That is
an analogy of form, not an identity: nothing here is a between-subpopulation
variance, and the chapter's own $F_{ST}$ column (@sec-era-step6) is the Nei
$G_{ST}$ of `gd_partition()`, a different statistic on a different partition.
$F_{drift}$ is exactly the quantity constrained when `G_VR2` goes into the
optimiser.

It protects against random change in phenotypic value for traits **outside** the
selection objective, against rare deleterious recessives rising to dangerous
frequencies, and against irreversible loss of rare alleles.

**It must not be read as a measure of *this programme's* selection intensity,
even though @meuwissen2020 offer exactly that reading for a locus under
directional selection.** They call $\delta p^2/[p_0(1-p_0)]$ "an approximation
to the squared total intensity ($i^2$) applied to the marker" (citing Liu &
Woolliams 2010 for the underlying non-linearity), and where the only thing
moving frequencies is truncation selection on a trait, that reading is theirs
to make. But $F_{drift}$ is accumulated over **every** source of displacement a
scheme has -- selection, pure drift, and a diversity constraint that pushes
frequencies on its own -- and the three are not separable from the statistic
alone. @sec-era-step4 runs a crossing block with **no selection of any kind**
and $F_{drift}$ climbs over twelve cycles to a value the step reports, which
under an unconditional intensity reading would be a large intensity applied by
nothing; @sec-era-step2's `max_div` plan moves $F_{drift}$ by minimising
coancestry, not by selecting on any trait. Even where selection is the only
driver, the displacement is not a function of frequency alone: it depends on
the average effect of a gene substitution at that locus, $a + d\,(q - p)$
[@falconer1996, Ch. 7] -- classically written $\alpha$, which in this book is
the diversity-loss statistic instead (@sec-heterosis says the same) -- and on
the phenotypic standard deviation of the trait under selection, neither of
which appears anywhere in $F_{drift}$. So an intensity read off `F_drift` alone,
with no independent knowledge of what moved it, cannot be trusted, and this
chapter does not attempt one. What constraining `F_drift` does constrain is the
squared *displacement* of the genome from its base, whatever moved it.

A programme running six cycles of genomic selection with a VanRaden
$\mathbf{G}$ and reporting $\Delta F \approx 1\%$ has reported
$\Delta F_{drift}$ on the training panel. Quantitative disease-resistance loci
not yet prioritised may have hitch-hiked to extreme frequencies alongside yield
QTL, and none of that appears in the reported metric.

## The classical equivalence, and the covariance that destroys it

Under purely random drift, with $E[\delta p_t \mid p_0] = 0$ for every $p_0$,

$$\frac{\mathrm{var}(p_{t,k})}{p_0(1-p_0)} = 1 - \frac{H_t}{H_0}
  \;\;\Longrightarrow\;\; F_{drift} = F_{hom}.$$

The paper's central result quantifies the departure:

$$F_{hom} - F_{drift} \;=\; 2\,\mathrm{cov}\!\left(
  \frac{\delta p_{t,k}}{\sqrt{p_{0,k}(1-p_{0,k})}},\;
  \frac{p_{0,k} - 1/2}{\sqrt{p_{0,k}(1-p_{0,k})}}\right)$$

[@meuwissen2020, their Eq. 3], where $\delta p_{t,k} = p_{t,k} - p_{0,k}$ and
the covariance is taken over markers with divisor $N_{SNP}$ -- which is why
`freq_metrics()` computes it as `2 * mean(x * y)` and not with `cov()`. The two
differ for two independent reasons: `stats::cov()` divides by $N_{SNP} - 1$,
and $\overline{xy} = \mathrm{cov}_{N}(x, y) + \bar x\,\bar y$, so the product
term vanishes only if *one of the two means* is zero -- either will do, and
neither is zero here.

In words: **if the direction of frequency change is correlated with the starting
frequency, the two measures diverge.** The non-obvious part is that this
covariance is induced by the *diversity management itself*, not by directional
selection -- @meuwissen2020 demonstrate this by running the scheme with random
GEBVs, and the divergence persists.

### Verifying the identity

`freq_metrics()` returns all three quantities in one marker sweep. The identity
is asserted in `tests/testthat/test-lenses.R`; here it is on live data:

```{r}
#| label: era-identity
set.seed(31)
idx <- sample.int(ctx$N, 60)
fm <- freq_metrics(idx, ctx)
fmt(fm[c("F_hom", "F_drift", "cov_diag")], 6)
```

```{r}
#| label: era-identity-residual
stopifnot(isTRUE(all.equal(fm[["cov_diag"]], fm[["F_hom"]] - fm[["F_drift"]])))
# The two standardised marker vectors of Eq. 3, on the markers polymorphic in
# the base, so `2 * mean(x * y)` can be put next to `2 * cov(x, y)`.
keep <- ctx$p0 > 1e-6 & ctx$p0 < 1 - 1e-6
sdv  <- sqrt(ctx$p0[keep] * (1 - ctx$p0[keep]))
x    <- (colMeans(ctx$X[idx, keep, drop = FALSE]) - ctx$p0[keep]) / sdv
y    <- (ctx$p0[keep] - 0.5) / sdv
stopifnot(isTRUE(all.equal(2 * mean(x * y), unname(fm[["cov_diag"]]))))
c(F_hom_minus_F_drift = unname(fm[["F_hom"]] - fm[["F_drift"]]),
  cov_diag            = unname(fm[["cov_diag"]]),
  residual            = unname(fm[["cov_diag"]] - (fm[["F_hom"]] - fm[["F_drift"]])),
  mean_x              = mean(x),
  mean_y              = mean(y),
  two_times_R_cov     = 2 * stats::cov(x, y))
```

Exact to machine precision, and the two `stopifnot()`s mean this page cannot
render if it stops being. The last three entries are the point of the previous
paragraph: neither marker-wise mean is zero, so `2 * cov(x, y)` returns
`r sprintf("%.6f", 2 * stats::cov(x, y))` where the quantity Eq. 3 defines is
`r sprintf("%.6f", fm[["cov_diag"]])`. And when the selection *is* the base,
both lenses must read zero:

```{r}
#| label: era-base-zero
fmt(freq_metrics(seq_len(ctx$N), ctx)[c("F_hom", "F_drift", "cov_diag")], 12)
```

### Why each matrix generates a sign

```{r}
#| label: era-matrix-signs
#| echo: false
knitr::kable(data.frame(
  Matrix = c("G_VR2", "G_0.5", "G_VR1", "G_i(p)"),
  Covariance = c("positive", "negative", "positive, attenuated", "negative"),
  Mechanism = c(
    "the constraint penalises delta-p^2, so moving an allele to the *far* extreme costs more than to the near one; the optimiser pushes rare alleles to 0 and common ones to 1",
    "frequencies are measured as deviations from 0.5, so the optimiser buys 'diversity' by pushing everything to intermediate frequencies",
    "G_VR2 reweighted by 2p0(1-p0), giving more weight to already-intermediate loci and less room at the extremes",
    "uses accumulated intensity instead of delta-p^2; since di/dp = [p(1-p)]^-1/2, moves toward the extremes get more expensive"),
  check.names = FALSE),
  caption = "The sign of the bias is a property of the matrix, not of the data. Mechanisms as set out by Meuwissen et al. (2020), their Discussion and Supplementary Information 1; the di/dp expression in the last row is theirs, after Liu and Woolliams (2010).")
```

::: {.callout-warning}
## Your MAF spectrum decides how bad this is

@meuwissen2020 state this as an expectation rather than as a measured
comparison: "the covariance generated is expected to be larger for WGS data
where allele frequencies are extreme with typical MAF close to 0, than for SNP
(chip) panels where allele frequencies are generally closer to 1/2" (their
second Conclusion bullet), having observed that the strength of the covariance
"depends on the distribution of $(p_0 - \tfrac12)$" (their Discussion). Their own
simulation used four panels, all random samples of the same whole-genome
sequence data (an SNP-BLUP panel, a neutral panel, a QTL panel and a second
management panel) -- none a SNP chip -- so the comparison across a WGS panel
and a chip panel is an inference from the algebra of Eq. 3 and not a result of
their experiment either.

The algebra can be made a number, which is worth doing because it decides whether
`G_VR2` is usable. Apply one deterministic management rule -- move every frequency
5% of the way toward 0.5, the caricature of a heterozygosity-maximising
constraint used again in @sec-era-step4 -- to a sequence-like panel whose folded
site-frequency spectrum has density proportional to $1/[p(1-p)]$, and then to
MAF-filtered subsets of the *same* draw:

```{r}
#| label: era-maf-weight
set.seed(21)
# Folded site-frequency spectrum with density proportional to 1/[p(1-p)] on
# [1e-3, 1 - 1e-3]: the logit of p is uniform. One draw, filtered four ways.
p_seq <- plogis(runif(2e5, qlogis(1e-3), qlogis(1 - 1e-3)))
maf_row <- function(floor, pull = 0.05) {
  p0 <- p_seq[pmin(p_seq, 1 - p_seq) >= floor]
  p1 <- p0 + pull * (0.5 - p0)                 # the rule, applied exactly
  s2 <- p0 * (1 - p0)
  c(maf_floor = floor, markers = length(p0),
    W = mean((p0 - 0.5)^2 / s2),               # the panel's own weight
    F_hom = 1 - mean(p1 * (1 - p1) / s2),
    F_drift = mean((p1 - p0)^2 / s2),
    cov_diag = 2 * mean((p0 - 0.5) * (p1 - p0) / s2))
}
mafw <- as.data.frame(t(sapply(c(0.001, 0.01, 0.05, 0.20), maf_row)))
# Under this rule all three lenses are multiples of one panel statistic, W.
stopifnot(isTRUE(all.equal(mafw$F_drift, 0.05^2 * mafw$W)),
          isTRUE(all.equal(mafw$cov_diag, -2 * 0.05 * mafw$W)))
keep0  <- ctx$p0 > 1e-6 & ctx$p0 < 1 - 1e-6        # as freq_metrics() drops them
W_book <- mean((ctx$p0[keep0] - 0.5)^2 / (ctx$p0[keep0] * (1 - ctx$p0[keep0])))
fmt(mafw, 4)
```

Under a rule of this form $\delta p = c\,(0.5 - p_0)$ the algebra is exact:
$F_{drift} = c^2 W$ and `cov_diag` $= -2cW$ with
$W = \overline{(p_0 - 0.5)^2 / [p_0(1-p_0)]}$, the panel statistic the
`stopifnot()` above verifies. Everything about the panel enters through $W$, and
$W$ is `r fmt(mafw$W[1] / mafw$W[3], 1)` times larger at a MAF floor of 0.001
than at 0.05 on the same draw: `cov_diag` moves from
`r fmt(mafw$cov_diag[1], 3)` to `r fmt(mafw$cov_diag[3], 3)` for the same 5%
pull. For reference, this book's own hybrid panel has $W =$ `r fmt(W_book, 2)`
over the `r sum(keep0)` markers polymorphic in the base, which sits at the
chip-like end of that table -- a consequence of the $U(0.05, 0.95)$ ancestral
draw of `simulate_lines()`, not of anything biological.

**Translated to plants, as a stated expectation and not as a measurement on real
germplasm:** GBS or skim-seq retaining the full spectrum puts a programme at the
high-$W$ end; a 20-50k chip filtered to MAF > 0.05 puts it at the low-$W$ end. The
rule above is a caricature, not an optimiser, so the *ratio* between panels is
the transferable part and the absolute values are not.
:::

## The other matrices, briefly

**`F_IBD` via `G_LA`.** Instead of assuming a 50/50 transmission probability for
each parental allele (as the pedigree matrix $\mathbf{A}$ does), linkage
analysis over the markers infers which parental haplotype was actually
transmitted along the chromosome. It requires pedigree **and** markers. It
measures IBD relative to a base explicitly defined as unrelated and
non-inbred, so frequency change is determined by the base population rather
than by linkage disequilibrium generated during selection -- and the
problematic covariance disappears. It was the **only scheme tested that met
both the $\Delta F_{hom}$ and $\Delta F_{drift}$ targets**, on the managed panel
and at unmonitored neutral loci. On gain it was not the highest in absolute
terms; what @meuwissen2020 report is that at the *same* averaged inbreeding --
interpolating each scheme's gain to the mean of $F_{hom}$ and $F_{drift}$ --
`G_LA` beat $\mathbf{A}$ in 65 of 100 replicates and `G_ROH` in 62 of 100
(their Genetic Gain section).

For a hybrid programme this is the point of greatest friction, and also the one
where the news is best. In biparental populations and backcross families
`G_LA` is natural and cheap -- phase is known and the number of meioses since
the base is small. In a maize recurrent-selection population with a complete
recycling pedigree, which commercial programmes generally have because every DH
line records its F1 parents, it is feasible. In a germplasm panel *without*
pedigree it is inapplicable, and the honest alternative is segment-based IBD.

**`F_ROH`.** A run of homozygosity is an uninterrupted stretch of homozygous
markers; the logic is that many consecutive markers are unlikely to be IBS by
chance, so the stretch is probably IBD. The fundamental ambiguity is that
`F_ROH` is IBD when runs are **long** (recent inbreeding) and homozygosity when
they are **short** (ancient inbreeding, accumulated IBS). The minimum-length
threshold is therefore a knob that selects the *age* of inbreeding you are
measuring, and there is no well-defined reference base.

Empirically, `G_ROH` behaved more like `G_0.5` than like an IBD measure. It also
is **not guaranteed positive semi-definite** -- its elements are computed
pairwise -- and required substantial diagonal additions for inversion, which
made it diagonally dominant and artificially inflated the number of parents
selected.

In an inbred parent line the entire genome is one ROH and $F_{ROH} \approx 1$:
the metric saturates for exactly the reason assumption A2 makes every
individual-level coefficient useless. Applied instead to the *hybrids*, or to
the segregating crossing block from which DH lines are induced, it is useful for
diagnosing erosion of an elite pool. For *management* inside an optimiser, check
positive definiteness carefully -- @sec-era-step7 shows the check.

### Side by side

```{r}
#| label: era-side-by-side
#| echo: false
knitr::kable(data.frame(
  Matrix = c("A", "G_VR2", "G_VR1", "G_0.5", "G_i(p)", "G_LA", "G_ROH"),
  Class = c("IBD (expectation)", "drift", "drift", "homozygosity", "drift",
            "IBD", "homozygosity, read as IBD"),
  `Behaviour in the optimiser` = c(
    "high gain, but exceeded targets by ~40% by not accounting for within-family LD",
    "meets the drift target *only on the managed panel*; F_hom runs away",
    "reweighted version; smaller bias, smaller gain",
    "misses the drift target by the widest margin of the eight schemes; loses genetic variance early",
    "inverts the sign of the bias; constraint only indirectly tied to F",
    "the only one meeting both targets; highest gain at equal averaged inbreeding",
    "high gain, uncontrolled drift, numerical problems"),
  check.names = FALSE),
  caption = "Seven matrices in the same optimiser (Meuwissen et al. 2020, their Tables 2 and 3, Figure 5 and Discussion). The Behaviour column is qualitative and no number from either panel is transcribed in this table; where a rate is invoked (the reference-results table below), it is always the *unmonitored neutral* panel N. The one exception is G_VR2's 'only on the managed panel', a paraphrase of their Discussion that reports a Panel M outcome without a Panel M number appearing anywhere in this chapter. Class is the lens the matrix is *computed* from, which is why G_ROH sits with homozygosity -- the Hybrid row of the first table is the same matrix labelled by its intended *interpretation*.")
```

## Reference results

Target $\Delta F = 0.005$ per generation ($N_e = 100$). Values at the
**unmonitored neutral loci** (their Panel N), generation 20, as reported by
@meuwissen2020. Four columns are transcribed and one is derived:

- `dF_hom`, `dF_drift` and `gap` are the three Panel N columns of their Table 2.
  `gap` is the *level* $F_{hom} - F_{drift}$ reached in generation 20, not a
  rate and not a difference of the two rate columns; it is `cov_diag` of Eq. 3
  accumulated over the 20 generations. Their reported standard errors are
  $< 2.5 \times 10^{-5}$ on the rates and $< 2.2 \times 10^{-4}$ on `gap`.
- `gain` is their Table 3: cumulative genetic gain after 20 generations,
  expressed in units of the **initial genetic standard deviation**, standard
  errors $\le 0.004$.
- `gain_per_dF` is computed here, not transcribed, and is **not** the device
  @meuwissen2020 use: their own interpolation (Genetic Gain section, their
  Figure 5) compares gains at a common generation-20 *level* of averaged
  inbreeding, $(F_{hom} + F_{drift})/2$ from their Table 3 -- a stock -- not at
  a common mean of the two *rate* columns used here. `gain_per_dF` is offered
  only as a same-chapter summary of the reference table above, and reading 1
  below states explicitly which ordering it does and does not support.

```{r}
#| label: era-reference-results
#| echo: false
ref_res <- data.frame(
  Scheme = c("G_VR2(M,M)", "G_VR2(M,D)", "G_VR1(M,M)", "G_i(p)(M,M)",
             "G_0.5(M,M)", "G_ROH(M,M)", "G_LA(M,M)", "A(M,~)"),
  Class = c("drift", "drift", "drift", "drift", "homozygosity",
            "homozygosity", "IBD", "IBD"),
  dF_hom = c(0.0103, 0.0101, 0.0080, 0.0065, 0.0073, 0.0054, 0.0043, 0.0070),
  dF_drift = c(0.0068, 0.0068, 0.0069, 0.0077, 0.0176, 0.0088, 0.0049, 0.0084),
  gap = c(0.054, 0.050, 0.021, -0.031, -0.170, -0.070, -0.010, -0.021),
  gain = c(7.124, 7.107, 6.680, 7.111, 6.734, 9.099, 7.188, 9.890),
  check.names = FALSE)
ref_res$gain_per_dF <- round(ref_res$gain /
                               ((ref_res$dF_hom + ref_res$dF_drift) / 2))
# The one qualitative claim the previous table makes about a rate: G_0.5 is the
# scheme furthest from the 0.005 drift target of the eight transcribed here.
stopifnot(ref_res$Scheme[which.max(ref_res$dF_drift)] == "G_0.5(M,M)")
knitr::kable(ref_res,
  caption = "Both rate targets are 0.005. Only G_LA meets both. gain_per_dF is derived in the chunk, the other columns are transcribed from their Tables 2 and 3.")
```

Four readings:

1. **No IBS scheme met both targets.** `G_LA` came closest: within about 2% of
   the 0.005 target on `dF_drift` on both panels M and N (their own figure for
   that lens), but 14% below target on `dF_hom` (0.0043 against 0.005) -- the
   2% is a drift-lens precision, not a joint one. It also has
   the largest `gain_per_dF`, `r max(ref_res$gain_per_dF)` against
   `r ref_res$gain_per_dF[ref_res$Scheme == "A(M,~)"]` for $\mathbf{A}$ and
   `r ref_res$gain_per_dF[ref_res$Scheme == "G_ROH(M,M)"]` for `G_ROH` -- but
   that ordering is not robust to the denominator: against `dF_hom` alone
   `G_ROH` is marginally ahead
   (`r round(ref_res$gain[ref_res$Scheme == "G_ROH(M,M)"] / ref_res$dF_hom[ref_res$Scheme == "G_ROH(M,M)"])`
   against
   `r round(ref_res$gain[ref_res$Scheme == "G_LA(M,M)"] / ref_res$dF_hom[ref_res$Scheme == "G_LA(M,M)"])`).
   The claim that survives is the replicate-level one in reading 4, not a ratio
   of table means.
2. Pedigree $\mathbf{A}$ gave the highest absolute gain but overshot by ~40% --
   neutral loci hitch-hiking with QTL through within-family LD, which the
   pedigree expectation of IBD cannot capture.
3. Using a separate panel for management instead of the GEBV panel changed gain
   by **less than 0.3%**. Not worth the complexity as a *gain* measure; it is
   still worth it as an *audit* measure, which is a different use and is the
   subject of @sec-era-step5 below.
4. At equal mean inbreeding, `G_LA` beat $\mathbf{A}$ in 65 of 100 replicates
   and `G_ROH` in 62 of 100.

Those are animal-breeding numbers under the assumptions of
@sec-era-transposition, and none of them is reproduced here. What follows is.

## The monitoring protocol {#sec-era-protocol}

Seven steps, in the order a cycle report is assembled. Each one names the
function to call, the quantity to check, and what a failure looks like.

### Step 1 -- Freeze $p_0$ at the founding cycle {#sec-era-step1}

Both lenses are ratios to a base, so the base is the first thing a cycle report
has to pin down. In this book it is `ctx$p0`, the candidate-pool allele
frequencies, set once in `build_ctx()` and carried forward unchanged by
`run_cycles()`.

At the hybrid level the base is not a free choice: on the full factorial it is
the average of the two pool bases, exactly.

```{r}
#| label: era-p0-identity
stopifnot(isTRUE(all.equal(ctx$p0, as.vector((P0["A", ] + P0["B", ]) / 2))))
# Unequal pool sizes, still a complete factorial: each pool-A line appears in
# n_pool_B hybrids and each pool-B line in n_pool_A, so the internal weight
# stays uniform within each pool and the identity is untouched.
cfg_u <- utils::modifyList(book_cfg, list(n_pool_A = 10, n_pool_B = 40))
ctx_u <- simulate_data(cfg_u)
Pu    <- pool_freqs(simulate_lines(cfg_u) / 2,
                    rep(c("A", "B"), c(cfg_u$n_pool_A, cfg_u$n_pool_B)))
# A balanced INCOMPLETE factorial -- not the full n_A*n_B hybrids, but each
# line still used equally often within its own pool (a cyclic 1-to-3 design,
# the shape of a real NCII): the identity does not need completeness.
k <- 3
idx_bal <- unlist(lapply(seq_len(book_cfg$n_pool_A), function(a) {
  bpart <- ((a - 1 + 0:(k - 1)) %% book_cfg$n_pool_B) + 1
  which(ctx$ped$a == a & ctx$ped$b %in% (bpart + book_cfg$n_pool_A))
}))
p_bal <- colMeans(ctx$X[idx_bal, , drop = FALSE])
# An UNbalanced candidate set with concentrated line usage: it is not.
p_sub <- colMeans(ctx$X[sel_truncation(ctx, 60), , drop = FALSE])
c(complete_balanced     = max(abs(ctx$p0 - (P0["A", ] + P0["B", ]) / 2)),
  complete_unequal_pools = max(abs(ctx_u$p0 - (Pu["A", ] + Pu["B", ]) / 2)),
  incomplete_balanced   = max(abs(p_bal - (P0["A", ] + P0["B", ]) / 2)),
  truncation_subset     = max(abs(p_sub - (P0["A", ] + P0["B", ]) / 2)))
```

Each pool-A line appears in `r book_cfg$n_pool_B` hybrids and each pool-B line
in `r book_cfg$n_pool_A`, so a column mean over the factorial is
$(p_A + p_B)/2$ with no weighting left over. Two consequences a breeder should
read off this identity. First, freezing $p_0$ at the hybrid level freezes both
pool bases *jointly*, so the hybrid-level lenses cannot attribute erosion to one
pool -- that is @sec-era-step6's job. Second -- and this is the part that is
easy to state wrongly -- what the identity requires is **uniform line usage
within each pool**, not completeness of the factorial as such. With
`r cfg_u$n_pool_A` lines in one pool against `r cfg_u$n_pool_B` in the other,
still complete, the identity holds to
`r sprintf("%.1e", max(abs(ctx_u$p0 - (Pu["A", ] + Pu["B", ]) / 2)))`. It holds
just as exactly (`r sprintf("%.1e", max(abs(p_bal - (P0["A", ] + P0["B", ]) / 2)))`)
on a **balanced but incomplete** factorial -- `r length(idx_bal)` of the
`r book_cfg$n_pool_A * book_cfg$n_pool_B` possible hybrids, each line used
exactly `r k` times, the shape of a real NCII design -- because each pool's
lines are still used equally often *within* that pool. What re-weights the two
pools inside the base is unequal **line usage**, which a balanced partial
factorial does not have and a selected subset typically does: the column means
of the 60-hybrid truncation set depart from $(p_A + p_B)/2$ by up to
`r fmt(max(abs(p_sub - (P0["A", ] + P0["B", ]) / 2)), 3)` on a single marker,
because truncation concentrates hybrids on a few high-value lines in each pool.
A base computed on a set with unequal line usage is therefore a different
base -- which is the failure mode below, and not a property of the design
being a factorial or not.

Now the failure mode. `build_ctx()` recomputes `p0 <- colMeans(X)` on every
call, so rebuilding the context from the *selected* hybrids each cycle -- which
is what re-running a `G`-construction function per cycle amounts to -- re-anchors
the base on the population you were about to measure:

```{r}
#| label: era-rebase
i_md   <- sel_greedy(ctx, 60, w = 0)
Xs     <- ctx$X[i_md, , drop = FALSE]
rebase <- build_ctx(X = Xs, f = molecular_coancestry(Xs), G = NULL,
                    traits = ctx$traits[i_md, , drop = FALSE],
                    ped = ctx$ped[i_md, ], n_lines = ctx$n_lines,
                    pool = ctx$pool, fL = ctx$fL, cfg = NULL)
fmt(rbind(frozen_base   = freq_metrics(i_md, ctx)[c("F_hom", "F_drift", "cov_diag")],
          rebased       = freq_metrics(seq_along(i_md), rebase)[c("F_hom", "F_drift", "cov_diag")]), 6)
```

Not attenuated -- **identically zero, both lenses**. A rebased report is not a
conservative report or a noisy one; it is a report that cannot detect erosion at
all, and it will look clean for as many cycles as you run it.

The check has to be on the base itself, and it has to be the right check. If the
base is stored per cycle, comparing each cycle's vector with cycle 0's catches
the failure; collapsing the whole history into a `range()` does not, because
`range()` discards the cycle dimension and returns the marker-wise spread of
$p_0$, the same number whether the base moved or not -- here
`r fmt(diff(range(ctx$p0)), 2)`, the whole unit interval, because
`r sum(ctx$p0 <= 1e-6 | ctx$p0 >= 1 - 1e-6)` of the `r ctx$m` markers are fixed
in the candidate pool:

```{r}
#| label: era-p0-frozen-check
chk <- function(h) {
  data.frame(
    all_identical_to_cycle0 = all(vapply(h, identical, logical(1), h[[1]])),
    max_abs_deviation       = max(vapply(h, function(p) max(abs(p - h[[1]])), 0)),
    collapsed_range         = diff(range(unlist(h))))
}
fmt(rbind(frozen  = chk(list(ctx$p0, ctx$p0, ctx$p0)),
          rebased = chk(list(ctx$p0, ctx$p0, rebase$p0))), 4)
```

The first two columns separate the two histories; the third is the same number
in both rows, which is the whole point.

**Call.** `build_ctx()` once at cycle 0; pass `p0` forward (`run_cycles()` does
this through its internal `cycle_ctx()`).
**Check.** `identical(ctx$p0, p0_cycle0)` at every cycle -- one `TRUE`, with no
tolerance. With a stored history,
`max(vapply(p0_history, function(p) max(abs(p - p0_history[[1]])), 0))`, which
must be exactly 0. Never a `range()` over the flattened history.
**Warning sign.** $F_{drift}$ near zero in a cycle where line usage was visibly
concentrated, or an $F_{drift}$ series that does not accumulate across cycles.
**Threshold.** There is none to tune: the correct value of the base drift is
exactly zero.

### Step 2 -- Compute both lenses in one marker sweep {#sec-era-step2}

`freq_metrics()` is the single sweep: it computes $He$, `alleles_lost`,
$F_{hom}$, $F_{drift}$, `cov_diag`, `cor_dp_p0` and `rare_retained` from one
`colMeans(X[idx, ])`. Anything else needing the selected frequencies belongs
inside it, not in a second function with its own sweep.

The pure max-diversity greedy is the `G_0.5` scheme in miniature: it constrains
molecular coancestry and nothing else. Its signature has to be read against a
**resampled** baseline rather than against one random draw, for the reason the
numbers below make explicit. The baseline is the package's own
`null_distribution()`, 1000 resampled subsets of the same size, and the z-scores
come from `discriminatory_power()`, so the comparison is the same machinery
every scenario in @sec-case-study is reported through:

```{r}
#| label: era-signature
# `sels` is the set of four selections built in @sec-era-fhom.
cols   <- c("He", "F_hom", "F_drift", "cov_diag", "rare_retained")
null60 <- null_distribution(ctx, 60, B = 2000, B_full = 1000, seed = 7)
panel  <- t(sapply(sels, freq_metrics, ctx = ctx))[, cols]
fmt(rbind(null_mean = colMeans(null60$full[, cols]),
          null_sd   = apply(null60$full[, cols], 2, stats::sd),
          panel), 4)
```

```{r}
#| label: era-signature-z
z <- discriminatory_power(evaluate(sels, ctx, null60$gd_ref), null60)[, cols]
fmt(as.data.frame(z), 2)
```

Result. The max-diversity plan drives $F_{hom}$ negative,
`r fmt(panel["max_div", "F_hom"], 4)`, which is
`r fmt(abs(z["max_div", "F_hom"]), 1)` s.d. below a null mean of
`r fmt(mean(null60$full[, "F_hom"]), 4)`: the selection is more heterozygous on
the monitored panel than the frozen base. Its `cov_diag` is
`r fmt(panel["max_div", "cov_diag"], 4)`, `r fmt(abs(z["max_div", "cov_diag"]), 1)`
s.d. below a null centred on
`r sprintf("%.4f", mean(null60$full[, "cov_diag"]))`. Its $F_{drift}$, however,
is `r fmt(panel["max_div", "F_drift"], 4)` against a null mean of
`r fmt(mean(null60$full[, "F_drift"]), 4)`: a z of
`r fmt(z["max_div", "F_drift"], 2)`, and
`r fmt(100 * mean(null60$full[, "F_drift"] > panel["max_div", "F_drift"]), 0)`%
of the 1000 resampled subsets pay **more** drift than it does. Truncation is the
row with the drift bill: $F_{drift}$ = `r fmt(panel["truncation", "F_drift"], 4)`,
z = `r fmt(z["truncation", "F_drift"], 1)`.

This is where one draw would have misled. The `random` row is a single subset,
and its $F_{drift}$ sits `r fmt(abs(z["random", "F_drift"]), 1)` s.d. below the
resampled mean -- *further* below it than `max_div` is -- so a two-row
`random`-versus-`max_div` comparison reports max-diversity as paying more drift
than random, which the 1000-subset null says it does not. The failure is not the
metric, it is the baseline: with a null s.d. of
`r sprintf("%.4f", sd(null60$full[, "F_drift"]))` on a quantity whose value is
`r fmt(panel["max_div", "F_drift"], 4)`, no single draw can serve as a control.

Interpretation, and what it does not license. At this scale, on a single
selection from a fixed candidate set, **minimising** molecular coancestry (the
`max_div` plan, `w = 0`, which maximises gene diversity `GD = 1 - theta`) buys a
measurable divergence between the two lenses -- the negative covariance that the
`G_0.5` row of the matrix-sign table predicts -- and **no** measurable drift
excess over random sampling. The drift excess @meuwissen2020 report for `G_0.5`
($\Delta F_{drift} = 0.0176$ against a 0.005 target) is a 20-generation
*cumulative* result under recurrent optimal contributions, where each cycle's
frequency change compounds on the last; one subset drawn once from one candidate
pool is not that experiment, and the honest way to run it here is the
multi-cycle arm of @sec-era-step4, where the same pull toward 0.5 does produce
drift at more than twice the neutral rate. What survives at one cycle is the
covariance, and a report carrying only $F_{hom}$ would not show it.

One thing the table quietly says about the *weight* rather than the metric:

```{r}
#| label: era-w1-equals-truncation
identical(sort(sel_greedy(ctx, 60, w = 1)), sort(sel_truncation(ctx, 60)))
```

At this scale `w = 1` is not a compromise between index and coancestry -- it
returns the truncation set exactly, because the index is standardised to unit
variance (`sd(ctx$index)` = `r fmt(sd(ctx$index), 3)`) while the spread of
$\theta$ across candidate subsets of the same size is
`r sprintf("%.2e", sd(null60$cheap[, "theta"]))`, the standard deviation of the
`theta` column over the `r nrow(null60$cheap)` cheap-metric resamples of the
same null. That is a ratio of
`r fmt(sd(ctx$index) / sd(null60$cheap[, "theta"]), 0)`: three orders of
magnitude, not two. A linear weight has to be calibrated to that ratio, which
is why `sel_greedy_budget()` sweeps a *logarithmic* grid of `w` instead of
trusting any single value.

**Call.** `freq_metrics(idx, ctx)` per cycle; `metrics_cheap()` if you also need
the coancestry block, `metrics_full()` post-hoc for the expensive ones.
**Check.** Both lenses present in the report, side by side, with the sign of
$F_{hom}$ preserved (never take an absolute value or clamp at zero).
**Warning sign.** A single "inbreeding" column in the cycle report. If you
cannot tell which lens it is, it is the one your constraint used, and the other
one is unmeasured.
**Threshold.** Not the $\Delta F$ target. Both columns here are *stocks*
accumulated against the frozen base, and a stock cannot be compared with a rate
-- @sec-era-step4 opens on exactly that point. Within one cycle the reference is
the resampled null in the same table, i.e. the z columns above, read on **both**
lenses separately and never on their average. The comparison with the
programme's $\Delta F$ budget requires a trajectory and happens only through the
regression of @sec-era-step4.

### Step 3 -- Report the covariance, not just the two lenses {#sec-era-step3}

`cov_diag` is the gap between the two lenses, and by Eq. 3 it is not an
approximation of the gap -- it *is* the gap, which the `stopifnot()` above
enforces on every render. The operational point is that `cov_diag` is the only
column that tells you *why* the two lenses disagree, and its sign identifies the
mechanism from the matrix-sign table above: negative means the management pushed
frequencies toward 0.5, positive means it pushed them to the extremes.

**Call.** `freq_metrics(idx, ctx)[["cov_diag"]]`, already computed by step 2 at
no extra cost; `cor_dp_p0` from the same call is the scale-free version.
**Check.** `cov_diag == F_hom - F_drift` to machine precision. If it does not,
the base and the frequencies were computed on different marker sets.
**Warning sign.** $|$`cov_diag`$|$ of the same order as the stock it is meant to
explain. In the panel of @sec-era-step2 the max-diversity plan reports
`cov_diag` = `r fmt(panel["max_div", "cov_diag"], 4)` against an
$F_{drift}$ of `r fmt(panel["max_div", "F_drift"], 4)`: the discrepancy between
the two lenses is `r fmt(abs(panel["max_div", "cov_diag"] / panel["max_div", "F_drift"]), 1)`
times the quantity being reported.
**Threshold.** No absolute one, and not the $\Delta F$ budget -- `cov_diag` is a
stock like the two lenses it separates. Compare it with $F_{drift}$ in the same
row, as just done, and with the resampled null: the `null_sd` row of
@sec-era-step2's table puts the max-diversity plan
`r fmt(abs(z["max_div", "cov_diag"]), 1)` s.d. out, which is the statement that
the covariance was induced by the management rather than by the sampling.

### Step 4 -- Regress $\log(1 - F)$ on cycle, and read $R^2$ {#sec-era-step4}

A single cycle's $F$ is a stock and cannot be compared to a rate target. The
rate comes from the trajectory. Before trusting the regression on a real
programme, calibrate it on a case whose answer is known.

Let $n_{gam}$ be the number of allele copies the pool samples per cycle. Drawing
them at random from the previous cycle's frequencies gives
$\mathbb{E}[H_t] = H_0\,(1 - 1/n_{gam})^t$ exactly, so
$\Delta F = 1/n_{gam}$ and $N_e = n_{gam}/2$; with $n_{gam} = 2N$ this is the
idealised population of [@falconer1996, Ch. 3] and $\Delta F = 1/(2N)$. In a
two-pool programme $n_{gam}$ is not a matter of taste, and the two natural
regimes differ by a factor of two:

- a **crossing block** of $n$ plants intermated at random passes $2n$ gametes,
  so $\Delta F = 1/(2n)$ and $N_e = n$;
- a pool **re-derived as $n$ doubled-haploid lines** passes only $n$ distinct
  allele copies, because each DH is homozygous and contributes one allele
  identity per locus, so $\Delta F = 1/n$ and $N_e = n/2$. This is literally
  what `next_pool()` does in @sec-cycles: one Bernoulli draw per locus per new
  line.

```{r}
#| label: era-traj-fun
# One heterotic pool, n_gam allele copies sampled per cycle, n_cyc cycles,
# `reps` independent replicates. `pull` first moves every frequency a fixed
# fraction toward 0.5: a caricature of a heterozygosity-maximising constraint,
# not an optimiser.
pool_traj <- function(n_gam, m, n_cyc, pull = 0, reps = 20, seed = 11) {
  out <- vector("list", reps)
  for (r in seq_len(reps)) {
    set.seed(seed + r)
    p0 <- runif(m, 0.05, 0.95)
    s2 <- p0 * (1 - p0)
    p  <- p0
    rows <- vector("list", n_cyc + 1L)
    for (t in 0:n_cyc) {
      rows[[t + 1L]] <- data.frame(
        rep = r, cycle = t,
        F_hom    = 1 - mean(p * (1 - p) / s2),
        F_drift  = mean((p - p0)^2 / s2),
        cov_diag = 2 * mean((p0 - 0.5) * (p - p0) / s2))
      p <- rbinom(m, n_gam, p + pull * (0.5 - p)) / n_gam
    }
    out[[r]] <- do.call(rbind, rows)
  }
  do.call(rbind, out)
}

per_cycle <- function(tr, n_gam) {
  a <- aggregate(cbind(F_hom, F_drift, cov_diag) ~ cycle, tr, mean)
  a$expected <- 1 - (1 - 1 / n_gam)^a$cycle
  # One s.d. per lens: the two lenses are not equally noisy across replicates,
  # so a deviation has to be compared with the s.d. of the lens it came from.
  a$sd_hom   <- aggregate(F_hom ~ cycle, tr, sd)$F_hom
  a$sd_drift <- aggregate(F_drift ~ cycle, tr, sd)$F_drift
  a
}

# dF from the slope, not from its first-order approximation.
dF_fit <- function(a, col) {
  d <- a[a$cycle > 0, ]                 # cycle 0 is identically zero by construction
  f <- lm(log(1 - d[[col]]) ~ cycle, d)
  b <- coef(f)[["cycle"]]
  c(dF = 1 - exp(b), Ne = 1 / (2 * (1 - exp(b))), R2 = summary(f)$r.squared)
}

n_line <- 50
n_cyc  <- 12
block  <- per_cycle(pool_traj(2 * n_line, book_cfg$m, n_cyc), 2 * n_line)
dh     <- per_cycle(pool_traj(n_line, book_cfg$m, n_cyc), n_line)
fmt(block, 5)
```

The `expected` column is the closed form, not a fit. Both lenses agree with it
and with each other, in both regimes:

```{r}
#| label: era-traj-drift-fit
dev <- function(a) max(abs(c(a$F_hom, a$F_drift) - rep(a$expected, 2)))
fit_n <- rbind(block_F_hom   = dF_fit(block, "F_hom"),
               block_F_drift = dF_fit(block, "F_drift"),
               dh_F_hom      = dF_fit(dh, "F_hom"),
               dh_F_drift    = dF_fit(dh, "F_drift"))
fmt(rbind(crossing_block = c(target_dF = 1 / (2 * n_line), target_Ne = n_line,
                             max_deviation = dev(block),
                             sd_hom_final = block$sd_hom[nrow(block)],
                             sd_drift_final = block$sd_drift[nrow(block)]),
          dh_pool        = c(1 / n_line, n_line / 2, dev(dh),
                             dh$sd_hom[nrow(dh)], dh$sd_drift[nrow(dh)])), 5)
fmt(fit_n, 5)
```

For the crossing block the fitted rates are `r fmt(fit_n["block_F_hom", "dF"], 5)`
and `r fmt(fit_n["block_F_drift", "dF"], 5)` against a target of
`r fmt(1 / (2 * n_line), 5)`, giving $N_e$ of
`r fmt(fit_n["block_F_hom", "Ne"], 1)` and
`r fmt(fit_n["block_F_drift", "Ne"], 1)` against the true `r n_line`. For the
DH-derived pool of the same `r n_line` lines they are
`r fmt(fit_n["dh_F_hom", "dF"], 5)` and `r fmt(fit_n["dh_F_drift", "dF"], 5)`
against a target of `r fmt(1 / n_line, 5)`, giving
$N_e \approx$ `r fmt(fit_n["dh_F_drift", "Ne"], 0)` against the true
`r n_line / 2`. In both regimes the largest deviation of either lens from the
closed form (`r fmt(dev(block), 4)` and `r fmt(dev(dh), 4)`) falls on the
homozygosity lens, and is within *its* between-replicate standard deviation at
the last cycle (`r fmt(block$sd_hom[nrow(block)], 4)` and
`r fmt(dh$sd_hom[nrow(dh)], 4)`, 20 replicates); the drift lens is the tighter
of the two on both counts, deviating by at most
`r sprintf("%.4f", max(abs(block$F_drift - block$expected)))` and
`r sprintf("%.4f", max(abs(dh$F_drift - dh$expected)))`. $R^2 \ge 0.999$ on all
four fits.

Two things to carry out of that. The regression recovers $N_e$ to within the
sampling noise of 20 replicates of `r book_cfg$m` loci, which is what licenses
the $R^2$ threshold used below -- *at this scale only*: 20 replicates,
`r book_cfg$m` loci, 12 cycles, one pool, one seed family. And a pool maintained
as doubled-haploid lines drifts at twice the rate of a crossing block of the same
count, so a programme that recycles `r n_line` DH lines per pool per cycle is
running at $N_e \approx$ `r n_line / 2`, not `r n_line`. That factor of two is
structural, not a simulation artefact: it is the homozygosity of the DH itself.

A third thing, which is what the $F_{drift}$ section above needed and could not
compute at that point in the chapter. At cycle `r n_cyc` the unmanaged crossing
block stands at $F_{drift} =$ `r fmt(block$F_drift[nrow(block)], 4)` and the DH
pool at `r fmt(dh$F_drift[nrow(dh)], 4)`, with **no selection applied in any
cycle of either arm**: `pool_traj(pull = 0)` is a binomial resampling of the
previous cycle's frequencies and nothing else. A statistic that reaches those
values under pure drift is not an accumulated selection intensity, and reading
it as one would attribute an intensity to a process that has none.

Now the crossing block again with the caricature constraint switched on: every
frequency moved 5% of the way toward 0.5 before each drift step, which is what a
heterozygosity-maximising objective does to the frequencies without needing an
optimiser to do it.

```{r}
#| label: era-traj-managed
managed <- per_cycle(pool_traj(2 * n_line, book_cfg$m, n_cyc, pull = 0.05),
                     2 * n_line)
fmt(managed[, c("cycle", "F_hom", "F_drift", "cov_diag", "sd_drift")], 5)
```

```{r}
#| label: era-traj-managed-fit
stopifnot(isTRUE(all.equal(managed$cov_diag, managed$F_hom - managed$F_drift)))
fit_m <- rbind(F_hom = dF_fit(managed, "F_hom"), F_drift = dF_fit(managed, "F_drift"))
fmt(fit_m, 5)
```

```{r}
#| label: fig-era-lens-divergence
#| fig-cap: "The two lenses on one heterotic pool over 12 cycles, averaged over 20 replicates, against a base frozen at cycle 0. Left: a crossing block of 50 plants (100 gametes per cycle). Centre: the same 50 lines re-derived as doubled haploids (50 allele copies per cycle), which drifts at twice the rate. In both, the closed form 1 - (1 - 1/n_gam)^t is the dashed line and the two lenses sit on it. Right: the crossing block with every frequency moved 5% toward 0.5 before each drift step, where the homozygosity lens reports a gain of heterozygosity against the frozen base while the drift lens reports more than twice the neutral rate; the dashed line is the same neutral closed form, here a reference both lenses leave. Nothing is selected in any panel; the divergence in the right-hand panel is produced by the management rule alone."
#| fig-width: 9
#| fig-height: 3.4
arms <- list(`crossing block` = block, `DH pool` = dh, `pull toward 0.5` = managed)
op <- par(mfrow = c(1, 3), mar = c(4, 4, 2.5, 1))
for (nm in names(arms)) {
  a <- arms[[nm]]
  plot(a$cycle, a$F_drift, type = "b", pch = 19, col = "#b2182b",
       ylim = c(-0.42, 0.26), xlab = "cycle", ylab = "F against frozen p0",
       main = nm)
  lines(a$cycle, a$F_hom, type = "b", pch = 17, col = "#2166ac")
  lines(a$cycle, a$expected, lty = 2, col = "grey40")
  abline(h = 0, col = "grey70")
  legend("bottomleft", c("F_drift", "F_hom", "closed form"), bty = "n",
         cex = 0.75, pch = c(19, 17, NA), lty = c(NA, NA, 2),
         col = c("#b2182b", "#2166ac", "grey40"))
}
par(op)
```

@fig-era-lens-divergence puts the three arms side by side. At cycle `r n_cyc` the
managed pool reports
$F_{hom} =$ `r fmt(managed$F_hom[nrow(managed)], 3)` and
$F_{drift} =$ `r fmt(managed$F_drift[nrow(managed)], 3)`, a gap of
`r fmt(managed$cov_diag[nrow(managed)], 3)` which is exactly `cov_diag`. The
fitted drift rate is `r fmt(fit_m["F_drift", "dF"], 4)`, or
`r fmt(fit_m["F_drift", "dF"] / fit_n["block_F_drift", "dF"], 2)` times the
unmanaged crossing-block rate, implying $N_e =$
`r fmt(fit_m["F_drift", "Ne"], 1)` instead of `r n_line`. The fit on
$\log(1 - F_{hom})$ returns a **negative** rate,
`r fmt(fit_m["F_hom", "dF"], 4)`, and therefore a negative $N_e$, which is the
arithmetic telling you the series it was handed is not an inbreeding trajectory
at all. $R^2$ falls to `r fmt(fit_m["F_hom", "R2"], 3)` on that lens against
`r fmt(fit_n["block_F_hom", "R2"], 5)` in the unmanaged arm.

The inference, and its limits: a rule that pushes frequencies toward 0.5
generates exactly the negative covariance that the `G_0.5` row of the
matrix-sign table predicts, and it does so with no selection whatsoever, which
is the point @meuwissen2020 make with random GEBVs. What this arm does **not**
establish is the magnitude -- `pull = 0.05` is a stipulated caricature, not a
fitted optimiser, and the factor of
`r fmt(fit_m["F_drift", "dF"] / fit_n["block_F_drift", "dF"], 2)`
would move with `pull`, with $n_{gam}$ and with the ancestral frequency
distribution.
The direction is algebra; the size is this parameterisation.

**Call.** `lm(log(1 - F) ~ cycle)` on cycles $1..t$, separately for each lens;
$\Delta F = 1 - e^{\text{slope}}$; @sec-cycles does this on `run_cycles()`
output.
**Check.** $R^2$ on both fits, and the sign of $\Delta F$.
**Warning sign.** A negative fitted $\Delta F$ (the series is not an inbreeding
series), or an $R^2$ that is high while the two lenses disagree in sign -- a
smooth curve through the wrong quantity.
**Threshold.** Calibrated above: a neutral trajectory at this scale gives
$R^2 \ge 0.999$, so treat $R^2 < 0.995$ on an averaged series as a non-constant
rate. $R^2$ says *that* the rate is not constant, not *why*: a decelerating,
saturating series depresses it too, and @sec-cycles shows that case -- there the
diagnostic fires on the **unconstrained** run.

### Step 5 -- Audit on loci that did not enter the management {#sec-era-step5}

The constraint is met on the panel it is computed on. Whether it is met anywhere
else is a separate measurement, and it needs loci the optimiser never saw. Split
the marker panel in two, manage on one half, and read the report on the other:

```{r}
#| label: era-audit
mgmt  <- seq(1, ctx$m, by = 2)
audit <- seq(2, ctx$m, by = 2)
sub_ctx <- function(k) {
  build_ctx(X = ctx$X[, k], f = molecular_coancestry(ctx$X[, k]), G = NULL,
            traits = ctx$traits, ped = ctx$ped, n_lines = ctx$n_lines,
            pool = ctx$pool, fL = ctx$fL, cfg = NULL)
}
c_mgmt <- sub_ctx(mgmt)
c_audit <- sub_ctx(audit)
# The base of each half is the corresponding slice of the full base, and the
# index is untouched: only the marker set differs.
stopifnot(isTRUE(all.equal(c_mgmt$p0, ctx$p0[mgmt])),
          isTRUE(all.equal(c_mgmt$index, ctx$index)))

i_div <- sel_greedy(c_mgmt, 60, w = 0)     # maximise diversity on the mgmt half
i_bal <- sel_greedy(c_mgmt, 60, w = 1)
cols <- c("He", "F_hom", "F_drift", "cov_diag")
aud <- rbind(max_div_managed = freq_metrics(i_div, c_mgmt)[cols],
             max_div_audit   = freq_metrics(i_div, c_audit)[cols],
             index_managed   = freq_metrics(i_bal, c_mgmt)[cols],
             index_audit     = freq_metrics(i_bal, c_audit)[cols])
fmt(as.data.frame(aud), 4)
```

```{r}
#| label: era-audit-gaps
gaps <- rbind(max_div = aud["max_div_audit", ] - aud["max_div_managed", ],
              index   = aud["index_audit", ]   - aud["index_managed", ])
fmt(as.data.frame(gaps[, c("F_hom", "F_drift")]), 4)
```

Result. The plan that maximises diversity on the management half reports
$F_{hom} =$ `r sprintf("%.4f", aud["max_div_managed", "F_hom"])` there and
$F_{hom} =$ `r sprintf("%.4f", aud["max_div_audit", "F_hom"])` on the half it never saw:
the sign reverses. Its drift readings are
`r fmt(aud["max_div_managed", "F_drift"], 4)` and
`r fmt(aud["max_div_audit", "F_drift"], 4)` -- practically the same number on
both halves, a gap of `r sprintf("%.4f", gaps["max_div", "F_drift"])` against
`r sprintf("%.4f", gaps["max_div", "F_hom"])` on the homozygosity lens. The
index-weighted plan, which optimises nothing about heterozygosity, has an
$F_{hom}$ gap of only `r sprintf("%.4f", gaps["index", "F_hom"])`.

Interpretation. The homozygosity lens does not generalise off the panel it was
maximised on, while the drift lens does; the gap is a property of having
maximised, not of the marker set, which is why it nearly vanishes in the
index-weighted row. Two limits on that inference. First, in this simulation loci
are in linkage equilibrium within pool (assumption A5), so the two halves are as
independent as a split panel can be. With real linkage disequilibrium the audit
half is correlated with the management half, the measured gap shrinks, and the
audit becomes *less* sensitive: the gap measured here is close to an upper bound
on what a real split panel would report, not a typical value. Second, this is a
one-cycle, single-seed measurement at the teaching scale; the cycle-by-cycle
version of the same audit is @sec-cycles.

**Call.** Two contexts over disjoint marker sets, both anchored on slices of the
same frozen $p_0$; `freq_metrics()` on each.
**Check.** The $F_{hom}$ and $F_{drift}$ gaps between managed and audit halves.
**Warning sign.** A sign reversal in $F_{hom}$, as above. Also a *large* gap in
$F_{drift}$, which would mean the management is not even controlling drift off
its own panel.
**Threshold.** The gap should be within the sampling noise of the split. A cheap
empirical reference: the `index` row of the same table, which had no
heterozygosity objective at all.

### Step 6 -- Read the two pools separately, and never $F_{ST}$ alone {#sec-era-step6}

Everything so far is a single-population reading. A hybrid programme has two,
and they must be constrained in opposite directions: within pool, limit
$\Delta F$; between pools, *preserve* divergence, because that is the heterosis
engine (@sec-heterosis proves the identity, @sec-stages Trap 3 states the
prescription). A single global
$\tfrac{1}{2}\mathbf{c}'\mathbf{G}\mathbf{c}$ cannot express that.

```{r}
#| label: era-pools
pool_row <- function(i) {
  c(GD = gene_diversity(i, ctx), index = mean(ctx$index[i]),
    theta_pools(i, ctx)[c("theta_A", "theta_B")],
    gd_partition(i, ctx)[c("GD_T", "GD_BS", "F_ST")],
    ne_lines_pool(i, ctx), Ne_parents = ne_parents(i, ctx),
    max_line = max_line_use(i, ctx))
}
set.seed(31)
pr <- rbind(random     = pool_row(sample.int(ctx$N, 60)),
            truncation = pool_row(sel_truncation(ctx, 60)),
            max_div    = pool_row(sel_greedy(ctx, 60, w = 0)),
            capped     = pool_row(sel_greedy(ctx, 60, w = 1, max_use = 4)))
stopifnot(isTRUE(all.equal(unname(pr[, "GD"]), unname(pr[, "GD_T"]))))
fmt(as.data.frame(pr), 4)
```

Two gene-diversity columns sit in that table and they are the **same** quantity
reached by two routes, which the `stopifnot()` in the chunk enforces on every
render: `GD` is $1 - \theta$ from $\mathbf{f}$ over the selected hybrids
(@sec-catalogue), `GD_T` is $1 - \bar f$ from the line-usage weights of the
Caballero & Toro partition, and both reduce to
$\overline{\bar p_k^2 + (1 - \bar p_k)^2}$ with $\bar p$ the usage-weighted
merged frequency, because under A2 a hybrid's marker row is the average of its
two parents'. It is the same identity as $He = GD_T$ in @sec-era-fhom, and
`GD`, `GD_T` and Nei's $He$ are one number on this germplasm. The columns that
*are* a different quantity are $GD_{BS}$ and $F_{ST}$: both are built from
$\tilde f = (\theta_A + \theta_B)/2$, which no hybrid-level average can
reproduce, and both are read below.

Result. Truncation on the index costs
`r fmt(100 * (1 - pr["truncation", "GD"] / pr["random", "GD"]), 1)`% of gene
diversity against the random baseline -- which is the $\alpha$ of
@sec-catalogue, computed here on a single random draw rather than on the
resampled `gd_ref` -- and `r fmt(100 * (1 - pr["truncation", "GD"] / pr["max_div", "GD"]), 1)`%
against the max-diversity ceiling. Meanwhile $\theta_A$ rises to
`r fmt(pr["truncation", "theta_A"], 4)` against
$\theta_B =$ `r fmt(pr["truncation", "theta_B"], 4)`, and the effective number
of lines used falls to `r fmt(pr["truncation", "Ne_lines_A"], 2)` in pool A
against `r fmt(pr["truncation", "Ne_lines_B"], 2)` in pool B. One line appears
in `r pr["truncation", "max_line"]` of the 60 hybrids.

The dangerous column is $F_{ST}$. It *rises* from
`r fmt(pr["max_div", "F_ST"], 4)` to
`r fmt(pr["truncation", "F_ST"], 4)` under truncation, and so does $GD_{BS}$.
Read as divergence, that looks like a gain in heterosis potential. It is not: by
the Caballero & Toro definitions `gd_partition()` implements
[@caballero2002, Eq. 7], $GD_{BS} = \tilde f - \bar f$ with
$\tilde f = (\theta_A + \theta_B)/2$ and
$\bar f = (\theta_A + 2\theta_{AB} + \theta_B)/4$, so concentrating usage inside
*one* pool raises $\tilde f$ twice as fast as it raises $\bar f$ and therefore
raises the between-pool component -- even though the two pools' lines are
unchanged and their coancestry matrix $\mathbf{f}^L$ is fixed. All that moved is
the weight the selection puts on each line. $F_{ST}$ then divides by a $GD_T$
that fell at the same time -- the `GD_T` column of the table above, from
`r fmt(pr["max_div", "GD_T"], 4)` to `r fmt(pr["truncation", "GD_T"], 4)` -- which
amplifies the artefact: the numerator rises by
`r fmt(100 * (pr["truncation", "GD_BS"] / pr["max_div", "GD_BS"] - 1), 0)`% while
the denominator falls by
`r fmt(100 * (1 - pr["truncation", "GD_T"] / pr["max_div", "GD_T"]), 0)`%. A
within-pool collapse and a genuine divergence are indistinguishable in $F_{ST}$,
and only the separate $\theta_A$, $\theta_B$ columns tell them apart.

The limitation to state here rather than hide: `freq_metrics()` computes its two
lenses on `ctx$X` against `ctx$p0`, and @sec-era-step1 showed that hybrid base to
be $(p_A + p_B)/2$. So $F_{hom}$ and $F_{drift}$ as this package computes them
are **pooled over the two pools** and cannot, on their own, tell you which pool
eroded. The pool-level attribution comes from three other places: $\theta_A$ and
$\theta_B$ (`theta_pools()`), the effective number of lines used per pool
(`ne_lines_pool()`), and the pool-level panel `s1_monitor()` returns from the
line matrix:

```{r}
#| label: era-s1-monitor
s1 <- s1_monitor(L / 2, pool, cycle = 0)
fmt(s1, 4)
```

The two panels are the same quantities computed from the two ends of the
contract, and on the full factorial -- where line usage is uniform -- they agree
exactly:

```{r}
#| label: era-s1-identity
gp_full <- gd_partition(seq_len(ctx$N), ctx)
stopifnot(isTRUE(all.equal(gp_full[["GD_T"]], s1$GD_total)),
          isTRUE(all.equal(gp_full[["F_ST"]], s1$Gst_Nei)))
c(gd_partition_GD_T = gp_full[["GD_T"]], s1_GD_total = s1$GD_total,
  gd_partition_F_ST = gp_full[["F_ST"]], s1_Gst_Nei = s1$Gst_Nei,
  s1_Fst_Wright = s1$Fst_Wright)
```

Note which two agree. `gd_partition()`'s `F_ST` is $GD_{BS}/GD_T$, which is the
Nei $G_{ST}$ form, and it equals `s1_monitor()`'s `Gst_Nei` to machine
precision; `Fst_Wright` is the other definition and is
`r fmt(s1$Fst_Wright / s1$Gst_Nei, 2)` times larger on the same germplasm. That
is @sec-stages Trap 2, and it is the reason to name the definition in the report
rather than the symbol.

**Call.** `theta_pools()`, `gd_partition()`, `ne_lines_pool()`,
`max_line_use()`, `ne_parents()` on the selection; `s1_monitor()` on the line
matrix of each cycle.
**Check.** $\theta_A$ and $\theta_B$ separately; $GD_{BS}$ held in a band, not
maximised; the number of distinct lines used per pool.
**Warning sign.** $F_{ST}$ rising while $GD_T$ falls -- almost always a
one-pool collapse, not divergence. An `Ne_lines` asymmetry between pools of the
size in the `truncation` row above
(`r fmt(pr["truncation", "Ne_lines_A"], 1)` against
`r fmt(pr["truncation", "Ne_lines_B"], 1)`).
**Threshold.** A per-line usage cap is the direct instrument: the `capped` row
is `sel_greedy(ctx, 60, w = 1, max_use = 4)`, which holds
$|\theta_A - \theta_B|$ at
`r fmt(abs(pr["capped", "theta_A"] - pr["capped", "theta_B"]), 4)` and both
pools above `r fmt(min(pr["capped", c("Ne_lines_A", "Ne_lines_B")]), 1)`
effective lines, against an index of `r fmt(pr["capped", "index"], 2)` where
truncation reached `r fmt(pr["truncation", "index"], 2)`. That index difference
is the price, and @sec-cycles is where it is paid back or not. `max_use`
defaults to `NULL` everywhere in this package, so the cap is opt-in.

> **The diagnostic nobody runs.** Schemes based on IBD selected many parents in
> @meuwissen2020; VanRaden-style schemes selected few. If your
> optimal-contribution run is delivering high gain with a suspiciously small
> number of parents and $\Delta F$ on target, compute $F_{hom}$ and $F_{drift}$
> separately on a panel **not used** in the management (@sec-era-step5). The
> difference between them is your hidden bill. This book's
> implementation is `ne_parents()`, which @sec-case-study reports alongside the
> two lenses for every scenario. The `era-pools` table above is the small
> version of the same diagnostic: between the `random` and `truncation` rows
> gene diversity falls by
> `r sprintf("%.1f", 100 * (1 - pr["truncation", "GD"] / pr["random", "GD"]))`%
> while the effective number of parents falls from
> `r sprintf("%.1f", pr["random", "Ne_parents"])` to
> `r sprintf("%.1f", pr["truncation", "Ne_parents"])`, which is
> `r sprintf("%.0f", 100 * (1 - pr["truncation", "Ne_parents"] / pr["random", "Ne_parents"]))`%
> -- the two relative drops differ by a factor of
> `r sprintf("%.0f", (1 - pr["truncation", "Ne_parents"] / pr["random", "Ne_parents"]) / (1 - pr["truncation", "GD"] / pr["random", "GD"]))`.
> A budget written on gene diversity alone is therefore a weak constraint on the
> parent count, and the `capped` row -- the same greedy with `max_use = 4` -- is
> what holding the parent count directly costs:
> `r sprintf("%.1f", pr["capped", "Ne_parents"])` effective parents at an index
> of `r fmt(pr["capped", "index"], 2)` where truncation reached
> `r fmt(pr["truncation", "index"], 2)`. Measured on one candidate set at the
> teaching scale, with a single random draw as the `random` row.

### Step 7 -- Check the matrix you are about to invert {#sec-era-step7}

::: {.callout-note}
## An implementation note that is routinely ignored

`G_VR1`, `G_VR2`, `G_0.5` and `G_i(p)` are all only **semi**-definite positive,
so every one of them needs $\alpha$ added to the diagonal ($\alpha = 0.01$ in
@meuwissen2020, their Methods) before it can be inverted and before
the optimal-contribution algorithm will run. $\mathbf{A}$ and `G_LA` are
positive definite by construction.

**The mechanism is not the same for all four, and the difference matters here.**
For the centred VanRaden forms it is the centring: $\mathbf{Z}$ is centred on the
*population's own* allele frequencies, so its columns sum to zero over
individuals, the constant vector lies in the null space and one degree of
freedom is gone by construction. For `G_0.5`, which is the molecular coancestry
this book calls $\mathbf{f}$, it is **not** the centring -- $\mathbf{f}$ is not
centred on anything and stays in $[0, 1]$ by design (@sec-coancestry). Its
deficiency has a different cause, derived below. Both statements are checkable:

```{r}
#| label: era-centering-vs-structure
Zh  <- ctx$X - 0.5
rkf <- qr(ctx[["f"]])$rank
rkG <- qr(ctx[["G"]])$rank
c(f_equals_half_plus_gram = max(abs(ctx[["f"]] -
                                      (2 * tcrossprod(Zh) / ctx$m + 0.5))),
  f_rowsum_min            = min(rowSums(ctx[["f"]])),
  f_rowsum_max            = max(rowSums(ctx[["f"]])),
  G_sum                   = sum(ctx[["G"]]),
  G_rowsum_max_abs        = max(abs(rowSums(ctx[["G"]]))),
  rank_f = rkf, rank_G = rkG, n_lines = ctx$n_lines, N = ctx$N)
```

$\mathbf{f} = 0.5 + \tfrac{2}{m}\mathbf{Z}_h\mathbf{Z}_h'$ with
$\mathbf{Z}_h = \mathbf{X} - 0.5$, to machine precision -- a Gram matrix plus the
positive rank-one term $0.5\,\mathbf{1}\mathbf{1}'$, hence positive
semi-definite, and with row sums nowhere near zero (`f_rowsum_min` above), so
neither $\mathbf{1}$ nor any other constant vector is in its null space.
VanRaden's
$\mathbf{G}$ on the same hybrids has `sum(G)` and every row sum at numerical
zero, which *is* the centring, and its rank comes out exactly one below
$\mathrm{rank}(\mathbf{f})$ on this data set: one dimension is what centring
costs, and it is the only part of either deficiency that centring explains. The
other `r ctx$N - rkf` missing dimensions have a different cause entirely, and the
rest of this step is that cause.
:::

In a hybrid programme the deficiency is far worse than one zero eigenvalue, and
it is structural rather than numerical. A hybrid's marker row is the average of
its two parents' rows, so $\mathbf{X} = \tfrac12 \mathbf{B}\mathbf{M}_L$ with
$\mathbf{B}$ the $N \times L$ parent-usage incidence matrix (a 1 in the two
columns of its parents) and $\mathbf{M}_L$ the $L \times m$ line matrix (the
`L / 2` object of `simulate_lines()`; the subscript keeps it apart from the
$\mathbf{G}_\bullet$ relationship matrices of this chapter). The hybrid
coancestry matrix is a Gram matrix of that,
$\mathbf{f} = [\mathbf{X}, 1 - \mathbf{X}][\mathbf{X}, 1 - \mathbf{X}]'/m$. Its
column space lies in the span of the columns of $\mathbf{X}$ plus the constant
vector, so its rank is bounded by
$\min(\mathrm{rank}(\mathbf{B}),\, m) + 1$ however many hybrids you have.
Measured:

```{r}
#| label: era-rank
incidence <- function(idx, ctx) {
  B <- matrix(0, length(idx), ctx$n_lines)
  B[cbind(seq_along(idx), ctx$ped$a[idx])] <- 1
  B[cbind(seq_along(idx), ctx$ped$b[idx])] <- 1
  B
}
set.seed(31)
# A separate set from the `sels` of @sec-era-fhom: these vary n_sel, which is
# the axis the rank depends on. Named apart so neither chunk shadows the other.
rk_sels <- list(full_factorial = seq_len(ctx$N),
                random_60      = sample.int(ctx$N, 60),
                truncation_60  = sel_truncation(ctx, 60),
                max_div_60     = sel_greedy(ctx, 60, w = 0),
                random_20      = sample.int(ctx$N, 20))
rk <- do.call(rbind, lapply(names(rk_sels), function(n) {
  i <- rk_sels[[n]]
  data.frame(selection = n, n_sel = length(i),
             lines_used = length(unique(c(ctx$ped$a[i], ctx$ped$b[i]))),
             rank_f = qr(ctx[["f"]][i, i])$rank,
             rank_incidence = qr(incidence(i, ctx))$rank)
}))
rk$deficiency <- rk$n_sel - rk$rank_f
fmt(rk)
```

```{r}
#| label: era-rank-check
stopifnot(all(rk$rank_f == rk$rank_incidence))
ev_min <- min(eigen(ctx[["f"]], symmetric = TRUE, only.values = TRUE)$values)
c(rank_full_factorial = rk$rank_f[1], n_lines = ctx$n_lines)
c(min_eigenvalue_f = signif(ev_min, 3))
```

Result. In all five selections the rank of $\mathbf{f}$ equals the rank of the
parent-usage incidence matrix exactly. On the full factorial that is
`r rk$rank_f[1]` $= L - 1$ for `r ctx$n_lines` lines, against `r ctx$N` hybrids:
a deficiency of `r rk$deficiency[1]`, with a smallest eigenvalue at
`r signif(ev_min, 2)`. A subset of 20 hybrids drawn from 50 lines is full rank.

Interpretation. The deficiency is not a rounding artefact and does not shrink
with more markers; it is the factorial structure. $N$ hybrids built from $L$
lines live in an at most $L$-dimensional space, so any hybrid-level optimiser
that needs $\mathbf{f}^{-1}$ must regularise, and the amount of regularisation
required is predictable *before* any eigendecomposition: compute
$\mathrm{rank}(\mathbf{B})$, which costs one `qr()` on an $N \times L$ 0/1
matrix. The equality `rank_f == rank_incidence` is a measurement over five
selections at one scale, not a theorem -- the algebra above bounds
$\mathrm{rank}(\mathbf{f})$ by $\mathrm{rank}(\mathbf{B}) + 1$, and the extra
dimension the bound allows is not attained in any of the five. It is also the
reason this book's selection machinery never inverts $\mathbf{f}$:
`metrics_cheap()` and the DE fitness only ever evaluate the quadratic form
$\mathbf{c}'\mathbf{f}\mathbf{c}$, which needs no inverse. The
quadratic-programming routes do need a definite matrix, and both
`optimal_contributions()` and `ocs_quadprog()` add `diag(1e-8, n)` for exactly
this reason -- four orders of magnitude above the numerical zero reported above,
and small enough not to move the optimum. See @sec-ocs and @sec-g-singular.

**Call.** `qr()` on the usage incidence matrix, or on $\mathbf{f}$ restricted to
the candidate set; `eigen(..., only.values = TRUE)` for the spectrum if you want
the size of the numerical zero.
**Check.** `rank(f) < n_sel`, and how far below.
**Warning sign.** A quadratic-programming solver that returns without error but
with contributions concentrated on a handful of hybrids: a singular matrix makes
the objective flat along $N - \mathrm{rank}$ directions, and the solver will pick
a vertex.
**Threshold.** The added diagonal $\alpha$ must exceed the largest magnitude of
the numerically-zero eigenvalues, here about `r signif(abs(ev_min), 1)`, by
several orders of magnitude; $\alpha = 0.01$ on a matrix whose diagonal averages
`r fmt(mean(diag(ctx[["f"]])), 2)` is a perturbation of about
`r fmt(100 * 0.01 / mean(diag(ctx[["f"]])), 1)`% and must be documented as such,
because it changes the optimum.

## What not to do

Four recommendations that do not need their own step. Each one is labelled with
what backs it, because they are not backed equally: one is measured in this
chapter, three are positions taken by @meuwissen2020 and carried forward here as
*their* argument, and none of the three has been tested on plant germplasm
anywhere in this book.

**Do not maximise marker heterozygosity as an objective.** *Measured here:*
raising heterozygosity on the monitored panel does not raise it at unmonitored
loci -- @sec-era-step5 measures the sign reversal, under linkage equilibrium,
which makes it a best case. *Their argument, not measured here:* doing so also
raises IBD-based inbreeding, because "changes in allele frequency result from
multiple copies of a subset of base generation alleles, so increasing frequency
is promoting IBD based inbreeding" [@meuwissen2020, Discussion, Management of
Diversity]. This book has no IBD route at all -- every matrix in it is
identity-by-state -- so that half is a reason to distrust the objective, not a
result reproduced here.

**Region-specific diversity targets rarely earn their complexity.** *Their
argument.* @meuwissen2020 give two grounds in the same paragraph: that causative
alleles of quantitative traits are quite evenly distributed across the genome
(their Discussion, citing Wood et al. 2014 on human stature -- an observation
about one species and one trait, not about maize), and that "what happens in a
sample of loci does not necessarily predict what happens at loci outside that
subset", which is the claim @sec-era-step5 does test, at the marker level. They
name the major histocompatibility complex as the case that *does* earn a
constraint of its own; the plant analogue offered here is R-gene-type resistance
loci, where allelic diversity has direct and known value. That analogy is
offered as such, and nothing in this book measures it.

**Deleterious recessives are a load problem, not a diversity problem.** *Their
judgement:* for regions carrying known recessive defects, "direct inclusion of
the known defects in the breeding goal seems more effective in controlling their
frequencies" than prioritising the region for diversity management
[@meuwissen2020, Discussion]. Neither they nor this book compare the two
routes quantitatively, so read it as the expectation of the authors of the
scheme, not as a measured efficiency. In a hybrid programme there is a second
and independent reason, which follows from A1 and A2 rather than from anyone's
judgement: a recessive carried at low frequency in pool A is masked in every
$A \times B$ hybrid unless pool B carries it too, so the load is expressed in
the crossing block and in the DH induction, not in the product.

**In a pure conservation scheme the problem is expected to be *worse*, not
better.** *Their expectation, stated as one:* "the potential covariance between
the change in allele frequency and the initial frequency is expected to
increase, since the inbreeding management term is more important in pure
conservation schemes" [@meuwissen2020, Discussion], which is why they argue the
IBD recommendation is stronger there. Their simulations all carry selection, so
no conservation scheme was run to test it; the sign of the mechanism is
nevertheless the same algebra as everything above -- a constraint that is the
only thing moving frequencies is a constraint that generates the whole
covariance.

## Limits of the animal-to-plant transposition {#sec-era-transposition}

```{r}
#| label: era-transposition
#| echo: false
knitr::kable(data.frame(
  `Assumption in the paper` = c(
    "discrete generations, two sexes, one contribution per individual",
    "reliable, deep pedigree",
    "base = simulated founder population at mutation-drift equilibrium",
    "target Ne = 100 over 20 generations",
    "one population, one target",
    "heterozygosity is always desirable"),
  `Situation in a hybrid crop programme` = c(
    "same line reused for years, cloning of parents through DH maintenance, overlapping cycles, one line used in dozens of crosses",
    "recycling pedigree usually complete inside the programme, absent for introduced germplasm and genebank accessions",
    "base is an administrative choice (cycle 0) with pre-existing heterotic structure already in it",
    "shorter cycles with GS; this book measures no commercial elite Ne, but @sec-era-step6's own teaching-scale example shows Ne_parents falling well below the pool size under plain truncation (39.3 to 13.0)",
    "two heterotic pools with opposite-signed constraints, several target environments",
    "desirable in the product, undesirable in the parents, irrelevant at the individual level once lines are fixed -- and in none of the three cases is it what F_hom measures, which is merged-pool gene diversity (@sec-era-fhom)"),
  check.names = FALSE))
```

None of this invalidates the central conclusion. It is an **algebraic** property
of the optimiser, not a biological property of the species. What it changes is
*which matrix is practicable*, and -- for a two-pool programme -- that one global
constraint is never enough.

## Operational checklist

The protocol above, in one column, with the step that implements each item.

| Check | Step |
|---|---|
| Is the base $p_0$ frozen at the founding cycle, and documented? | @sec-era-step1 |
| Is the hybrid-level base the balanced average of the two pool bases? | @sec-era-step1 |
| Are `F_hom` and `F_drift` computed **separately** and reported together? | @sec-era-step2 |
| Is it recorded that `F_hom` is merged-pool gene diversity and **not** F1 heterozygosity, with $GD_{BS}$ reported next to it? | @sec-era-fhom |
| Is every single-cycle value compared with a **resampled** null rather than with one random draw? | @sec-era-step2 |
| Is the sign of `F_hom` preserved rather than clamped? | @sec-era-step2 |
| Is `cov_diag` reported as a diagnostic, with its sign? | @sec-era-step3 |
| Is $\log(1 - F)$ regressed on cycle, with $R^2$ shown, for both lenses? | @sec-era-step4 |
| Is $\Delta F$ read as $1 - e^{\text{slope}}$ rather than $-\text{slope}$? | @sec-era-step4 |
| Is there a panel of loci **not used in management**, for audit? | @sec-era-step5 |
| Are $\theta_A$ and $\theta_B$ reported separately, never only $F_{ST}$? | @sec-era-step6 |
| Is the number of parent lines used per pool biologically plausible? | @sec-era-step6 |
| Is $GD_{BS}$ held in a band rather than maximised? | @sec-era-step6 |
| Is the matrix's rank checked before any inversion, and the added $\alpha$ documented? | @sec-era-step7 |
| Is true genetic variance (or the best estimate of it) being tracked? | @sec-response |
| Where possible, is an IBD alternative compared in parallel? | @sec-era-transposition |
