# Diversity metrics in the genomic era {#sec-genomic-era}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
```

In classical theory, inbreeding measured as **loss of heterozygosity** and
inbreeding measured as **drift of allele frequencies** are the same thing.
@meuwissen2020 show that inside genomic optimal-contribution schemes they stop
being the same thing, and can diverge badly, whenever diversity is controlled
through **identity by state** on a marker panel. Only when control is through
**identity by descent** is the classical equivalence restored.

The practical consequence is uncomfortable:

> You hit exactly the target you wrote into the optimiser's constraint, and you
> pay an unmonitored price in the metric you did not write.

For a plant breeder: if you constrain $\tfrac{1}{2}\mathbf{c}'\mathbf{G}\mathbf{c}$
using a VanRaden $\mathbf{G}$, you are constraining **drift**. The loss of
heterozygosity in your germplasm may be running at twice the rate you believe
you are controlling.

## Three lenses, not three synonyms

```{r}
#| echo: false
knitr::kable(data.frame(
  Lens = c("Drift", "Homozygosity", "IBD", "Hybrid"),
  `What the metric actually sees` = c(
    "squared allele-frequency change relative to a base",
    "proportion of heterozygosity remaining",
    "segments inherited by descent from a defined base",
    "haplotype homozygosity used as a proxy for IBD"),
  `Inbreeding metric` = c("F_drift", "F_hom", "F_IBD", "F_ROH"),
  `Relationship matrix` = c("G_VR2, G_VR1, G_i(p)", "G_0.5 (molecular coancestry)",
                            "A, G_LA", "G_ROH"),
  check.names = FALSE),
  caption = "The three lenses. G_0.5 is the matrix this book calls f -- see @sec-coancestry.")
```

::: {.callout-important}
## Every inbreeding metric is a ratio to a reference population

Without explicitly declaring your base -- cycle 0 of the recurrent programme? a
landrace panel? the founding heterotic pool? -- none of the numbers below mean
anything. This is the commonest error in plant applications: using the *current*
generation's frequencies as the base, which zeroes the historical memory of
erosion.
:::

## The metrics, one at a time

### Rate of inbreeding and effective size

What you manage is not $F$ but its rate of accumulation. $F$ is a stock;
$\Delta F$ is a flow.

$$\Delta F = \frac{F_t - F_{t-1}}{1 - F_{t-1}}, \qquad N_e = \frac{1}{2\Delta F}$$

Since $1 - F_t = (1 - \Delta F)^t$, a regression of $\log(1 - F_t)$ on $t$ has
slope $\approx -\Delta F$. **The linearity of that regression is itself a
diagnostic:** in the paper, $\log(1 - F_{drift})$ was approximately linear under
every scheme, while $\log(1 - F_{hom})$ showed marked curvature under some,
notably the ROH-based one. Curvature means a non-constant rate -- your $N_e$
target is not being met steadily.

A tropical maize recurrent programme targeting $N_e = 50$ per cycle needs
$\Delta F = 0.01$. Measuring $F$ only at cycle 8 and dividing by 8 hides the
possibility that cycles 1-3 were conservative and cycles 6-8 collapsed the base.
Always fit the log-linear regression and look at the residual.

### `F_hom`, homozygosity-based

Wright defined an inbreeding coefficient to run from 0 to 1 as heterozygosity
under random mating falls from its initial state to zero. So the metric is the
fraction of heterozygosity already lost relative to the base:

$$F_{hom} = 1 - \frac{1}{N_{SNP}}\sum_k \frac{2p_{t,k}(1 - p_{t,k})}{2p_{0,k}(1 - p_{0,k})}
          = 1 - \frac{1}{N_{SNP}}\sum_k \frac{H_{t,k}}{H_{0,k}}$$

**It can be negative.** If management pushes frequencies toward 0.5,
heterozygosity rises above the base and the "inbreeding coefficient" goes below
zero. That is not a numerical error: it is the expected behaviour of any scheme
that maximises heterozygosity on the monitored panel.

It protects against inbreeding depression and the expression of deleterious
recessives in homozygosis. It is the right lens when the dominant risk is
*vigour*.

In maize, heterozygosity is the product itself. Within a heterotic pool under
reciprocal recurrent selection, `F_hom` computed **within pool** anticipates the
decline in line-per-se vigour; it says nothing about between-pool divergence,
which is what generates heterosis in the hybrid. In selfing species, `F_hom`
measured on fixed lines is $\approx 1$ by construction and is **useless for
management** -- it has to be computed on allele frequencies in the crossing
population, not on the homozygosity of individuals.

### `F_drift`, drift-based

$$F_{drift} = \frac{1}{N_{SNP}}\sum_k \frac{(p_{t,k} - p_{0,k})^2}{p_{0,k}(1 - p_{0,k})}$$

**Never negative.** It is formally analogous to Holsinger & Weir's $F_{ST}$
applied to *one population through time* rather than to a set of populations in
space. It is exactly the quantity constrained when `G_VR2` goes into the
optimiser.

It protects against random change in phenotypic value for traits **outside** the
selection objective, against rare deleterious recessives rising to dangerous
frequencies, and against irreversible loss of rare alleles.

There is also a selection-intensity reading: the term
$\delta p^2 / [p_0(1-p_0)]$ approximates the **squared total intensity** applied
at that marker, with $i \approx \delta p / \sqrt{p_0(1-p_0)}$. Constraining
`F_drift` constrains how much effective selection the whole genome underwent --
including at loci you never meant to select.

A wheat programme running six cycles of genomic selection with a VanRaden
$\mathbf{G}$ and reporting $\Delta F \approx 1\%$ has reported
$\Delta F_{drift}$ on the training panel. Quantitative disease-resistance loci
not yet prioritised may have hitch-hiked to extreme frequencies alongside yield
QTL, and none of that appears in the reported metric.

## The classical equivalence, and the covariance that destroys it

Under purely random drift, with $E[\delta p_t \mid p_0] = 0$ for every $p_0$,

$$\frac{\mathrm{var}(p_{t,k})}{p_0(1-p_0)} = 1 - \frac{H_t}{H_0}
  \;\;\Longrightarrow\;\; F_{drift} = F_{hom}.$$

The paper's central result quantifies the departure:

$$F_{hom} - F_{drift} \;=\; 2\,\mathrm{cov}\!\left(
  \frac{\delta p_{t,k}}{\sqrt{p_{0,k}(1-p_{0,k})}},\;
  \frac{p_{0,k} - 1/2}{\sqrt{p_{0,k}(1-p_{0,k})}}\right)$$

In words: **if the direction of frequency change is correlated with the starting
frequency, the two measures diverge.** The non-obvious part is that this
covariance is induced by the *diversity management itself*, not by directional
selection -- the paper demonstrates this by running the scheme with random
GEBVs, and the divergence persists.

### Verifying the identity

`freq_metrics()` returns all three quantities in one marker sweep. The identity
is asserted in the test suite; here it is on live data:

```{r}
set.seed(31)
idx <- sample.int(ctx$N, 60)
fm <- freq_metrics(idx, ctx)
fmt(fm[c("F_hom", "F_drift", "cov_diag")], 6)
```

```{r}
c(F_hom_minus_F_drift = unname(fm[["F_hom"]] - fm[["F_drift"]]),
  cov_diag            = unname(fm[["cov_diag"]]),
  residual            = unname(fm[["cov_diag"]] - (fm[["F_hom"]] - fm[["F_drift"]])))
```

Exact to machine precision. And when the selection *is* the base, both lenses
must read zero:

```{r}
fmt(freq_metrics(seq_len(ctx$N), ctx)[c("F_hom", "F_drift", "cov_diag")], 12)
```

### Why each matrix generates a sign

```{r}
#| echo: false
knitr::kable(data.frame(
  Matrix = c("G_VR2", "G_0.5", "G_VR1", "G_i(p)"),
  Covariance = c("positive", "negative", "positive, attenuated", "negative"),
  Mechanism = c(
    "the constraint penalises delta-p^2, so moving an allele to the *far* extreme costs more than to the near one; the optimiser pushes rare alleles to 0 and common ones to 1",
    "frequencies are measured as deviations from 0.5, so the optimiser buys 'diversity' by pushing everything to intermediate frequencies",
    "G_VR2 reweighted by 2p0(1-p0), giving more weight to already-intermediate loci and less room at the extremes",
    "uses accumulated intensity instead of delta-p^2; since di/dp = [p(1-p)]^-1/2, moves toward the extremes get more expensive"),
  check.names = FALSE),
  caption = "The sign of the bias is a property of the matrix, not of the data.")
```

::: {.callout-warning}
## Your MAF spectrum decides how bad this is

The magnitude depends on the minor-allele-frequency spectrum. The effect is
dramatic with sequence data dominated by rare alleles, and much milder with SNP
chips designed for intermediate MAF.

**Directly translated to plants:** GBS or skim-seq with the full MAF spectrum
puts you in the severe scenario; a 20-50k chip filtered to MAF > 0.05 puts you in
the mild one. This is the parameter that decides whether you can live with
`G_VR2`.
:::

### The signature, on live data

The pure max-diversity greedy is the `G_0.5` scheme in miniature: it constrains
molecular coancestry and nothing else. It should drive `F_hom` negative while
paying the largest drift bill.

```{r}
panel <- rbind(
  random      = freq_metrics(sample.int(ctx$N, 60), ctx),
  max_div     = freq_metrics(sel_greedy(ctx, 60, w = 0), ctx),
  balanced    = freq_metrics(sel_greedy(ctx, 60, w = 1), ctx),
  truncation  = freq_metrics(sel_truncation(ctx, 60), ctx)
)
fmt(as.data.frame(panel[, c("He", "F_hom", "F_drift", "cov_diag", "rare_retained")]), 4)
```

Read the `F_hom` and `F_drift` columns together. Maximising diversity on the
homozygosity lens produces a *negative* `F_hom` -- the selection is more
heterozygous than the base -- while `F_drift` is strictly positive and larger
than under random sampling. The bill is real, it is being paid, and a report
carrying only `F_hom` would not show it.

## The other matrices, briefly

**`F_IBD` via `G_LA`.** Instead of assuming a 50/50 transmission probability for
each parental allele (as the pedigree matrix $\mathbf{A}$ does), linkage
analysis over the markers infers which parental haplotype was actually
transmitted along the chromosome. It requires pedigree **and** markers. It
measures IBD relative to a base explicitly defined as unrelated and
non-inbred, so frequency change is determined by the base population rather
than by linkage disequilibrium generated during selection -- and the
problematic covariance disappears. It was the **only scheme tested that met
both the $\Delta F_{hom}$ and $\Delta F_{drift}$ targets**, on the managed panel
and at unmonitored neutral loci, and it converted inbreeding into gain most
efficiently.

For plants this is the point of greatest friction. In biparental populations and
backcross families `G_LA` is natural and cheap -- phase is known and the number
of meioses since the base is small. In recurrent-selection populations with full
pedigree (common in commercial maize) it is feasible. In germplasm panels
*without* pedigree it is inapplicable, and the honest alternative is
segment-based IBD.

**`F_ROH`.** A run of homozygosity is an uninterrupted stretch of homozygous
markers; the logic is that many consecutive markers are unlikely to be IBS by
chance, so the stretch is probably IBD. The fundamental ambiguity is that
`F_ROH` is IBD when runs are **long** (recent inbreeding) and homozygosity when
they are **short** (ancient inbreeding, accumulated IBS). The minimum-length
threshold is therefore a knob that selects the *age* of inbreeding you are
measuring, and there is no well-defined reference base.

Empirically, `G_ROH` behaved more like `G_0.5` than like an IBD measure. It also
is **not guaranteed positive semi-definite** -- its elements are computed
pairwise -- and required substantial diagonal additions for inversion, which
made it diagonally dominant and artificially inflated the number of parents
selected.

In a soybean inbred the entire genome is one ROH: `F_ROH` $\approx 1$ and the
metric saturates. In maize it is useful for diagnosing erosion of elite
heterotic pools; for *management* inside an optimiser, check positive
definiteness carefully.

### Side by side

```{r}
#| echo: false
knitr::kable(data.frame(
  Matrix = c("A", "G_VR2", "G_VR1", "G_0.5", "G_i(p)", "G_LA", "G_ROH"),
  Class = c("IBD (expectation)", "drift", "drift", "homozygosity", "drift",
            "IBD", "hybrid"),
  `Behaviour in the optimiser` = c(
    "high gain, but exceeded targets by ~40% by not accounting for within-family LD",
    "meets the drift target *only on the managed panel*; F_hom runs away",
    "reweighted version; smaller bias, smaller gain",
    "F_hom near zero, drift 4x the target; loses genetic variance early",
    "inverts the sign of the bias; constraint only indirectly tied to F",
    "the only one meeting both targets; best gain per unit of inbreeding",
    "high gain, uncontrolled drift, numerical problems"),
  check.names = FALSE),
  caption = "Seven matrices in the same optimiser (Meuwissen et al. 2020).")
```

::: {.callout-note}
## An implementation note that is routinely ignored

`G_VR1`, `G_VR2`, `G_0.5` and `G_i(p)` are only **semi**-definite positive (a
zero eigenvalue from the centering). You must add $\alpha$ to the diagonal
($\alpha = 0.01$ in the paper) to invert them and for the optimal-contribution
algorithm to work. $\mathbf{A}$ and `G_LA` are positive definite by
construction.
:::

## Reference results

Target $\Delta F = 0.005$ per generation ($N_e = 100$). Values at the
**unmonitored neutral loci**, generation 20.

```{r}
#| echo: false
knitr::kable(data.frame(
  Scheme = c("G_VR2(M,M)", "G_VR2(M,D)", "G_VR1(M,M)", "G_i(p)(M,M)",
             "G_0.5(M,M)", "G_ROH(M,M)", "G_LA(M,M)", "A(M,~)"),
  Class = c("drift", "drift", "drift", "drift", "homozygosity",
            "homozygosity", "IBD", "IBD"),
  dF_hom = c(0.0103, 0.0101, 0.0080, 0.0065, 0.0073, 0.0054, 0.0043, 0.0070),
  dF_drift = c(0.0068, 0.0068, 0.0069, 0.0077, 0.0176, 0.0088, 0.0049, 0.0084),
  gap = c(0.054, 0.050, 0.021, -0.031, -0.170, -0.070, -0.010, -0.021),
  gain = c(7.124, 7.107, 6.680, 7.111, 6.734, 9.099, 7.188, 9.890),
  check.names = FALSE),
  caption = "Both rate targets are 0.005. Only G_LA meets both.")
```

Four readings:

1. **No IBS scheme met both targets.** `G_LA` did, within about 2%.
2. Pedigree $\mathbf{A}$ gave the highest absolute gain but overshot by ~40% --
   neutral loci hitch-hiking with QTL through within-family LD, which the
   pedigree expectation of IBD cannot capture.
3. Using a separate panel for management instead of the GEBV panel changed gain
   by **less than 0.3%**. Not worth the complexity.
4. At equal mean inbreeding, `G_LA` beat $\mathbf{A}$ in 65 of 100 replicates
   and `G_ROH` in 62 of 100.

### The diagnostic nobody runs

Schemes based on IBD selected many parents; VanRaden-style schemes selected few.

> **Rule of thumb.** If your optimal-contribution run is delivering high gain
> with a suspiciously small number of parents and $\Delta F$ on target, compute
> `F_hom` and `F_drift` separately on a panel **not used** in the management.
> The difference between them is your hidden bill.

The book's implementation of that diagnostic is `ne_parents()`, reported
alongside the two lenses in every scenario -- see @sec-case-study, where a
selection meeting its diversity target while collapsing from 40 effective
parents to 12 is visible in one row of the output.

## What to carry into plant breeding

**Declare the base and never redefine it.** Redefining the reference frequencies
every generation is common in plant pipelines because it is the *default* in
several $\mathbf{G}$-construction functions. Since frequency changes in
successive cycles are positively correlated, constraining the variance *within*
each cycle lets the *accumulated* variance overshoot the target. Freeze $p_0$
from the founding cycle and propagate it. In this book that is `ctx$p0`, set
once in `build_ctx()` and never recomputed.

**Do not maximise marker heterozygosity as an objective.** Raising
heterozygosity on the monitored panel does not raise it at unmonitored loci --
and it increases real IBD, because raising an allele's frequency means
multiplying copies of a subset of base alleles.

**Heterotic pools need a two-directional constraint.** Within pool you want to
limit $\Delta F$. Between pools you want to *preserve* divergence, which is
directed drift and is desirable. A single global
$\tfrac{1}{2}\mathbf{c}'\mathbf{G}\mathbf{c}$ cannot express that. Use multiple
matrices with simultaneous constraints, or per-subpopulation constraints -- which
is what `gd_partition()` measures and @sec-stages operationalises.

**Selfing species need inbreeding by selfing separated from inbreeding by
drift.** In soybean, wheat or rice, individual $F$ is $\approx 1$ by design and
carries no management information. The manageable quantity is allele frequency
**in the crossing population** and coancestry among the chosen parents. Apply
the whole `F_hom`/`F_drift` framework at population level, never at line level.

**Region-specific targets rarely earn their complexity.** Causal variants are
well distributed and what happens at a subset of loci does not predict what
happens outside it. The legitimate plant exception is R-gene-type resistance
loci, where allelic diversity has direct and known value.

**Deleterious recessives are a load problem, not a diversity problem.** If you
have mapped the defect, putting it in the selection index is more efficient than
controlling it indirectly through a diversity constraint.

**In pure conservation schemes the problem is *worse*, not better.** Without
directional selection the inbreeding-management term dominates the optimiser and
the covariance tends to be larger. The IBD recommendation is stronger there.

## Limits of the animal-to-plant transposition

```{r}
#| echo: false
knitr::kable(data.frame(
  `Assumption in the paper` = c(
    "discrete generations, two sexes, one contribution per individual",
    "reliable, deep pedigree",
    "base = simulated founder population at mutation-drift equilibrium",
    "target Ne = 100 over 20 generations",
    "one population, one target",
    "heterozygosity is always desirable"),
  `Situation in plants` = c(
    "hermaphroditism, selfing, cloning, overlapping cycles, repeated use of the same parent",
    "often truncated, or absent in genebanks",
    "base is an administrative choice (cycle 0) with pre-existing structure",
    "shorter cycles with GS, but effective elite Ne often below 30",
    "multiple target environments, G x E, separate heterotic pools",
    "in selfers transitory; in hybrids the final product; in lines undesirable"),
  check.names = FALSE))
```

None of this invalidates the central conclusion. It is an **algebraic** property
of the optimiser, not a biological property of the species. What it changes is
*which matrix is practicable*.

## Operational checklist

- Is the base $p_0$ frozen at the founding cycle, and documented?
- Are `F_hom` and `F_drift` computed **separately** and reported together?
- Is there a panel of loci **not used in management**, for audit?
- Is the regression of $\log(1 - F)$ on cycle linear (constant rate)?
- Is the covariance $2\,\mathrm{cov}(\delta p/s,\, (p_0 - \tfrac12)/s)$ reported
  as a diagnostic?
- Is the number of selected parents biologically plausible?
- Is true genetic variance (or the best estimate of it) being tracked?
- Is $\mathbf{G}$ positive definite before inversion, and is the added $\alpha$
  documented?
- Where possible, is an IBD alternative compared in parallel?
