# Heterosis, dominance, and what divergence buys {#sec-heterosis}

```{r}
#| include: false
source("_common.R")
ctx  <- simulate_data(book_cfg)
L    <- simulate_lines(book_cfg)
pool <- rep(c("A", "B"), c(book_cfg$n_pool_A, book_cfg$n_pool_B))
P    <- pool_freqs(L / 2, pool)
pA   <- P["A", ]
pB   <- P["B", ]
nA   <- book_cfg$n_pool_A
U    <- L[seq_len(nA), ] / 2
V    <- L[nA + seq_len(book_cfg$n_pool_B), ] / 2
set.seed(4)
d    <- rgamma(ctx$m, 2, 4)
```

Since @sec-subdivided this book has repeated a claim without proving it: that the
divergence between the heterotic pools is the engine of heterosis, and is therefore
to be maintained rather than minimised. @sec-catalogue will decompose diversity on
that basis and @sec-stages will build a monitoring panel around it. The claim has
carried a lot of weight for an assertion.

This chapter derives it, and then corrects it. The derivation is short and the
identity is exact. The correction is that the usual reading of the identity --
*push the pools apart and you will get more heterosis* -- is false, and the
simulation says so unambiguously.

Then it goes one step further than the mean. Expected heterosis is what a pool
pair is worth on average; what a breeder selects on is how heterosis is
*distributed* across the crosses in the factorial. Those are different questions
with different answers, and the second one is where general and specific combining
ability come from.

The model used throughout is dominance only, which @labroo2021 review as one of
several non-exclusive mechanisms proposed for heterosis. Nothing below depends on
dominance being the whole story; it depends on there being *some* per-locus
advantage to heterozygosity, and $d_k$ is the name for whatever that is.

## The dominance model, minimally

### The one-locus scale

The classical parameterisation puts the origin at the midpoint of the two
homozygotes. One homozygote is then worth $+a$, the other $-a$, and the
heterozygote is worth $d$. Under random mating at frequency $p$, with $q = 1 - p$,
the population mean is

$$M = a\,(p - q) + 2pq\,d$$

The first term is additive and the second is the dominance contribution. Only the
second depends on how many heterozygotes there are, and it is the only one that a
cross between two inbred parents can move.

::: {.callout-note}
## $a$ appears in this section and nowhere else

The symbols in that display are the textbook ones, local to this section. This
book's additive marker effects are $\beta_k$ (@sec-genomic-prediction), and its
$\alpha$ is the diversity-loss statistic of @sec-catalogue, not the classical
average effect of a substitution. Nothing downstream of this section carries an
$a$.
:::

### What $d_k$ is, and what it is not

$d_k$ is *functional* dominance: the heterozygote's departure from the homozygote
midpoint at locus $k$. It is a property of the locus. It does not move when allele
frequencies move.

Fisher's *statistical* dominance deviation is a different object -- the residual
left after projecting genotypic values onto the best additive model **in a
particular population**. It is a property of the population as much as of the
locus, and it changes when frequencies change.

Both appear in this chapter, and keeping them apart is the whole trick.
$d_k$ is the input. The specific combining ability of @sec-heterosis-gca-sca is
the statistical dominance deviation that a given $d_k$ *induces* over a given pair
of pools. One is a constant; the other is an answer that depends on where the
pools happen to sit.

The distinction also buys the model its generality. Dominance, overdominance
(a heterozygote outside both homozygotes, $|d| > |a|$) and pseudo-overdominance
(two dominant favourable alleles in repulsion phase, close enough that no marker
separates them) are three different biological stories, and @labroo2021 review the
evidence for each. At marker resolution they are the same arithmetic: a locus at
which the F1 gains something the mid-parent does not. A single $d_k$ absorbs all
three, and nothing below can distinguish them -- nor needs to.

### The mid-parent difference

Write the genotypic value of a hybrid as an additive part plus a dominance
deviation. The parents are fully inbred, so every parent is homozygous at every
locus, and their additive contributions average exactly to the F1's own additive
value. They cancel from the mid-parent difference. What survives is the dominance
deviation at the loci where the F1 is heterozygous:

$$H_i = \sum_k d_k \, \mathbb{1}\{x_{ik} = 0.5\}$$

The additive effects never appear. That is the result, not an omission -- and it
is why `heterosis_value()` takes no additive-effect argument:

```{r}
heterosis_value
```

Against the literal oracle of @sec-verification, which loops over loci and
compares the two parents' genotypes one at a time:

```{r}
max(abs(heterosis_value(ctx$X[1:25, ], d) -
        ref_heterosis(L, head(ctx$ped, 25), d)))
```

The cancellation is worth seeing directly. Give the loci additive effects as well,
score a real F1 and its two parents, and subtract:

```{r}
a   <- rnorm(ctx$m, 0, 0.2)
val <- function(x) sum(a * (2 * x - 1)) + sum(d * (x == 0.5))
i   <- 7
mid <- (val(L[ctx$ped$a[i], ] / 2) + val(L[ctx$ped$b[i], ] / 2)) / 2
c(mid_parent_difference = val(ctx$X[i, ]) - mid,
  heterosis_value       = heterosis_value(ctx$X[i, , drop = FALSE], d))
```

The additive effects were drawn at random and never entered the second number.

## Expected heterosis over a pool pair

A locus is heterozygous in the F1 when the two gametes carry different alleles,
which for a cross drawn at random between the pools happens with probability
$p_A(1 - p_B) + (1 - p_A)p_B$. So the expected heterosis over all $A \times B$
crosses follows from the pool frequencies alone:

$$\mathbb{E}[H] = \sum_k d_k \left[\, p_{A k} + p_{B k} - 2 p_{A k} p_{B k} \right]$$

The candidate pool here *is* the full factorial, so this is exact rather than
approximate, and the residual is at machine precision:

```{r}
abs(heterosis_expected(pA, pB, d) - mean(heterosis_value(ctx$X, d)))
```

That is a mean over $N =$ `r ctx$N` crosses, and a mean is not a ranking. Hold on
to the question of how the individual crosses are spread around it;
@sec-heterosis-gca-sca answers it exactly.

## The identity {#sec-heterosis-identity}

Now split that bracket. Writing $\bar p = (p_A + p_B)/2$ and $y = p_A - p_B$, the
per-locus F1 heterozygosity separates exactly:

$$p_A + p_B - 2 p_A p_B \;=\; \underbrace{2 \bar p (1 - \bar p)}_{\text{shared}}
                            \;+\; \underbrace{y^2 / 2}_{\text{divergence}}$$

The first term is the heterozygosity a single merged pool at frequency $\bar p$
would have produced. The second is what keeping the pools apart adds on top of it.
Averaged over loci, with $d_k \equiv 1$ so that the weighting does not obscure the
algebra, those two terms are exactly the $GD_T$ and $GD_{BS}$ of the Caballero &
Toro partition that @sec-subdivided derived and `gd_partition()` computes:

```{r}
hp <- heterosis_partition(pA, pB, rep(1, ctx$m)) / ctx$m
gp <- gd_partition(seq_len(ctx$N), ctx)
fmt(data.frame(term       = c("shared", "divergence"),
               heterosis  = c(hp[["shared"]], hp[["divergence"]]),
               partition  = c(gp[["GD_T"]],   gp[["GD_BS"]]),
               residual   = c(abs(hp[["shared"]]     - gp[["GD_T"]]),
                              abs(hp[["divergence"]] - gp[["GD_BS"]]))), 12)
```

> $GD_{BS}$ weighted by dominance deviations **is** the divergence half of expected
> heterosis. "Between-pool divergence is the heterosis engine" is an identity, not
> a slogan.

### A third form: one minus the between-pool coancestry

The sum of those two components has a name of its own. $\theta_{AB} = \mathbf{w}_A'
\mathbf{f}^L \mathbf{w}_B$ is the between-pool coancestry of @sec-catalogue -- the
probability that an allele drawn from pool A and an allele drawn from pool B are
identical in state. Its complement is the whole bracket:

$$GD_T + GD_{BS} \;=\; 1 - \theta_{AB}$$

```{r}
tp <- theta_pools(seq_len(ctx$N), ctx)
c(shared_plus_divergence = gp[["GD_T"]] + gp[["GD_BS"]],
  one_minus_theta_AB     = 1 - tp[["theta_AB"]])
```

Which is the shortest statement the chapter has to make: **expected F1
heterozygosity is one minus between-pool coancestry.** Everything else here is a
weighting of that quantity, or a way of splitting it.

## A second identity, at the hybrid level

@sec-catalogue reports `GD_WI_hyb`, the heterozygosity actually carried inside the
F1s, separately from the line-level partition, because with inbred parents
`GD_WI` is zero by construction and carries no signal. The two are linked:

$$2 \cdot GD_{WI\text{-}hyb} = GD_T + GD_{BS}$$

```{r}
c(lhs = 2 * gp[["GD_WI_hyb"]], rhs = gp[["GD_T"]] + gp[["GD_BS"]],
  GD_WI = gp[["GD_WI"]])
```

Exact on the full factorial, because every $A \times B$ pairing is present exactly
once. On a *selected* subset it holds only approximately -- the gap is the pairing
structure the selection imposed, and `tests/testthat/test-dominance.R` pins it at
both scales.

## Inbreeding depression, and why it does not price $\alpha$ {#sec-heterosis-id}

The same model run backwards gives inbreeding depression. A locus at frequency
$p$ in a population with inbreeding coefficient $F$ is heterozygous with
probability $2pq(1-F)$, so the mean declines linearly in $F$:

$$ID \;=\; 2F\sum_k d_k \, p_k q_k$$

At $F = 1$ that is exactly the *shared* term of `heterosis_partition()`. So the
depression of a pool driven to complete homozygosity and the expected heterosis
of a cross made *within* that same pool are one quantity seen from two sides --
Falconer's symmetry, and it needs no new machinery here:

```{r}
c(inbreeding_depression = inbreeding_depression(pA, d),
  shared_term           = heterosis_partition(pA, pA, d)[["shared"]],
  deviation             = abs(inbreeding_depression(pA, d) -
                                heterosis_partition(pA, pA, d)[["shared"]]))
```

It is checked against simulated genotypes rather than against its own algebra,
which is what `ref_inbreeding_depression()` exists for -- Monte Carlo, so the
agreement is to sampling error:

```{r}
k <- 1:300
c(closed_form = inbreeding_depression(pA[k], d[k], 0.6),
  simulated   = ref_inbreeding_depression(pA[k], d[k], 0.6, n = 20000, seed = 2))
```

Now the part that matters, because the obvious next step is wrong.

It is very tempting to say: $\alpha$ budgets a loss of within-pool gene
diversity, gene diversity is $\sum p q$, inbreeding depression is $\sum d\,pq$,
therefore spending $\alpha$ costs yield in the current cycle and *that* is what
prices the budget. The algebra does not support it.

```{r}
c(max_dominance_contribution_of_an_inbred_line = max(abs(heterosis_value(U, d))))
```

Zero, exactly. A fully inbred line has no heterozygous loci, so its dominance
contribution is zero **whatever the pool's allele frequencies are**. In a
programme whose parents are inbred lines the depression is a constant, paid in
full at the moment the lines were made, and it does not move when the pool
subsequently loses diversity.

::: {.callout-important}
## Inbreeding depression is not the current-cycle price of the diversity budget

This is the same fact that makes $GD_{WI}$ exactly zero in @sec-catalogue's
partition, and it has the same cause: with inbred parents there is no
within-individual variation left to lose. Two consequences.

The cost of $\alpha$ in this system remains what @sec-cycles measures -- an
option value, realised cycles later through the variance available to select on
(@sec-response), not a yield loss booked this season.

And inbreeding depression **is** the live quantity wherever the population is
only partially inbred: F2, F3 and F4 in @sec-sizing, where $F$ is still climbing
and the depression is still accumulating. That is the chapter it belongs to, not
this one.
:::

Note also what this does *not* say. Expected heterosis in the product is a
separate question, and @sec-heterosis-flat answers it: under neutral drift the
shared term lost and the divergence term gained cancel exactly, so hybrid
heterosis is invariant. Neither the parent side nor the product side turns
diversity loss into an immediate dominance cost. The bill arrives later, and it
arrives in the additive variance.

## When genetic distance predicts heterosis exactly {#sec-heterosis-distance}

Before the corrections start, one special case is worth having, because it
explains a whole literature.

Set $d_k \equiv d$, constant across loci. Then $H$ stops being a weighted count and
becomes a plain one: $d$ times the number of loci at which the two parents differ.
For inbred lines that count is exactly what the coancestry kernel of @sec-stages
measures, so for a cross between lines $a$ and $b$

$$H_{ab} \;=\; d\,m\,(1 - f^L_{ab}) \;=\; d\,m\,\mathrm{MRD}_{ab}^2$$

```{r}
h1 <- heterosis_value(ctx$X, rep(1, ctx$m))
fL <- molecular_coancestry(L / 2)
ab <- cbind(ctx$ped$a, ctx$ped$b)
c(vs_coancestry = max(abs(h1 - ctx$m * (1 - fL[ab]))),
  vs_mrd        = max(abs(h1 - ctx$m * mrd_matrix(L / 2)[ab]^2)))
```

Both residuals are at machine precision. Under flat dominance, heterosis and
molecular coancestry are one statistic seen twice -- which is Trap 1 of
@sec-stages read from the other side, and the reason the diversity kernel this
whole book runs on turns out to be the right object for the heterosis question
as well.

That is also the exact condition under which marker distance predicts heterosis.
It is a strong condition, and it fails as soon as $d$ varies. With the
gamma-distributed deviations used everywhere else in this chapter the relationship
survives but stops being an identity:

```{r}
c(cor_uniform_d = cor(h1, 1 - fL[ab]),
  cor_real_d    = cor(heterosis_value(ctx$X, d), 1 - fL[ab]))
```

A correlation of `r round(cor(heterosis_value(ctx$X, d), 1 - fL[ab]), 2)` is a
useful predictor and a poor identity. The empirical record on distance-versus-heterosis
is mixed for exactly this reason: distance is a sufficient statistic for heterosis
precisely when dominance is flat across loci, and @sec-heterosis-alignment is about
what happens when it is not.

## Divergence is not a bonus {#sec-heterosis-flat}

Read `divergence` as free money and you will reach for $F_{ST}$. Do not.

Sweep the pools apart and watch all three terms. Nothing here is selected; the
pools simply drift further from their common ancestor as `fst` rises:

```{r}
sweep_het <- function(fst) {
  cf <- modifyList(book_cfg, list(fst = fst, seed = 5))
  Lf <- simulate_lines(cf)
  pl <- rep(c("A", "B"), c(cf$n_pool_A, cf$n_pool_B))
  Pf <- pool_freqs(Lf / 2, pl)
  hp <- heterosis_partition(Pf["A", ], Pf["B", ], rep(1, cf$m)) / cf$m
  cm <- pool_complementarity(Lf / 2, pl)
  data.frame(fst = fst, shared = hp[["shared"]], divergence = hp[["divergence"]],
             total = hp[["total"]],
             frac_opposite_fixed = cm[["frac_opposite_fixed"]],
             mean_abs_delta_p = cm[["mean_abs_delta_p"]],
             mean_sq_delta_p = cm[["mean_sq_delta_p"]])
}
sw <- do.call(rbind, lapply(c(0.02, 0.05, 0.10, 0.20, 0.30, 0.45), sweep_het))
fmt(sw, 4)
```

The divergence term rises by a factor of `r round(max(sw$divergence)/min(sw$divergence), 1)`.
The shared term falls by almost exactly as much. The total does not move.

### The two terms, term by term

The algebra is one line. Pool A and pool B drift *independently* away from a
common ancestral frequency $p$, so $\mathbb{E}[p_A] = \mathbb{E}[p_B] = p$ and
$\mathbb{E}[p_A p_B] = p^2$. Therefore

$$\mathbb{E}\!\left[p_A + p_B - 2 p_A p_B\right] = 2p(1-p),$$

which contains no $F_{ST}$ at all. With ancestral frequencies drawn uniformly on
$[0.05, 0.95]$ the predicted level is $\mathbb{E}[2p(1-p)] =$ `r round(2*(0.5 - (0.9^2/12 + 0.25)), 4)`,
against a simulated mean of `r round(mean(sw$total), 4)` across the whole sweep.

That the *sum* is flat is the headline, but the Balding-Nichols model of
@sec-notation predicts each half separately, which is a sharper statement. Under
it $\mathrm{Var}(p_A) = \mathrm{Var}(p_B) = F_{ST}\,p(1-p)$, so
$\mathrm{Var}(\bar p) = F_{ST}\,p(1-p)/2$ and $\mathrm{Var}(y) = 2F_{ST}\,p(1-p)$.
Substituting:

$$\mathbb{E}\!\left[2\bar p(1 - \bar p)\right] = 2p(1-p)\left(1 - \tfrac{F_{ST}}{2}\right),
\qquad
\mathbb{E}\!\left[y^2/2\right] = F_{ST}\,p(1-p)$$

The shared term loses $F_{ST}\,p(1-p)$ and the divergence term gains
$F_{ST}\,p(1-p)$. Not approximately, and not on average over some range: the same
expression, with opposite signs. Both predictions are parameter-free once the
ancestral distribution is fixed, so they can be laid straight on top of the sweep:

```{r}
Ep <- 2 * (0.5 - (0.9^2 / 12 + 0.25))      # E[2p(1-p)] for p ~ U(0.05, 0.95)
sw$pred_shared     <- Ep * (1 - sw$fst / 2)
sw$pred_divergence <- Ep * sw$fst / 2
fmt(sw[, c("fst", "shared", "pred_shared", "divergence", "pred_divergence")], 4)
```

```{r}
#| fig-cap: "Expected heterosis and its two components as the pools are driven apart by drift alone. Dashed lines are the Balding-Nichols predictions for each component separately, with no fitted parameter. The divergence term grows eightfold; the shared term pays for all of it, unit for unit; the total is flat. Nothing is selected here -- this is what neutral divergence does."
plot(sw$fst, sw$total, type = "b", pch = 19, col = "#1b7837", ylim = c(0, 0.40),
     xlab = "target fst between pools", ylab = "expected heterosis per locus (d = 1)")
lines(sw$fst, sw$shared, type = "b", pch = 17, col = "#2166ac")
lines(sw$fst, sw$divergence, type = "b", pch = 15, col = "#b2182b")
lines(sw$fst, sw$pred_shared, lty = 2, col = "#2166ac")
lines(sw$fst, sw$pred_divergence, lty = 2, col = "#b2182b")
abline(h = mean(sw$total), lty = 2, col = "grey45")
legend("right", c("total", "shared", "divergence", "predicted"), bty = "n",
       cex = 0.8, pch = c(19, 17, 15, NA), lty = c(NA, NA, NA, 2),
       col = c("#1b7837", "#2166ac", "#b2182b", "grey45"))
grid()
```

The one visible gap is at the low end, where the measured divergence sits above
its prediction. That is not drift, it is estimation: $y$ is computed from
`r book_cfg$n_pool_A` and `r book_cfg$n_pool_B` sampled lines, so the measured
$y^2$ carries a sampling term of $p(1-p)(1/n_A + 1/n_B)$ on top of the real
divergence. At `fst = 0.02` that term is larger than the signal. A pool pair's
*apparent* $F_{ST}$ always overstates its true divergence, and by more the smaller
the pools.

:::: {.callout-important}
## Every unit of divergence is paid for out of the shared term

Under neutral divergence the trade is exactly one for one. Drifting the pools
apart does not create heterosis; it relocates it from the shared term to the
divergence term. A programme that maximises $F_{ST}$ has bought nothing and spent
its within-pool diversity to do it.

This is the mechanism behind @sec-stages Trap 3, which observed the symptom --
"$F_{ST}$ keeps rising at loci that are already divergent, which buys no
additional heterosis" -- and prescribed the right treatment without being able to
say why.
::::

## What actually buys heterosis {#sec-heterosis-alignment}

If divergence per se is worth nothing, why do heterotic groups work?

Because real pools are not diverged at random with respect to $d$. The divergence
term is $\sum_k d_k y_k^2 / 2$, a *weighted* sum. It grows when the loci that
diverge are the loci that carry dominance -- and under drift the two are
independent, which is exactly why the total was flat.

### Alignment is a covariance

That last sentence can be made into a number rather than left as a word. A
weighted sum is a mean times a mean plus a covariance, so

$$\sum_k d_k\,y_k^2 / 2 \;=\; \frac{m}{2}\Big[\, \bar d \; \overline{y^2}
  \;+\; \mathrm{cov}(d,\, y^2) \,\Big]$$

```{r}
y2 <- (pA - pB)^2
c(divergence = sum(d * y2 / 2),
  from_parts = ctx$m / 2 * (mean(d) * mean(y2) + cov(d, y2) * (ctx$m - 1) / ctx$m))
```

The first bracketed term is what $F_{ST}$ can see: average divergence times
average dominance. The second is what it cannot. **Alignment is
$\mathrm{cov}(d, y^2)$**, and no statistic computed from allele frequencies alone
can contain it, because $d$ does not appear in one.

### The scenarios

Hold the pools fixed, hold $F_{ST}$ fixed, hold the multiset of dominance
deviations fixed, and change only which locus gets which $d_k$:

```{r}
mk <- function(dec) { v <- numeric(ctx$m); v[order(y2)] <- sort(d, decreasing = dec); v }
scen <- list(`anti-aligned` = mk(TRUE), `independent` = d, `aligned` = mk(FALSE))
al <- do.call(rbind, lapply(names(scen), function(n) {
  h <- heterosis_partition(pA, pB, scen[[n]])
  data.frame(scenario = n, cor_d_y2 = cor(scen[[n]], y2),
             shared = h[["shared"]], divergence = h[["divergence"]], total = h[["total"]])
}))
al$relative <- al$total / al$total[al$scenario == "independent"]
fmt(al, 3)
```

```{r}
#| fig-cap: "The same two pools, the same F_ST, and the same set of dominance deviations in all three bars -- only their placement across loci differs. Expected heterosis moves by half again between the extremes. F_ST cannot see any of this, because d does not enter it."
bp <- barplot(rbind(al$shared, al$divergence), beside = FALSE,
              names.arg = al$scenario, col = c("#2166ac", "#b2182b"),
              ylab = "expected heterosis", border = NA)
legend("topleft", c("shared", "divergence"), fill = c("#2166ac", "#b2182b"),
       bty = "n", cex = 0.8, border = NA)
text(bp, al$total, sprintf("%.2fx", al$relative), pos = 3, cex = 0.8, xpd = NA)
grid(nx = NA, ny = NULL)
```

$F_{ST}$ is identical in all three columns -- it is a function of the frequencies
only, and $d$ never enters it. Expected heterosis ranges over a factor of
`r round(max(al$total)/min(al$total), 2)`.

### Isolating the divergence channel

One honest qualification to that demonstration. Sorting $d$ against $y^2$ moves
the *shared* term as well, because $y^2$ is not independent of
$2\bar p(1 - \bar p)$: a locus with a large frequency difference tends to sit at
an intermediate mean frequency. Some of the effect above is therefore arriving
through the shared half, not the divergence half.

The channel can be separated. Permute $d$ only *within* groups of loci that share
a value of the shared weight. The shared term is then invariant by construction --
exactly, not approximately -- and whatever moves is the divergence channel alone:

```{r}
hz  <- 2 * ((pA + pB) / 2) * (1 - (pA + pB) / 2)
grp <- split(seq_len(ctx$m), round(hz, 9))
strat <- function(dec) {
  v <- numeric(ctx$m)
  for (i in grp) v[i[order(y2[i])]] <- sort(d[i], decreasing = dec)
  v
}
st <- do.call(rbind, lapply(c(aligned = FALSE, `anti-aligned` = TRUE), function(dec)
  as.data.frame(as.list(heterosis_partition(pA, pB, strat(dec))))))
fmt(st, 4)
```

The shared column is identical to the digit, and the divergence term still moves
by a factor of `r round(st$divergence[1] / st$divergence[2], 1)`. The alignment
result survives its own control.

## Reading the S1 panel again

@sec-stages logs `frac_opposite_fixed` and `mean_abs_delta_p` every cycle as
stand-ins for heterosis potential. The sweep above says how good they are.

`mean_sq_delta_p` is not a proxy at all -- it is the divergence term, exactly,
up to the factor of two:

```{r}
c(half_mean_sq = mean(sw$mean_sq_delta_p) / 2,
  divergence   = mean(sw$divergence))
```

`frac_opposite_fixed` is a different matter. At the divergence a real maize pool
pair actually shows it is essentially zero, and it stays near zero across most of
the range where breeding decisions are made:

```{r}
fmt(sw[, c("fst", "divergence", "frac_opposite_fixed", "mean_abs_delta_p",
           "mean_sq_delta_p")], 4)
```

At `fst = 0.10` it is exactly zero: not one marker in `r book_cfg$m` qualifies.
It counts *near-fixed* opposite loci, and at realistic divergence there are almost
none -- the divergence all lives at intermediate frequency differences, which the
threshold discards. It becomes informative only above `fst = 0.30`, well outside
the range @sec-stages takes a dent-versus-flint pair to occupy.

**Log `mean_sq_delta_p`.** It is exact, it is monotone across the whole range, and
it costs one subtraction more than the statistic already being logged.
`pool_complementarity()` and `s1_monitor()` now return it.

## General and specific combining ability {#sec-heterosis-gca-sca}

Everything so far has been about $\mathbb{E}[H]$. Now the spread.

The same algebra that produced the shared/divergence split produces the
combining-ability decomposition, with no new machinery and no model fitting. Take
one locus, write the two parents' genotypes as their pool frequency plus a
deviation, $u = p_A + \delta u$ and $v = p_B + \delta v$, and expand the per-locus
heterozygosity $u + v - 2uv$:

$$u + v - 2uv \;=\; \underbrace{p_A + p_B - 2p_Ap_B}_{\text{mean}}
  \;+\; \underbrace{(1 - 2p_B)\,\delta u}_{\text{parent A}}
  \;+\; \underbrace{(1 - 2p_A)\,\delta v}_{\text{parent B}}
  \;-\; \underbrace{2\,\delta u\, \delta v}_{\text{both}}$$

Four terms: one constant, one linear in each parent, one bilinear. Weight by $d_k$
and sum over loci and the classical decomposition falls out,

$$H_{ij} = \mu + g^A_i + g^B_j + s_{ij}$$

with

$$g^A_i = \sum_k d_k (1 - 2p_{Bk})\,\delta u_{ik}, \qquad
  s_{ij} = -2 \sum_k d_k\, \delta u_{ik}\, \delta v_{jk}$$

What is linear in one parent is general combining ability -- it is carried into
every cross that parent enters. What is bilinear depends on *which two* lines met,
and is specific combining ability. `heterosis_gca_sca()` is that expansion and
nothing more:

```{r}
heterosis_gca_sca
```

It reconstructs every hybrid in the factorial exactly, and all three components
are mean zero:

```{r}
cs <- heterosis_gca_sca(U, V, d)
H  <- heterosis_value(ctx$X, d)
ia <- ctx$ped$a
ib <- ctx$ped$b - nA
c(max_residual = max(abs(H - (cs$mu + cs$gca_A[ia] + cs$gca_B[ib] +
                              cs$sca[cbind(ia, ib)]))),
  mu_vs_mean   = abs(cs$mu - mean(H)),
  mean_gca_A   = mean(cs$gca_A), mean_sca = mean(cs$sca))
```

This is not a decomposition that resembles the two-way analysis of variance. It
*is* the two-way analysis of variance -- the $s_{ij}$ matrix is the interaction
residual of the additive fit, to machine precision:

```{r}
max(abs(resid(lm(H ~ factor(ia) + factor(ib))) - cs$sca[cbind(ia, ib)]))
```

@comstock1949 designed reciprocal recurrent selection around the distinction
between these two components. The algebra above says what each one is made of.

### GCA is an average effect in the other pool's background

Look at the weight on $\delta u$: it is $d_k(1 - 2p_{Bk})$, and $1 - 2p_B$ is
$q_B - p_B$. That is precisely the frequency factor in the classical average
effect of a gene substitution, $a + d\,(q - p)$ -- except that the frequencies are
**pool B's**, not pool A's.

This is the whole reason combining ability is not a property of a line. A pool-A
line's alleles are always substituted into a pool-B background, so its general
combining ability is a dominance effect *evaluated at the tester pool's
frequencies*. Change the tester pool and every number changes. Two consequences
follow immediately:

- A locus where the opposite pool sits at $p_B = 0.5$ contributes **exactly zero**
  GCA and pure SCA. Maximum uncertainty in the tester makes the locus's
  contribution unpredictable from either parent alone.
- A locus where the opposite pool is fixed contributes its full $d_k$ to GCA and
  no SCA at all. The cross is then predictable from the pool-A parent by itself.

Between those two limits sits every real programme, and @sec-heterosis-relocation
is about which limit divergence pushes it towards.

### SCA is minus twice a weighted parental covariance

The bilinear term is $s_{ij} = -2\sum_k d_k\,\delta u_{ik}\,\delta v_{jk}$: minus
twice a $d$-weighted inner product of the two parents' *centred* genotypes. Two
parents that deviate from their pools in the same direction have a positive
covariance and therefore a **negative** SCA.

That is the formal version of a rule of thumb the literature has carried for
decades. Parents that resemble each other -- after removing what is typical of
each pool -- cross badly, and the resemblance that matters is weighted by $d$. It
is also the same object as @sec-heterosis-distance: a covariance between centred
genotypes is a genetic distance in disguise, which is why distance predicts SCA
about as well as it predicts heterosis, and fails in the same circumstances.

### The variance components

Because the three components are orthogonal, the variance of heterosis over the
factorial splits with them, and under linkage equilibrium each piece has a closed
form in the pool frequencies:

$$\sigma^2_{GCA_A} = \sum_k d_k^2\,(1 - 2p_{Bk})^2\,p_{Ak}q_{Ak},
\qquad
\sigma^2_{SCA} = 4 \sum_k d_k^2\,p_{Ak}q_{Ak}\,p_{Bk}q_{Bk}$$

```{r}
qa <- pA * (1 - pA)
qb <- pB * (1 - pB)
fmt(data.frame(
  component = c("GCA_A", "GCA_B", "SCA"),
  observed  = c(var(cs$gca_A), var(cs$gca_B), var(as.vector(cs$sca))),
  closed_form = c(sum(d^2 * (1 - 2 * pB)^2 * qa),
                  sum(d^2 * (1 - 2 * pA)^2 * qb),
                  4 * sum(d^2 * qa * qb))), 2)
```

These agree to the sampling error of a variance computed on
`r book_cfg$n_pool_A` lines, which is around
`r round(100 * sqrt(2 / book_cfg$n_pool_A))`%. The closed forms are the exact
statements; the observed columns are one realisation of them.

Note what each one requires. $\sigma^2_{SCA}$ carries $p_Aq_A\,p_Bq_B$: it needs
**both** pools to be polymorphic. Fix either pool at a locus and that locus
contributes no specific combining ability at all, whatever its $d_k$. Within-pool
diversity is not merely something the diversity constraint defends -- it is the
raw material of SCA.

## Divergence relocates dominance from SCA into GCA {#sec-heterosis-relocation}

Put the two closed forms next to the drift argument of @sec-heterosis-flat and a
second result appears, one the mean could not have shown.

Under Balding-Nichols, $\mathbb{E}[p_Aq_A] = p(1-p)(1 - F_{ST})$ and
$\mathbb{E}[(1 - 2p_B)^2] = (1-2p)^2 + 4F_{ST}\,p(1-p)$. So as the pools diverge

$$\sigma^2_{SCA} \propto (1 - F_{ST})^2,
\qquad
\sigma^2_{GCA} \text{ picks up a } +\,4F_{ST}\,p(1-p) \text{ term}$$

SCA variance falls quadratically; GCA variance rises. Expected heterosis, we
already know, does not move at all. Sweeping the same pools apart and evaluating
the closed forms:

```{r}
reloc <- function(fst) {
  cf <- modifyList(book_cfg, list(fst = fst, seed = 5))
  Pf <- pool_freqs(simulate_lines(cf) / 2,
                   rep(c("A", "B"), c(cf$n_pool_A, cf$n_pool_B)))
  a <- Pf["A", ]; b <- Pf["B", ]
  va <- sum(d^2 * (1 - 2 * b)^2 * a * (1 - a))
  vb <- sum(d^2 * (1 - 2 * a)^2 * b * (1 - b))
  vs <- 4 * sum(d^2 * a * (1 - a) * b * (1 - b))
  data.frame(fst = fst, var_GCA = va + vb, var_SCA = vs,
             frac_SCA = vs / (va + vb + vs))
}
rl <- do.call(rbind, lapply(c(0.02, 0.05, 0.10, 0.15, 0.20, 0.30, 0.45), reloc))
fmt(rl, 3)
```

```{r}
#| fig-cap: "The same neutral divergence sweep, read through the combining-ability decomposition rather than through the mean. Expected heterosis is flat across this entire range (previous figure); its decomposition is not. Specific combining ability falls by two thirds while general combining ability rises, so the share of hybrid variation that is specific drops from about 0.60 to 0.26. Divergence does not create heterosis -- it moves heterosis into the half of the model an additive predictor can express."
plot(rl$fst, rl$var_GCA, type = "b", pch = 19, col = "#1b7837",
     ylim = range(0, rl$var_GCA, rl$var_SCA),
     xlab = "target fst between pools",
     ylab = "variance of heterosis over the factorial")
lines(rl$fst, rl$var_SCA, type = "b", pch = 15, col = "#b2182b")
legend("right", c("GCA (both pools)", "SCA"), bty = "n", cex = 0.8,
       pch = c(19, 15), col = c("#1b7837", "#b2182b"))
grid()
```

The share of variation that is specific falls from
`r round(rl$frac_SCA[1], 3)` to `r round(rl$frac_SCA[nrow(rl)], 3)` across the
sweep, with the mean fixed throughout.

> Divergence does not buy heterosis. It buys **predictability** of heterosis. The
> dominance that was expressed as SCA at low divergence is expressed as GCA at
> high divergence, and GCA is the half that an additive model can express.

This is the answer to a question @sec-genomic-prediction had to leave open. That
chapter fits one additive effect per marker and notes that such a model cannot
represent SCA at all. The relocation result says how much that costs, and when:
against poorly separated pools an additive model is missing most of the
variation; against well separated ones it is missing a minority of it. Well
defined heterotic groups are what make hybrid performance predictable from
parents, and @technow2014 report exactly the regime -- established, clearly
separated dent and flint pools -- in which additive hybrid prediction works well.

:::: {.callout-warning}
## Relocation is not creation, and this is still not a licence to maximise $F_{ST}$

Three things stay true at every point on that sweep. Expected heterosis does not
change. The within-pool diversity spent to buy the divergence is spent for good.
And $\sigma^2_{SCA} \to 0$ is the limit in which no cross is better than its
parents predict -- there is nothing left to *find* by testing combinations, which
is the activity the whole factorial exists to support.

So this is a second, independent reason for @sec-stages Trap 3's prescription to
hold $F_{ST}$ in a band rather than at a maximum. @sec-heterosis-flat gave the
diversity-side reason. This is the prediction-side one, and it points the same
way.
::::

### What this says about ranking crosses

@sec-stages S2 shows that for crosses between fixed inbred lines the progeny
variance collapses to zero, so the usefulness criterion loses its variance term
and cross ranking reduces to ranking means. That is true, and it is why this book
ranks hybrids on an index rather than on $\mu + i h \sigma$.

It does not follow that the ranking is easy. The *mean* of an F1 contains an SCA
term that an additive $\mathbf{G}$ omits entirely, and @sec-heterosis-relocation
just put a size on it. The progeny variance collapses; the dominance deviation
does not.

## What this does not change

:::: {.callout-warning}
## The diversity constraint stays additive

Nothing in this chapter reopens the optimisation. $\theta$, $GD$ and $\alpha$ are
computed on molecular coancestry and are functions of the frequencies alone; no
dominance term enters them, and @sec-ocs, @sec-combinatorial and @sec-advanced are
untouched. The combining-ability split of @sec-heterosis-gca-sca sits outside the
constraint for the same reason: it is a statement about the objective, not about
the feasible set.

A dominance model changes the **mean-performance ranking**, not the diversity
term. @sec-case-study states the same separation from the other side. Do not
conflate them, and do not read this chapter as licence to put a heterosis term
inside the constraint.
::::

What it does change is the interpretation of the constraint's output. Two plans
with the same $\alpha$ and the same $GD_{BS}$ can differ in expected heterosis by
half again, if one of them concentrated its divergence on the dominance loci. The
partition cannot distinguish them. Only an estimate of $d$ can.

### What it would take to use $d$

@sec-catalogue leaves a weight $\lambda$ undetermined between within- and
between-pool diversity and observes that the ratio of the two heterosis terms is
what $\lambda$ is trying to express. This chapter can now say precisely what would
have to be estimated to set it, and it is not much:

- **$\lambda$ needs $\mathrm{cov}(d, y^2)$**, not $d$ itself. A programme does not
  need per-locus dominance deviations to weight the two diversity terms against
  each other; it needs to know whether its divergence is landing on its dominance
  loci. That is one number.
- **Ranking crosses needs more.** $s_{ij}$ is a $d$-weighted covariance between
  two specific parents, so it needs the deviations locus by locus. This is the
  extension @sec-case-study names and the book does not attempt.
- **The data to fit it exists in any programme with testcross records.** The
  decomposition above is the two-way layout; @comstock1949's design produces
  exactly that layout on purpose.

## Where this connects

@sec-subdivided supplied the partition and @sec-catalogue will use it; this chapter
supplies the reason the $GD_{BS}$ component is singled out for protection, and the
reason protecting it is not the same as maximising it. @sec-stages Trap 3 already
recommended holding $F_{ST}$ in a band -- @sec-heterosis-flat is why a band, and
@sec-heterosis-alignment is what to watch instead of the band.
@sec-heterosis-distance closes the loop with the coancestry kernel of
@sec-coancestry: under flat dominance, the diversity matrix and the heterosis
prediction are the same object.

@sec-genomic-prediction is the chapter this one supplies a missing term to. It
predicts breeding values with an additive model on a trait that is additive by
construction, and lists specific combining ability first among what it does not
model. @sec-heterosis-gca-sca is where that term comes from and
@sec-heterosis-relocation is how large it is; both are derived here rather than
fitted, because on a simulated $d$ they can be.

It is also why heterotic groups are described in terms of *patterns* -- which
pool crosses well onto which -- rather than by how far apart the pools have
drifted. @white2020 track exactly that distinction across decades of North
American proprietary dent germplasm.

Forward, the alignment result is the argument for @sec-cycles. Selection on cross
performance is the only process in the programme that searches for the alignment
between divergence and dominance, which means the heterotic pattern is not a
standing asset but something each cycle either builds or erodes. A one-cycle
constraint cannot see that, and neither can $F_{ST}$. @sec-alphasimr runs that
loop on a dominance trait and is the pipeline this chapter's algebra describes.
