# Heterosis, dominance, and what divergence buys {#sec-heterosis}

```{r}
#| include: false
source("_common.R")
ctx  <- simulate_data(book_cfg)
L    <- simulate_lines(book_cfg)
pool <- rep(c("A", "B"), c(book_cfg$n_pool_A, book_cfg$n_pool_B))
P    <- pool_freqs(L / 2, pool)
pA   <- P["A", ]
pB   <- P["B", ]
set.seed(4)
d    <- rgamma(ctx$m, 2, 4)
```

Since @sec-subdivided this book has repeated a claim without proving it: that the
divergence between the heterotic pools is the engine of heterosis, and is therefore
to be maintained rather than minimised. @sec-catalogue will decompose diversity on
that basis and @sec-stages will build a monitoring panel around it. The claim has
carried a lot of weight for an assertion.

This chapter derives it, and then corrects it. The derivation is short and the
identity is exact. The correction is that the usual reading of the identity --
*push the pools apart and you will get more heterosis* -- is false, and the
simulation says so unambiguously.

The model used throughout is dominance only, which @labroo2021 review as one of
several non-exclusive mechanisms proposed for heterosis. Nothing below depends on
dominance being the whole story; it depends on there being *some* per-locus
advantage to heterozygosity, and $d_k$ is the name for whatever that is.

## The dominance model, minimally

Write the genotypic value of a hybrid as an additive part plus a dominance
deviation. The parents are fully inbred, so every parent is homozygous at every
locus, and their additive contributions average exactly to the F1's own additive
value. They cancel from the mid-parent difference. What survives is the dominance
deviation at the loci where the F1 is heterozygous:

$$H_i = \sum_k d_k \, \mathbb{1}\{x_{ik} = 0.5\}$$

The additive effects never appear. That is the result, not an omission -- and it
is why `heterosis_value()` takes no additive-effect argument:

```{r}
heterosis_value
```

Against the literal oracle of @sec-verification, which loops over loci and
compares the two parents' genotypes one at a time:

```{r}
max(abs(heterosis_value(ctx$X[1:25, ], d) -
        ref_heterosis(L, head(ctx$ped, 25), d)))
```

The cancellation is worth seeing directly. Give the loci additive effects as well,
score a real F1 and its two parents, and subtract:

```{r}
a   <- rnorm(ctx$m, 0, 0.2)
val <- function(x) sum(a * (2 * x - 1)) + sum(d * (x == 0.5))
i   <- 7
mid <- (val(L[ctx$ped$a[i], ] / 2) + val(L[ctx$ped$b[i], ] / 2)) / 2
c(mid_parent_difference = val(ctx$X[i, ]) - mid,
  heterosis_value       = heterosis_value(ctx$X[i, , drop = FALSE], d))
```

The additive effects were drawn at random and never entered the second number.

## Expected heterosis over a pool pair

A locus is heterozygous in the F1 when the two gametes carry different alleles,
which for a cross drawn at random between the pools happens with probability
$p_A(1 - p_B) + (1 - p_A)p_B$. So the expected heterosis over all $A \times B$
crosses follows from the pool frequencies alone:

$$\mathbb{E}[H] = \sum_k d_k \left[\, p_{A k} + p_{B k} - 2 p_{A k} p_{B k} \right]$$

The candidate pool here *is* the full factorial, so this is exact rather than
approximate, and the residual is at machine precision:

```{r}
abs(heterosis_expected(pA, pB, d) - mean(heterosis_value(ctx$X, d)))
```

## The identity {#sec-heterosis-identity}

Now split that bracket. Writing $\bar p = (p_A + p_B)/2$ and $y = p_A - p_B$, the
per-locus F1 heterozygosity separates exactly:

$$p_A + p_B - 2 p_A p_B \;=\; \underbrace{2 \bar p (1 - \bar p)}_{\text{shared}}
                            \;+\; \underbrace{y^2 / 2}_{\text{divergence}}$$

The first term is the heterozygosity a single merged pool at frequency $\bar p$
would have produced. The second is what keeping the pools apart adds on top of it.
Averaged over loci, with $d_k \equiv 1$ so that the weighting does not obscure the
algebra, those two terms are exactly the $GD_T$ and $GD_{BS}$ of the Caballero &
Toro partition that @sec-subdivided derived and `gd_partition()` computes:

```{r}
hp <- heterosis_partition(pA, pB, rep(1, ctx$m)) / ctx$m
gp <- gd_partition(seq_len(ctx$N), ctx)
fmt(data.frame(term       = c("shared", "divergence"),
               heterosis  = c(hp[["shared"]], hp[["divergence"]]),
               partition  = c(gp[["GD_T"]],   gp[["GD_BS"]]),
               residual   = c(abs(hp[["shared"]]     - gp[["GD_T"]]),
                              abs(hp[["divergence"]] - gp[["GD_BS"]]))), 12)
```

> $GD_{BS}$ weighted by dominance deviations **is** the divergence half of expected
> heterosis. "Between-pool divergence is the heterosis engine" is an identity, not
> a slogan.

That settles what the book has been asserting. The next two sections are about
what it does *not* license.

## A second identity, at the hybrid level

@sec-catalogue reports `GD_WI_hyb`, the heterozygosity actually carried inside the
F1s, separately from the line-level partition, because with inbred parents
`GD_WI` is zero by construction and carries no signal. The two are linked:

$$2 \cdot GD_{WI\text{-}hyb} = GD_T + GD_{BS}$$

```{r}
c(lhs = 2 * gp[["GD_WI_hyb"]], rhs = gp[["GD_T"]] + gp[["GD_BS"]],
  GD_WI = gp[["GD_WI"]])
```

Exact on the full factorial, because every $A \times B$ pairing is present exactly
once. On a *selected* subset it holds only approximately -- the gap is the pairing
structure the selection imposed, and `tests/testthat/test-dominance.R` pins it at
both scales.

## Divergence is not a bonus {#sec-heterosis-flat}

Read `divergence` as free money and you will reach for $F_{ST}$. Do not.

Sweep the pools apart and watch all three terms. Nothing here is selected; the
pools simply drift further from their common ancestor as `fst` rises:

```{r}
sweep_het <- function(fst) {
  cf <- modifyList(book_cfg, list(fst = fst, seed = 5))
  Lf <- simulate_lines(cf)
  pl <- rep(c("A", "B"), c(cf$n_pool_A, cf$n_pool_B))
  Pf <- pool_freqs(Lf / 2, pl)
  hp <- heterosis_partition(Pf["A", ], Pf["B", ], rep(1, cf$m)) / cf$m
  cm <- pool_complementarity(Lf / 2, pl)
  data.frame(fst = fst, shared = hp[["shared"]], divergence = hp[["divergence"]],
             total = hp[["total"]],
             frac_opposite_fixed = cm[["frac_opposite_fixed"]],
             mean_abs_delta_p = cm[["mean_abs_delta_p"]],
             mean_sq_delta_p = cm[["mean_sq_delta_p"]])
}
sw <- do.call(rbind, lapply(c(0.02, 0.05, 0.10, 0.20, 0.30, 0.45), sweep_het))
fmt(sw, 4)
```

The divergence term rises by a factor of `r round(max(sw$divergence)/min(sw$divergence), 1)`.
The shared term falls by almost exactly as much. The total does not move.

```{r}
#| fig-cap: "Expected heterosis and its two components as the pools are driven apart by drift alone. The divergence term grows eightfold; the shared term pays for all of it; the total is flat. Nothing is selected here -- this is what neutral divergence does."
plot(sw$fst, sw$total, type = "b", pch = 19, col = "#1b7837", ylim = c(0, 0.40),
     xlab = "target fst between pools", ylab = "expected heterosis per locus (d = 1)")
lines(sw$fst, sw$shared, type = "b", pch = 17, col = "#2166ac")
lines(sw$fst, sw$divergence, type = "b", pch = 15, col = "#b2182b")
abline(h = mean(sw$total), lty = 2, col = "grey45")
legend("right", c("total", "shared", "divergence"), bty = "n", cex = 0.8,
       pch = c(19, 17, 15), col = c("#1b7837", "#2166ac", "#b2182b"))
grid()
```

The algebra is one line. Pool A and pool B drift *independently* away from a
common ancestral frequency $p$, so $\mathbb{E}[p_A] = \mathbb{E}[p_B] = p$ and
$\mathbb{E}[p_A p_B] = p^2$. Therefore

$$\mathbb{E}\!\left[p_A + p_B - 2 p_A p_B\right] = 2p(1-p),$$

which contains no $F_{ST}$ at all. With ancestral frequencies drawn uniformly on
$[0.05, 0.95]$ the predicted level is $\mathbb{E}[2p(1-p)] =$ `r round(2*(0.5 - (0.9^2/12 + 0.25)), 4)`,
against a simulated mean of `r round(mean(sw$total), 4)` across the whole sweep.

:::: {.callout-important}
## Every unit of divergence is paid for out of the shared term

Under neutral divergence the trade is exactly one for one. Drifting the pools
apart does not create heterosis; it relocates it from the shared term to the
divergence term. A programme that maximises $F_{ST}$ has bought nothing and spent
its within-pool diversity to do it.

This is the mechanism behind @sec-stages Trap 3, which observed the symptom --
"$F_{ST}$ keeps rising at loci that are already divergent, which buys no
additional heterosis" -- and prescribed the right treatment without being able to
say why.
::::

## What actually buys heterosis {#sec-heterosis-alignment}

If divergence per se is worth nothing, why do heterotic groups work?

Because real pools are not diverged at random with respect to $d$. The divergence
term is $\sum_k d_k y_k^2 / 2$, a *weighted* sum. It grows when the loci that
diverge are the loci that carry dominance -- and under drift the two are
independent, which is exactly why the total was flat.

Hold the pools fixed, hold $F_{ST}$ fixed, hold the multiset of dominance
deviations fixed, and change only which locus gets which $d_k$:

```{r}
y2 <- (pA - pB)^2
mk <- function(dec) { v <- numeric(ctx$m); v[order(y2)] <- sort(d, decreasing = dec); v }
scen <- list(`anti-aligned` = mk(TRUE), `independent` = d, `aligned` = mk(FALSE))
al <- do.call(rbind, lapply(names(scen), function(n) {
  h <- heterosis_partition(pA, pB, scen[[n]])
  data.frame(scenario = n, cor_d_y2 = cor(scen[[n]], y2),
             shared = h[["shared"]], divergence = h[["divergence"]], total = h[["total"]])
}))
al$relative <- al$total / al$total[al$scenario == "independent"]
fmt(al, 3)
```

```{r}
#| fig-cap: "The same two pools, the same F_ST, and the same set of dominance deviations in all three bars -- only their placement across loci differs. Expected heterosis moves by half again between the extremes. F_ST cannot see any of this, because d does not enter it."
bp <- barplot(rbind(al$shared, al$divergence), beside = FALSE,
              names.arg = al$scenario, col = c("#2166ac", "#b2182b"),
              ylab = "expected heterosis", border = NA)
legend("topleft", c("shared", "divergence"), fill = c("#2166ac", "#b2182b"),
       bty = "n", cex = 0.8, border = NA)
text(bp, al$total, sprintf("%.2fx", al$relative), pos = 3, cex = 0.8, xpd = NA)
grid(nx = NA, ny = NULL)
```

$F_{ST}$ is identical in all three columns -- it is a function of the frequencies
only, and $d$ never enters it. Expected heterosis ranges over a factor of
`r round(max(al$total)/min(al$total), 2)`.

> $F_{ST}$ does not predict heterosis. The alignment between divergence and
> dominance does. Divergence at the dominance loci buys everything; divergence
> anywhere else buys nothing.

It is also why heterotic groups are described in terms of *patterns* -- which
pool crosses well onto which -- rather than by how far apart the pools have
drifted. @white2020 track exactly that distinction across decades of North
American proprietary dent germplasm.

This is also why reciprocal recurrent selection builds heterotic groups and
geographic isolation does not. Selection on *cross* performance is a direct search
for that alignment; drift is a random walk with respect to it. @sec-cycles puts
that loop in motion.

## Reading the S1 panel again

@sec-stages logs `frac_opposite_fixed` and `mean_abs_delta_p` every cycle as
stand-ins for heterosis potential. The sweep above says how good they are.

`mean_sq_delta_p` is not a proxy at all -- it is the divergence term, exactly,
up to the factor of two:

```{r}
c(half_mean_sq = mean(sw$mean_sq_delta_p) / 2,
  divergence   = mean(sw$divergence))
```

`frac_opposite_fixed` is a different matter. At the divergence a real maize pool
pair actually shows it is essentially zero, and it stays near zero across most of
the range where breeding decisions are made:

```{r}
fmt(sw[, c("fst", "divergence", "frac_opposite_fixed", "mean_abs_delta_p",
           "mean_sq_delta_p")], 4)
```

At `fst = 0.10` it is exactly zero: not one marker in `r book_cfg$m` qualifies.
It counts *near-fixed* opposite loci, and at realistic divergence there are almost
none -- the divergence all lives at intermediate frequency differences, which the
threshold discards. It becomes informative only above `fst = 0.30`,
which is outside the range @sec-plant-literature reports for dent-versus-flint
pools.

**Log `mean_sq_delta_p`.** It is exact, it is monotone across the whole range, and
it costs one subtraction more than the statistic already being logged.
`pool_complementarity()` and `s1_monitor()` now return it.

## Specific combining ability

The same algebra names GCA and SCA without any new machinery. The additive part of
a line's value is transmitted to every cross it enters -- that is general combining
ability, and it is already inside `ctx$index`. The dominance deviation depends on
*which two* lines met, so it is specific combining ability, and an additive model
cannot express it.

This is what @sec-collapse's result does and does not say. For crosses between
fixed inbred lines the progeny variance collapses to zero, so the usefulness
criterion loses its variance term and cross ranking reduces to ranking means. That
is true. It does not follow that the ranking is easy: the *mean* of an F1 contains
an SCA term that an additive $\mathbf{G}$ omits entirely. The variance collapses;
the dominance deviation does not.

## What this does not change

:::: {.callout-warning}
## The diversity constraint stays additive

Nothing in this chapter reopens the optimisation. $\theta$, $GD$ and $\alpha$ are
computed on molecular coancestry and are functions of the frequencies alone; no
dominance term enters them, and @sec-ocs, @sec-combinatorial and @sec-advanced are
untouched.

A dominance model changes the **mean-performance ranking**, not the diversity
term. @sec-case-study states the same separation from the other side. Do not
conflate them, and do not read this chapter as licence to put a heterosis term
inside the constraint.
::::

What it does change is the interpretation of the constraint's output. Two plans
with the same $\alpha$ and the same $GD_{BS}$ can differ in expected heterosis by
half again, if one of them concentrated its divergence on the dominance loci. The
partition cannot distinguish them. Only an estimate of $d$ can.

## Where this connects

@sec-subdivided supplied the partition and @sec-catalogue will use it; this chapter
supplies the reason the $GD_{BS}$ component is singled out for protection, and the
reason protecting it is not the same as maximising it. @sec-stages Trap 3 already
recommended holding $F_{ST}$ in a band -- @sec-heterosis-flat is why a band, and
@sec-heterosis-alignment is what to watch instead of the band.

Forward, the alignment result is the argument for @sec-cycles. Selection on cross
performance is the only process in the programme that searches for the alignment
between divergence and dominance, which means the heterotic pattern is not a
standing asset but something each cycle either builds or erodes. A one-cycle
constraint cannot see that, and neither can $F_{ST}$.
