# Heterosis, dominance, and what divergence buys {#sec-heterosis}

```{r}
#| include: false
source("_common.R")
ctx  <- simulate_data(book_cfg)
L    <- simulate_lines(book_cfg)
pool <- rep(c("A", "B"), c(book_cfg$n_pool_A, book_cfg$n_pool_B))
P    <- pool_freqs(L / 2, pool)
pA   <- P["A", ]
pB   <- P["B", ]
nA   <- book_cfg$n_pool_A
U    <- L[seq_len(nA), ] / 2
V    <- L[nA + seq_len(book_cfg$n_pool_B), ] / 2
set.seed(4)
d    <- rgamma(ctx$m, 2, 4)
```

Since @sec-subdivided this book has repeated a claim without proving it: that the
divergence between the heterotic pools is the engine of heterosis, and is therefore
to be maintained rather than minimised. @sec-catalogue will decompose diversity on
that basis and @sec-stages will build a monitoring panel around it. The claim has
carried a lot of weight for an assertion.

This chapter derives it, and then corrects it. The derivation is short and the
identity is exact. The correction is that the usual reading of the identity --
*push the pools apart and you will get more heterosis* -- is false: under neutral
divergence expected heterosis does not move at all, and the simulation shows it.

The chapter then goes one step past the mean. Expected heterosis is what a pool
pair is worth on average; what a breeder selects on is how heterosis is
*distributed* across the crosses of the factorial. Those are different questions
with different answers, and the second one is where general and specific combining
ability come from.

Three premises carry everything below. They are stated here once, because every
result in the chapter uses all three and two of them are restrictive.

**Parents are fully inbred.** Every parent line is homozygous at every locus, so
an F1's marker row is the average of its two parents' rows and a locus is
heterozygous in the F1 exactly when the parents differ. This is what makes the
additive effects cancel from the mid-parent difference and what makes `GD_WI`
zero. Nothing here transfers to a programme crossing non-inbred parents without
re-derivation.

**The genotypic model is dominance within loci and additivity between them.**
There is no epistatic term anywhere below: the genotypic value is a sum over
loci, $H_i$ is a sum of per-locus indicators, and the combining-ability expansion
of @sec-heterosis-gca-sca is bilinear only because no term couples two loci.
@labroo2021 review epistasis alongside dominance as a proposed mechanism for
heterosis; this model excludes it, and no result below should be read as evidence
about it. What $d_k$ does absorb is every *within-locus* story, as the next
section sets out.

**The marker panel is the set of dominance loci.** $d_k$ is indexed by the same
$k$ as the markers, so every locus carrying dominance is genotyped and the
frequency statistics are computed on exactly the loci that generate heterosis.
This is the strongest assumption of the three, and it is the one that carries the
distance results of @sec-heterosis-distance. A real panel tags causal loci
through linkage disequilibrium rather than containing them, which weakens every
statement below about markers predicting heterosis but does not change the
algebra, since the algebra is written over whatever set $k$ indexes.

## The dominance model, minimally

### The one-locus scale

The classical parameterisation puts the origin at the midpoint of the two
homozygotes: one homozygote is worth $+a$, the other $-a$, and the heterozygote
$d$. Under random mating at frequency $p$, with $q = 1 - p$, the population mean
is [@falconer1996, Eq. 7.2]

$$M = a\,(p - q) + 2pq\,d$$

The first term is the additive contribution, and the mid-parent of any cross
already carries it. The second is the dominance contribution. It is the only part
of $M$ that depends on how many heterozygotes there are, and therefore the only
part a cross can add over its own mid-parent.

::: {.callout-note}
## $a$ is a scale, not a quantity this chapter computes

The symbols in that display are the textbook ones. This book's additive marker
effects are $\beta_k$ (@sec-genomic-prediction), and its $\alpha$ is the
diversity-loss statistic of @sec-catalogue, not the classical average effect of a
substitution. The per-locus $a_k$ reappears three times below and never as a
result: in the genotypic-value model used to show that it cancels, in the
definition of overdominance as $|d| > |a|$, and in the average effect of a
substitution $a + d(q - p)$. No quantity this chapter reports depends on it.
:::

### What $d_k$ is, and what it is not

$d_k$ is *functional* dominance: the departure of the heterozygote from the
homozygote midpoint at locus $k$. It is a property of the locus. It does not move
when allele frequencies move.

Fisher's *statistical* dominance deviation is a different object: the residual
left after projecting genotypic values onto the best additive model **in a
particular population**. It is a property of the population as much as of the
locus, and it changes when frequencies change.

Both appear in this chapter, and keeping them apart is the whole trick.
$d_k$ is the input. The specific combining ability of @sec-heterosis-gca-sca is
the statistical dominance deviation that a given $d_k$ *induces* over a given pair
of pools. One is a constant; the other is an answer that depends on where the
pools happen to sit.

The distinction also buys the model what generality it has, and marks where that
generality stops. Dominance, overdominance (a heterozygote outside both
homozygotes, $|d| > |a|$) and pseudo-overdominance (two dominant favourable
alleles in repulsion phase, too close for any marker to separate) are three
different biological stories, and @labroo2021 review the evidence for each. At
marker resolution they are the same arithmetic: a locus at which the F1 gains
something the mid-parent does not. A single $d_k$ absorbs all three, and nothing
below can distinguish them -- nor needs to. Epistasis is the fourth story on
@labroo2021's list and $d_k$ does not absorb it, because a single per-locus
number cannot represent a term that depends on two loci at once. That is the
second premise above, and it is an exclusion rather than an approximation.

### The mid-parent difference

Code a genotype by $x_k \in \{0, 0.5, 1\}$, the frequency of the reference allele
within the individual, and take the genotypic value to be the additive part plus
the dominance deviation at every heterozygous locus:

$$G(\mathbf{x}) = \sum_k a_k\,(2x_k - 1) + \sum_k d_k\,\mathbb{1}\{x_k = 0.5\}$$

The parents are fully inbred, so each is homozygous at every locus and the F1's
row is the average of its two parents' rows. The additive part is linear in
$\mathbf{x}$, so the two parents' additive values average exactly to the F1's own
and cancel from the mid-parent difference. What survives is the dominance
deviation at the loci where the F1 is heterozygous:

$$H_i = \sum_k d_k \, \mathbb{1}\{x_{ik} = 0.5\}$$

The additive effects never appear. That is the result, not an omission, and it is
why `heterosis_value()` takes no additive-effect argument:

```{r}
heterosis_value
```

Against the literal oracle of @sec-verification, which loops over loci and
compares the two parents' genotypes one at a time:

```{r}
max(abs(heterosis_value(ctx$X[1:25, ], d) -
        ref_heterosis(L, head(ctx$ped, 25), d)))
```

The cancellation is worth seeing directly. Draw additive effects at random, score
a real F1 and its two parents under $G(\mathbf{x})$, and subtract:

```{r}
a   <- rnorm(ctx$m, 0, 0.2)
val <- function(x) sum(a * (2 * x - 1)) + sum(d * (x == 0.5))
i   <- 7
mid <- (val(L[ctx$ped$a[i], ] / 2) + val(L[ctx$ped$b[i], ] / 2)) / 2
c(mid_parent_difference = val(ctx$X[i, ]) - mid,
  heterosis_value       = heterosis_value(ctx$X[i, , drop = FALSE], d))
```

The additive effects entered the first number and cancelled out of it; they never
entered the second.

## Expected heterosis over a pool pair

Write $p_{Ak}$ and $p_{Bk}$ for the reference-allele frequencies at locus $k$ in
the two pools, each computed over that pool's lines. A locus is heterozygous in
the F1 when the two gametes carry different alleles, which for a cross drawn at
random between the pools happens with probability $p_A(1 - p_B) + (1 - p_A)p_B$.
The expected heterosis over all $A \times B$ crosses therefore follows from the
pool frequencies alone:

$$\mathbb{E}[H] = \sum_k d_k \left[\, p_{A k} + p_{B k} - 2 p_{A k} p_{B k} \right]$$

The candidate pool here *is* the full factorial, every $A \times B$ pairing
exactly once, and the frequencies are the means of those same lines. The closed
form is then the mean of the realised values as an identity rather than an
approximation, and the residual is at machine precision:

```{r}
abs(heterosis_expected(pA, pB, d) - mean(heterosis_value(ctx$X, d)))
```

That is a mean over $N =$ `r ctx$N` crosses, and a mean is not a ranking. Hold on
to the question of how the individual crosses are spread around it;
@sec-heterosis-gca-sca answers it exactly.

## The identity {#sec-heterosis-identity}

Now split the bracket. Writing $\bar p = (p_A + p_B)/2$ and $y = p_A - p_B$, the
per-locus F1 heterozygosity separates exactly:

$$p_A + p_B - 2 p_A p_B \;=\; \underbrace{2 \bar p (1 - \bar p)}_{\text{shared}}
                            \;+\; \underbrace{y^2 / 2}_{\text{divergence}}$$

The first term is the heterozygosity a single merged pool at frequency $\bar p$
would produce under random mating. The second is what keeping the pools apart
adds on top of it. Average both over loci with $d_k \equiv 1$, so that the
weighting does not obscure the algebra, and they are exactly the $GD_T$ and
$GD_{BS}$ of the Caballero & Toro partition that @sec-subdivided derived and
`gd_partition()` computes:

```{r}
hp <- heterosis_partition(pA, pB, rep(1, ctx$m)) / ctx$m
gp <- gd_partition(seq_len(ctx$N), ctx)
fmt(data.frame(term       = c("shared", "divergence"),
               heterosis  = c(hp[["shared"]], hp[["divergence"]]),
               partition  = c(gp[["GD_T"]],   gp[["GD_BS"]]),
               residual   = c(abs(hp[["shared"]]     - gp[["GD_T"]]),
                              abs(hp[["divergence"]] - gp[["GD_BS"]]))), 12)
```

> $GD_{BS}$ weighted by dominance deviations **is** the divergence half of expected
> heterosis. "Between-pool divergence is the heterosis engine" is an identity, not
> a slogan.

One thing to be clear about before reading on. $GD_T$ is the gene diversity of
the *merged* population, and by @sec-subdivided it already contains $GD_{BS}$:
with inbred lines, $GD_T = GD_{BI} + GD_{BS}$. Expected F1 heterozygosity is
therefore $GD_{BI} + 2\,GD_{BS}$, the between-pool component counted twice --
once inside the shared term, once as the divergence term. That factor of two is
where the classical result sits. Falconer & Mackay's heterosis of a cross
between two *random-mating* populations is $\sum_k d_k y_k^2$
[@falconer1996, Ch. 14], exactly twice the divergence term here. The two do not
conflict; they use different mid-parents. Theirs is two random-mating
populations, each carrying its own within-pool heterozygosity, so their $H$ is
the total above $\sum_k d_k (p_{Ak}q_{Ak} + p_{Bk}q_{Bk})$. Ours is two inbred
lines carrying none, so the whole F1 heterozygosity is heterosis, and the shared
term is the mean of the two pools' random-mating heterozygosities plus the half
of $\sum_k d_k y_k^2$ that $GD_T$ already holds.

### A third form: one minus the between-pool coancestry

The sum of the two components has a name of its own. $\theta_{AB} = \mathbf{w}_A'
\mathbf{f}^L \mathbf{w}_B$ is the between-pool coancestry of @sec-catalogue: on
the full factorial, where the line weights are uniform, it is the probability
that an allele drawn from a pool-A line and an allele drawn from a pool-B line
are identical in state. Its complement is the whole bracket:

$$GD_T + GD_{BS} \;=\; 1 - \theta_{AB}$$

```{r}
tp <- theta_pools(seq_len(ctx$N), ctx)
c(shared_plus_divergence = gp[["GD_T"]] + gp[["GD_BS"]],
  one_minus_theta_AB     = 1 - tp[["theta_AB"]])
```

This one is algebra rather than simulation. `gd_partition()` defines
$GD_T = 1 - \bar f$ and $GD_{BS} = \tilde f - \bar f$ with
$\bar f = (\theta_A + 2\theta_{AB} + \theta_B)/4$ and
$\tilde f = (\theta_A + \theta_B)/2$, and the sum collapses to $1 - \theta_{AB}$.
Which is the shortest statement the chapter has to make: **expected F1
heterozygosity is one minus between-pool coancestry.** Everything else here is a
weighting of that quantity, or a way of splitting it.

## A second identity, at the hybrid level

@sec-catalogue reports `GD_WI_hyb`, the within-individual gene diversity actually
carried by the F1s, separately from the line-level partition, because with inbred
parents `GD_WI` is zero by construction and carries no signal. A hybrid's
self-coancestry is $1 - h_i/2$, with $h_i$ its heterozygous fraction, so
`GD_WI_hyb` is half the mean F1 heterozygosity and the two levels are linked:

$$2 \cdot GD_{WI\text{-}hyb} = GD_T + GD_{BS}$$

```{r}
c(lhs = 2 * gp[["GD_WI_hyb"]], rhs = gp[["GD_T"]] + gp[["GD_BS"]],
  GD_WI = gp[["GD_WI"]])
```

Exact on the full factorial, because every $A \times B$ pairing is present exactly
once. On a *selected* subset it holds only approximately. There $GD_T$ and
$GD_{BS}$ are computed from the line-usage weights, which is what a random
re-pairing of the selected parents would give, while `GD_WI_hyb` scores the pairs
actually chosen. The gap is the pairing structure of the subset.
`tests/testthat/test-dominance.R` asserts the identity to `1e-12` on the full
factorial and bounds the gap at `5e-3` on a *random* 40-hybrid subset, at one
scale (12 lines per pool, 800 markers). A subset chosen by an optimiser, which
concentrates line usage deliberately, is not covered by that bound and has not
been measured here.

## Inbreeding depression, and why it does not price $\alpha$ {#sec-heterosis-id}

The same model run backwards gives inbreeding depression. A locus at frequency
$p$ in a population with inbreeding coefficient $F$ is heterozygous with
probability $2pq(1-F)$, so the mean declines linearly in $F$
[@falconer1996, Ch. 14]:

$$ID \;=\; 2F\sum_k d_k \, p_k q_k$$

At $F = 1$ that is exactly the *shared* term of `heterosis_partition()` evaluated
at $p_A = p_B = p$. So the depression of a pool driven to complete homozygosity
and the expected heterosis of a cross made *within* that same pool are one
quantity seen from two sides. @mackay2020 treat heterosis and inbreeding
depression together on that basis, alongside transgressive segregation, which
this chapter does not take up. The identity needs no new machinery:

```{r}
c(inbreeding_depression = inbreeding_depression(pA, d),
  shared_term           = heterosis_partition(pA, pA, d)[["shared"]],
  deviation             = abs(inbreeding_depression(pA, d) -
                                heterosis_partition(pA, pA, d)[["shared"]]))
```

The closed form is checked against simulated genotypes rather than against its
own algebra, which is what `ref_inbreeding_depression()` exists for: it draws
each genotype as identical by descent with probability $F$ and as two independent
alleles otherwise, so the agreement is to Monte Carlo error:

```{r}
k <- 1:300
c(closed_form = inbreeding_depression(pA[k], d[k], 0.6),
  simulated   = ref_inbreeding_depression(pA[k], d[k], 0.6, n = 20000, seed = 2))
```

Now the part that matters, because the obvious next step is wrong.

It is tempting to argue: $\alpha$ budgets a loss of within-pool gene diversity;
gene diversity is $\sum_k p_kq_k$; inbreeding depression is $\sum_k d_k\,p_kq_k$;
therefore spending $\alpha$ costs yield in the current cycle, and *that* is the
price of the budget. The algebra does not support it.

```{r}
c(max_dominance_contribution_of_an_inbred_line = max(abs(heterosis_value(U, d))))
```

Zero, exactly. A fully inbred line has no heterozygous loci, so its dominance
contribution is zero **whatever the pool's allele frequencies are**. In a
programme whose parents are inbred lines the depression was paid in full when the
lines were made; it is a constant, and it does not move when the pool
subsequently loses diversity. The formula above describes the gap between a
hypothetical random-mating pool and its inbred derivatives, and that gap
*shrinks* as $p_kq_k$ shrinks -- but nothing in the programme stands at the
random-mating end of it.

::: {.callout-important}
## Inbreeding depression is not the current-cycle price of the diversity budget

This is the same fact that makes $GD_{WI}$ exactly zero in @sec-catalogue's
partition, and it has the same cause: with inbred parents there is no
within-individual variation left to lose. Two consequences.

The cost of $\alpha$ in this system remains what @sec-cycles measures -- an
option value, realised cycles later through the variance available to select on
(@sec-response), not a yield loss booked this season.

And inbreeding depression **is** the live quantity wherever the population is
only partially inbred: F2, F3 and F4 in @sec-sizing, where $F$ is still climbing
and the depression is still accumulating. That is the chapter it belongs to, not
this one.
:::

Note also what this does *not* say. Expected heterosis in the product is a
separate question. @sec-heterosis-flat answers it for the neutral case: under
drift the shared term lost and the divergence term gained cancel exactly, so
hybrid heterosis is invariant. Under selection the frequencies move
directionally, and what happens to expected heterosis is then the alignment
question of @sec-heterosis-alignment, which can go either way. In neither case is
the diversity loss itself an immediate dominance cost. The bill arrives later,
and it arrives in the additive variance.

## When genetic distance predicts heterosis exactly {#sec-heterosis-distance}

Before the corrections start, one special case is worth having, because it is the
only case in which marker distance and heterosis are the same number.

Set $d_k \equiv d$, constant across loci. Then $H$ stops being a weighted count and
becomes a plain one: $d$ times the number of loci at which the two parents differ.
For inbred lines that count is exactly what the coancestry kernel of
@sec-coancestry measures, so for a cross between lines $a$ and $b$

$$H_{ab} \;=\; d\,m\,(1 - f^L_{ab}) \;=\; d\,m\,\mathrm{MRD}_{ab}^2$$

```{r}
h1 <- heterosis_value(ctx$X, rep(1, ctx$m))
fL <- molecular_coancestry(L / 2)
ab <- cbind(ctx$ped$a, ctx$ped$b)
c(vs_coancestry = max(abs(h1 - ctx$m * (1 - fL[ab]))),
  vs_mrd        = max(abs(h1 - ctx$m * mrd_matrix(L / 2)[ab]^2)))
```

Both residuals are at machine precision. Under flat dominance, heterosis and
molecular coancestry are one statistic seen twice -- Trap 1 of @sec-stages read
from the other side -- and the diversity kernel this whole book runs on turns out
to be the right object for the heterosis question as well.

Flat $d$ is also precisely the condition under which $H$ is a function of marker
distance alone. It is a strong condition, and it fails as soon as $d$ varies: two
crosses at the same distance then differ by *which* loci they differ at. With the
gamma-distributed deviations used everywhere else in this chapter the relationship
survives but stops being an identity:

```{r}
c(cor_uniform_d = cor(h1, 1 - fL[ab]),
  cor_real_d    = cor(heterosis_value(ctx$X, d), 1 - fL[ab]))
```

A correlation of `r round(cor(heterosis_value(ctx$X, d), 1 - fL[ab]), 2)` is a
useful predictor and a poor identity. Two things move it, and both are properties
of the germplasm rather than of the statistic: how $d$ is spread over the loci at
which the parents differ, which is what @sec-heterosis-alignment is about, and --
once the third premise is relaxed -- how well the panel tags the causal loci.
Under flat $d$ on a panel that *is* the causal set the correlation is 1 by the
identity above. Departing from either condition generally lowers it -- not
always, since a $d$ that varies only across loci at which no pair of parents
differs changes no $H_i$ at all -- and by how much is something this chapter does
not attempt to predict.

## Divergence is not a bonus {#sec-heterosis-flat}

Read `divergence` as free money and you will reach for $F_{ST}$. Do not.

Draw the pools further apart and watch all three terms. Nothing here is selected.
Each point re-draws both pools from the *same* ancestral frequencies -- the
ancestral vector is fixed by `seed = 5` and shared across the sweep, while the two
pool-frequency vectors are fresh Balding-Nichols draws at each `fst`
(@sec-notation) -- so what varies along the sweep is how far the draw scatters
and nothing else:

```{r}
sweep_het <- function(fst) {
  cf <- modifyList(book_cfg, list(fst = fst, seed = 5))
  Lf <- simulate_lines(cf)
  pl <- rep(c("A", "B"), c(cf$n_pool_A, cf$n_pool_B))
  Pf <- pool_freqs(Lf / 2, pl)
  hp <- heterosis_partition(Pf["A", ], Pf["B", ], rep(1, cf$m)) / cf$m
  cm <- pool_complementarity(Lf / 2, pl)
  data.frame(fst = fst, shared = hp[["shared"]], divergence = hp[["divergence"]],
             total = hp[["total"]],
             frac_opposite_fixed = cm[["frac_opposite_fixed"]],
             mean_abs_delta_p = cm[["mean_abs_delta_p"]],
             mean_sq_delta_p = cm[["mean_sq_delta_p"]])
}
sw <- do.call(rbind, lapply(c(0.02, 0.05, 0.10, 0.20, 0.30, 0.45), sweep_het))
fmt(sw, 4)
```

Across the sweep the divergence term rises by a factor of
`r round(max(sw$divergence)/min(sw$divergence), 1)`. The comparison that matters
is in absolute units, not in ratios: divergence gains
`r round(sw$divergence[nrow(sw)] - sw$divergence[1], 4)` per locus and the shared
term loses `r round(sw$shared[1] - sw$shared[nrow(sw)], 4)`. The total does not
move.

### The two terms, term by term

The algebra is one line. Pool A and pool B are drawn *independently* around a
common ancestral frequency $p$, so $\mathbb{E}[p_A] = \mathbb{E}[p_B] = p$ and
$\mathbb{E}[p_A p_B] = p^2$. Therefore

$$\mathbb{E}\!\left[p_A + p_B - 2 p_A p_B\right] = 2p(1-p),$$

which contains no $F_{ST}$ at all. With ancestral frequencies drawn uniformly on
$[0.05, 0.95]$ the predicted level is $\mathbb{E}[2p(1-p)] =$ `r round(2*(0.5 - (0.9^2/12 + 0.25)), 4)`,
against a simulated mean of `r round(mean(sw$total), 4)` across the whole sweep.

That the *sum* is flat is the headline, but the Balding-Nichols model predicts
each half separately, which is the sharper statement. Under it
$\mathrm{Var}(p_A) = \mathrm{Var}(p_B) = F_{ST}\,p(1-p)$, so
$\mathrm{Var}(\bar p) = F_{ST}\,p(1-p)/2$ and $\mathrm{Var}(y) = 2F_{ST}\,p(1-p)$.
Substituting:

$$\mathbb{E}\!\left[2\bar p(1 - \bar p)\right] = 2p(1-p)\left(1 - \tfrac{F_{ST}}{2}\right),
\qquad
\mathbb{E}\!\left[y^2/2\right] = F_{ST}\,p(1-p)$$

The shared term loses $F_{ST}\,p(1-p)$ and the divergence term gains
$F_{ST}\,p(1-p)$: the same expression with opposite signs, not approximately and
not on average over some range. Both predictions are parameter-free once the
ancestral distribution is fixed, so they can be laid straight on top of the sweep:

```{r}
Ep <- 2 * (0.5 - (0.9^2 / 12 + 0.25))      # E[2p(1-p)] for p ~ U(0.05, 0.95)
sw$pred_shared     <- Ep * (1 - sw$fst / 2)
sw$pred_divergence <- Ep * sw$fst / 2
fmt(sw[, c("fst", "shared", "pred_shared", "divergence", "pred_divergence")], 4)
```

```{r}
#| fig-cap: "Expected heterosis and its two components as the pools are driven apart by drift alone. Dashed lines are the Balding-Nichols predictions for each component separately, with no fitted parameter. Each measured component is offset from its prediction -- divergence above, shared below -- by sampling error in the pool frequencies; that offset narrows along the axis, and the text quantifies both it and the measured gap. The divergence term grows several-fold; the shared term pays for all of it, unit for unit; the total is flat. Nothing is selected here -- this is what neutral divergence does."
plot(sw$fst, sw$total, type = "b", pch = 19, col = "#1b7837", ylim = c(0, 0.40),
     xlab = "target fst between pools", ylab = "expected heterosis per locus (d = 1)")
lines(sw$fst, sw$shared, type = "b", pch = 17, col = "#2166ac")
lines(sw$fst, sw$divergence, type = "b", pch = 15, col = "#b2182b")
lines(sw$fst, sw$pred_shared, lty = 2, col = "#2166ac")
lines(sw$fst, sw$pred_divergence, lty = 2, col = "#b2182b")
abline(h = mean(sw$total), lty = 2, col = "grey45")
legend("right", c("total", "shared", "divergence", "predicted"), bty = "n",
       cex = 0.8, pch = c(19, 17, 15, NA), lty = c(NA, NA, NA, 2),
       col = c("#1b7837", "#2166ac", "#b2182b", "grey45"))
grid()
```

The measured divergence sits above its prediction at every `fst`, and the measured
shared term about the same distance below it. That is not drift, it is
estimation. The
pool frequencies are estimated from `r book_cfg$n_pool_A` and
`r book_cfg$n_pool_B` sampled lines, and a frequency estimated from $n$ lines
carries sampling variance $V = pq/n$. Conditional on the true pool frequencies,
$\mathbb{E}[\hat y^2/2] = y^2/2 + (V_A + V_B)/2$ and
$\mathbb{E}[2\hat{\bar p}(1 - \hat{\bar p})] = 2\bar p(1 - \bar p) - (V_A + V_B)/2$:
equal and opposite, so the total estimated from sample frequencies is unbiased
while each half is not.

The offset is not constant along the sweep. It is proportional to
$\mathbb{E}[p_Aq_A + p_Bq_B]$, which the pools themselves erode as they diverge,
so it shrinks with $(1 - F_{ST})$: about
`r round(Ep * (1 - sw$fst[1]) * (1 / book_cfg$n_pool_A + 1 / book_cfg$n_pool_B) / 4, 4)`
per locus at `fst =` `r sw$fst[1]`, down to
`r round(Ep * (1 - sw$fst[nrow(sw)]) * (1 / book_cfg$n_pool_A + 1 / book_cfg$n_pool_B) / 4, 4)`
at `fst =` `r sw$fst[nrow(sw)]`, against measured gaps of
`r round(sw$divergence[1] - sw$pred_divergence[1], 4)` and
`r round(sw$divergence[nrow(sw)] - sw$pred_divergence[nrow(sw)], 4)`.

Those two pairs do not agree equally well. At `fst =` `r sw$fst[1]` prediction
and measurement are within
`r round(100 * abs((sw$divergence[1] - sw$pred_divergence[1]) / (Ep * (1 - sw$fst[1]) * (1 / book_cfg$n_pool_A + 1 / book_cfg$n_pool_B) / 4) - 1))`%
of each other; at `fst =` `r sw$fst[nrow(sw)]` the measured gap is
`r round(100 * (1 - (sw$divergence[nrow(sw)] - sw$pred_divergence[nrow(sw)]) / (Ep * (1 - sw$fst[nrow(sw)]) * (1 / book_cfg$n_pool_A + 1 / book_cfg$n_pool_B) / 4)))`%
*below* the predicted offset. The formula gives an expectation; each row of the
sweep is one draw of `r book_cfg$m` loci at one seed, and the scatter of that
draw is not constant either. $\mathrm{Var}(y)$ is $2F_{ST}\,p(1-p)$, so the
per-locus divergence $y^2/2$ has a variance that grows with $F_{ST}$ while the
number of loci averaging it stays fixed -- the high-`fst` rows are the noisy
ones, and the agreement at the bottom of the sweep is the more informative of the
two. The offset is the right order and the right sign at both ends; it is not a
per-point prediction, and nothing below rests on it being one.

What changes by an order of magnitude along the sweep is not the offset but the
signal it sits on: at `fst = 0.02` the offset exceeds the predicted divergence of
`r round(sw$pred_divergence[1], 4)`, and by `fst = 0.45` it is a small fraction of
it. A pool pair's *apparent* divergence overstates its true divergence, most
severely when the true divergence is small and the pools are few.

:::: {.callout-important}
## Every unit of divergence is paid for out of the shared term

Under neutral divergence the trade is exactly one for one. Drifting the pools
apart does not create heterosis; it relocates it from the shared term to the
divergence term. A programme that maximises $F_{ST}$ has bought nothing and spent
its within-pool diversity to do it.

This is the mechanism behind @sec-stages Trap 3. That section states the symptom
-- "$F_{ST}$ keeps rising as pools drift apart at loci that are *already*
divergent, which buys no additional heterosis" -- and points here for the reason.
The two displays above are that reason.
::::

## What actually buys heterosis {#sec-heterosis-alignment}

If divergence per se is worth nothing, why do heterotic groups work?

Because real pools are not diverged at random with respect to $d$. The divergence
term is $\sum_k d_k y_k^2 / 2$, a *weighted* sum. It grows when the loci that
diverge are the loci that carry dominance; under drift the two are independent,
which is exactly why the total was flat.

### Alignment is a covariance

That sentence can be a number rather than a word. A sum of products is $m$ times
a mean times a mean, plus $m$ times a covariance -- the population covariance
with divisor $m$, which is why the chunk rescales `cov()`:

$$\sum_k d_k\,y_k^2 / 2 \;=\; \frac{m}{2}\Big[\, \bar d \; \overline{y^2}
  \;+\; \mathrm{cov}(d,\, y^2) \,\Big]$$

```{r}
y2 <- (pA - pB)^2
c(divergence = sum(d * y2 / 2),
  from_parts = ctx$m / 2 * (mean(d) * mean(y2) + cov(d, y2) * (ctx$m - 1) / ctx$m))
```

The first bracketed term is what $F_{ST}$ can see: average divergence times
average dominance. The second is what it cannot. **Alignment is
$\mathrm{cov}(d, y^2)$**, and no statistic computed from allele frequencies alone
can contain it, because $d$ does not appear in one.

### The scenarios

Hold the pools fixed, hold $F_{ST}$ fixed, hold the multiset of dominance
deviations fixed, and change only which locus gets which $d_k$. `aligned` gives
the largest $d_k$ to the largest $y_k^2$, `anti-aligned` does the reverse, and
`independent` is the draw as made:

```{r}
mk <- function(dec) { v <- numeric(ctx$m); v[order(y2)] <- sort(d, decreasing = dec); v }
scen <- list(`anti-aligned` = mk(TRUE), `independent` = d, `aligned` = mk(FALSE))
al <- do.call(rbind, lapply(names(scen), function(n) {
  h <- heterosis_partition(pA, pB, scen[[n]])
  data.frame(scenario = n, cor_d_y2 = cor(scen[[n]], y2),
             shared = h[["shared"]], divergence = h[["divergence"]], total = h[["total"]])
}))
al$relative <- al$total / al$total[al$scenario == "independent"]
fmt(al, 3)
```

```{r}
#| fig-cap: "The same two pools, the same F_ST, and the same set of dominance deviations in all three bars -- only their placement across loci differs. Expected heterosis moves by half again between the extremes. F_ST cannot see any of this, because d does not enter it."
bp <- barplot(rbind(al$shared, al$divergence), beside = FALSE,
              names.arg = al$scenario, col = c("#2166ac", "#b2182b"),
              ylab = "expected heterosis", border = NA)
legend("topleft", c("shared", "divergence"), fill = c("#2166ac", "#b2182b"),
       bty = "n", cex = 0.8, border = NA)
text(bp, al$total, sprintf("%.2fx", al$relative), pos = 3, cex = 0.8, xpd = NA)
grid(nx = NA, ny = NULL)
```

$F_{ST}$ is identical in all three columns -- it is a function of the frequencies
only, and $d$ never enters it. Expected heterosis ranges over a factor of
`r round(max(al$total)/min(al$total), 2)`.

### Isolating the divergence channel

One honest qualification. Sorting $d$ against $y^2$ moves the *shared* term as
well, because $y^2$ is not independent of $2\bar p(1 - \bar p)$: a locus with a
large frequency difference tends to sit at an intermediate mean frequency. Part
of the effect above therefore arrives through the shared half, not the divergence
half.

The channel can be separated. Permute $d$ only *within* groups of loci that share
a value of the shared weight $2\bar p(1 - \bar p)$. The shared term is then
invariant by construction -- exactly, not approximately -- and whatever moves is
the divergence channel alone:

```{r}
hz  <- 2 * ((pA + pB) / 2) * (1 - (pA + pB) / 2)
grp <- split(seq_len(ctx$m), round(hz, 9))
strat <- function(dec) {
  v <- numeric(ctx$m)
  for (i in grp) v[i[order(y2[i])]] <- sort(d[i], decreasing = dec)
  v
}
st <- do.call(rbind, lapply(c(aligned = FALSE, `anti-aligned` = TRUE), function(dec)
  as.data.frame(as.list(heterosis_partition(pA, pB, strat(dec))))))
fmt(st, 4)
```

The shared column is identical to the digit; the divergence term still moves by
a factor of `r round(st$divergence[1] / st$divergence[2], 1)` and the total by
`r round(100 * (st$total[1] / st$total[2] - 1))`%. The alignment result survives
its own control.

## Reading the S1 panel again

@sec-stages logs `frac_opposite_fixed` and `mean_abs_delta_p` every cycle as
stand-ins for heterosis potential. The sweep above says how good they are.

`mean_sq_delta_p` is not a proxy at all. It is $\overline{y^2}$, so half of it is
the unweighted divergence term $GD_{BS}$, exactly:

```{r}
c(half_mean_sq = mean(sw$mean_sq_delta_p) / 2,
  divergence   = mean(sw$divergence))
```

What it cannot carry is the weighting by $d$, which is the previous section's
point; the frequency half of the divergence term it reports without loss.

`frac_opposite_fixed` is a different matter:

```{r}
fmt(sw[, c("fst", "divergence", "frac_opposite_fixed", "mean_abs_delta_p",
           "mean_sq_delta_p")], 4)
```

Up to `fst = 0.10` it is exactly zero -- not one marker in `r book_cfg$m`
qualifies -- and it first exceeds 1% of markers at `fst = 0.30`, near the top of
the sweep. It counts *near-fixed* opposite loci, and the divergence lives at
intermediate frequency differences, which the threshold discards.

**Log `mean_sq_delta_p`.** It is exact, it is monotone across the whole sweep,
and it costs one squaring more than the statistic already being logged.
`pool_complementarity()` and `s1_monitor()` return it.

## General and specific combining ability {#sec-heterosis-gca-sca}

Everything so far has been about $\mathbb{E}[H]$. Now the spread.

The same expansion that produced the shared/divergence split produces the
combining-ability decomposition, with no new machinery and no model fitting. Let
$u_{ik} \in \{0, 1\}$ be the genotype of pool-A line $i$ at locus $k$ and
$v_{jk}$ that of pool-B line $j$, and write each as its pool frequency plus a
deviation, $u = p_A + \delta u$ and $v = p_B + \delta v$. The per-locus
heterozygosity of the cross, $u + v - 2uv$, expands to

$$u + v - 2uv \;=\; \underbrace{p_A + p_B - 2p_Ap_B}_{\text{mean}}
  \;+\; \underbrace{(1 - 2p_B)\,\delta u}_{\text{parent A}}
  \;+\; \underbrace{(1 - 2p_A)\,\delta v}_{\text{parent B}}
  \;-\; \underbrace{2\,\delta u\, \delta v}_{\text{both}}$$

Four terms: one constant, one linear in each parent, one bilinear. Weight by $d_k$,
sum over loci, and the classical decomposition falls out,

$$H_{ij} = \mu + g^A_i + g^B_j + s_{ij}$$

with

$$g^A_i = \sum_k d_k (1 - 2p_{Bk})\,\delta u_{ik}, \qquad
  s_{ij} = -2 \sum_k d_k\, \delta u_{ik}\, \delta v_{jk}$$

and $g^B_j$ symmetrically. What is linear in one parent is general combining
ability: it is carried into every cross that parent enters. What is bilinear
depends on *which two* lines met, and is specific combining ability. Because the
deviations are taken from each pool's own line frequencies they sum to zero over
that pool's lines, which is what makes the three components mean zero and
mutually orthogonal. `heterosis_gca_sca()` is that expansion and nothing more:

```{r}
heterosis_gca_sca
```

It reconstructs every hybrid in the factorial exactly, and all three components
are mean zero:

```{r}
cs <- heterosis_gca_sca(U, V, d)
H  <- heterosis_value(ctx$X, d)
ia <- ctx$ped$a
ib <- ctx$ped$b - nA
c(max_residual = max(abs(H - (cs$mu + cs$gca_A[ia] + cs$gca_B[ib] +
                              cs$sca[cbind(ia, ib)]))),
  mu_vs_mean   = abs(cs$mu - mean(H)),
  mean_gca_A   = mean(cs$gca_A), mean_sca = mean(cs$sca))
```

This is not a decomposition that resembles the two-way analysis of variance. It
*is* the two-way analysis of variance: $\mu$, $g^A$ and $g^B$ are the grand mean
and the row and column effects of the $H_{ij}$ table, and $s_{ij}$ is the
interaction residual of the additive fit, to machine precision:

```{r}
max(abs(resid(lm(H ~ factor(ia) + factor(ib))) - cs$sca[cbind(ia, ib)]))
```

@comstock1949 designed reciprocal recurrent selection around the distinction
between these two components. The algebra above says what each one is made of --
and the next section says which of the two their scheme actually selects on.

### GCA is an average effect in the other pool's background

Look at the weight on $\delta u$: it is $d_k(1 - 2p_{Bk})$, and $1 - 2p_B$ is
$q_B - p_B$. That is the frequency factor of the classical average effect of a
gene substitution, $a + d\,(q - p)$ [@falconer1996, Ch. 7], with two differences:
the additive part $a$ is absent, because it belongs to the parental mean and not
to $H$, and the frequencies are **pool B's**, not pool A's.

This is the whole reason combining ability is not a property of a line. A pool-A
line's alleles are always substituted into a pool-B background, so its general
combining ability for heterosis is a dominance effect *evaluated at the tester
pool's frequencies*. Change the tester pool and every number changes. Two limits
follow immediately:

- A locus at which the tester pool sits at $p_B = 0.5$ contributes **exactly
  zero** to $g^A$, whatever $\delta u_{ik}$ is. It still contributes to $g^B$,
  through $(1 - 2p_A)\,\delta v_{jk}$, and to $s_{ij}$; what it cannot do is
  discriminate between pool-A parents.
- A locus at which the tester pool is fixed, $p_B \in \{0, 1\}$, has
  $\delta v \equiv 0$, so it contributes $\pm d_k\,\delta u_{ik}$ to $g^A$ and
  nothing to either $g^B$ or $s_{ij}$. The cross is then predictable from the
  pool-A parent by itself.

Between those two limits sits every real programme, and @sec-heterosis-relocation
is about which limit divergence pushes it towards.

This is also the exact sense in which @comstock1949's scheme is built on the
decomposition. Reciprocal recurrent selection evaluates each pool-A candidate on
its cross performance against a *sample of the opposite pool*, and symmetrically
for pool B. Averaging over testers removes $s_{ij}$, so of the heterosis in the
decomposition above only $g^A_i$ survives as a contrast between pool-A
candidates. Note what that does *not* say: cross performance is a genotypic
value, not a mid-parent difference, so it also carries the candidate's own
additive contribution -- the part that cancelled out of $H$ in the mid-parent
derivation above and is not in this decomposition at all.
Selection on testcross means therefore acts on the additive part and $g^A_i$
together. What the algebra here delimits is the dominance share of that response,
and it says the share is general combining ability against the tester pool.

### SCA is minus twice a weighted parental inner product

The bilinear term is $s_{ij} = -2\sum_k d_k\,\delta u_{ik}\,\delta v_{jk}$: minus
twice the $d$-weighted inner product of the two parents' *centred* genotypes, a
covariance up to the factor $1/m$. For $d_k > 0$, two parents that deviate from
their pools in the same direction have a positive inner product and therefore a
**negative** SCA.

Read in words: two parents that resemble each other -- after removing what is
typical of each pool -- cross badly, and the resemblance that counts is weighted
by $d$, not by marker count. Similarity enters negatively throughout this chapter
-- $1 - \theta_{AB}$, $d\,m\,(1 - f^L_{ab})$ -- but everywhere else it is
similarity measured against a fixed base. Here it is measured against each pool's
own mean, which is what makes the term specific to a *pair* rather than a
property either parent carries into all its crosses.

It is also @sec-heterosis-distance again. Since $\mu$, $g^A$ and $g^B$ are the
grand, row and column means of the $H_{ij}$ table, $s_{ij}$ is the double-centred
$H$ matrix. Under flat $d$, $H_{ij} = d\,m\,(1 - f^L_{ij})$, and double-centring
annihilates the constant and flips the sign of what remains:

$$s_{ij} \;=\; -\,d\,m \cdot \mathrm{DC}\big(f^L\big)_{ij},$$

with $\mathrm{DC}(M)_{ij} = M_{ij} - \bar M_{i\cdot} - \bar M_{\cdot j} + \bar
M_{\cdot\cdot}$ over the $A \times B$ block. The factor $-dm$ is not cosmetic:
$s_{ij}$ is on the trait scale and $f^L$ is a probability, and the minus is the
same minus as in $-2\sum_k d_k \delta u\, \delta v$ -- more coancestry than the
two margins predict means *less* SCA. So the coancestry kernel determines SCA
under exactly the condition it determines heterosis, up to that constant, and
stops doing so under exactly the same condition: a $d$ that varies across loci.

### The variance components

Because the three components are orthogonal, the variance of heterosis over the
factorial splits with them. Under the sampling model of the simulation -- each
line's genotype an independent Bernoulli($p_{Ak}$) draw at every locus, so
unrelated lines and linkage equilibrium within each pool -- each piece has a
closed form in the pool frequencies:

$$\sigma^2_{GCA_A} = \sum_k d_k^2\,(1 - 2p_{Bk})^2\,p_{Ak}q_{Ak},
\qquad
\sigma^2_{SCA} = 4 \sum_k d_k^2\,p_{Ak}q_{Ak}\,p_{Bk}q_{Bk}$$

```{r}
qa <- pA * (1 - pA)
qb <- pB * (1 - pB)
fmt(data.frame(
  component = c("GCA_A", "GCA_B", "SCA"),
  observed  = c(var(cs$gca_A), var(cs$gca_B), var(as.vector(cs$sca))),
  closed_form = c(sum(d^2 * (1 - 2 * pB)^2 * qa),
                  sum(d^2 * (1 - 2 * pA)^2 * qb),
                  4 * sum(d^2 * qa * qb))), 2)
```

The closed forms take the sample frequencies as the pool frequencies; the observed
column is one realisation of them. The GCA rows are variances over
`r book_cfg$n_pool_A` lines, whose relative sampling error is about
$\sqrt{2/n} \approx$ `r round(100 * sqrt(2 / book_cfg$n_pool_A))`%, and the
agreement is within that.

Note what each one requires. $\sigma^2_{SCA}$ carries $p_Aq_A\,p_Bq_B$: it needs
**both** pools to be polymorphic. Fix either pool at a locus and that locus
contributes no specific combining ability at all, whatever its $d_k$. Within-pool
diversity is not merely something the diversity constraint defends; it is the
raw material of SCA.

## Divergence destroys SCA faster than it builds GCA {#sec-heterosis-relocation}

Put the two closed forms next to the drift argument of @sec-heterosis-flat and
the variance tells a different story from the mean. The mean was flat; the
variance is not, and it does not merely change shape.

Under Balding-Nichols, $\mathbb{E}[p_Aq_A] = p(1-p)(1 - F_{ST})$ and
$\mathbb{E}[(1 - 2p_B)^2] = (1-2p)^2 + 4F_{ST}\,p(1-p)$, and $p_A$ and $p_B$ are
independent given the ancestral $p$. Per locus, in expectation over the draw,

$$\mathbb{E}\big[\sigma^2_{SCA,k}\big] = 4d_k^2\,p^2(1-p)^2\,(1 - F_{ST})^2,
\qquad
\mathbb{E}\big[\sigma^2_{GCA_A,k}\big] = d_k^2\,p(1-p)\,(1 - F_{ST})
  \left[(1-2p)^2 + 4F_{ST}\,p(1-p)\right]$$

Read the two expressions before reading the sweep, because they already say what
it will show.

Both carry the erosion factor $(1 - F_{ST})$, and both are **zero at
$F_{ST} = 1$**: a pool fixed at every locus has no two lines that differ, so it
has no combining-ability variance of either kind. Divergence therefore cannot
transfer variance from one component to the other indefinitely; it destroys both
in the limit. SCA carries the factor twice, because it needs both pools
polymorphic at the same locus. GCA carries it once, multiplied by the tester
factor $(1 - 2p_B)^2$, whose expectation *grows* with $F_{ST}$.

The two expressions therefore behave differently in $F_{ST}$. The SCA expression
is convex and decreasing on $[0, 1]$: it is a positive constant times
$(1 - F_{ST})^2$, so it only ever falls. The GCA expression is a *concave*
parabola, its $F_{ST}^2$ coefficient being $-4d_k^2p_k^2(1-p_k)^2$, and
differentiating it gives the only turnover in this section. GCA variance rises
from $F_{ST} = 0$ at a locus if and only if $8p(1-p) > 1$, that is
$\tfrac12(1 - 1/\sqrt2) < p < \tfrac12(1 + 1/\sqrt2)$, and turns over at

$$F^\star_k = \frac{8p_k(1-p_k) - 1}{8p_k(1-p_k)},$$

which is negative outside that interval -- at such loci GCA variance falls from
the start. A sum of concave parabolas is a concave parabola, and its maximum
works out to the $d_k^2\,(p_kq_k)^2$-weighted mean of the $F^\star_k$. Whether
that mean lands inside $[0, 1]$ is a property of the ancestral frequency
distribution, not a theorem: pools founded mostly at extreme frequencies would
put it below zero and GCA variance would decline monotonically. Under the
$U(0.05, 0.95)$ ancestral draw used here enough loci sit in the interval to place
it inside the range, which is why the sweep below straddles a peak. The sweep's
grid can show that the peak is interior; it cannot locate it.

What is monotone is the ratio,

$$\frac{\mathbb{E}[\sigma^2_{SCA,k}]}{\mathbb{E}[\sigma^2_{GCA_A,k}]}
  = \frac{4p(1-p)\,(1 - F_{ST})}{(1-2p)^2 + 4F_{ST}\,p(1-p)},$$

which falls in $F_{ST}$ at every locus. So the defensible statement is about the
*share*, not about a transfer. Expected heterosis, we already know, does not move
at all. Sweeping the pools apart -- same ancestral frequencies, fresh
Balding-Nichols draws at each `fst`, as in @sec-heterosis-flat -- and evaluating
the closed forms on the realised frequencies:

```{r}
reloc <- function(fst) {
  cf <- modifyList(book_cfg, list(fst = fst, seed = 5))
  Pf <- pool_freqs(simulate_lines(cf) / 2,
                   rep(c("A", "B"), c(cf$n_pool_A, cf$n_pool_B)))
  a <- Pf["A", ]; b <- Pf["B", ]
  va <- sum(d^2 * (1 - 2 * b)^2 * a * (1 - a))
  vb <- sum(d^2 * (1 - 2 * a)^2 * b * (1 - b))
  vs <- 4 * sum(d^2 * a * (1 - a) * b * (1 - b))
  data.frame(fst = fst, var_GCA = va + vb, var_SCA = vs,
             frac_SCA = vs / (va + vb + vs))
}
rl <- do.call(rbind, lapply(c(0.02, 0.05, 0.10, 0.15, 0.20, 0.30, 0.45), reloc))
fmt(rl, 3)
```

```{r}
#| fig-cap: "The same neutral divergence sweep, read through the combining-ability decomposition rather than through the mean. Expected heterosis is flat across this entire range (previous figure); its decomposition is not. Specific combining ability falls steeply throughout. General combining ability rises, by much less than SCA loses, and itself turns over inside the range -- both components vanish at F_ST = 1. The two curves do not sum to a constant: most of the SCA variance lost is destroyed rather than relocated, and the exact percentages are in the text."
plot(rl$fst, rl$var_GCA, type = "b", pch = 19, col = "#1b7837",
     ylim = range(0, rl$var_GCA, rl$var_SCA),
     xlab = "target fst between pools",
     ylab = "variance of heterosis over the factorial")
lines(rl$fst, rl$var_SCA, type = "b", pch = 15, col = "#b2182b")
legend("right", c("GCA (both pools)", "SCA"), bty = "n", cex = 0.8,
       pch = c(19, 15), col = c("#1b7837", "#b2182b"))
grid()
```

Take the three columns in order, because the headline is in the arithmetic
between them.

SCA variance falls by
`r round(100 * (1 - rl$var_SCA[nrow(rl)] / rl$var_SCA[1]))`% across the sweep, a
drop of `r round(rl$var_SCA[1] - rl$var_SCA[nrow(rl)], 1)` units. GCA variance
ends `r round(100 * (rl$var_GCA[nrow(rl)] / rl$var_GCA[1] - 1))`% above where it
started, a gain of `r round(rl$var_GCA[nrow(rl)] - rl$var_GCA[1], 1)` units, and
peaks at an `fst` of `r rl$fst[which.max(rl$var_GCA)]` -- interior, as the
ancestral distribution used here leads one to expect, though a grid whose
neighbouring points are `r rl$fst[which.max(rl$var_GCA) - 1]` and
`r rl$fst[which.max(rl$var_GCA) + 1]` locates the peak only to that interval and
cannot resolve it further. So the GCA gain recovers about
`r round(100 * (rl$var_GCA[nrow(rl)] - rl$var_GCA[1]) / (rl$var_SCA[1] - rl$var_SCA[nrow(rl)]))`%
of what SCA lost. The rest is gone: total variance in heterosis over the
factorial falls from `r round(rl$var_GCA[1] + rl$var_SCA[1], 1)` to
`r round(rl$var_GCA[nrow(rl)] + rl$var_SCA[nrow(rl)], 1)`, a
`r round(100 * (1 - (rl$var_GCA[nrow(rl)] + rl$var_SCA[nrow(rl)]) / (rl$var_GCA[1] + rl$var_SCA[1])))`%
decline, while the mean stays fixed throughout.

Read those four percentages as one realisation, not as constants. They come from
a single pair of seeds -- `seed = 5` for the pool frequencies, `set.seed(4)` for
$d$ -- with no replication and no dispersion reported, and they will move if
either changes. What does not depend on the seed is the direction, which is
already in the two closed forms above: $\sigma^2_{SCA}$ is a positive multiple of
$(1 - F_{ST})^2$ and can only fall, and both components vanish at $F_{ST} = 1$,
so no amount of neutral divergence can hold the total up. The ordering of the
columns is algebra; their magnitudes are this draw.

The share of that variance which is specific falls from
`r round(rl$frac_SCA[1], 3)` to `r round(rl$frac_SCA[nrow(rl)], 3)`. That
statement -- about the share -- is the one the sweep supports; "the dominance
expressed as SCA reappears as GCA" is not, and the two columns above are why.

The share result itself is not new. @reif2007 prove that the ratio of dominance
to additive variance in a hybrid population falls as its two parent populations
diverge, and conclude that superior hybrids then become predictable from general
combining ability effects. That half of this section is theirs. What the sweep
adds is the denominator, which a ratio hides: a ratio can fall because its
numerator falls, because its denominator rises, or both, and the columns above
say which. Here the numerator collapses and the denominator recovers only part of
it, so the share falls and the total falls with it.

> *Neutral* divergence -- drift, the only kind on this sweep -- does not buy
> expected heterosis, and it does not move heterosis from one combining-ability
> component into the other. It **shrinks the part of hybrid variation that only a
> cross can reveal**, destroying SCA variance faster than it builds GCA variance.
> An additive model looks better against diverged pools mainly because there is
> less left for it to miss. Divergence that is *aligned* with dominance is a
> different matter and @sec-heterosis-alignment is where it is measured; the
> sweep here holds alignment at zero by construction.

This sharpens a question @sec-genomic-prediction had to leave open. That chapter
fits one additive effect per marker and notes that such a model cannot represent
SCA at all. The sweep says how large the unrepresentable share is, and how it
moves: against poorly separated pools most of the variance in heterosis is
specific; against well separated ones a minority is. These are shares of the
variance in *heterosis*, not of total hybrid value, whose parental-mean part is
additive at every $F_{ST}$. @technow2014 report high accuracy for additive hybrid
prediction in established, clearly separated dent and flint pools, which is the
regime where this model puts the specific share at its lowest -- consistent with
the mechanism, though a single observational study in one crop cannot test it.

:::: {.callout-warning}
## A smaller SCA share is not a better programme, and this is not a licence to maximise $F_{ST}$

Three things stay true at every point on that sweep. Expected heterosis does not
change. The within-pool diversity spent to buy the divergence is spent for good.
And $\sigma^2_{SCA} \to 0$ is the limit in which no cross is better than its
parents predict -- there is nothing left to *find* by testing combinations, which
is the activity the whole factorial exists to support. Past the peak of the GCA
curve even the predictable half shrinks: at $F_{ST} = 1$ there is no
combining-ability variance of either kind, because no two lines in a pool differ.

The falling SCA share is therefore a fact about what a predictor can express, not
a gain. A programme reading it as one would be congratulating itself on having
less to predict.

So this is a second, independent reason for @sec-stages Trap 3's prescription to
hold $F_{ST}$ in a band rather than at a maximum. @sec-heterosis-flat gave the
diversity-side reason. This is the prediction-side one, and it points the same
way.
::::

### What this says about ranking crosses

The F1 of two fixed lines is genetically uniform, so the progeny-variance term
that drives the usefulness criterion at @sec-stages S2 is zero for the hybrid
itself, and cross ranking reduces to ranking means. That is true, and it is why
this book ranks hybrids on an index rather than on $\mu + i h \sigma$.

It does not follow that the ranking is easy. The *mean* of an F1 contains an SCA
term that an additive $\mathbf{G}$ omits entirely, and @sec-heterosis-relocation
sized its share of the variance in heterosis: a majority of it against weakly
separated pools, a minority against well separated ones, and zero only at the
degenerate limit where one pool is fixed at every locus. The
progeny variance collapses; the dominance deviation does not.

## What this does not change

:::: {.callout-warning}
## The diversity constraint stays additive

Nothing in this chapter reopens the optimisation. $\theta$, $GD$ and $\alpha$ are
computed on molecular coancestry and are functions of the frequencies alone; no
dominance term enters them, and @sec-ocs, @sec-combinatorial and @sec-advanced are
untouched. The combining-ability split of @sec-heterosis-gca-sca sits outside the
constraint for the same reason: it is a statement about the objective, not about
the feasible set.

A dominance model changes the **mean-performance ranking**, not the diversity
term. @sec-case-study states the same separation from the other side. Do not
conflate them, and do not read this chapter as licence to put a heterosis term
inside the constraint.
::::

What it does change is the interpretation of the constraint's output. Two plans
with the same $\alpha$ and the same $GD_{BS}$ can differ in expected heterosis by
half again, if one of them concentrated its divergence on the dominance loci --
that is the scenario table of @sec-heterosis-alignment, whose three columns share
every frequency-based statistic. The partition cannot distinguish them. Only an
estimate of $d$ can.

### What it would take to use $d$

@sec-catalogue leaves a weight $\lambda$ undetermined between within- and
between-pool diversity and observes that the ratio of the two heterosis terms is
what $\lambda$ is trying to express. This chapter can now say what would have to
be estimated to set it, and it is not much:

- **$\lambda$ needs two covariances, not $d$ itself.** Each heterosis term is $m$
  times a mean times a mean plus a covariance, so their ratio needs
  $\mathrm{cov}(d, y^2)$ and $\mathrm{cov}(d, 2\bar p(1 - \bar p))$, each relative
  to $\bar d$, on top of what the frequencies already give. A programme does not
  need per-locus dominance deviations to weight the two diversity terms against
  each other; it needs to know whether its divergence is landing on its dominance
  loci.
- **Ranking crosses needs more.** $s_{ij}$ is a $d$-weighted inner product between
  two specific parents, so it needs the deviations locus by locus. This is the
  extension @sec-case-study names and the book does not attempt.
- **Fitting it needs a factorial, and a testcross file is not one.** The
  decomposition above is the two-way layout with *identified* rows and columns:
  each pool-A line crossed to several named pool-B lines and each pool-B line to
  several named pool-A lines, which is what makes $g^A$, $g^B$ and $s_{ij}$
  separately estimable and what the `lm()` above fits. Testcrosses against a
  single common tester give only a column of $\mu + g^A_i + \text{(one tester's
  contribution)}$ -- no SCA is identifiable, because there is no second tester to
  contrast against. Programmes that already cross two pools in a partial
  factorial for hybrid evaluation have the layout; programmes that testcross
  against one inbred tester per pool do not.

## Where this connects

@sec-subdivided supplied the partition and @sec-catalogue will use it; this chapter
supplies the reason the $GD_{BS}$ component is singled out for protection, and the
reason protecting it is not the same as maximising it. @sec-stages Trap 3 already
recommended holding $F_{ST}$ in a band: @sec-heterosis-flat is why a band, and
@sec-heterosis-alignment is what to watch instead of the band.
@sec-heterosis-distance closes the loop with the coancestry kernel of
@sec-coancestry: under flat dominance, the diversity matrix and the heterosis
prediction are the same object.

@sec-genomic-prediction is the chapter this one supplies a missing term to. It
predicts breeding values with an additive model on a trait that is additive by
construction, and lists specific combining ability first among what it does not
model. @sec-heterosis-gca-sca is where that term comes from and
@sec-heterosis-relocation is what share of the variance in heterosis it holds;
both are derived here rather than fitted, because on a simulated $d$ they can be.

It also gives a reason why a heterotic *pattern* -- which pool crosses well onto
which -- is not recoverable from how far apart the pools have drifted:
@sec-heterosis-alignment's three scenarios share every frequency-based statistic
and differ by half again in expected heterosis. @white2020 describe heterotic
patterns and diversity separately in North American proprietary dent germplasm;
this chapter says why the two need describing separately.

Forward, the alignment result is the argument for @sec-cycles. Drift moves
frequencies independently of $d$, and selection on a line's own performance
cannot respond to $d$ either, because an inbred line's dominance contribution is
exactly zero (@sec-heterosis-id) -- which leaves selection on *cross* performance
as the only process in this model that can move frequencies in a way that depends
on where the dominance is. The alignment between divergence and dominance is
therefore something each cycle either builds or erodes, not a standing asset. A
one-cycle constraint cannot see that, and neither can $F_{ST}$. @sec-alphasimr
runs that loop on a dominance trait and is the pipeline this chapter's algebra
describes.
