# A catalogue of metrics on `f` {#sec-catalogue}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
n_sel <- 60
set.seed(41)
idx <- sample.int(ctx$N, n_sel)
```

Every metric in this chapter has the same signature, `f(idx, ctx) -> scalar`,
and every one operates on the molecular coancestry matrix from
@sec-coancestry. That uniformity is not tidiness: it is what makes metrics
interchangeable inside an optimiser, and it is what lets @sec-choosing benchmark
them against each other on equal terms.

The chapter also settles the question of what "no loss" means, which turns out
to be less obvious than it sounds.

## Group coancestry, and the off-diagonal problem

The group coancestry [@cockerham1967] of a set $S$ of $n$ individuals is the
mean of **all** $n^2$ elements, diagonal included:

$$\theta_S \;=\; \frac{1}{n^2}\sum_{i \in S}\sum_{j \in S} f_{ij}
            \;=\; \frac{1}{n^2}\,\mathbf{1}'\mathbf{f}_S\mathbf{1}.$$

It is the probability that two alleles sampled at random *from the whole set*
are identical -- including the possibility that both come from the same
individual. That is why the diagonal enters: it is the chance of sampling the
same individual twice, which is entirely real when the set is used as a crossing
pool.

Splitting the diagonal mean $\bar f_d$ from the off-diagonal mean $\bar f_o$,

$$\theta_S \;=\; \frac{\bar f_d}{n} \;+\; \frac{n-1}{n}\,\bar f_o . \tag{1}$$

So the off-diagonal mean -- the metric many programmes already report -- is one
of two terms. When $\bar f_d$ varies little across individuals, which is the
case in an all-F1 population, the two rank almost identically:

```{r}
sels <- list(random     = sel_random(ctx, n_sel),
             trunc_cap  = sel_truncation_cap(ctx, n_sel),
             greedy_w1  = sel_greedy(ctx, n_sel, w = 1),
             greedy_div = sel_greedy(ctx, n_sel, w = 0),
             truncation = sel_truncation(ctx, n_sel))

cmp <- t(sapply(sels, function(s)
  c(theta = theta_group(s, ctx), offdiag = mean_offdiag(s, ctx))))
fmt(as.data.frame(cmp), 5)
c(rank_correlation = cor(cmp[, "theta"], cmp[, "offdiag"], method = "spearman"))
```

The off-diagonal metric ranks correctly. What it does not give you is an
interpretable scale or a defensible baseline -- which is what the rest of this
chapter supplies.

## Gene diversity: $\theta$ and $H_e$ are the same quantity {#sec-identity}

This is the result that simplifies everything downstream. Starting from the
definition and swapping the order of summation, the double sum factors:

$$\frac{1}{n^2}\sum_i\sum_j x_{ik}x_{jk}
  = \Big(\frac{1}{n}\sum_i x_{ik}\Big)^2 = \bar p_k^2,$$

with $\bar p_k$ the **group's** allele frequency. Hence

$$\theta_S = \frac{1}{m}\sum_k \big(\bar p_k^2 + \bar q_k^2\big),
  \qquad
  GD_S = 1 - \theta_S = \frac{1}{m}\sum_k 2\,\bar p_k \bar q_k \;\equiv\; H_e,$$

which is exactly Nei's gene diversity [@nei1973]. **Group coancestry and
expected heterozygosity are not two independent checks. They are one quantity
written two ways.**

```{r}
c(via_submatrix = theta_group(idx, ctx),
  via_frequency = theta_from_freq(idx, ctx),
  difference    = theta_group(idx, ctx) - theta_from_freq(idx, ctx))

c(GD = gene_diversity(idx, ctx), He_Nei = he_nei(idx, ctx))
```

Equality to machine precision is the implementation's strongest sanity check.
If it fails, `f` is centered, rescaled, or built on the wrong marker coding.

Three practical consequences:

1. One fewer metric to justify to a technical committee.
2. "Losing 5% of diversity" acquires an exact meaning: 5% of the set's expected
   heterozygosity.
3. Two computation routes with the same answer and very different costs. The
   matrix route is $O(n^2)$; the frequency route is $O(nm)$. For $n = 200$ and
   $m = 25000$ that is forty thousand operations against five million.

```{r}
reps <- 40
t_mat <- system.time(for (i in seq_len(reps)) theta_group(idx, ctx))[["elapsed"]]
t_frq <- system.time(for (i in seq_len(reps)) theta_from_freq(idx, ctx))[["elapsed"]]
c(us_matrix = 1e6 * t_mat / reps, us_frequency = 1e6 * t_frq / reps,
  ratio = t_frq / t_mat)
```

The matrix route is the one that belongs in an optimiser's inner loop.
@sec-collapse makes it cheaper still.

## Status number $N_s$ {#sec-ns}

@lindgren1997 express group coancestry on the scale of individual counts:

$$N_s = \frac{1}{2\theta_S}.$$

$N_s$ is the number of **unrelated, non-inbred** individuals that would have the
same gene diversity as the observed set. The edge cases fix the intuition:

```{r}
fake <- ctx
fake$f <- matrix(0.5, 5, 5)      # unrelated, p = 0.5
c(theta = theta_group(1:5, fake), Ns = status_number(1:5, fake))

fake$f <- matrix(1, 5, 5)        # all identical and homozygous
c(theta = theta_group(1:5, fake), Ns = status_number(1:5, fake))
```

::: {.callout-warning}
## $N_s$ is a descriptor, not a drift projection

@toro2020 show that the **rate-based** effective size $N_e = 1/(2\Delta F)$
loses meaning under molecular diversity management: optimisation increases
diversity in early generations, producing a negative $\Delta F$ and therefore a
negative $N_e$, with no long-term asymptote.

This does not invalidate $N_s$. $N_s$ is a monotonic reparametrisation of
$\theta$ -- a *static descriptor of a set* -- while $N_e$ is a *rate between
generations*. Use $N_s$ to compare sets; do not extrapolate it to project future
drift. For the same reason, do not substitute the linkage-disequilibrium $N_e$
here: that estimates population history, not the composition of a selected
sample.
:::

And, from @sec-genomic-era: $\theta$ tracks homozygosity, not drift. $GD$ and
$N_s$ answer "how much heterozygosity does this selection preserve", which is
the right question for next season's hybrid vigour. They do not certify that
allele frequencies are moving as unmanaged drift would have moved them, which
matters more across multiple recurrent cycles than for a single round of hybrid
selection.

## Effective number of parents

A metric independent of $\mathbf{f}$, computed purely from the pedigree. With
$p_i$ the frequency with which line $i$ appears as a parent among the selected
hybrids,

$$N_{e(\text{par})} = \frac{1}{\sum_i p_i^2},$$

the inverse Simpson index. It captures precisely what coancestry blurs: a set of
200 hybrids can have an acceptable $\theta$ while drawing on only 12 lines, if
those 12 happen to be genetically distant from each other.

```{r}
c(sapply(sels, ne_parents, ctx = ctx), theoretical_max = ctx$n_lines)
```

Truncation collapses the parent base far more than it moves $\theta$. Its cost
is a `tabulate` call, which makes it a natural candidate to enter the objective
function alongside $\theta$ -- and in @sec-choosing it does.

## Allelic richness {#sec-richness}

$\theta$ and $H_e$ are averages across markers and are dominated by
intermediate-frequency alleles. A lost rare allele barely moves them, yet the
loss is irreversible. So count markers that were polymorphic in the population
and became rare or fixed in the selection:

$$A_{\text{lost}} = \#\{k : \text{MAF}^{(S)}_k < 0.05 \;\wedge\;
                            \text{MAF}^{(\text{pop})}_k \ge 0.05\}.$$

```{r}
sapply(sels, alleles_lost, ctx = ctx)
```

It is the only metric in this catalogue **genuinely non-redundant** with
$\theta$ -- a claim @sec-choosing tests rather than asserts.

@lopezcortegano2019 compared, in a subdivided simulated population, strategies
maximising *expected heterozygosity* ($H_T$, the same family as $GD$) against
strategies maximising *allelic diversity* ($A_T$). Maximising $A_T$ retained
more alleles **and** controlled molecular inbreeding better than maximising
$H_T$. The mechanism connects directly to @sec-genomic-era:
heterozygosity-maximising management pushes allele frequencies toward 0.5, which
favours rare alleles in the short run but does not stop them drifting to loss
once they are no longer the rarest; allele-count management defends every
segregating allele directly.

Their recommendation for structured conservation programmes was to make $A_T$,
not $H_T$, the primary target. In this pipeline's terms: `alleles_lost` deserves
a place next to $\theta$ as a *tracked outcome*, not only as a post-hoc sanity
check, whenever the selection feeds several future cycles rather than a single
hybrid release.

## Distances: E-NE and A-NE

With distance $d_{ij} = 1 - f_{ij}$, Core Hunter 3 [@debeukelaer2018] defines
two measures capturing opposite aspects that $\theta$ alone cannot separate:

$$\text{E-NE} = \frac{1}{n}\sum_{i \in S} \min_{j \in S,\, j \ne i} d_{ij},
  \qquad
  \text{A-NE} = \frac{1}{N}\sum_{i=1}^{N} \min_{j \in S} d_{ij}.$$

- **E-NE**, entry-to-nearest-entry, measures **non-redundancy within the
  selection**. It distinguishes "200 individuals spread out" from "100
  near-identical pairs spread out" -- configurations that can share the same
  $\theta$.
- **A-NE**, accession-to-nearest-entry, measures **representativeness of the
  population**. It detects a selection that optimises $\theta$ by concentrating
  in one region of genetic space and ignoring the rest.

```{r}
fmt(t(sapply(sels, function(s) c(ENE = ene(s, ctx), ANE = ane(s, ctx)))), 4)
```

Note the direction: A-NE is a distance to the *nearest selected entry*, so
**lower is better** for representativeness.

## Effective dimensionality

Over the eigenvalues of the submatrix, the participation ratio

$$D_{\text{eff}} = \frac{(\sum_i \lambda_i)^2}{\sum_i \lambda_i^2}$$

counts how many independent directions of genetic variation remain: 1 if all
variation lies on a single axis, $n$ if spread evenly.

```{r}
fmt(sapply(sels, eff_dim, ctx = ctx), 2)
```

The submatrix is **double-centered** before the eigendecomposition. On a raw
all-positive `f` the leading eigenvalue is just $n\theta$, so an uncentered
version would only re-measure $\theta$ under a more expensive name -- a mistake
worth naming because it is easy to make and produces a plausible-looking number.

## Decomposition by heterotic group

In hybrid maize the diversity **between** pools is the engine of heterosis and
cannot be spent; the diversity **within** each pool is the fuel for future
progress. @sec-heterosis derives the first half of that sentence -- `GD_BS`
weighted by dominance deviations *is* the divergence half of expected heterosis --
and qualifies it: divergence is worth having where it lands on the dominance loci,
and worth nothing where it does not. @sec-cycles demonstrates the second half.
Following @caballero2002, compute molecular coancestry between lines
and weight it by each line's usage among the selected hybrids:

$$\theta_A = \mathbf{w}_A' \mathbf{f}^{L} \mathbf{w}_A, \qquad
  \theta_B = \mathbf{w}_B' \mathbf{f}^{L} \mathbf{w}_B, \qquad
  \theta_{AB} = \mathbf{w}_A' \mathbf{f}^{L} \mathbf{w}_B.$$

```{r}
fmt(t(sapply(sels, theta_pools, ctx = ctx)), 4)
```

Feeding those into the Caballero & Toro partition of @sec-subdivided:

```{r}
fmt(t(sapply(sels, gd_partition, ctx = ctx)), 4)
```

Two things to read here. `GD_WI` is **exactly zero**: the parent lines are
inbred, so there is no within-individual diversity at line level -- the algebra
of @sec-subdivided predicted this, and it is why `GD_WI_hyb`, the heterozygosity
actually carried inside the F1s, is reported separately. And `GD_BS` *rises*
under truncation: selection concentrates on the most complementary line pairs,
which pushes the pools further apart.

::: {.callout-important}
## A blind optimiser erodes one pool while global $\theta$ looks fine

Without this decomposition that goes unnoticed, and the consequence appears only
in the following cycle. `ne_lines_pool()` is the cheap version of the same
alarm -- it needs no `fL`, only the pedigree:
:::

```{r}
t(sapply(sels, ne_lines_pool, ctx = ctx))
```

### Where this could go next

$\theta_A$, $\theta_B$ and $\theta_{AB}$ as computed here are **diagnostic**:
reported, not fed back into the objective. The subdivided-population literature
offers a ready-made way to promote them into a tunable target.
@caballerorodriguezramilo2010, used by @lopezcortegano2019 to manage two
heterotic-like pools simultaneously, partition total diversity as

$$H_T = D_G + \lambda H_S, \qquad A_T = D_A + \lambda A_S,$$

where $H_S$ and $A_S$ are the within-pool components, $D_G$ and $D_A$ the
between-pool components, and $\lambda \in [0, \infty)$ is a weight chosen by the
breeder. $\lambda = 0$ optimises only between-pool divergence -- protects
heterosis, ignores each pool's internal health. $\lambda \to \infty$ optimises
only within-pool diversity -- protects each pool's future options, ignores the
cross.

That maps directly onto the two constraints a maize breeder already has in mind
but applies only qualitatively: *don't blend the pools* versus *don't paint
either pool into a corner*. Turning $\lambda$ into a second explicit knob
alongside $\alpha_{\max}$ is the natural next step if pool erosion shows up in
practice. The current implementation reports the ingredients and leaves the
weighting to the analyst.

$\lambda$ need not stay a matter of taste. @sec-heterosis shows that the value of
between-pool divergence is not a free parameter but a $d$-weighted quantity: the
divergence term contributes $\sum_k d_k y_k^2 / 2$ against the shared term's
$\sum_k d_k \, 2\bar p_k (1 - \bar p_k)$. Given an estimate of the dominance
deviations, the ratio of those two sums is what $\lambda$ is trying to express,
and the reason a purely frequency-based partition cannot pin it down is that $d$
does not appear in the partition at all.

## The correct baseline {#sec-baseline}

### Why not compare against the whole population

The natural comparison -- coancestry of the selected set against that of the
whole population -- is **biased by construction**. Applying equation (1) to
both:

$$\theta_S - \theta_N
  = \left(\frac{1}{n} - \frac{1}{N}\right)\left(\bar f_d - \bar f_o\right) > 0,$$

since $\bar f_d > \bar f_o$ whenever individuals are not clones. A **random**
sample already looks less diverse than the population it came from, with no
selection whatsoever. The bias is pure sampling and grows as the sample shrinks.

```{r}
fd <- mean(diag(ctx$f)); fo <- mean(ctx$f[upper.tri(ctx$f)])
predicted_bias <- (1 / n_sel - 1 / ctx$N) * (fd - fo)

set.seed(11)
observed_bias <- mean(replicate(400, theta_group(sample.int(ctx$N, n_sel), ctx))) -
                 theta_group(seq_len(ctx$N), ctx)

c(predicted = predicted_bias, observed = observed_bias)
```

The formula is exact; the residual is Monte Carlo error in the resampling.

Note also that the **off-diagonal** mean of a random subset *is* an unbiased
estimator of the population's off-diagonal mean. So the metric many programmes
already use is free of this particular bias -- but it is also not comparable
across sample sizes on the $N_s$ scale.

### Operational definition of loss

The "0% loss" reference is the **distribution of random subsets of the same
size**:

$$\alpha \;=\; \frac{GD_{\text{random}(n)} - GD_S}{GD_{\text{random}(n)}}.$$

$\alpha = 0$ means "as diverse as a random sample of the same size";
$\alpha < 0$ means more diverse than chance. Resampling also delivers the
standard deviation, and hence a z-score for any strategy.

```{r}
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
c(GD_population = gene_diversity(seq_len(ctx$N), ctx),
  GD_reference  = null$gd_ref,
  sd            = null$gd_sd,
  apparent_loss_if_wrong_baseline =
    (gene_diversity(seq_len(ctx$N), ctx) - null$gd_ref) /
     gene_diversity(seq_len(ctx$N), ctx))
```

That last number is what a programme would report as a diversity loss if it
compared against the full pool. It is entirely a sampling artefact.

```{r}
fmt(sapply(sels, alpha_loss, ctx = ctx, gd_ref = null$gd_ref), 5)
```

Negative values are selections *more* diverse than chance, which is exactly what
the max-diversity greedy is for.

### How to set the $\alpha$ limit

The acceptable limit **should not be chosen a priori.** Pure truncation -- the
greediest possible selection -- defines the maximum attainable $\alpha$. If that
ceiling is below your intended limit, the constraint restricts nothing and you
have bought a false sense of control.

```{r}
attainable <- alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref)
c(attainable_alpha_max = attainable,
  intended_limit       = 0.05,
  constraint_is_active = attainable > 0.05)
```

At the full simulated scale that ceiling comes out near **3.6%**, so a 5% limit
would be inoperative. The correct procedure is to trace the $\alpha$-versus-gain
frontier and choose its knee, which is @sec-combinatorial.

## Every metric against the null

A metric without a baseline says nothing. The z-score of each metric against the
random-subset null is how this book reports all of them:

```{r}
tab <- evaluate(sels, ctx, null$gd_ref)
fmt(as.data.frame(discriminatory_power(tab, null)[
  , c("GD", "Ns", "Ne_parents", "F_hom", "F_drift", "alleles_lost", "ENE", "ANE")]), 1)
```

Read the rows against each other. Truncation is many standard deviations from
random on `F_drift` and on `Ne_parents` while being far less extreme on `GD` --
the constraint that would catch it is not the one most programmes apply. That
observation is what @sec-choosing turns into a recommendation.
