# Predicting breeding values: GBLUP, BayesA, BayesB {#sec-genomic-prediction}

```{r}
#| include: false
source("_common.R")
acc  <- read.csv(book_data("genomic_prediction_accuracy.csv"))
sens <- read.csv(book_data("genomic_prediction_pi_sensitivity.csv"))
```

Three earlier chapters flagged the same gap and moved on. @sec-simulating-crosses
built `ctx$traits` from *known* marker effects and called the result "an
optimistic ceiling." @sec-choosing's Finding 5 showed what happens once the index
carries real error, but manufactured that error rather than fitting a model to
produce it. @sec-case-study repeats the caveat and still moves on. This chapter
is where the gap closes: it fits actual prediction models to a training
population with real phenotypic noise, and hands the result to the same
`stage1_build()` contract every later chapter already consumes.

`vanraden_G()` has said since @sec-coancestry that its matrix "is the right
matrix for GBLUP." Nothing in the package has used it that way until now.

The chapter also has a second job, which only became apparent once the models
were fitted: an accuracy number means nothing until you say what was hidden from
the model. @sec-hybrid-cv shows that the obvious split -- hold out a random 40%
of the hybrids -- answers the easiest of the three questions a hybrid programme
actually asks, and that the fit itself says so, before the truth is consulted.

## A training population with real noise

Everywhere else in this book, `ctx$index` *is* the truth. Here it plays a
different role: it is the value a real breeding programme does not get to see,
and can only estimate from phenotypes recorded on a subset of individuals.

```{r}
ctx <- simulate_data(book_cfg)
N   <- ctx$N
g   <- ctx$traits[, 1]         # the true genetic value -- known only to us
```

A phenotype is the genetic value plus environmental noise, $y = g + e$, and
heritability is the fraction of phenotypic variance the genetic part accounts
for, $h^2 = \mathrm{Var}(g) / (\mathrm{Var}(g) + \mathrm{Var}(e))$. `traits`
is built by `scale()`, so $\mathrm{Var}(g) = 1$ exactly, and the noise variance
that delivers a target $h^2$ is $(1-h^2)/h^2$:

```{r}
h2 <- 0.5
set.seed(9)
e <- rnorm(N, 0, sqrt((1 - h2) / h2))
y <- g + e
```

Split the population into a training set, whose phenotypes the model gets to
see, and a candidate set -- the hybrids actually being selected among -- whose
phenotypes it does not:

```{r}
set.seed(9)
train <- sample.int(N, floor(0.6 * N))
cand  <- setdiff(seq_len(N), train)
c(train = length(train), candidates = length(cand))
```

## The truth this chapter is predicting {#sec-genpred-truth}

Two properties of $g$ decide everything that follows, and both are consequences
of how @sec-simulating-crosses built it rather than choices made here.

### The generative model is GBLUP's own assumption

`ctx$traits` is `scale(Z %*% B)` with the marker effects $\mathbf{B}$ drawn
independently. For a single trait that is $\mathbf{g} = \mathbf{Z}\boldsymbol\beta$
with $\mathrm{Var}(\boldsymbol\beta) = \mathbf{I}\sigma^2_\beta$, so

$$\mathrm{Var}(\mathbf{g}) = \mathbf{Z}\mathbf{Z}'\sigma^2_\beta
  = \mathbf{G}\,\underbrace{2\textstyle\sum_k p_k(1-p_k)\,\sigma^2_\beta}_{\sigma^2_u},$$

because $\mathbf{G}$ is *defined* as $\mathbf{Z}\mathbf{Z}'$ over that same
divisor (@sec-coancestry). The second equality is exact, not approximate:

```{r}
vr <- vanraden_G(ctx$X * 2)
s2 <- 2 * sum(vr$p * (1 - vr$p))
max(abs(tcrossprod(vr$Z) / s2 - ctx$G))
```

So the covariance GBLUP *assumes* for the breeding values is, up to the scalar
$\sigma^2_u$, the covariance the breeding values actually have. This simulation
is the infinitesimal model, exactly and not approximately, and that single fact
is most of the answer to "which model wins" below. It is also the sense in which
every accuracy in this chapter is a best case: no model here is misspecified.

### A hybrid's value is the mean of two parental values

The hybrids are F1s of homozygous lines, so a hybrid's marker row is the average
of its two parents' rows, and averaging is linear. For *any* additive effect
vector whatsoever, the hybrid's genetic value is therefore the mean of the two
parental line values -- general combining ability, with no specific combining
ability left over:

```{r}
L  <- simulate_lines(book_cfg)
bb <- rnorm(ctx$m)                    # any effects at all; the identity is linear
vL <- as.numeric((L / 2) %*% bb)      # the parental line values they imply
max(abs(as.numeric(ctx$X %*% bb) - (vL[ctx$ped$a] + vL[ctx$ped$b]) / 2))
```

The residual is at machine precision because the statement is an identity. Two
consequences run through the rest of the chapter. The mild one is that a
"hybrid-specific" prediction has nothing to predict here: the dominance
deviations of @sec-heterosis are absent from `ctx$traits` by construction, so
SCA is not merely unmodelled, it is not in the data. The sharp one is
@sec-g-singular.

## GBLUP: prediction from the relationship matrix

GBLUP treats the breeding value as a random effect $\mathbf{u}$ with covariance
proportional to the genomic relationship matrix, $\mathrm{Var}(\mathbf{u}) =
\mathbf{G}\,\sigma_u^2$, and solves the mixed model equations.

### Where the mixed model equations come from

The model is $\mathbf{y} = \mathbf{1}\mu + \mathbf{Z}\mathbf{u} + \mathbf{e}$
with $\mathbf{u} \sim (\mathbf{0}, \mathbf{G}\sigma^2_u)$ and $\mathbf{e} \sim
(\mathbf{0}, \mathbf{I}\sigma^2_e)$ independent. Treating $\mathbf{u}$ as a
random variable rather than a parameter to be estimated, @henderson1975
maximises the *joint* density of what is observed and what is not,

$$-2\log p(\mathbf{y}, \mathbf{u}) \;\propto\;
  \frac{(\mathbf{y} - \mathbf{1}\mu - \mathbf{Z}\mathbf{u})'
        (\mathbf{y} - \mathbf{1}\mu - \mathbf{Z}\mathbf{u})}{\sigma^2_e}
  \;+\; \frac{\mathbf{u}'\mathbf{G}^{-1}\mathbf{u}}{\sigma^2_u},$$

and the two stationarity conditions, multiplied through by $\sigma^2_e$, *are*

$$\begin{bmatrix}\mathbf{1}'\mathbf{1} & \mathbf{1}'\mathbf{Z} \\
                  \mathbf{Z}'\mathbf{1} & \mathbf{Z}'\mathbf{Z} + \mathbf{G}^{-1}\lambda
    \end{bmatrix}
    \begin{bmatrix}\hat\mu \\ \hat{\mathbf{u}}\end{bmatrix} =
    \begin{bmatrix}\mathbf{1}'\mathbf{y} \\ \mathbf{Z}'\mathbf{y}\end{bmatrix},
\qquad \lambda = \sigma_e^2/\sigma_u^2,$$

with $\mathbf{Z}$ the incidence matrix mapping training records to individuals.
Three things fall out of that derivation that the equations alone do not say.

$\mathbf{u}$ is defined over *all* $N$ individuals, training and candidates
together -- it enters through the prior, not through the data -- so one solve
produces predictions for both, and a candidate with no record of its own is
predicted entirely by its $\mathbf{G}$ entries against individuals that do have
one. Everything @sec-hybrid-cv finds is a consequence of that sentence.

$\lambda$ is a variance *ratio*, not a tuning parameter: $\sigma^2_u = 1$ because
`traits` is standardised, and `e` was built with variance $(1-h^2)/h^2$, so
$\lambda = (1-h^2)/h^2$ is not a coincidence but the same number entered twice.
Nothing in this chapter estimates it; a real programme would, by REML.

And under normality the joint maximiser is the conditional mean,
$\hat{\mathbf{u}} = \mathbb{E}[\mathbf{u} \mid \mathbf{y}]$, which is why BLUP
minimises mean squared error rather than merely being a sensible shrinkage rule.
Dropping normality it remains best among *linear* predictors, which is what the
acronym claims.

### Why the ridge blend is not optional {#sec-g-singular}

@sec-why-not-g proves $\mathbf{1}'\mathbf{G}\mathbf{1} = 0$ and reads it as the
reason $\mathbf{G}$ cannot measure diversity. It has a second consequence, for
prediction. $\mathbf{G}$ is positive semi-definite, and a quadratic form of a PSD
matrix vanishes only on its null space, so $\mathbf{1}'\mathbf{G}\mathbf{1} = 0$
forces

$$\mathbf{G}\mathbf{1} = \mathbf{0}.$$

$\mathbf{G}$ is singular for *every* population, whatever $N$ and $m$ are. The
blend is not insurance against an awkward case; without it `solve()` has nothing
to invert.

Here it is much worse than one lost dimension. By the identity of
@sec-genpred-truth every hybrid row is an average of two line rows, so the whole
$N \times m$ matrix lives in the span of the $L$ line genotypes. Each hybrid uses
exactly one line from each pool, which makes the pool-A indicator columns sum to
the pool-B ones and costs a further dimension, and centering costs one more:

```{r}
c(max_abs_G_times_one = max(abs(ctx$G %*% rep(1, N))))

rank_G <- function(n_per_pool) {
  cfg <- utils::modifyList(book_cfg, list(n_pool_A = n_per_pool, n_pool_B = n_per_pool))
  G   <- simulate_data(cfg)$G
  c(lines = 2 * n_per_pool, hybrids = nrow(G),
    rank = sum(eigen(G, only.values = TRUE)$values > 1e-8))
}
```

```{r}
#| echo: false
knitr::kable(as.data.frame(t(sapply(c(10, 20, 25), rank_G))),
  caption = "The rank of the hybrid G is the number of parent lines minus two, whatever N and m are.")
```

$\mathrm{rank}(\mathbf{G}) = L - 2$, independent of both the number of hybrids
and the number of markers. At the book's scale that is 48 dimensions carrying
625 hybrids: 577 of the eigenvalues are exactly zero, and `0.95 * G + 0.05 * I`
is what makes the equations solvable at all. Fifty lines were genotyped and
crossed into 625 combinations, but there were only ever 48 independent
dimensions to measure -- and no amount of phenotyping the 625 adds a
forty-ninth.

```{r}
G <- ctx$G
gblup <- function(y, train, h2) {
  Gstar <- 0.95 * G + 0.05 * diag(N)
  n_t <- length(train)
  Z  <- matrix(0, n_t, N); Z[cbind(seq_len(n_t), train)] <- 1
  X1 <- matrix(1, n_t, 1); lambda <- (1 - h2) / h2
  LHS <- rbind(cbind(crossprod(X1), crossprod(X1, Z)),
               cbind(crossprod(Z, X1), crossprod(Z) + solve(Gstar) * lambda))
  RHS <- rbind(crossprod(X1, y[train]), crossprod(Z, y[train]))
  C   <- solve(LHS)                                  # kept: it carries the PEVs
  list(u   = (C %*% RHS)[-1],
       rel = 1 - diag(C)[-1] * lambda / diag(Gstar))
}
fit   <- gblup(y, train, h2)
u_hat <- fit$u
cat("GBLUP accuracy (candidates):", cor(u_hat[cand], g[cand]), "\n")
```

### Reliability: the accuracy you can compute without the truth

Inverting the coefficient matrix rather than just solving against it costs one
more decomposition and returns something a breeding programme can act on. The
$\mathbf{u}$ block of that inverse holds the prediction error variances,
$\mathrm{PEV}_i = \sigma^2_e\,[\mathbf{C}^{uu}]_{ii}$, and the reliability of an
individual's prediction is the fraction of its prior variance the data removed:

$$r_i^2 \;=\; 1 - \frac{\mathrm{PEV}_i}{\sigma^2_u\,\mathbf{G}^{*}_{ii}}.$$

This is the same $r$ as the accuracy reported everywhere in this chapter, with
one difference that is the entire point: $g$ does not appear in it. A programme
can compute it for candidates it has never phenotyped and never will.

```{r}
c(mean_reliability = mean(fit$rel[cand]),
  implied_accuracy = sqrt(mean(fit$rel[cand])),
  realised_accuracy = cor(u_hat[cand], g[cand]))
```

The model under-states itself by about 0.06. Some of that is the blend's doing:
$0.95\mathbf{G} + 0.05\mathbf{I}$ tells the equations that every candidate
carries 5% of variance no relative can explain, and each $r_i^2$ is computed
against that inflated prior. Dialling the blend down says how much:

```{r}
blend <- function(w) {
  Gstar <- w * G + (1 - w) * diag(N)
  n_t <- length(train)
  Z  <- matrix(0, n_t, N); Z[cbind(seq_len(n_t), train)] <- 1
  X1 <- matrix(1, n_t, 1); lambda <- (1 - h2) / h2
  C  <- solve(rbind(cbind(crossprod(X1), crossprod(X1, Z)),
                    cbind(crossprod(Z, X1), crossprod(Z) + solve(Gstar) * lambda)))
  sqrt(mean((1 - diag(C)[-1] * lambda / diag(Gstar))[cand]))
}
c(implied_at_0.95 = blend(0.95), implied_at_0.999 = blend(0.999),
  realised = cor(u_hat[cand], g[cand]))
```

Roughly half the gap, then, and the other half survives however little identity
is mixed in. A mean reliability and a correlation taken across a *set* of
candidates are not the same quantity: the second is helped by the spread of true
values among the candidates, which the first knows nothing about. Both halves
point the same way -- the reliability is conservative -- and @sec-hybrid-cv shows
that the gap is this small only because this particular split was easy.

## What the random split was measuring {#sec-hybrid-cv}

The split at the top of this chapter took a random 60% of the *hybrids*. There
are 625 hybrids but only 50 lines, so a sample that size contains every line
several times over: each of the 250 candidates has **both of its parents**
sitting in the training set, in other hybrids with recorded phenotypes. Given
@sec-genpred-truth -- a hybrid's value is the mean of its two parental values --
that split asks the model to interpolate inside a factorial whose margins it has
already seen. It is not a hard question.

@burgueno2012 named the schemes that distinguish the cases, and @technow2014
applied them to exactly this design, a factorial between two heterotic pools.
Holding out *lines* rather than hybrids splits the candidates three ways: both
parents seen (the random split, CV2), one parent unseen (CV1), neither parent
seen (CV0). Removing 8 lines from each pool leaves 289 training hybrids and
gives 272 CV1 and 64 CV0 candidates; the CV2 row below draws a random training
set of the same 289, so the only thing that differs across the three rows is
*what* was hidden, not how much.

```{r}
cv_split <- function(seed, k = 8) {
  set.seed(seed)
  hoA <- sample(unique(ctx$ped$a), k)
  hoB <- sample(unique(ctx$ped$b), k)
  seen_a <- !(ctx$ped$a %in% hoA)
  seen_b <- !(ctx$ped$b %in% hoB)
  lin <- gblup(y, which(seen_a & seen_b), h2)         # lines held out
  rnd <- sample.int(N, sum(seen_a & seen_b))          # same size, random hybrids
  rd  <- gblup(y, rnd, h2)
  i2  <- setdiff(seq_len(N), rnd)
  i1  <- which(xor(seen_a, seen_b))
  i0  <- which(!seen_a & !seen_b)
  c(CV2 = cor(rd$u[i2],  g[i2]),  r2 = sqrt(mean(rd$rel[i2])),
    CV1 = cor(lin$u[i1], g[i1]),  r1 = sqrt(mean(lin$rel[i1])),
    CV0 = cor(lin$u[i0], g[i0]),  r0 = sqrt(mean(lin$rel[i0])))
}
cv <- t(sapply(1:20, cv_split))
```

```{r}
#| echo: false
knitr::kable(fmt(data.frame(
  scheme      = c("CV2 -- random hybrids held out",
                  "CV1 -- one parent unseen",
                  "CV0 -- both parents unseen"),
  accuracy    = colMeans(cv)[c("CV2", "CV1", "CV0")],
  sd          = apply(cv, 2, sd)[c("CV2", "CV1", "CV0")],
  reliability = colMeans(cv)[c("r2", "r1", "r0")],
  row.names   = NULL), 3),
  caption = "GBLUP over 20 line-holdout replicates at h2 = 0.5, training set fixed at 289 hybrids. `accuracy` is the realised correlation with the truth; `reliability` is the model's own estimate of it, computed without the truth.")
```

The accuracy the chapter opened with survives only in the first row. Ask the
model about a hybrid with one unseen parent and it drops by a third; ask about
one with two unseen parents and what is left is 0.12, with a spread across
replicates nearly twice that -- small, and not reliably distinguishable from
nothing. @habier2007 separate the two things
a genomic prediction can be using -- the relationship between candidates and
training individuals, and the marker effects estimated in their own right -- and
this table is that separation, measured. The first row is almost entirely
relatedness. The third row is what is left when relatedness is removed, and it
is the honest measure of what the 2000 markers bought.

That the third row collapses is not a defect of the fit. It is
$\mathrm{rank}(\mathbf{G}) = L - 2$ again: the 289 training records carry at most
34 lines' worth of independent information, and asking them to pin down 2000
marker effects well enough to score a genotype no relative shares is asking for
something that was never in the data. @daetwyler2008 give the deterministic
accuracy for exactly that regime, $r = \sqrt{n h^2 / (n h^2 + M_e)}$, and with
the units the simulation actually has -- $n$ the independent *lines*, and $M_e$
the marker count, because these markers are drawn independently and carry no
linkage to bundle them into segments -- it agrees:

```{r}
n_lines_train <- 2 * (book_cfg$n_pool_A - 8)
c(deterministic = sqrt(n_lines_train * h2 / (n_lines_train * h2 + book_cfg$m)),
  measured_CV0  = mean(cv[, "CV0"]))
```

The most useful column is the last one in the table. The reliabilities computed
from the mixed model equations -- with no access to $g$ -- track the realised
accuracies down the collapse, 0.64 against 0.65 for CV1 and 0.13 against 0.12
for CV0. The model knows. A programme that reports reliability alongside every
GEBV does not need to discover this the way this chapter did; the warning is in
the same solve that produced the prediction.

::: {.callout-warning}
## Report which parents the training set had seen

An accuracy figure without its cross-validation scheme is not interpretable.
The same fit, the same data and the same trait give 0.93, 0.65 or 0.12 here
depending only on which hybrids were withheld. In a hybrid programme this is not
a technicality: CV2 is the accuracy for filling in a partly-observed factorial,
CV1 for a cross onto a new line from one pool, CV0 for a cross between two new
lines. They are different questions, and only the first one is easy.
:::

## BayesA and BayesB: prediction from marker effects

GBLUP assumes every marker contributes an equal, infinitesimal share of the
genetic variance. The Bayesian alphabet [@meuwissen2001; @delcampos2013]
relaxes that one assumption at a time. **BayesA** gives every marker its own
variance, drawn from a scaled-inverse-$\chi^2$ prior -- effects are shrunk
individually rather than uniformly, which matters when a few markers carry
much more signal than the rest. **BayesB** adds a point mass at zero: each
marker has prior probability $\pi$ of being excluded entirely, so the model
performs variable selection as it fits. Both reduce to GBLUP's ridge-regression
solution as a special case; they differ only in how much of the genetic
architecture they let the data shape.

### The priors, explicitly

All three models write $\mathbf{g} = \mathbf{Z}\boldsymbol\beta$ and differ only
in what they assume about $\beta_k$. Ridge regression, and therefore GBLUP, uses
one variance for every marker; BayesA gives each marker its own, drawn from a
scaled inverse chi-square; BayesB puts a spike at zero in front of it:

$$\text{ridge, and so GBLUP:}\quad \beta_k \sim N(0, \sigma^2_\beta)
  \qquad\qquad
  \text{BayesA:}\quad \beta_k \sim N(0, \sigma^2_k), \;\; \sigma^2_k \sim \chi^{-2}(\nu, S)$$

$$\text{BayesB:}\quad \beta_k \sim
  \begin{cases} 0 & \text{with probability } \pi \\[2pt]
                N(0, \sigma^2_k) & \text{otherwise.}\end{cases}$$

The scaled inverse chi-square is the conjugate prior for a normal variance, which
is the whole reason it appears: it keeps the Gibbs sampler in closed form. Its
two hyperparameters are the arguments the fits below pass. $\nu$ is `df = 5`, the
prior weight in pseudo-observations -- small, so the prior stays vague, but not
zero, because the posterior is improper at $\nu = 0$. $S$ is set from
`R2 = 0.5`, the share of phenotypic variance the prior expects the markers to
explain, spread over the markers by their observed variation. Neither is fitted;
both are choices, and `R2` in particular is a statement about heritability made
before seeing the data.

This book fits both with `bWGR::BayesA()`/`bWGR::BayesB()`
[@xavier2019bwgr] rather than by hand -- unlike GBLUP, a Gibbs sampler over
marker effects is not the few lines the mixed-model solve above was, and
`bWGR` is exactly the tool `vanraden_G()`'s docstring pointed toward without
naming:

```{r}
library(bWGR)
M <- ctx$X * 2   # dosage 0/1/2, what bWGR expects
fitA <- BayesA(y[train], M[train, ], it = 1500, bi = 500, df = 5, R2 = 0.5)
fitB <- BayesB(y[train], M[train, ], it = 1500, bi = 500, pi = 0.95, df = 5, R2 = 0.5)
predA <- as.numeric(M[cand, ] %*% fitA$b)
predB <- as.numeric(M[cand, ] %*% fitB$b)
c(BayesA = cor(predA, g[cand]), BayesB = cor(predB, g[cand]))
```

Both fits produce a marker-effect vector `$b`; a candidate's GEBV is just its
genotypes dotted with those effects. GBLUP produced the same quantity from the
relationship matrix instead of the markers directly -- two routes to the same
thing, checked next.

### What the sampler actually selects

Both fits also return the quantities the priors above are about -- BayesA's
per-marker variances in `$vb`, BayesB's posterior inclusion probabilities in
`$d` -- and on the polygenic trait there is nothing in them to look at. Rebuild
the oligogenic architecture the cached sweep below uses, three markers of
`r book_cfg$m` given effects and everything else exactly zero, and there is:

```{r}
set.seed(book_cfg$seed + 42)
qtl  <- sample.int(book_cfg$m, 3)
Bq   <- rep(0, book_cfg$m); Bq[qtl] <- rnorm(3)
g_ol <- as.numeric(scale(vr$Z %*% Bq))
set.seed(31)
y_ol <- g_ol + rnorm(N, 0, sqrt((1 - h2) / h2))
round(apply(vr$Z[, qtl], 2, var) * Bq[qtl]^2 /
      sum(apply(vr$Z[, qtl], 2, var) * Bq[qtl]^2), 3)   # share of variance each
```

Worth knowing before reading anything into the sweep: the three effects were
drawn from `rnorm(3)`, and this draw put 96% of the genetic variance on one of
them. "Three causal markers" describes the construction; the trait is
effectively monogenic, and the true $\pi$ is nearer $1 - 1/2000$ than the
$1 - 3/2000$ the caption below quotes.

```{r}
oA <- BayesA(y_ol[train], M[train, ], it = 1500, bi = 500, df = 5, R2 = 0.5)
oB <- BayesB(y_ol[train], M[train, ], it = 1500, bi = 500, pi = 0.95, df = 5, R2 = 0.5)
big <- qtl[which.max(abs(Bq[qtl]))]
c(BayesB_d_at_QTL   = oB$d[big],  BayesB_d_median  = median(oB$d),
  BayesA_vb_at_QTL  = oA$vb[big], BayesA_vb_median = median(oA$vb))
```

```{r}
#| label: fig-genpred-selection
#| fig-cap: "The two priors on the same oligogenic trait, with the marker carrying 96% of the genetic variance circled. Both find it. Only BayesB can then remove the other 1999 from the model."
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
for (v in list(list(oB$d, "BayesB: posterior inclusion"),
               list(oA$vb, "BayesA: marker variance"))) {
  plot(v[[1]], pch = 16, cex = 0.3, col = "grey60", xlab = "marker", ylab = "",
       main = v[[2]], cex.main = 0.95)
  points(big, v[[1]][big], pch = 1, cex = 2, lwd = 2)
}
```

### What BayesB does that BayesA cannot

Both panels find the marker, which is worth saying plainly because the usual
telling of this comparison implies BayesA does not: its variance for the QTL is
about four times the median marker's and sits well clear of the background band.
What separates the two priors is not detection but what each can do afterwards.

BayesA estimates a private variance $\sigma^2_k$ for every marker, and the only
data bearing on $\sigma^2_k$ is the single effect $\beta_k$ drawn from it. One
observation never overwhelms a prior, so each $\sigma^2_k$ stays near the
$\chi^{-2}(\nu, S)$ it started from however many individuals are phenotyped --
@gianola2009 make this the central objection to describing the Bayesian alphabet
as "learning the genetic architecture". The cost is not the missed QTL. It is
that the other 1999 markers keep a variance they have not earned, and go on
absorbing signal that belongs to the one marker that has. BayesB's spike deletes
them, and it can afford to because it learns a single shared $\pi$ from all 2000
markers at once rather than 2000 private variances from one observation each.

So the advantage is real but bounded, and shows up in accuracy rather than in
the figure: on this architecture at $h^2 = 0.5$ the table below gives BayesB
0.952 against BayesA's 0.931 and GBLUP's 0.930. BayesA sits with GBLUP, not with
BayesB, which is the practical summary of the paragraph above.

## Verification: two routes agree

GBLUP with a flat prior on every marker's variance and the un-selected member
of the Bayesian alphabet, Bayesian ridge regression (`iv = FALSE, pi = 0` in
`bWGR::wgr()`), are the same model solved two different ways -- one through
the $N \times N$ relationship matrix, the other through the $N \times m$
marker matrix.

They are the same model in a stronger sense than "they should agree". Substitute
$\mathbf{u} = \mathbf{Z}\boldsymbol\beta$ and $\mathbf{G} = \mathbf{Z}\mathbf{Z}'
/ 2\sum_k p_k(1-p_k)$ into the equations above and the two solutions are related
by an identity: with $t$ the training rows,

$$\hat{\mathbf{u}} = \mathbf{G}_{\cdot t}
    \left(\mathbf{G}_{tt} + \mathbf{I}\lambda\right)^{-1}(\mathbf{y}_t - \hat\mu)
  \;=\; \mathbf{Z}\hat{\boldsymbol\beta},
  \qquad
  \hat{\boldsymbol\beta} = \mathbf{Z}_t'
    \left(\mathbf{Z}_t\mathbf{Z}_t' + \mathbf{I}\lambda_\beta\right)^{-1}
    (\mathbf{y}_t - \hat\mu),$$

with the two penalties differing by exactly the divisor that defines
$\mathbf{G}$: $\lambda_\beta = \lambda \cdot 2\sum_k p_k(1-p_k)$. Getting that
scalar wrong is the usual way this equivalence is reported as "approximate". It
is not approximate:

```{r}
lambda <- (1 - h2) / h2
yc  <- y[train] - mean(y[train])
u_G <- G[, train] %*% solve(G[train, train] + diag(lambda, length(train)), yc)
u_M <- vr$Z %*% crossprod(vr$Z[train, ],
         solve(tcrossprod(vr$Z[train, ]) + diag(lambda * s2, length(train)), yc))
max(abs(u_G - u_M))
```

Machine precision, over all `r N` individuals. Note this route uses the raw
$\mathbf{G}$: it needs no inverse of it, so @sec-g-singular's blend never enters,
which is also why it is the cleaner statement of the identity.

The Gibbs sampler should land in the same place, and does, though only to
sampling error -- it is estimating $\lambda$ from the data rather than being told
it, and the blend used in `gblup()` has no exact marker-space equivalent:

```{r}
yobs <- rep(NA_real_, N); yobs[train] <- y[train]
fit_brr <- wgr(yobs, M, iv = FALSE, pi = 0, it = 1500, bi = 500)
c(gblup_vs_gibbs = cor(u_hat, fit_brr$hat - mean(fit_brr$hat)))
```

That correlation, `r round(cor(u_hat, fit_brr$hat - mean(fit_brr$hat)), 3)`, is
the weaker of the two checks and is included because it exercises the code path
the Bayesian fits above actually use. The identity is what proves the two routes
are one model.

## Which model wins, and when

`ctx$traits` is built from thousands of markers with independent, similarly
sized effects -- the textbook case GBLUP's infinitesimal-model assumption was
designed for. A trait controlled by a handful of large-effect loci is a
different regime, and it is the one BayesB's point-mass prior exists for.
`book/scripts/regenerate.R genpred` reruns the fits above 15 times each, at
two heritabilities, under both `ctx$traits`'s polygenic architecture and a
book-local variant with genetic variance concentrated on three markers out of
`r book_cfg$m`:

```{r}
#| echo: false
knitr::kable(fmt(aggregate(cbind(GBLUP, BayesA, BayesB) ~ arch + h2, acc, mean), 3),
  caption = "Mean prediction accuracy (correlation, candidates only) over 15 independent training/candidate splits.")
```

```{r}
#| label: fig-genpred-accuracy
#| fig-cap: "Prediction accuracy by architecture and heritability, 15 replicate splits each."
sub <- reshape(acc, direction = "long", varying = c("GBLUP", "BayesA", "BayesB"),
               v.names = "accuracy", timevar = "method",
               times = c("GBLUP", "BayesA", "BayesB"))
sub$grp <- interaction(sub$arch, sub$h2, sep = "\nh2=")
boxplot(accuracy ~ method + grp, sub, las = 2, cex.axis = 0.7,
        col = rep(c("grey85", "grey65", "grey45"), 4),
        ylab = "accuracy (candidates)")
```

At $h^2 = 0.5$ under the oligogenic architecture, BayesB's mean accuracy is
`r round(with(subset(acc, arch=="oligogenic" & h2==0.5), mean(BayesB) - mean(GBLUP)), 3)`
higher than GBLUP's -- the gap the model was built to close. Under the
polygenic architecture the three methods are indistinguishable within
replicate noise: with genetic variance spread over thousands of markers,
BayesB's variable selection has nothing to select, and its point-mass prior
buys nothing GBLUP was not already doing.

That result is not evidence that the alphabet rarely helps. It is
@sec-genpred-truth restated: on the polygenic arm the trait was *generated* from
one shared marker variance, so GBLUP's prior is the true one and no competitor
can do better than match it. What the comparison establishes is the weaker and
more useful claim -- that BayesA and BayesB cost nothing when they are wrong, and
recover a real gap when they are right.

### `pi` was fixed, not tuned {#sec-genpred-pi}

A real analyst does not know the true sparsity in advance. `genpred`'s second
sweep holds the oligogenic architecture and $h^2 = 0.5$ fixed and only moves
`pi`:

```{r}
#| echo: false
knitr::kable(fmt(aggregate(BayesB ~ pi, sens, mean), 3),
  caption = "BayesB accuracy as `pi` moves from far too lenient toward the true sparsity (3 causal markers of 2000, pi = 0.9985).")
```

Accuracy improves monotonically as `pi` approaches the truth, but the
degradation from being off is gradual, not catastrophic -- `pi` is worth
cross-validating, not agonising over.

The obvious alternative is to stop choosing it. BayesC$\pi$ puts a prior on $\pi$
itself and samples it alongside the effects, which ought to make the sweep above
unnecessary. Fit it to both architectures:

```{r}
cpi_poly <- BayesCpi(y[train],    M[train, ], it = 1500, bi = 500, df = 5, R2 = 0.5)
cpi_olig <- BayesCpi(y_ol[train], M[train, ], it = 1500, bi = 500, df = 5, R2 = 0.5)
c(pi_hat_polygenic  = cpi_poly$pi,   # truth: 0
  pi_hat_oligogenic = cpi_olig$pi,   # truth: ~0.9995
  acc_oligogenic    = cor(as.numeric(M[cand, ] %*% cpi_olig$b), g_ol[cand]),
  acc_BayesB_at_095 = cor(as.numeric(M[cand, ] %*% oB$b), g_ol[cand]))
```

It returns essentially the same $\pi$ for a trait with 2000 causal markers as for
one that is effectively monogenic, and predicts the oligogenic trait slightly
worse than BayesB told the right answer. With 375 records and 2000 markers there is
little in the data that speaks to $\pi$ specifically -- @sec-g-singular's 48
dimensions again -- so what comes back is close to the prior. Estimating $\pi$ is
not a substitute for choosing it here; the sweep above remains the honest route.

## From GEBVs to `stage1_build()`

Everything above ends where @sec-simulating-crosses's contract begins. The
candidates' predicted values, whichever model produced them, are exactly the
`traits` argument `stage1_build()` expects -- one column per trait, one row
per hybrid:

```{r}
hybrids_traits <- data.frame(hybrid = cand, trait1 = predB)
head(hybrids_traits)
```

Repeat the fit once per trait in the multi-trait index and this data frame is
what feeds `stage1_build(X, ped, traits, fL)`. The second consumer is
`ocs_quadprog(f, gebv, lambda, ...)` of @sec-ocs, whose argument is literally
named `gebv` and has until now been handed the truth.

Carry the reliabilities with them. @sec-hybrid-cv's table is the argument for
storing `fit$rel` next to every predicted value rather than discarding it: a
candidate whose GEBV is unreliable is not a bad candidate, it is an unmeasured
one, and the difference matters to a selection decision.

## Back to Finding 5: a measured accuracy, not an assumed one

@sec-choosing's Finding 5 assumed a prediction accuracy `r` and manufactured
error correlated with relatedness to match it. The GBLUP fit above at
$h^2 = 0.5$ gives a *measured* accuracy of
`r round(cor(u_hat[cand], g[cand]), 2)` on the polygenic architecture, at the
top of the range Finding 5 explored ($r \in \{0.5, 0.7, 0.9, 1\}$) -- but
@sec-hybrid-cv is the reason not to read that as the typical case. It is the
CV2 number, the accuracy for a candidate whose two parents are already
phenotyped in other combinations. The CV1 row of that table, 0.65, sits near the
*bottom* of Finding 5's range, and the CV0 row falls well below anything it
explored. Between them the two chapters bracket the same interval from opposite
ends, and Finding 5's pessimistic cases are the realistic ones.

The errors are, in fact, structured by relatedness rather than white:

```{r}
err <- g[cand] - u_hat[cand]
fcand <- ctx$f[cand, cand]
ut <- upper.tri(fcand)
cor(outer(err, err)[ut], fcand[ut])
```

Pairs of candidates that are more related have more correlated prediction
errors -- the mechanism Finding 5 assumed, confirmed here from an actual
fitted model rather than injected. It is a modest correlation, not the
extreme case Finding 5 stress-tested, because $h^2 = 0.5$ with `r
length(train)` training records is a comparatively favourable regime; a
smaller or less related training set would push it higher.

::: {.callout-note}
## What this chapter still does not model

Closing the "traits are given" gap opens several more, named here so they are
not silently dropped:

- **Specific combining ability.** @sec-genpred-truth showed there is none in
  `ctx$traits` to find: the trait is additive by construction, so a hybrid's
  value is exactly the mean of its parents'. SCA is not merely unmodelled here,
  it is absent from the data. A real hybrid programme has it --
  @sec-heterosis's dominance deviations are where it comes from -- and would
  fit GCA and SCA jointly rather than one additive effect per marker.
- **Estimated variance components.** $h^2$ is supplied to `gblup()`, never
  estimated. A real fit gets $\lambda$ from REML on the training data, and the
  reliabilities of @sec-hybrid-cv inherit that estimate's uncertainty.
- **Multivariate prediction.** `ctx$traits` has three correlated columns; this
  chapter predicted one. A multi-trait Bayesian model borrows information
  across traits and should out-predict fitting each one separately, especially
  for the traits with less training data.
- **Forward validation.** @sec-hybrid-cv splits by line, which is the right
  axis for a factorial, but every scheme in it still samples training and
  candidates from one generation. A programme selecting for the future
  validates forward in time, on a later cycle than it trained on.
- **Retraining across cycles.** @sec-cycles runs the *selection* side of a
  multi-cycle programme with a fixed, perfect index. A model trained once and
  never updated goes stale as the population it predicts drifts from the
  population it was trained on.
- **Genotyping cost and marker density.** Accuracy was measured here at a
  fixed `r book_cfg$m` markers; a real programme trades marker density against
  genotyping cost per individual, not just against training-set size.
:::
