# Predicting breeding values: GBLUP, BayesA, BayesB {#sec-genomic-prediction}

```{r}
#| include: false
source("_common.R")
acc  <- read.csv(book_data("genomic_prediction_accuracy.csv"))
sens <- read.csv(book_data("genomic_prediction_pi_sensitivity.csv"))
```

Three earlier chapters flagged the same gap and moved on. @sec-simulating-crosses
built `ctx$traits` from *known* marker effects and called the result "an
optimistic ceiling." @sec-choosing's Finding 5 showed what happens once the index
carries real error, but manufactured that error rather than fitting a model to
produce it. @sec-case-study repeats the caveat and still moves on. This chapter
is where the gap closes: it fits actual prediction models to a training
population with real phenotypic noise, and hands the result to the same
`stage1_build()` contract every later chapter already consumes.

`vanraden_G()` has said since @sec-coancestry that its matrix "is the right
matrix for GBLUP." Nothing in the package has used it that way until now.

## A training population with real noise

Everywhere else in this book, `ctx$index` *is* the truth. Here it plays a
different role: it is the value a real breeding programme does not get to see,
and can only estimate from phenotypes recorded on a subset of individuals.

```{r}
ctx <- simulate_data(book_cfg)
N   <- ctx$N
g   <- ctx$traits[, 1]         # the true genetic value -- known only to us
```

A phenotype is the genetic value plus environmental noise, $y = g + e$, and
heritability is the fraction of phenotypic variance the genetic part accounts
for, $h^2 = \mathrm{Var}(g) / (\mathrm{Var}(g) + \mathrm{Var}(e))$. `traits`
is built by `scale()`, so $\mathrm{Var}(g) = 1$ exactly, and the noise variance
that delivers a target $h^2$ is $(1-h^2)/h^2$:

```{r}
h2 <- 0.5
set.seed(9)
e <- rnorm(N, 0, sqrt((1 - h2) / h2))
y <- g + e
```

Split the population into a training set, whose phenotypes the model gets to
see, and a candidate set -- the hybrids actually being selected among -- whose
phenotypes it does not:

```{r}
set.seed(9)
train <- sample.int(N, floor(0.6 * N))
cand  <- setdiff(seq_len(N), train)
c(train = length(train), candidates = length(cand))
```

## GBLUP: prediction from the relationship matrix

GBLUP treats the breeding value as a random effect $\mathbf{u}$ with covariance
proportional to the genomic relationship matrix, $\mathrm{Var}(\mathbf{u}) =
\mathbf{G}\,\sigma_u^2$, and solves the mixed model equations

$$\begin{bmatrix}\mathbf{1}'\mathbf{1} & \mathbf{1}'\mathbf{Z} \\
                  \mathbf{Z}'\mathbf{1} & \mathbf{Z}'\mathbf{Z} + \mathbf{G}^{-1}\lambda
    \end{bmatrix}
    \begin{bmatrix}\hat\mu \\ \hat{\mathbf{u}}\end{bmatrix} =
    \begin{bmatrix}\mathbf{1}'\mathbf{y} \\ \mathbf{Z}'\mathbf{y}\end{bmatrix},
\qquad \lambda = \sigma_e^2/\sigma_u^2,$$

with $\mathbf{Z}$ the incidence matrix mapping training records to individuals.
$\mathbf{u}$ is defined over *all* $N$ individuals, training and candidates
together, so one solve produces predictions for both. $\lambda$ is exactly
$(1-h^2)/h^2$ once traits are standardised, the same ratio used to build `e`
above. `vanraden_G()`'s $\mathbf{G}$ is blended a small amount with the identity
before inverting -- standard practice, and necessary here because `$G` can be
ill-conditioned at $N < m$:

```{r}
G <- ctx$G
gblup <- function(y, train, h2) {
  Gstar <- 0.95 * G + 0.05 * diag(N)
  n_t <- length(train)
  Z  <- matrix(0, n_t, N); Z[cbind(seq_len(n_t), train)] <- 1
  X1 <- matrix(1, n_t, 1); lambda <- (1 - h2) / h2
  LHS <- rbind(cbind(crossprod(X1), crossprod(X1, Z)),
               cbind(crossprod(Z, X1), crossprod(Z) + solve(Gstar) * lambda))
  RHS <- rbind(crossprod(X1, y[train]), crossprod(Z, y[train]))
  solve(LHS, RHS)[-1]
}
u_hat <- gblup(y, train, h2)
cat("GBLUP accuracy (candidates):", cor(u_hat[cand], g[cand]), "\n")
```

## BayesA and BayesB: prediction from marker effects

GBLUP assumes every marker contributes an equal, infinitesimal share of the
genetic variance. The Bayesian alphabet [@meuwissen2001; @delcampos2013]
relaxes that one assumption at a time. **BayesA** gives every marker its own
variance, drawn from a scaled-inverse-$\chi^2$ prior -- effects are shrunk
individually rather than uniformly, which matters when a few markers carry
much more signal than the rest. **BayesB** adds a point mass at zero: each
marker has prior probability $\pi$ of being excluded entirely, so the model
performs variable selection as it fits. Both reduce to GBLUP's ridge-regression
solution as a special case; they differ only in how much of the genetic
architecture they let the data shape.

This book fits both with `bWGR::BayesA()`/`bWGR::BayesB()`
[@xavier2019bwgr] rather than by hand -- unlike GBLUP, a Gibbs sampler over
marker effects is not the few lines the mixed-model solve above was, and
`bWGR` is exactly the tool `vanraden_G()`'s docstring pointed toward without
naming:

```{r}
library(bWGR)
M <- ctx$X * 2   # dosage 0/1/2, what bWGR expects
fitA <- BayesA(y[train], M[train, ], it = 1500, bi = 500, df = 5, R2 = 0.5)
fitB <- BayesB(y[train], M[train, ], it = 1500, bi = 500, pi = 0.95, df = 5, R2 = 0.5)
predA <- as.numeric(M[cand, ] %*% fitA$b)
predB <- as.numeric(M[cand, ] %*% fitB$b)
c(BayesA = cor(predA, g[cand]), BayesB = cor(predB, g[cand]))
```

Both fits produce a marker-effect vector `$b`; a candidate's GEBV is just its
genotypes dotted with those effects. GBLUP produced the same quantity from the
relationship matrix instead of the markers directly -- two routes to the same
thing, checked next.

## Verification: two routes agree

GBLUP with a flat prior on every marker's variance and the un-selected member
of the Bayesian alphabet, Bayesian ridge regression (`iv = FALSE, pi = 0` in
`bWGR::wgr()`), are the same model solved two different ways -- one through
the $N \times N$ relationship matrix, the other through the $N \times m$
marker matrix. They should agree, and at this book's scale checking that costs
one more fit:

```{r}
yobs <- rep(NA_real_, N); yobs[train] <- y[train]
fit_brr <- wgr(yobs, M, iv = FALSE, pi = 0, it = 1500, bi = 500)
c(worst_case_disagreement = cor(u_hat, fit_brr$hat - mean(fit_brr$hat)))
```

The correlation between the two routes, computed over all `r N` individuals,
is `r round(cor(u_hat, fit_brr$hat - mean(fit_brr$hat)), 3)` -- not exact,
because the Gibbs sampler is stochastic and the relationship-blend used above
has no exact marker-space equivalent, but close enough to confirm neither
route has a bug the other would not share.

## Which model wins, and when

`ctx$traits` is built from thousands of markers with independent, similarly
sized effects -- the textbook case GBLUP's infinitesimal-model assumption was
designed for. A trait controlled by a handful of large-effect loci is a
different regime, and it is the one BayesB's point-mass prior exists for.
`book/scripts/regenerate.R genpred` reruns the fits above 15 times each, at
two heritabilities, under both `ctx$traits`'s polygenic architecture and a
book-local variant with genetic variance concentrated on three markers out of
`r book_cfg$m`:

```{r}
#| echo: false
knitr::kable(fmt(aggregate(cbind(GBLUP, BayesA, BayesB) ~ arch + h2, acc, mean), 3),
  caption = "Mean prediction accuracy (correlation, candidates only) over 15 independent training/candidate splits.")
```

```{r}
#| fig-cap: "Prediction accuracy by architecture and heritability, 15 replicate splits each."
sub <- reshape(acc, direction = "long", varying = c("GBLUP", "BayesA", "BayesB"),
               v.names = "accuracy", timevar = "method",
               times = c("GBLUP", "BayesA", "BayesB"))
sub$grp <- interaction(sub$arch, sub$h2, sep = "\nh2=")
boxplot(accuracy ~ method + grp, sub, las = 2, cex.axis = 0.7,
        col = rep(c("grey85", "grey65", "grey45"), 4),
        ylab = "accuracy (candidates)")
```

At $h^2 = 0.5$ under the oligogenic architecture, BayesB's mean accuracy is
`r round(with(subset(acc, arch=="oligogenic" & h2==0.5), mean(BayesB) - mean(GBLUP)), 3)`
higher than GBLUP's -- the gap the model was built to close. Under the
polygenic architecture the three methods are indistinguishable within
replicate noise: with genetic variance spread over thousands of markers,
BayesB's variable selection has nothing to select, and its point-mass prior
buys nothing GBLUP was not already doing.

**The `pi = 0.95` above was fixed, not tuned.** A real analyst does not know
the true sparsity in advance. `genpred`'s second sweep holds the oligogenic
architecture and $h^2 = 0.5$ fixed and only moves `pi`:

```{r}
#| echo: false
knitr::kable(fmt(aggregate(BayesB ~ pi, sens, mean), 3),
  caption = "BayesB accuracy as `pi` moves from far too lenient toward the true sparsity (3 causal markers of 2000, pi = 0.9985).")
```

Accuracy improves monotonically as `pi` approaches the truth, but the
degradation from being off is gradual, not catastrophic -- `pi` is worth
worth cross-validating, not agonising over -- see the closing note below.

## From GEBVs to `stage1_build()`

Everything above ends where @sec-simulating-crosses's contract begins. The
candidates' predicted values, whichever model produced them, are exactly the
`traits` argument `stage1_build()` expects -- one column per trait, one row
per hybrid:

```{r}
hybrids_traits <- data.frame(hybrid = cand, trait1 = predB)
head(hybrids_traits)
```

Repeat the fit once per trait in the multi-trait index and this data frame is
what feeds `stage1_build(X, ped, traits, fL)`. Everything from @sec-ocs onward
already assumes exactly this shape; this chapter is only the part that used to
be silent about where it comes from.

## Back to Finding 5: a measured accuracy, not an assumed one

@sec-choosing's Finding 5 assumed a prediction accuracy `r` and manufactured
error correlated with relatedness to match it. The GBLUP fit above at
$h^2 = 0.5$ gives a *measured* accuracy of
`r round(cor(u_hat[cand], g[cand]), 2)` on the polygenic architecture -- near
the high end of the range Finding 5 explored ($r \in \{0.5, 0.7, 0.9, 1\}$), a
reminder that a well-powered training set at $h^2 = 0.5$ is a comparatively
easy case -- and its errors are, in fact, structured by relatedness rather
than white:

```{r}
err <- g[cand] - u_hat[cand]
fcand <- ctx$f[cand, cand]
ut <- upper.tri(fcand)
cor(outer(err, err)[ut], fcand[ut])
```

Pairs of candidates that are more related have more correlated prediction
errors -- the mechanism Finding 5 assumed, confirmed here from an actual
fitted model rather than injected. It is a modest correlation, not the
extreme case Finding 5 stress-tested, because $h^2 = 0.5$ with `r
length(train)` training records is a comparatively favourable regime; a
smaller or less related training set would push it higher.

::: {.callout-note}
## What this chapter still does not model

Closing the "traits are given" gap opens several more, named here so they are
not silently dropped:

- **Hybrid-specific prediction.** The GEBV above is additive, predicted from
  the hybrid's own dosage. It says nothing about specific combining ability --
  @sec-heterosis's dominance deviation $H_i$ is a separate quantity, and a
  full hybrid-prediction model would estimate GCA and SCA jointly rather than
  fitting one additive effect per marker.
- **Multivariate prediction.** `ctx$traits` has three correlated columns; this
  chapter predicted one. A multi-trait Bayesian model borrows information
  across traits and should out-predict fitting each one separately, especially
  for the traits with less training data.
- **Cross-validated accuracy, not a single split.** Every accuracy number above
  comes from one training/candidate partition (or `genpred`'s 15 replicates of
  one). A real programme estimates accuracy by k-fold or forward
  cross-validation before trusting a model with a selection decision.
- **Training-set design.** Accuracy here depends on training-set size and on
  how related the training records are to the candidates -- the same
  relatedness that makes $N_s$ (@sec-catalogue) and the founder-frequency
  discipline of @sec-genomic-era matter elsewhere in this book.
- **Retraining across cycles.** @sec-cycles runs the *selection* side of a
  multi-cycle programme with a fixed, perfect index. A model trained once and
  never updated goes stale as the population it predicts drifts from the
  population it was trained on.
- **Genotyping cost and marker density.** Accuracy was measured here at a
  fixed `r book_cfg$m` markers; a real programme trades marker density against
  genotyping cost per individual, not just against training-set size.
:::
