# Optimal contributions and the $\alpha$ constraint {#sec-ocs}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
n_sel <- 60
```

Everything so far has been measurement. This chapter turns the measurement into
a constraint an optimiser can respect, and connects it to the
optimal-contribution literature it specialises. `ctx$index` below is still the
known-marker-effects version built in @sec-simulating-crosses; @sec-genomic-prediction
shows how a real programme gets a `traits` matrix to hand `stage1_build()` in
the first place, and what accuracy it can expect once it does.

## Constraint, not a weighted sum

With $\mathbf{u}$ the multi-trait index and $S$ the selected set of size $n$:

$$\max_{S,\; |S| = n} \; \frac{1}{n}\sum_{i \in S} u_i
  \qquad \text{subject to} \qquad \alpha(S) \le \alpha_{\max}.$$

The common alternative, $\max\; \bar u - \lambda\,\theta_S$, requires
calibrating $\lambda$, which has no direct interpretation. The constraint form
preserves an operational meaning: *"I accept losing at most $\alpha_{\max}$ of
diversity."* That is a sentence a breeding committee can approve or reject.
"Diversity is worth $\lambda = 3.7$ index units" is not.

The two are not equivalent in practice either. A weighted sum lets the optimiser
buy index by spending diversity anywhere on the curve; a constraint fixes the
budget and makes the optimiser find the best plan *inside* it. Since the
acceptable loss is a policy decision and the exchange rate is not, the
constraint form puts the decision in the right hands.

```{r}
make_fitness
```

The violation enters as a *proportional* penalty, so a plan that overshoots by
10% of the budget is penalised in units the budget defines, whatever the
absolute value of $\alpha_{\max}$.

### This argument is older than it looks

The paragraphs above read as a claim about human decision-making: budgets are
approvable, exchange rates are not. That is true, and it undersells the case.
Kempthorne & Nordskog [-@kempthorne1959] posed the identical problem for
selection indices -- maximise aggregate gain subject to **zero expected change**
in a specified direction -- and solved it in closed form. @sec-response derives
their solution; `index_restricted()` implements it. The distinguishing property
is not softness:

```{r}
P <- matrix(c(4.0, 1.2, 1.2, 1.0), 2)
G <- matrix(c(2.0, 0.9, 0.9, 0.3), 2)
a <- c(1, 1); R <- cbind(c(0, 1))
c(unrestricted = drop(crossprod(R, G %*% index_smith_hazel(P, G, a))),
  restricted   = drop(crossprod(R, G %*% index_restricted(P, G, a, R))))
```

Exactly zero, not approximately. A penalty trades the restricted direction away
at some price; a restriction refuses to trade it at any price. That is the
distinction this section is drawing, and it has stood since 1959.

::: {.callout-note}
## What is actually new here

Not restriction over pricing. What is new is *what* is being restricted.
Kempthorne and Nordskog restricted a trait -- something with a genetic variance,
an economic weight and a row in $\mathbf{G}$. The $\alpha$ constraint restricts
the loss of molecular gene diversity, which has none of those: it is a property
of a matrix, not of a phenotype, and it is bounded relative to a resampling
baseline rather than held at zero.

The algebra of restriction carries over. The object being restricted does not,
which is why the rest of this chapter cannot simply cite 1959 and stop.
:::

## Relation to the standard OCS formulation

@gomezromano2016 write the general optimal-contribution problem, for a vector of
continuous contributions $\mathbf{c}$, as

$$\text{minimise} \quad \mathbf{c}'\mathbf{G}\mathbf{c}
  \qquad \text{subject to} \qquad
  \mathbf{c}'\mathbf{s} = 0.5,\;\; \mathbf{c}'\mathbf{d} = 0.5,\;\; c_i \ge 0,$$

with $\mathbf{s}$ and $\mathbf{d}$ indicating sex and $\mathbf{G}$ the coancestry
matrix chosen for managing diversity -- here, $\mathbf{f}$.

The formulation of this book is that problem **restricted to
$c_i \in \{0, 1/n\}$**: a fixed number $n$ of candidates contribute equally, none
partially. The minimisation of $\mathbf{c}'\mathbf{f}\mathbf{c}$ is turned into
a constraint and $\bar u$ is promoted to the objective.

That restriction to equal, all-or-nothing contributions is exactly what makes
the problem combinatorial, and what pushes the solution method from convex
quadratic programming to a metaheuristic. It is also not arbitrary: a commercial
hybrid programme plants a whole plot of a hybrid or does not plant it.

::: {.callout-note}
## The sex constraint, in plants

The per-sex sum constraint disappears in hermaphrodite and monoecious species
(a single sum to 1) and reappears in dioecious crops and in schemes using
cytoplasmic male sterility. In hybrid maize the natural problem is not OCS but
**optimal cross selection**: $\mathbf{c}$ stops being an individual contribution
and becomes a weight per cross, with the between-pool covariance entering
explicitly. And there is an extra restriction the animal literature does not
have -- the **maximum number of crosses executable in a field**, which makes the
problem mixed-integer.
:::

### The continuous relaxation as an upper bound {#sec-relaxation}

Dropping $c_i \in \{0, 1/n\}$ back to $0 \le c_i \le 1/n$ gives the continuous
relaxation of *this* restricted problem. Every discrete solution is feasible
under the relaxation, so the relaxation's optimum is a valid **upper bound**.

```{r}
project_capped_simplex
```

The cap is what makes the bound valid. Without $c_i \le 1/n$ the relaxation
would put all weight on the single best candidate and the bound would be
useless.

Turning the relaxation into a bound on the *constrained* problem needs one more
step, and turning it into an actual selection needs a rounding: both are in
@sec-advanced.

```{r}
lambdas <- 10^seq(-1, 2.5, length.out = 12)
relax <- ocs_relaxation(ctx, n_sel, lambdas)
fmt(as.data.frame(relax), 4)
```

Sweeping $\lambda$ traces the merit-versus-diversity frontier. Each row is the
best *continuous* plan at that exchange rate, and therefore an upper bound on
what any discrete plan of the same diversity can achieve.

```{r}
#| fig-cap: "The continuous relaxation traces an upper bound. Discrete strategies must sit below and to the left of it; how far below is the convergence diagnostic."
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
a_relax <- (null$gd_ref - relax[, "GD"]) / null$gd_ref

sels <- list(random     = sel_random(ctx, n_sel),
             trunc_cap  = sel_truncation_cap(ctx, n_sel),
             greedy_w1  = sel_greedy(ctx, n_sel, w = 1),
             greedy_w3  = sel_greedy(ctx, n_sel, w = 3),
             truncation = sel_truncation(ctx, n_sel))
pts <- t(sapply(sels, function(s)
  c(alpha = alpha_loss(s, ctx, null$gd_ref), index = mean_index(s, ctx))))

plot(a_relax * 100, relax[, "index"], type = "b", pch = 1, lty = 2, col = "grey45",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)",
     xlim = range(c(a_relax, pts[, "alpha"]) * 100),
     ylim = range(c(relax[, "index"], pts[, "index"])))
points(pts[, "alpha"] * 100, pts[, "index"], pch = 19, col = "#b2182b")
text(pts[, "alpha"] * 100, pts[, "index"], rownames(pts), pos = 2, cex = 0.7)
abline(v = 0, lty = 3)
legend("bottomright", c("continuous OCS relaxation (upper bound)", "discrete strategies"),
       col = c("grey45", "#b2182b"), pch = c(1, 19), lty = c(2, NA), bty = "n", cex = 0.8)
grid()
```

::: {.callout-warning}
## Too few iterations gives an "upper bound" below the discrete optimum

The relaxation is solved by projected gradient. Under-converged, it returns a
value *smaller* than the discrete optimum -- which is useless, and worse,
silently so. The default of 4000 iterations is the minimum that converged in
testing; re-check it whenever the dataset changes.
:::

```{r}
short <- ocs_relaxation(ctx, n_sel, lambdas = 10, iters = 50)
long  <- ocs_relaxation(ctx, n_sel, lambdas = 10, iters = 4000)
c(index_50_iters = short[, "index"], index_4000_iters = long[, "index"],
  greedy_w3 = mean_index(sels$greedy_w3, ctx))
```

## Setting $\alpha_{\max}$

The limit **should not be chosen a priori**. Pure truncation -- the greediest
possible selection -- defines the maximum attainable $\alpha$. If that ceiling
sits below your intended limit, the constraint restricts nothing.

```{r}
attainable <- alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref)
c(attainable_ceiling = attainable,
  a_5pct_limit_would_be_active = attainable > 0.05,
  a_1pct_limit_would_be_active = attainable > 0.01)
```

`stage2_select()` computes this ceiling automatically and warns about any
requested $\alpha$ above it, rather than silently returning an unconstrained
solution that looks constrained.

The correct procedure is to trace the frontier and choose its knee:

```{r}
#| fig-cap: "The greedy frontier, swept over the merit weight w. The knee is where further diversity costs index steeply."
ws <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5, 8, 12)
front <- do.call(rbind, lapply(ws, function(w) {
  s <- sel_greedy(ctx, n_sel, w = w)
  data.frame(w = w, alpha = alpha_loss(s, ctx, null$gd_ref),
             index = mean_index(s, ctx), Ne_par = ne_parents(s, ctx))
}))
plot(front$alpha * 100, front$index, type = "b", pch = 19, col = "#2166ac",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)")
abline(v = attainable * 100, lty = 2, col = "#b2182b")
text(attainable * 100, min(front$index), "truncation ceiling", pos = 2, cex = 0.7,
     col = "#b2182b")
grid()
```

```{r}
fmt(front, 4)
```

### How far $\alpha_{\max}$ moves when $\mathbf{f}$ moves

The ceiling above is quoted to four decimals, which invites more confidence than
the marker panel supports. @sec-genomic-era showed that the MAF spectrum decides
how badly the two inbreeding lenses disagree; the same spectrum feeds
$\mathbf{f}$, and therefore $\bar H_{\text{ref}}$ and the ceiling. It is worth
knowing which choices move the budget and which only look as though they should.

Rebuild everything downstream of the marker panel -- $\mathbf{f}$, the null, the
ceiling -- under a filter, a density and an ascertainment scenario:

```{r}
ceiling_for <- function(keep, label, B = 250) {
  Xs <- ctx$X[, keep, drop = FALSE]
  ck <- build_ctx(X = Xs, f = molecular_coancestry(Xs), G = NULL,
                  traits = ctx$traits, ped = ctx$ped, n_lines = ctx$n_lines,
                  pool = ctx$pool, fL = NULL, cfg = ctx$cfg)
  nl <- null_distribution(ck, n_sel, B = B, B_full = 20)
  data.frame(panel = label, markers = ncol(Xs), gd_ref = nl$gd_ref,
             alpha_max = alpha_loss(sel_truncation(ck, n_sel), ck, nl$gd_ref))
}
maf <- pmin(ctx$p0, 1 - ctx$p0)
polyA <- colMeans(simulate_lines(book_cfg)[seq_len(book_cfg$n_pool_A), ] / 2)
sens <- rbind(
  do.call(rbind, lapply(c(0, 0.01, 0.05, 0.10),
    function(t) ceiling_for(maf >= t, sprintf("MAF >= %.2f", t)))),
  ceiling_for(polyA > 0 & polyA < 1, "ascertained in pool A"))
fmt(sens, 4)
```

$\bar H_{\text{ref}}$ moves by
`r fmt(100 * (max(sens$gd_ref) / min(sens$gd_ref) - 1), 0)`% across these panels.
$\alpha_{\max}$ moves by
`r fmt(100 * (max(sens$alpha_max) / min(sens$alpha_max) - 1), 1)`%. That is not a
coincidence: $\alpha$ is a *ratio* of diversities measured on the same panel, so a
filter that shifts the absolute level shifts numerator and denominator together
and largely cancels out. **Filtering choices move the diversity level; they do not
move the budget.**

Density does not move it either, but it does something worse: it makes the
estimate noisy without making it wrong. Thinning the panel and averaging over six
independent draws of which markers survive:

```{r}
# Every subset is drawn BEFORE the loop. `null_distribution()` calls set.seed()
# internally, so a `sample.int()` written inside the loop would hand every
# replicate after the first the very same markers -- see the warning below.
set.seed(4)
subs <- lapply(c(250, 500, 1000),
               function(mm) lapply(1:6, function(i) sort(sample.int(ctx$m, mm))))
dens <- do.call(rbind, lapply(subs, function(sl) {
  v <- vapply(sl, function(k) ceiling_for(k, "", B = 200)$alpha_max, numeric(1))
  data.frame(markers = length(sl[[1]]), alpha_max = mean(v), sd = stats::sd(v))
}))
fmt(rbind(dens, data.frame(markers = ctx$m, alpha_max = attainable, sd = NA)), 4)
```

The means sit within one standard deviation of the full-panel ceiling all the way
down to 250 markers. The standard deviation does not: at 250 it is about
`r fmt(100 * dens$sd[1] / dens$alpha_max[1], 0)`% of the estimate itself. A budget
set from a thin panel is not biased, it is a coin flip -- and a single thin-panel
study has no way to tell which side it landed on.

So all three axes leave $\alpha_{\max}$ where it was. That is the useful finding,
and it is worth stating positively: **the budget is a more portable quantity than
the diversity level it is computed from.** $\bar H_{\text{ref}}$ is not
comparable across panels; $\alpha$ largely is, provided the panel is dense enough
for the estimate to be stable.

:::: {.callout-warning}
## `null_distribution()` reseeds, and will freeze a resampling loop

It calls `set.seed(seed)` on entry, with `seed = 7` by default, so the random
stream is reset every time it is invoked. Any `sample()` written *after* it inside
a loop therefore draws the same values on every pass but the first, and a
replication study silently becomes one replicate reported with a fabricated
standard deviation.

This is not hypothetical: the table above returned two different wrong answers --
once suggesting thin panels understate the ceiling, once that they overstate it --
before the subsets were hoisted out of the loop. Draw randomisation up front, or
pass an explicit varying `seed`.
::::

:::: {.callout-tip}
## Quote $\alpha_{\max}$ with the panel that produced it

A budget is not a property of the germplasm. It is a property of the germplasm
*and* the marker panel, and only the second of those changes when the genotyping
platform does. Re-derive the ceiling whenever the panel changes, and do not
compare an $\alpha$ across two studies genotyped differently -- @sec-choosing
Finding 3's caveat on transfer applies here with more force, because this one is a
constraint rather than a diagnostic.
::::

## A caveat carried over from the OCS literature

@gomezromano2016 show that the *predicted* rate of coancestry in the next
generation,
$E(\Delta f_{t+1}) = (\mathbf{c}'\mathbf{G}_t\mathbf{c} - f_t)/(1 - f_t)$, is
systematically **biased downward** relative to the coancestry actually realised
after offspring are generated -- by roughly a factor of two in their simulations
with small candidate pools, shrinking as the pool grows. The bias arises because
$\mathbf{c}'\mathbf{G}\mathbf{c}$ is an expectation over Mendelian segregation,
and the realised draw carries its own sampling variance on top of that
expectation.

This does **not** affect the $\theta_S$ computed in this book. $\mathbf{f}$ is
measured on real marker genotypes of an already-realised hybrid population;
there is no forward-prediction step to bias.

It becomes relevant the moment the pipeline is extended to *plan a further
generation* from the selected hybrids -- to predict *their* offspring's
coancestry from a contribution vector. At that point the naive
$\mathbf{c}'\mathbf{f}\mathbf{c}$ prediction should be expected to underestimate
realised coancestry, and the margin built into $\alpha_{\max}$ must account for
it.

## Quadratic programming, when contributions are continuous

Where the problem genuinely is continuous -- composing a synthetic, assembling a
core collection, setting seed proportions -- solve it as a quadratic programme
rather than with a metaheuristic. That is @sec-subdivided's
`optimal_contributions()`, with `ocs_quadprog()` adding the merit term:

```{r}
sub <- sample.int(ctx$n_lines, 20)
fLs <- ctx$fL[sub, sub]
merit <- as.vector(scale(rnorm(20)))

front_qp <- do.call(rbind, lapply(10^seq(-1, 2, length.out = 8), function(lam) {
  o <- ocs_quadprog(fLs, merit, lambda = lam, upper = 0.20)
  data.frame(lambda = lam, merit = o$merit, coancestry = o$coancestry,
             n_used = sum(o$c_i > 1e-6))
}))
fmt(front_qp, 4)
```

Note the `upper = 0.20` cap: an operational ceiling on how much any single line
may contribute. Caps of this kind are usually easier to defend than the
$\lambda$ that produced them, and they are the continuous analogue of the
per-line usage cap that @sec-combinatorial enforces in the decoder.

As $\lambda$ falls the solution concentrates on fewer sources -- the same
mechanism, in continuous form, that makes $N_{e,\text{par}}$ a necessary second
constraint in the discrete problem.

## What belongs in the inner loop

The optimiser performs on the order of $10^5$ to $10^6$ evaluations, so the cost
per call decides what goes into the objective and what waits for post-hoc
validation.

```{r}
fmt(cost_per_eval(ctx, n_sel, reps = 20), 1)
```

$\theta$ by the matrix route (or, better, the line route of @sec-collapse) and
$N_{e,\text{par}}$ go into the objective. Allelic richness, A-NE and $G_{st}$
are computed **once**, on the final solution. This is the same conclusion
@sec-choosing reached from the production-scale benchmark, reproduced here at
book scale so it can be re-run on your own data.
