# Optimal contributions and the $\alpha$ constraint {#sec-ocs}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
n_sel <- 60
```

Everything so far has been measurement. This chapter turns the measurement into
a constraint an optimiser can respect, and connects it to the
optimal-contribution literature it specialises.

## Constraint, not a weighted sum

With $\mathbf{u}$ the multi-trait index and $S$ the selected set of size $n$:

$$\max_{S,\; |S| = n} \; \frac{1}{n}\sum_{i \in S} u_i
  \qquad \text{subject to} \qquad \alpha(S) \le \alpha_{\max}.$$

The common alternative, $\max\; \bar u - \lambda\,\theta_S$, requires
calibrating $\lambda$, which has no direct interpretation. The constraint form
preserves an operational meaning: *"I accept losing at most $\alpha_{\max}$ of
diversity."* That is a sentence a breeding committee can approve or reject.
"Diversity is worth $\lambda = 3.7$ index units" is not.

The two are not equivalent in practice either. A weighted sum lets the optimiser
buy index by spending diversity anywhere on the curve; a constraint fixes the
budget and makes the optimiser find the best plan *inside* it. Since the
acceptable loss is a policy decision and the exchange rate is not, the
constraint form puts the decision in the right hands.

```{r}
make_fitness
```

The violation enters as a *proportional* penalty, so a plan that overshoots by
10% of the budget is penalised in units the budget defines, whatever the
absolute value of $\alpha_{\max}$.

## Relation to the standard OCS formulation

@gomezromano2016 write the general optimal-contribution problem, for a vector of
continuous contributions $\mathbf{c}$, as

$$\text{minimise} \quad \mathbf{c}'\mathbf{G}\mathbf{c}
  \qquad \text{subject to} \qquad
  \mathbf{c}'\mathbf{s} = 0.5,\;\; \mathbf{c}'\mathbf{d} = 0.5,\;\; c_i \ge 0,$$

with $\mathbf{s}$ and $\mathbf{d}$ indicating sex and $\mathbf{G}$ the coancestry
matrix chosen for managing diversity -- here, $\mathbf{f}$.

The formulation of this book is that problem **restricted to
$c_i \in \{0, 1/n\}$**: a fixed number $n$ of candidates contribute equally, none
partially. The minimisation of $\mathbf{c}'\mathbf{f}\mathbf{c}$ is turned into
a constraint and $\bar u$ is promoted to the objective.

That restriction to equal, all-or-nothing contributions is exactly what makes
the problem combinatorial, and what pushes the solution method from convex
quadratic programming to a metaheuristic. It is also not arbitrary: a commercial
hybrid programme plants a whole plot of a hybrid or does not plant it.

::: {.callout-note}
## The sex constraint, in plants

The per-sex sum constraint disappears in hermaphrodite and monoecious species
(a single sum to 1) and reappears in dioecious crops and in schemes using
cytoplasmic male sterility. In hybrid maize the natural problem is not OCS but
**optimal cross selection**: $\mathbf{c}$ stops being an individual contribution
and becomes a weight per cross, with the between-pool covariance entering
explicitly. And there is an extra restriction the animal literature does not
have -- the **maximum number of crosses executable in a field**, which makes the
problem mixed-integer.
:::

### The continuous relaxation as an upper bound {#sec-relaxation}

Dropping $c_i \in \{0, 1/n\}$ back to $0 \le c_i \le 1/n$ gives the continuous
relaxation of *this* restricted problem. Every discrete solution is feasible
under the relaxation, so the relaxation's optimum is a valid **upper bound**.

```{r}
project_capped_simplex
```

The cap is what makes the bound valid. Without $c_i \le 1/n$ the relaxation
would put all weight on the single best candidate and the bound would be
useless.

Turning the relaxation into a bound on the *constrained* problem needs one more
step, and turning it into an actual selection needs a rounding: both are in
@sec-advanced.

```{r}
lambdas <- 10^seq(-1, 2.5, length.out = 12)
relax <- ocs_relaxation(ctx, n_sel, lambdas)
fmt(as.data.frame(relax), 4)
```

Sweeping $\lambda$ traces the merit-versus-diversity frontier. Each row is the
best *continuous* plan at that exchange rate, and therefore an upper bound on
what any discrete plan of the same diversity can achieve.

```{r}
#| fig-cap: "The continuous relaxation traces an upper bound. Discrete strategies must sit below and to the left of it; how far below is the convergence diagnostic."
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
a_relax <- (null$gd_ref - relax[, "GD"]) / null$gd_ref

sels <- list(random     = sel_random(ctx, n_sel),
             trunc_cap  = sel_truncation_cap(ctx, n_sel),
             greedy_w1  = sel_greedy(ctx, n_sel, w = 1),
             greedy_w3  = sel_greedy(ctx, n_sel, w = 3),
             truncation = sel_truncation(ctx, n_sel))
pts <- t(sapply(sels, function(s)
  c(alpha = alpha_loss(s, ctx, null$gd_ref), index = mean_index(s, ctx))))

plot(a_relax * 100, relax[, "index"], type = "b", pch = 1, lty = 2, col = "grey45",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)",
     xlim = range(c(a_relax, pts[, "alpha"]) * 100),
     ylim = range(c(relax[, "index"], pts[, "index"])))
points(pts[, "alpha"] * 100, pts[, "index"], pch = 19, col = "#b2182b")
text(pts[, "alpha"] * 100, pts[, "index"], rownames(pts), pos = 2, cex = 0.7)
abline(v = 0, lty = 3)
legend("bottomright", c("continuous OCS relaxation (upper bound)", "discrete strategies"),
       col = c("grey45", "#b2182b"), pch = c(1, 19), lty = c(2, NA), bty = "n", cex = 0.8)
grid()
```

::: {.callout-warning}
## Too few iterations gives an "upper bound" below the discrete optimum

The relaxation is solved by projected gradient. Under-converged, it returns a
value *smaller* than the discrete optimum -- which is useless, and worse,
silently so. The default of 4000 iterations is the minimum that converged in
testing; re-check it whenever the dataset changes.
:::

```{r}
short <- ocs_relaxation(ctx, n_sel, lambdas = 10, iters = 50)
long  <- ocs_relaxation(ctx, n_sel, lambdas = 10, iters = 4000)
c(index_50_iters = short[, "index"], index_4000_iters = long[, "index"],
  greedy_w3 = mean_index(sels$greedy_w3, ctx))
```

## Setting $\alpha_{\max}$

The limit **should not be chosen a priori**. Pure truncation -- the greediest
possible selection -- defines the maximum attainable $\alpha$. If that ceiling
sits below your intended limit, the constraint restricts nothing.

```{r}
attainable <- alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref)
c(attainable_ceiling = attainable,
  a_5pct_limit_would_be_active = attainable > 0.05,
  a_1pct_limit_would_be_active = attainable > 0.01)
```

`stage2_select()` computes this ceiling automatically and warns about any
requested $\alpha$ above it, rather than silently returning an unconstrained
solution that looks constrained.

The correct procedure is to trace the frontier and choose its knee:

```{r}
#| fig-cap: "The greedy frontier, swept over the merit weight w. The knee is where further diversity costs index steeply."
ws <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5, 8, 12)
front <- do.call(rbind, lapply(ws, function(w) {
  s <- sel_greedy(ctx, n_sel, w = w)
  data.frame(w = w, alpha = alpha_loss(s, ctx, null$gd_ref),
             index = mean_index(s, ctx), Ne_par = ne_parents(s, ctx))
}))
plot(front$alpha * 100, front$index, type = "b", pch = 19, col = "#2166ac",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)")
abline(v = attainable * 100, lty = 2, col = "#b2182b")
text(attainable * 100, min(front$index), "truncation ceiling", pos = 2, cex = 0.7,
     col = "#b2182b")
grid()
```

```{r}
fmt(front, 4)
```

## A caveat carried over from the OCS literature

@gomezromano2016 show that the *predicted* rate of coancestry in the next
generation,
$E(\Delta f_{t+1}) = (\mathbf{c}'\mathbf{G}_t\mathbf{c} - f_t)/(1 - f_t)$, is
systematically **biased downward** relative to the coancestry actually realised
after offspring are generated -- by roughly a factor of two in their simulations
with small candidate pools, shrinking as the pool grows. The bias arises because
$\mathbf{c}'\mathbf{G}\mathbf{c}$ is an expectation over Mendelian segregation,
and the realised draw carries its own sampling variance on top of that
expectation.

This does **not** affect the $\theta_S$ computed in this book. $\mathbf{f}$ is
measured on real marker genotypes of an already-realised hybrid population;
there is no forward-prediction step to bias.

It becomes relevant the moment the pipeline is extended to *plan a further
generation* from the selected hybrids -- to predict *their* offspring's
coancestry from a contribution vector. At that point the naive
$\mathbf{c}'\mathbf{f}\mathbf{c}$ prediction should be expected to underestimate
realised coancestry, and the margin built into $\alpha_{\max}$ must account for
it.

## Quadratic programming, when contributions are continuous

Where the problem genuinely is continuous -- composing a synthetic, assembling a
core collection, setting seed proportions -- solve it as a quadratic programme
rather than with a metaheuristic. That is @sec-subdivided's
`optimal_contributions()`, with `ocs_quadprog()` adding the merit term:

```{r}
sub <- sample.int(ctx$n_lines, 20)
fLs <- ctx$fL[sub, sub]
merit <- as.vector(scale(rnorm(20)))

front_qp <- do.call(rbind, lapply(10^seq(-1, 2, length.out = 8), function(lam) {
  o <- ocs_quadprog(fLs, merit, lambda = lam, upper = 0.20)
  data.frame(lambda = lam, merit = o$merit, coancestry = o$coancestry,
             n_used = sum(o$c_i > 1e-6))
}))
fmt(front_qp, 4)
```

Note the `upper = 0.20` cap: an operational ceiling on how much any single line
may contribute. Caps of this kind are usually easier to defend than the
$\lambda$ that produced them, and they are the continuous analogue of the
per-line usage cap that @sec-combinatorial enforces in the decoder.

As $\lambda$ falls the solution concentrates on fewer sources -- the same
mechanism, in continuous form, that makes $N_{e,\text{par}}$ a necessary second
constraint in the discrete problem.

## What belongs in the inner loop

The optimiser performs on the order of $10^5$ to $10^6$ evaluations, so the cost
per call decides what goes into the objective and what waits for post-hoc
validation.

```{r}
fmt(cost_per_eval(ctx, n_sel, reps = 20), 1)
```

$\theta$ by the matrix route (or, better, the line route of @sec-collapse) and
$N_{e,\text{par}}$ go into the objective. Allelic richness, A-NE and $G_{st}$
are computed **once**, on the final solution. This is the same conclusion
@sec-choosing reached from the production-scale benchmark, reproduced here at
book scale so it can be re-run on your own data.
