# Combinatorial selection {#sec-combinatorial}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
n_sel <- 60
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
de <- read.csv(book_data("de_constraints.csv"))
bench <- read.csv(book_data("search_benchmark.csv"))
# Mean index per method and scale in the head-to-head benchmark of @sec-advanced,
# so this chapter quotes the same numbers that chapter measured.
bmean <- function(scale, method)
  mean(bench$index[bench$scale == scale & bench$method == method])
bsec <- function(scale, method)
  mean(bench$secs[bench$scale == scale & bench$method == method])
```

The problem is: choose $n$ of $N$ hybrids, maximising mean index subject to a
diversity budget. This chapter builds up the strategies that solve it, from the
two that need no optimiser at all to the differential evolution that beats them
-- and explains why the DE only beats them under one specific condition.

The decision is a concrete one, and it is the last of a cycle in a
cross-pollinated crop. Two heterotic pools, A and B, hold $L_A$ and $L_B$
inbred (or doubled-haploid) lines. Every candidate is one cross
$A_a \times B_b$, so the candidate set is the factorial $N = L_A \times L_B$ --
in maize, the set of single crosses that could be made from the current elite
lines without introducing anything new. The $n$ you keep are the crosses that
go forward: into multi-location testing, into seed increase, and -- through
their parents -- into the next recycling cycle. Keeping too few distinct
parents is not a statistical inconvenience, it is a decision to narrow the pool
that the programme will breed from next.

**What has to be in hand before any line of this chapter runs** is the
three-object contract of @sec-stages:

- `X`, the $N \times m$ hybrid marker matrix, coded $\{0, 0.5, 1\}$ as the
  average of the two homozygous parents;
- `f`, the $N \times N$ molecular coancestry of @caballero2002, values in
  $[0, 1]$, built from `X` and never from a centred genomic relationship matrix
  (@sec-why-not-g);
- `hybrids`, carrying each cross's parents and its predicted traits, from which
  `build_ctx()` forms the single index `ctx$index`, standardised to unit
  variance over the candidate pool (@sec-genomic-prediction).

**Assumptions, declared once and not violated afterwards.** Parent lines are
homozygous, so a hybrid's genotype is the mean of its parents and `X` takes
only the three values above. Markers are treated as neutral. The index is
*frozen* before selection: nothing is re-predicted inside the optimiser, which
matters because the index carries error that is correlated with relatedness
(@sec-choosing, finding 5). The allele-frequency base `ctx$p0` is the candidate
pool and is set once in `build_ctx()`. Diversity loss $\alpha$ is always
measured against a random subset of *the same size* `n_sel` (@sec-baseline),
never against the whole pool. Dominance appears nowhere in the objective: it is
already inside the predicted hybrid performance, and the between-pool
divergence that generates heterosis is a quantity to *preserve*, not to
minimise (@sec-heterosis).

The instance this chapter runs on:

```{r}
#| label: instance
c(candidates_N      = ctx$N,
  lines_A           = sum(ctx$pool == "A"),
  lines_B           = sum(ctx$pool == "B"),
  markers_m         = ctx$m,
  hybrids_kept      = n_sel,
  uses_per_line     = 2 * n_sel / ctx$n_lines,
  gd_ref            = null$gd_ref,
  gd_ref_sd         = null$gd_sd,
  alpha_ceiling_pct = 100 * alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref))
```

Read it as a breeder would. `r n_sel` crosses are kept out of `r ctx$N`, which
is `r round(100 * n_sel / ctx$N, 1)`% of the factorial. The `r n_sel` crosses
consume $2n =$ `r 2 * n_sel` parent slots spread over `r ctx$n_lines` lines, so
a perfectly even plan would use each line `r 2 * n_sel / ctx$n_lines` times --
the scale against which any per-line cap has to be set. `gd_ref` is the gene
diversity of a random subset of exactly this size, with its resampling standard
deviation next to it, and the last row is the **attainable ceiling**: pure
truncation loses `r fmt(100 * alpha_loss(sel_truncation(ctx, n_sel), ctx, null$gd_ref), 2)`%
of it. Any $\alpha_{\max}$ above that ceiling constrains nothing at all, which
is the first thing to check and the subject of @sec-ocs.

## Four baselines before any optimiser

Four families -- random, truncation, truncation under a per-line cap, and
greedy -- in six rows, because greedy is shown at three merit weights.

```{r}
sels <- list(
  random     = sel_random(ctx, n_sel),
  truncation = sel_truncation(ctx, n_sel),
  trunc_cap  = sel_truncation_cap(ctx, n_sel),
  greedy_div = sel_greedy(ctx, n_sel, w = 0),
  greedy_w1  = sel_greedy(ctx, n_sel, w = 1),
  greedy_w3  = sel_greedy(ctx, n_sel, w = 3)
)
fmt(t(sapply(sels, function(s) c(
  index      = mean_index(s, ctx),
  alpha_pct  = 100 * alpha_loss(s, ctx, null$gd_ref),
  Ns         = status_number(s, ctx),
  Ne_parents = ne_parents(s, ctx),
  max_line   = max_line_use(s, ctx)))), 3)
```

```{r}
#| label: greedy-collapses
c(w1_is_truncation = setequal(sels[["greedy_w1"]], sels[["truncation"]]),
  w3_is_truncation = setequal(sels[["greedy_w3"]], sels[["truncation"]]))
```

Three of those six rows are the same plan. `greedy_w1` and `greedy_w3` return
*exactly* the truncation set -- same index, same $\alpha$, same `max_line` --
because both weights sit far above the crossover measured in the `w-units`
chunk below, where the merit term swamps the coancestry term and the greedy
argmax degenerates to the index ranking. Read them as one row until that
measurement has been made; it is the reason a $w$ grid of round numbers is a
trap.

Read the columns in this order, because each answers a different question and
three of them are routinely skipped:

1. `index`, in standard deviations of the candidate pool -- what the plan is
   worth.
2. `alpha_pct`, the loss of gene diversity relative to a random subset of the
   same size -- what it costs, on the lens the constraint will pin. Negative
   means *more* diverse than random.
3. `Ns`, the status number $1/(2\theta)$ (@sec-ns) -- the same information as
   $\alpha$ on an absolute scale, kept because it is what the animal literature
   quotes.
4. `Ne_parents`, the effective number of parent lines, which is computed from
   counts alone and is the metric that catches "$\theta$ looks fine but twelve
   lines did all the work".
5. `max_line`, the largest number of selected crosses sharing one parent --
   the operational number, because that line is a seed parent and has to be
   increased, maintained and, in a CMS system, converted.

**Truncation** is the ceiling on gain and the floor on diversity. It is what
every programme does by default, and the interesting question is not whether it
loses diversity but how *little* index a constrained plan has to give up
relative to it. Its `max_line` here is `r max_line_use(sels$truncation, ctx)`:
one line parents that many of the `r n_sel` selected crosses, against
`r 2 * n_sel / ctx$n_lines` under even use. That is the concentration a cap is
meant to remove, and it is visible without any coancestry matrix.

**Truncation with a per-line cap** is the cheap baseline that deserves more
attention than it gets. It needs no coancestry matrix at all -- only counting:

```{r}
sel_truncation_cap
```

The procedure, stated as a breeder would run it:

1. Rank all `r ctx$N` candidate crosses by index, best first.
2. Walk down the ranking. Admit a cross only if *both* its parents are still
   below the cap.
3. Stop when `n_sel` crosses have been admitted, or when the ranking is
   exhausted -- in which case the returned vector is **shorter than `n_sel`**
   and the cap was infeasible for this set size.
4. Feasibility is arithmetic, not luck: with $L_A$ lines in pool A a cap of $c$
   admits at most $L_A c$ crosses, so the condition is
   $n \le \min(L_A, L_B)\,c$ -- here `r n_sel` against
   `r min(sum(ctx$pool == "A"), sum(ctx$pool == "B")) * (ceiling(2 * n_sel / ctx$n_lines) + 1)`.

```{r}
c(index_truncation = mean_index(sels$truncation, ctx),
  index_capped     = mean_index(sels$trunc_cap, ctx),
  index_retained_pct = 100 * mean_index(sels$trunc_cap, ctx) /
                             mean_index(sels$truncation, ctx),
  Ne_par_truncation = ne_parents(sels$truncation, ctx),
  Ne_par_capped     = ne_parents(sels$trunc_cap, ctx))
```

```{r}
#| label: cap-diagnostics
cap <- ceiling(2 * n_sel / ctx$n_lines) + 1          # the function's default
ord <- order(ctx$index, decreasing = TRUE)
lines_used <- function(s) sum(tabulate(c(ctx$ped$a[s], ctx$ped$b[s]),
                                       ctx$n_lines) > 0)
c(cap                    = cap,
  reached_n_sel          = length(sels[["trunc_cap"]]) == n_sel,
  candidates_scanned     = max(match(sels[["trunc_cap"]], ord)),
  shared_with_truncation = length(intersect(sels[["truncation"]],
                                            sels[["trunc_cap"]])),
  lines_used_truncation  = lines_used(sels[["truncation"]]),
  lines_used_capped      = lines_used(sels[["trunc_cap"]]))
```

The default cap is `ceiling(2 * n_sel / ctx$n_lines) + 1`, one above even use,
which is `r ceiling(2 * n_sel / ctx$n_lines) + 1` here. Filling `r n_sel` slots
under it required scanning
`r max(match(sels[["trunc_cap"]], order(ctx$index, decreasing = TRUE)))` of the
`r ctx$N` ranked candidates, and only
`r length(intersect(sels[["truncation"]], sels[["trunc_cap"]]))` of the
resulting crosses are also in the truncation set: the cap does not trim the
top of the ranking, it reshuffles most of it. The last two rows count the same
reshuffling in lines rather than in crosses, and they are the reason
$N_{e,\text{par}}$ is not a line count: the capped plan touches
`r lines_used(sels[["trunc_cap"]])` distinct parent lines against truncation's
`r lines_used(sels[["truncation"]])`, out of `r ctx$n_lines` available -- to
within one line, the same set size -- while its effective number of parents is
`r fmt(ne_parents(sels[["trunc_cap"]], ctx), 1)` against
`r fmt(ne_parents(sels[["truncation"]], ctx), 1)`. What the cap changes is how
evenly the `r 2 * n_sel` parent slots are spread over the lines that are used,
not how many lines are used at all.

What that buys and what it costs, separately. The cap cuts the diversity loss
from `r fmt(100 * alpha_loss(sels$truncation, ctx, null$gd_ref), 2)`% to
`r fmt(100 * alpha_loss(sels$trunc_cap, ctx, null$gd_ref), 2)`% and raises the
effective number of parents from `r fmt(ne_parents(sels$truncation, ctx), 1)`
to `r fmt(ne_parents(sels$trunc_cap, ctx), 1)`, at
`r fmt(100 * mean_index(sels$trunc_cap, ctx) / mean_index(sels$truncation, ctx), 0)`%
of truncation's index. Whether that is a good trade is a programme decision,
not a statistical one -- but note that the capped plan is *feasible* at a 1%
budget while truncation is not, and it was found by counting alone.

Before reaching for an optimiser, run this. If a per-line cap already meets your
constraint at an acceptable index, you are done. The failure mode at this step
is silent: a returned vector shorter than `n_sel`, which every downstream
metric will happily evaluate on the wrong set size -- `length()` it.

## Greedy, with incremental updates

Adding hybrid $j$ to an existing set changes the submatrix total by
$2 s_j + f_{jj}$, where $s_j$ is the sum of $f_{ij}$ over the already-chosen.
That makes each step $O(N)$ instead of $O(n^2)$, and the whole construction
affordable.

```{r}
#| label: greedy-increment
S <- sel_greedy(ctx, n_sel = 10, w = 1)
j <- setdiff(seq_len(ctx$N), S)[1]
f <- ctx[["f"]]
recomputed  <- sum(f[c(S, j), c(S, j)])                 # O(n^2), from scratch
incremental <- sum(f[S, S]) + 2 * sum(f[S, j]) + f[j, j] # O(n), the update
stopifnot(isTRUE(all.equal(recomputed, incremental)))
c(recomputed = recomputed, incremental = incremental)
```

```{r}
sel_greedy
```

One step $k$, in the order the code performs it:

1. `theta_new <- (total + 2 * s + d) / k^2` is the group coancestry the set
   *would* have, for every candidate at once, with `total` the current
   submatrix sum, `s` the running vector of coancestry sums against the chosen
   set and `d = diag(f)`. The normaliser is $k^2$, the size after the addition.
2. `score <- w * ctx$index - theta_new` -- merit against relatedness.
3. Candidates already chosen are set to `-Inf`; with `max_use` set, so is every
   hybrid of a line that has reached its cap. The cap is therefore a
   **feasibility test on the move**, never a penalty on the result -- the
   partition-matroid treatment of @sec-advanced.
4. The argmax is taken, `total` and `s` are updated in $O(N)$, and the parent
   counts incremented.
5. If no finite score remains, the loop exits early and the returned vector is
   short -- the same check as for the capped truncation.

The score is $w \cdot \text{index} - \theta$. At $w = 0$ this is pure diversity
-- the ceiling on diversity and near-zero gain. As $w \to \infty$ it converges to
truncation. Sweeping $w$ traces a frontier -- but **only if the grid is in the
right units**, and that is the trap at this step.

$w$ is an exchange rate: index standard deviations per unit of group
coancestry. The two quantities it converts between differ by four orders of
magnitude, and that ratio is measurable before any sweep is run:

```{r}
#| label: w-units
S30 <- sel_greedy(ctx, 30, w = 0)
theta_if_added <- (sum(f[S30, S30]) + 2 * colSums(f[S30, ]) + diag(f)) / 31^2
c(index_spread = diff(range(ctx$index)),
  theta_spread = diff(range(theta_if_added)),
  w_crossover  = diff(range(theta_if_added)) / diff(range(ctx$index)))
```

`w_crossover` is the weight at which the two terms of the score have equal
range. Below it the sweep returns the diversity solution; a decade or two above
it, every point collapses onto truncation. A grid of
$w \in \{0.25, 0.5, \dots, 30\}$ -- the obvious one, and the one this chapter
used before it was measured -- sits entirely above the transition and returns
the *same* selection at every point, which makes greedy look an order of
magnitude worse than it is. The grid below spans six decades around the
crossover and is the one `book/scripts/regenerate.R` uses for the benchmark of
@sec-advanced, so the two chapters report the same greedy.

```{r}
#| label: greedy-frontier
#| fig-cap: "The greedy frontier. w = 0 is maximum diversity; large w converges to truncation. The useful range of w spans roughly 1e-5 to 1e-2 on this dataset, the same window the prose and the sweep table below quote: below it the sweep returns the maximum-diversity plan, above it the truncation plan, and every row outside the window is a duplicate of one of those two."
ws <- c(0, 10^seq(-5, 1, length.out = 19))
gf <- do.call(rbind, lapply(ws, function(w) {
  s <- sel_greedy(ctx, n_sel, w = w)
  data.frame(w = w, alpha = alpha_loss(s, ctx, null$gd_ref),
             index = mean_index(s, ctx))
}))
plot(gf$alpha * 100, gf$index, type = "b", pch = 19, col = "#2166ac",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)")
points(alpha_loss(sels$truncation, ctx, null$gd_ref) * 100,
       mean_index(sels$truncation, ctx), pch = 4, cex = 1.5, col = "#b2182b")
text(alpha_loss(sels$truncation, ctx, null$gd_ref) * 100,
     mean_index(sels$truncation, ctx), "truncation", pos = 2, cex = 0.75)
abline(v = 0, lty = 3); grid()
```

```{r}
#| label: greedy-sweep-table
fmt(gf[gf$w <= 0.01, ], 5)
step <- which.max(gf$alpha > 0)          # first grid point that loses diversity
jump <- gf[c(step - 1L, step), ]
fmt(jump, 5)
# The upper edge of the useful window: the smallest w whose plan is already the
# truncation plan, above which every row of the sweep is a duplicate.
flat_from <- min(gf$w[gf$index >= max(gf$index) - 1e-12])
c(flat_from = flat_from, rows_above = sum(gf$w > flat_from))
```

Two readings, and the second is the argument for everything that follows.
First, the whole frontier lives in the three decades between $w = 10^{-5}$ --
where the sweep is still moving, `r fmt(gf$index[gf$w == 0], 2)` at $w = 0$
against `r fmt(gf$index[2], 2)` at the first non-zero grid point -- and
$w \approx 10^{-2}$, by which the plan is within
`r fmt(100 * (1 - gf$index[gf$w <= 0.01][sum(gf$w <= 0.01)] / max(gf$index)), 2)`%
of truncation's index. From $w =$ `r fmt(flat_from, 4)` upward every one of the
remaining `r sum(gf$w > flat_from)` rows is the *identical* selection, and a
sweep whose rows are all the same selection is a sweep in the wrong units, not
a flat frontier. Second, the sweep is
parameterised by $w$ and *the breeder's constraint is not*. Between the two
adjacent grid points printed above, $\alpha$ jumps from
`r fmt(100 * jump$alpha[1], 2)`% to `r fmt(100 * jump$alpha[2], 2)`% and the
index from `r fmt(jump$index[1], 2)` to `r fmt(jump$index[2], 2)`; a budget
falling inside that interval cannot be requested of a greedy sweep at all, only
stumbled upon. An optimiser parameterised by $\alpha_{\max}$ can be asked for
it directly.

Greedy solutions are also what warm-start the differential evolution. That is
not an optimisation detail -- it is the difference between the DE working and
not working.

## The encoding problem

An $N$-dimensional 0/1 encoding is infeasible for differential evolution at this
scale: at production size that is 5041 binary dimensions. Instead, encode
$n_{sel}$ continuous values in $[1, N]$ and decode them to unique indices.

```{r}
decode
```

The decode, step by step:

1. `as.integer(par)` truncates each coordinate toward zero, so the real
   interval $[i, i+1)$ maps to candidate $i$. `sel_de()` searches the box
   $[1,\, N + 0.999]$, so candidates $1$ to $N-1$ each own a full unit
   interval and candidate $N$ owns $[N,\, N + 0.999]$ -- narrower by one part
   in a thousand, because the upper bound stops just short of $N + 1$ instead
   of reaching it.
2. Coordinates outside the box are clamped to the ends.
3. `duplicated()` marks every coordinate after the first that lands on an
   already-used candidate.
4. Each duplicate is replaced by the lowest index not otherwise in use, so the
   result always has exactly `n_sel` distinct candidates.

A worked case, small enough to check by hand:

```{r}
#| label: decode-example
d <- decode(c(3.9, 3.2, 1.1, 7.5), N = 10, n_sel = 4)
stopifnot(all(d == c(3, 2, 1, 7)))
d
```

Truncation gives $(3, 3, 1, 7)$; the second coordinate is the duplicate; the
lowest index not in $\{3, 1, 7\}$ is 2; the repair returns $(3, 2, 1, 7)$. The
same mechanism is what lets a warm start be injected exactly: a seed is written
into the population as `index + 0.5`, the middle of its own interval, and
decodes back to itself.

```{r}
#| label: decode-roundtrip
s0 <- sel_greedy(ctx, n_sel, w = 1)
c(box_lower      = 1,
  box_upper      = ctx$N + 0.999,
  width_typical  = 2 - 1,                       # candidate 1 owns [1, 2)
  width_last     = (ctx$N + 0.999) - ctx$N,     # candidate N owns [N, N+0.999]
  seed_roundtrip = all(decode(s0 + 0.5, ctx$N, n_sel) == s0))
```

Duplicates are repaired **deterministically**. A stochastic repair would make
the fitness noisy, and a noisy fitness stops DE converging -- the same parameter
vector must always score the same.

```{r}
set.seed(3)
p1 <- runif(n_sel, 1, ctx$N + 1)
p2 <- runif(n_sel, 1, ctx$N + 1)
d1 <- decode(p1, ctx$N, n_sel)
d2 <- decode(p2, ctx$N, n_sel)
c(unique_1 = length(unique(d1)), unique_2 = length(unique(d2)),
  in_range = all(c(d1, d2) >= 1 & c(d1, d2) <= ctx$N),
  collisions_repaired = sum(duplicated(as.integer(p1))),
  deterministic = identical(d1, decode(p1, ctx$N, n_sel)))
```

::: {.callout-note}
## A known bias in the repair

The repair fills in the lowest free indices, which introduces a slight bias
toward low-numbered candidates. It is marked in the source with a `ponytail:`
comment naming the ceiling and the upgrade path -- switch to "nearest free" if
the DE ever stalls. It has not mattered at these scales, but it is the kind of
thing that should be written down rather than discovered later.
:::

### A third encoding: one weight per line

@sec-collapse showed that $\theta$ depends only on the line-usage vector. That
makes a third encoding available, and one that changes what the coordinates
*mean*: one weight per **line**, dimension $L$. Against the $N$-dimensional key
vector that is a collapse from 3600 coordinates to 120 in the run below.
Against the $n_{sel}$ coordinates `sel_de()` actually searches it is close to a
wash, and the numbers below say which of those two comparisons is which.

```{r}
decode_lines
```

The line weight shifts each hybrid's score by the weights of its two parents,
and hybrids are taken greedily subject to the per-line usage cap -- so the cap is
satisfied **by construction** rather than by penalty. In detail:

1. Each candidate scores `ctx$index + w_div * (p[a] + p[b])`, with `p` the
   line-weight vector being searched and `w_div = 3` by default. A line's
   weight therefore moves every cross it parents, which is what makes the
   decision a *line* decision.
2. Candidates are taken in decreasing score, skipping any whose parents have
   reached `max_use`.
3. If the cap makes `n_sel` unreachable, the selection is topped up ignoring
   it, and the fitness function's penalty term then sees the violation -- so
   an infeasible cap degrades gracefully instead of returning a short vector.

At `p = 0` the decoder is exactly capped truncation, which is the natural zero
of the encoding; @sec-advanced verifies that identity numerically where the
encoding is actually driven.

The claim this chapter opened with -- that an $N$-dimensional encoding is
infeasible for DE at this scale -- is measured rather than asserted, and the
measurement is cached in `data/de_constraints.csv`. It is worth being exact
about what it compares, because **neither arm of it is `sel_de()`**. Both are
driven through `make_fitness_fast()`, on a pool of $N = 3600$ crosses from
$L = 120$ lines, under the four simultaneous restrictions of @sec-choosing:
diversity loss $\le 0.5\%$, $N_{e,\text{par}} \ge 50$, at most 6 uses per line,
and heterotic-pool balance within 5%. One arm is `decode_rank()` -- one random
key per *hybrid*, dimension 3600, the $N$-dimensional encoding under
indictment. The other is `decode_lines()`, dimension 120.

```{r}
#| echo: false
agg <- do.call(rbind, lapply(split(de, de$encoding), function(d) data.frame(
  encoding     = d$encoding[1],
  dim          = d$dim[1],
  index        = mean(d$index),
  index_min    = min(d$index),
  index_max    = max(d$index),
  `loss %`     = mean(d$loss_pct),
  `Ne parents` = mean(d$Ne_parents),
  `max uses`   = mean(d$max_use),
  seconds      = mean(d$secs),
  check.names  = FALSE)))
knitr::kable(agg, digits = 3, row.names = FALSE,
  caption = "60,000 evaluations per run, four simultaneous restrictions (@sec-choosing), three seeds. index_min and index_max are the seed-to-seed range, not a confidence interval.")
```

```{r}
#| label: encoding-spread
#| echo: false
hk <- de[de$encoding == "hybrid keys / penalty", "index"]
lw <- de[de$encoding == "line weights / repair", "index"]
fmt(c(hybrid_keys_mean = mean(hk), hybrid_keys_range = diff(range(hk)),
      line_weights_mean = mean(lw), line_weights_range = diff(range(lw)),
      ratio_of_means = mean(lw) / mean(hk)), 3)
```

At dimension 3600 the search reaches index `r fmt(mean(hk), 3)`, at dimension
120 it reaches `r fmt(mean(lw), 3)`, and the seed-to-seed range within each arm
(`r fmt(diff(range(hk)), 3)` and `r fmt(diff(range(lw)), 3)`) is an order of
magnitude smaller than the distance between them, so the ordering is not a seed
effect. Penalty and repair constraint handlers returned *numerically identical*
plans at dimension 3600, because at that dimension the search never travels far
enough from its random start to reach the boundary where the two handlers
differ.

Three things that result does **not** say, each of which it has been used to
say:

- It is not a comparison against the encoding this chapter runs. `sel_de()`
  searches $n_{sel}$ coordinates, not $N$; in the book's own head-to-head
  benchmark (@sec-advanced) `sel_de()` finishes *ahead* of the line-encoded
  `sel_de_fast()` at both scales -- `r fmt(bmean("book", "de_hybrid"), 3)`
  against `r fmt(bmean("book", "de_lines"), 3)` at book scale, and
  `r fmt(bmean("production", "de_hybrid"), 3)` against
  `r fmt(bmean("production", "de_lines"), 3)` at production scale. The line
  encoding beats the *N*-dimensional encoding; it does not beat this one.
- It is not a claim that one evaluation is cheaper. It is not, and
  @sec-scaling measures both.
- It is not a clean like-for-like. Four restrictions were active at once and
  the line decoder satisfies one of them -- the per-line cap -- by
  construction, so part of the margin is a penalty the other arm had to pay and
  the line arm did not.

The selected-set size of that cached run was not recorded alongside it, which
is worth naming rather than papering over: the table is reproducible in
direction, not to the digit.

> The encoding decides whether the optimiser works. The constraint handler does
> not. Neither decides which of two *viable* encodings wins -- that has to be
> measured, and @sec-advanced measures it.

## Differential evolution

The run, as a procedure:

1. Fix `n_sel` first, then obtain `gd_ref` from `null_distribution()` **at that
   same size**. A reference computed at another size is a different quantity
   and every $\alpha$ downstream is wrong by a fixed factor.
2. Choose $\alpha_{\max}$ below the attainable ceiling printed at the top of
   this chapter, by the procedure of @sec-ocs.
3. Call `sel_de()`. It builds its warm start, injects it into the initial
   population, and hands `DEoptim` the fitness
   $-(\bar u_S - 1000 \cdot \text{violation})$ to minimise.
4. Check the returned plan before using it: realised $\alpha$ against the
   budget, `max_line`, $N_{e,\text{par}}$, and the gap to the bound. The
   diversity constraint is a **penalty**, so a returned plan is not guaranteed
   feasible -- the violation is only made expensive.

```{r}
sel_de
```

```{r}
alpha_max <- 0.01
set.seed(9)
de_idx <- suppressWarnings(
  sel_de(ctx, n_sel, alpha_max = alpha_max, gd_ref = null$gd_ref,
         NP = 120, itermax = 250))

fmt(c(index      = mean_index(de_idx, ctx),
      alpha_pct  = 100 * alpha_loss(de_idx, ctx, null$gd_ref),
      budget_pct = 100 * alpha_max,
      Ns         = status_number(de_idx, ctx),
      Ne_parents = ne_parents(de_idx, ctx)), 3)
```

The realised loss sits just inside the budget, which is the sign that the
constraint is active and the penalty correctly scaled: with `penalty = 1000`
multiplying the *absolute* violation, one tenth of a percentage point of excess
loss costs a full index standard deviation. A plan that comes back well inside
the budget at a low index usually means the opposite -- that the penalty
dominated and the search never approached the boundary.

### The warm start is not optional

The strong form of this claim was measured at the package's own simulation
scale -- 2500 candidates from 50 + 50 lines, $n = 100$, $4.5 \times 10^5$
evaluations, recorded as finding 7 of `README.md` -- where a cold-started DE
finishes *below* the greedy solutions it was supposed to improve on. That run
is not reproduced here; this chapter runs 625 candidates and
$3 \times 10^4$ evaluations, and the conclusion at this scale is weaker. What
follows is the same comparison, measured rather than asserted.

```{r}
#| label: warm-start
#| fig-cap: "With and without a greedy warm start, at identical evaluation budget, at book scale. The cold start finishes below the best feasible greedy solution; the warm start finishes above the cold start but does not overtake greedy at this budget."
budget <- list(NP = 120, itermax = 250)
set.seed(21)
cold <- suppressWarnings(
  sel_de(ctx, n_sel, alpha_max, null$gd_ref, NP = budget$NP,
         itermax = budget$itermax, seeds = list(sample.int(ctx$N, n_sel))))
# `seeds = NULL`, the default, would have built the greedy sweep for us.

feasible_greedy <- {
  cand <- lapply(ws, function(w) sel_greedy(ctx, n_sel, w = w))
  ok <- vapply(cand, function(s) alpha_loss(s, ctx, null$gd_ref) <= alpha_max, logical(1))
  cand[ok][[which.max(vapply(cand[ok], mean_index, numeric(1), ctx = ctx))]]
}

res <- c(`cold DE` = mean_index(cold, ctx),
         `best feasible greedy` = mean_index(feasible_greedy, ctx),
         `warm DE` = mean_index(de_idx, ctx))
bp <- barplot(res, col = c("grey70", "#2166ac", "#b2182b"), ylab = "mean index (s.d.)",
              ylim = c(0, max(res) * 1.2))
text(bp, res, sprintf("%.3f", res), pos = 3, cex = 0.8)
```

```{r}
fmt(res, 4)
```

One run of each is not a comparison. Three seeds, with the cold and warm arms
branched off the *same* seed so that the two differ only in the initial
population:

```{r}
#| label: warm-vs-cold
reps <- do.call(rbind, lapply(c(9, 21, 33), function(sd_) {
  set.seed(sd_)
  warm <- suppressWarnings(
    sel_de(ctx, n_sel, alpha_max, null$gd_ref, NP = 120, itermax = 250))
  set.seed(sd_)
  cold <- suppressWarnings(
    sel_de(ctx, n_sel, alpha_max, null$gd_ref, NP = 120, itermax = 250,
           seeds = list(sample.int(ctx$N, n_sel))))
  data.frame(seed = sd_, warm = mean_index(warm, ctx),
             cold = mean_index(cold, ctx))
}))
fmt(reps, 4)
fmt(c(warm_mean = mean(reps$warm), warm_sd = sd(reps$warm),
      cold_mean = mean(reps$cold), cold_sd = sd(reps$cold)), 4)
```

Result: over three seeds the warm start averages
`r fmt(mean(reps$warm), 3)` with a standard deviation of
`r fmt(sd(reps$warm), 4)`, the cold start `r fmt(mean(reps$cold), 3)` with
`r fmt(sd(reps$cold), 4)`. Interpretation: the warm start is worth about
`r fmt(100 * (mean(reps$warm) / mean(reps$cold) - 1), 0)`% of mean index at
equal evaluation budget, and rather more importantly it shrinks the run-to-run
spread by an order of magnitude. A single cold run is not reportable: three
cold runs span `r fmt(diff(range(reps$cold)), 3)` index units against
`r fmt(diff(range(reps$warm)), 3)` for three warm ones -- a spread that is
`r fmt(100 * diff(range(reps$cold)) / (mean(reps$warm) - mean(reps$cold)), 0)`%
of the very difference being measured.

What the warm start actually contains at this scale is worth looking at,
because it is weaker than the name suggests:

```{r}
#| label: de-seeds
fit <- make_fitness(ctx, n_sel, alpha_max, null$gd_ref)
seed_w <- c(0, 0.25, 0.5, 1, 2, 4, 8, 16)        # the grid sel_de() hard-codes
seeds <- lapply(seed_w, function(w) sel_greedy(ctx, n_sel, w = w))
fmt(data.frame(
  w         = seed_w,
  index     = vapply(seeds, mean_index, numeric(1), ctx = ctx),
  alpha_pct = 100 * vapply(seeds, alpha_loss, numeric(1), ctx = ctx,
                           gd_ref = null$gd_ref),
  fitness   = vapply(seeds, function(s) fit(s + 0.5), numeric(1))), 4)
```

Seven of the eight seeds are the *same* selection -- truncation -- because the
hard-coded grid lies entirely above the crossover measured earlier, and at a 1%
budget all seven are infeasible, which the penalty converts into a fitness of
about `r fmt(fit(seeds[[2]] + 0.5), 0)` on a scale where a random row scores
near zero. The one feasible seed is the $w = 0$ diversity solution at index
`r fmt(mean_index(seeds[[1]], ctx), 3)`. So at this scale the warm start
installs one feasible plan and seven copies of an infeasible one, and is still
worth the difference measured above. Widening that grid is a change to the
package rather than to the call, and it is not made here: every number in this
book was produced with `sel_de()` as it stands.

The mechanism is dimensional. At roughly 60 integer dimensions the search space
is far too large for 30,000 evaluations to find anything a construction
heuristic has not already found. The warm start does not help the DE converge
faster -- it gives the DE something worth improving on.

`sel_de()` builds the warm start automatically from a greedy sweep, and
**removing it will silently degrade every result in this book.**

The limitation, stated where the result is: at book scale the warm-started DE
(`r fmt(mean_index(de_idx, ctx), 3)`) is still below the best feasible greedy
solution of the finely swept grid (`r fmt(mean_index(feasible_greedy, ctx), 3)`)
at this budget. The DE's advantage here is not raw index but that it is
parameterised by $\alpha_{\max}$, so it can be asked for a budget the greedy
grid does not happen to land on. The method that beats both on index, at a
fraction of the wall-clock, is in @sec-advanced.

### Checking convergence against the bound

@sec-relaxation gives a relaxation. Turning it into a *bound* takes one more
step, derived in @sec-bound: the relaxation at a given $\lambda$ maximises a
penalised objective rather than the index, so the best relaxation point inside
the budget is not an upper bound on the best plan inside the budget. The
Lagrangian combination is.

```{r}
relax <- ocs_relaxation(ctx, n_sel, lambdas = 10^seq(-1, 2.5, length.out = 10))
a_relax <- (null$gd_ref - relax[, "GD"]) / null$gd_ref

bound <- ocs_bound(ctx, n_sel, alpha_max, null$gd_ref)

c(DE_index = mean_index(de_idx, ctx),
  upper_bound = as.numeric(bound),
  gap_pct = 100 * (as.numeric(bound) - mean_index(de_idx, ctx)) / as.numeric(bound))
```

The four steps behind that one call, and the check that has to accompany it:

1. The budget is translated into group coancestry,
   $\theta_{\max} = 1 - \text{gd\_ref}(1 - \alpha_{\max})$ -- the largest
   $\theta$ a plan may have.
2. The relaxation is solved at each $\lambda$ on the grid, giving
   $V(\lambda)$, its optimal penalised value.
3. The bound is $\min_\lambda [V(\lambda) + \lambda\,\theta_{\max}]$. Every
   $\lambda$ gives a valid bound, so a coarse grid costs tightness, not
   validity.
4. `ocs_bound()` doubles its iteration count until the value stops moving and
   records whether it succeeded. **Read that flag before quoting any gap**: an
   under-converged relaxation reports $V(\lambda)$ too small, which makes the
   "bound" too small, and then it is not a bound at all (@sec-scaling).

```{r}
#| label: bound-certificate
c(lambda_star = attr(bound, "lambda"),
  iterations  = attr(bound, "iters"),
  theta_max   = 1 - null$gd_ref * (1 - alpha_max))
attr(bound, "converged")
```

A large gap means **the optimiser has not converged**, not that the problem is
intrinsically hard. A small gap means further optimiser effort is wasted. The
gap here is
`r fmt(100 * (as.numeric(bound) - mean_index(de_idx, ctx)) / as.numeric(bound), 1)`%,
against a bound the function certifies as converged; @sec-advanced closes most
of it at the same budget with a cheaper move, which settles which of the two
readings applies.

::: {.callout-tip}
## If the DE ever looks unconvincing on real data

@debeukelaer2018, optimising core-collection subsets with local search rather
than DE, rank three stochastic searches on the same objective across four crop
datasets -- rice, coconut, maize and pea -- in the Results section
"Optimizing E-NE and A-NE with local searches": parallel tempering first, a
genetic algorithm second, and random descent -- a plain stochastic
hill-climber -- third, with all twelve pairwise comparisons significant under a
Holm-adjusted Wilcoxon test. The genetic algorithm therefore *does* beat the
hill-climber on raw score. What it does not do is win, and the method that wins
is the one assembled from many simple local searches run in parallel, not the
one with the more elaborate update rule.

The inference, stated separately from that result and limited by it: extra
sophistication inside a *single-solution* algorithm bought a rank there and not
the win, which is a weak reason to spend complexity on the optimiser and no
reason at all to expect a DE to be the right one. It is one application, on an
allele-coverage objective rather than an index under a coancestry budget, so it
transfers as a prior about where effort pays and not as a result. That
direction is taken up in @sec-advanced, which builds the local search on the
`O(L)` swap this package already had and measures it against the DE at equal
wall-clock.
:::

## The full frontier

One optimisation per budget, which is the epsilon-constraint method: each point
is a plan that can be put in front of a committee, described by the budget it
respects rather than by an exchange rate nobody can interpret.

```{r}
#| label: full-frontier
#| fig-cap: "Everything together. The constrained DE, the greedy sweep, and the continuous OCS relaxation -- which is a relaxation, not a bound at matched diversity (@sec-bound)."
alphas <- c(0, 0.0025, 0.005, 0.01, 0.02)
set.seed(45)
front <- do.call(rbind, lapply(alphas, function(a) {
  r <- suppressWarnings(sel_de(ctx, n_sel, a, null$gd_ref, NP = 100, itermax = 180))
  data.frame(alpha_max = a, alpha = alpha_loss(r, ctx, null$gd_ref),
             index = mean_index(r, ctx), Ne_par = ne_parents(r, ctx))
}))

plot(front$alpha * 100, front$index, type = "b", pch = 19, lwd = 2, col = "#b2182b",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)",
     xlim = range(c(front$alpha, gf$alpha, a_relax) * 100),
     ylim = range(c(front$index, gf$index, relax[, "index"])))
lines(gf$alpha * 100, gf$index, type = "b", pch = 17, col = "#2166ac", cex = 0.7)
lines(sort(a_relax) * 100, relax[order(a_relax), "index"], type = "b", pch = 1,
      lty = 2, col = "grey55", cex = 0.7)
abline(v = 0, lty = 3)
legend("bottomright", c("constrained DE", "greedy sweep", "continuous OCS relaxation"),
       col = c("#b2182b", "#2166ac", "grey55"), pch = c(19, 17, 1),
       lty = c(1, 1, 2), lwd = c(2, 1, 1), bty = "n", cex = 0.8)
grid()
```

```{r}
fmt(front, 4)
```

```{r}
#| label: frontier-vs-greedy
best_greedy <- vapply(front$alpha_max, function(a) {
  ok <- gf$alpha <= a
  if (any(ok)) max(gf$index[ok]) else NA_real_
}, numeric(1))
fmt(data.frame(alpha_max = front$alpha_max, DE = front$index,
               greedy_sweep = best_greedy,
               DE_minus_greedy = front$index - best_greedy), 4)
```

Three things to read, in this order.

**The constraint does its job.** Realised `alpha` tracks `alpha_max` at every
budget: the optimiser spends the diversity it was allowed and no more. A plan
that lands far inside its budget has not been constrained by it, and whatever
else is limiting it -- the cap, the penalty scale, an unconverged search -- has
to be identified before the frontier is read as a price list.

**The comparison against greedy is budget-dependent.** The DE is ahead at
`r sum(front$index > best_greedy, na.rm = TRUE)` of the
`r length(alphas)` budgets and behind at the rest. Where it is ahead, the
greedy grid simply has no point near that budget -- between two consecutive
values of $w$ the sweep steps over the whole interval, and the best *feasible*
greedy plan is then one from far inside it. Where the grid does land well,
greedy is at least as good for a fraction of the cost. Interpretation: at this
scale the DE buys the ability to hit a requested budget, not a better plan; the
statement that it dominates greedy on index was an artefact of a coarse $w$
grid, and does not survive measurement.

**The dashed curve is a relaxation, not a ceiling.** It sits above the discrete
points over most of the range, which is exactly what makes it tempting to read
its height at a given $\alpha$ as an upper bound on what a plan there could be
worth. It is not one: the best relaxation point whose own $\alpha$ falls inside
the 1% budget reaches `r fmt(max(relax[a_relax <= alpha_max, "index"]), 3)`,
*below* both the DE plan and the best feasible greedy plan at that budget,
because each point maximises a penalised objective rather than the index at its
own diversity. @sec-bound falsifies that construction and gives the Lagrangian
statement that is a ceiling -- `r fmt(as.numeric(bound), 3)` here -- and the
vertical distance to *that* at $\alpha_{\max}$ is what @sec-advanced spends its
effort on. Note that this figure runs one DE per
budget with a fixed seed; the seed-to-seed spread measured above
(`r fmt(sd(reps$warm), 3)` for the warm start) is small against the steps
between budgets here, but it is not zero and it should be quoted whenever two
methods are compared at one budget.

## Which strategy to use

```{r}
#| echo: false
knitr::kable(data.frame(
  Situation = c("The constraint is inactive (ceiling below your limit)",
                "A per-line cap already meets the constraint",
                "One-off analysis, small pool",
                "Production scale, a budget and a per-line cap",
                "A restriction a swap cannot test (Ne floor, pool balance)"),
  Use = c("truncation -- nothing to optimise",
          "`sel_truncation_cap()` -- no matrix needed",
          "`sel_greedy()` swept over w",
          sprintf(paste("`sel_greedy(max_use = cap)` swept over w -- the highest index",
                        "of any *capped* route at both scales of @sec-advanced's benchmark",
                        "(%s at book scale, %s at production), and cheap at both: at book",
                        "scale it and `ocs_round(max_use = cap)` both cost milliseconds",
                        "(%s s and %s s, means of 3 seeds that single-call timing cannot",
                        "separate -- neither is reliably the cheaper), and at production it",
                        "is second cheapest (%s s, behind `ocs_round(max_use = cap)`'s %s s",
                        "in every seed).",
                        "`sel_local_search(max_use = cap, mode = \"tempering\")` -- the only",
                        "capped local-search mode in that benchmark -- comes close without",
                        "matching it: %s thousandths of an index unit short at book scale,",
                        "for roughly %sx the wall-clock, and %s thousandths short at production"),
                  fmt(bmean("book", "greedy_cap"), 3),
                  fmt(bmean("production", "greedy_cap"), 3),
                  fmt(bsec("book", "greedy_cap"), 3),
                  fmt(bsec("book", "ocs_round_cap"), 3),
                  fmt(bsec("production", "greedy_cap"), 3),
                  fmt(bsec("production", "ocs_round_cap"), 3),
                  fmt(1000 * (bmean("book", "greedy_cap") -
                                bmean("book", "tempering_cap")), 1),
                  round(bsec("book", "tempering_cap") / bsec("book", "greedy_cap")),
                  fmt(1000 * (bmean("production", "greedy_cap") -
                                bmean("production", "tempering_cap")), 1)),
          "`sel_de_fast()` -- the only route that accepts them; it trails the capped greedy on index (@sec-advanced)"),
  check.names = FALSE))
```

As an order of operations at the end of a cycle:

1. Build the contract objects and print the instance panel at the top of this
   chapter. Confirm `gd_ref` was computed at the `n_sel` you will actually use.
2. Compute the attainable ceiling from truncation. If it is below your intended
   $\alpha_{\max}$, stop: the constraint is inactive and truncation is the
   answer.
3. Run `sel_truncation_cap()`. It costs nothing and needs no matrix. If it
   meets the budget at an acceptable index, stop.
4. Sweep `sel_greedy()` over a $w$ grid spanning the crossover measured from
   your own data, not a grid of round numbers. Read the frontier and locate the
   knee.
5. Only if a specific budget has to be hit, or several restrictions apply at
   once, run an optimiser -- and warm-start it.
6. Whatever returned the plan, validate it on the second lens and on allelic
   richness before it leaves the analysis (@sec-case-study).

The honest summary is that an optimiser earns its place only in the last row,
where a restriction rules the cheap routes out. In the first four -- including
the capped production case, once the capped benchmark is read on capped methods
only -- something simpler is either exactly as good or good enough, and
@sec-choosing's cost table is the reason to check which row you are in before
writing any optimisation code.
