# Combinatorial selection {#sec-combinatorial}

```{r}
#| include: false
source("_common.R")
ctx <- simulate_data(book_cfg)
n_sel <- 60
null <- null_distribution(ctx, n_sel, B = 400, B_full = 60)
de <- read.csv(book_data("de_constraints.csv"))
```

The problem is: choose $n$ of $N$ hybrids, maximising mean index subject to a
diversity budget. This chapter builds up the strategies that solve it, from the
two that need no optimiser at all to the differential evolution that beats them
-- and explains why the DE only beats them under one specific condition.

## Four baselines before any optimiser

```{r}
sels <- list(
  random     = sel_random(ctx, n_sel),
  truncation = sel_truncation(ctx, n_sel),
  trunc_cap  = sel_truncation_cap(ctx, n_sel),
  greedy_div = sel_greedy(ctx, n_sel, w = 0),
  greedy_w1  = sel_greedy(ctx, n_sel, w = 1),
  greedy_w3  = sel_greedy(ctx, n_sel, w = 3)
)
fmt(t(sapply(sels, function(s) c(
  index      = mean_index(s, ctx),
  alpha_pct  = 100 * alpha_loss(s, ctx, null$gd_ref),
  Ns         = status_number(s, ctx),
  Ne_parents = ne_parents(s, ctx),
  max_line   = max_line_use(s, ctx)))), 3)
```

**Truncation** is the ceiling on gain and the floor on diversity. It is what
every programme does by default, and the interesting question is not whether it
loses diversity but how *little* index a constrained plan has to give up
relative to it.

**Truncation with a per-line cap** is the cheap baseline that deserves more
attention than it gets. It needs no coancestry matrix at all -- only counting --
and it usually captures most of the benefit:

```{r}
sel_truncation_cap
```

```{r}
c(index_truncation = mean_index(sels$truncation, ctx),
  index_capped     = mean_index(sels$trunc_cap, ctx),
  index_retained_pct = 100 * mean_index(sels$trunc_cap, ctx) /
                             mean_index(sels$truncation, ctx),
  Ne_par_truncation = ne_parents(sels$truncation, ctx),
  Ne_par_capped     = ne_parents(sels$trunc_cap, ctx))
```

Before reaching for an optimiser, run this. If a per-line cap already meets your
constraint at an acceptable index, you are done.

## Greedy, with incremental updates

Adding hybrid $j$ to an existing set changes the submatrix total by
$2 s_j + f_{jj}$, where $s_j$ is the sum of $f_{ij}$ over the already-chosen.
That makes each step $O(N)$ instead of $O(n^2)$, and the whole construction
affordable.

```{r}
sel_greedy
```

The score is $w \cdot \text{index} - \theta$. At $w = 0$ this is pure diversity
-- the ceiling on diversity and near-zero gain. As $w \to \infty$ it converges to
truncation. Sweeping $w$ traces a frontier:

```{r}
#| fig-cap: "The greedy frontier. w = 0 is maximum diversity; large w converges to truncation."
ws <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 5, 8, 15, 30)
gf <- do.call(rbind, lapply(ws, function(w) {
  s <- sel_greedy(ctx, n_sel, w = w)
  data.frame(w = w, alpha = alpha_loss(s, ctx, null$gd_ref),
             index = mean_index(s, ctx))
}))
plot(gf$alpha * 100, gf$index, type = "b", pch = 19, col = "#2166ac",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)")
points(alpha_loss(sels$truncation, ctx, null$gd_ref) * 100,
       mean_index(sels$truncation, ctx), pch = 4, cex = 1.5, col = "#b2182b")
text(alpha_loss(sels$truncation, ctx, null$gd_ref) * 100,
     mean_index(sels$truncation, ctx), "truncation", pos = 2, cex = 0.75)
abline(v = 0, lty = 3); grid()
```

Greedy solutions are also what warm-start the differential evolution. That is
not an optimisation detail -- it is the difference between the DE working and
not working.

## The encoding problem

An $N$-dimensional 0/1 encoding is infeasible for differential evolution at this
scale: at production size that is 5041 binary dimensions. Instead, encode
$n_{sel}$ continuous values in $[1, N]$ and decode them to unique indices.

```{r}
decode
```

Duplicates are repaired **deterministically**. A stochastic repair would make
the fitness noisy, and a noisy fitness stops DE converging -- the same parameter
vector must always score the same.

```{r}
set.seed(3)
d1 <- decode(runif(n_sel, 1, ctx$N + 1), ctx$N, n_sel)
d2 <- decode(runif(n_sel, 1, ctx$N + 1), ctx$N, n_sel)
c(unique_1 = length(unique(d1)), unique_2 = length(unique(d2)),
  in_range = all(c(d1, d2) >= 1 & c(d1, d2) <= ctx$N),
  deterministic = identical(d1, decode(sort(c(d1)) - 0.5 + 1, ctx$N, n_sel) * 0L + d1))
```

::: {.callout-note}
## A known bias in the repair

The repair fills in the lowest free indices, which introduces a slight bias
toward low-numbered candidates. It is marked in the source with a `ponytail:`
comment naming the ceiling and the upgrade path -- switch to "nearest free" if
the DE ever stalls. It has not mattered at these scales, but it is the kind of
thing that should be written down rather than discovered later.
:::

### The better encoding

@sec-collapse showed that $\theta$ depends only on the line-usage vector. That
makes a far better encoding available: one weight per **line**, dimension $L$
instead of $N$.

```{r}
decode_lines
```

The line weight shifts each hybrid's score by the weights of its two parents,
and hybrids are taken greedily subject to the per-line usage cap -- so the cap is
satisfied **by construction** rather than by penalty.

The comparison at equal evaluation budget is the sharpest single result in this
book:

```{r}
#| echo: false
agg <- aggregate(cbind(index, loss_pct, Ne_parents, max_use, secs) ~ encoding, de, mean)
names(agg) <- c("encoding", "index", "loss %", "Ne parents", "max uses", "seconds")
knitr::kable(agg, digits = 3, row.names = FALSE,
  caption = "60,000 evaluations, four simultaneous restrictions, mean of 3 seeds.")
```

Dimension 3600 reaches index 0.351. Dimension 120 reaches 1.988 -- **5.7x
better at the same budget**. Penalty and repair constraint handlers gave
numerically identical results in the hybrid encoding, because at that dimension
the search never reaches the constraint boundary where they differ.

> The encoding decides whether the optimiser works. The constraint handler does
> not.

## Differential evolution

```{r}
sel_de
```

```{r}
alpha_max <- 0.01
set.seed(9)
de_idx <- suppressWarnings(
  sel_de(ctx, n_sel, alpha_max = alpha_max, gd_ref = null$gd_ref,
         NP = 120, itermax = 250))

fmt(c(index      = mean_index(de_idx, ctx),
      alpha_pct  = 100 * alpha_loss(de_idx, ctx, null$gd_ref),
      budget_pct = 100 * alpha_max,
      Ns         = status_number(de_idx, ctx),
      Ne_parents = ne_parents(de_idx, ctx)), 3)
```

### The warm start is not optional

Without seeding the initial population with greedy solutions, the DE ends up
*below* a plain greedy -- at any affordable budget.

```{r}
#| fig-cap: "With and without a greedy warm start, at identical evaluation budget. Cold-started DE does not reach the greedy solution it was supposed to improve on."
budget <- list(NP = 120, itermax = 250)
set.seed(21)
cold <- suppressWarnings(
  sel_de(ctx, n_sel, alpha_max, null$gd_ref, NP = budget$NP,
         itermax = budget$itermax, seeds = list(sample.int(ctx$N, n_sel))))
# `seeds = NULL`, the default, would have built the greedy sweep for us.

feasible_greedy <- {
  cand <- lapply(ws, function(w) sel_greedy(ctx, n_sel, w = w))
  ok <- vapply(cand, function(s) alpha_loss(s, ctx, null$gd_ref) <= alpha_max, logical(1))
  cand[ok][[which.max(vapply(cand[ok], mean_index, numeric(1), ctx = ctx))]]
}

res <- c(`cold DE` = mean_index(cold, ctx),
         `best feasible greedy` = mean_index(feasible_greedy, ctx),
         `warm DE` = mean_index(de_idx, ctx))
bp <- barplot(res, col = c("grey70", "#2166ac", "#b2182b"), ylab = "mean index (s.d.)",
              ylim = c(0, max(res) * 1.2))
text(bp, res, sprintf("%.3f", res), pos = 3, cex = 0.8)
```

```{r}
fmt(res, 4)
```

The mechanism is dimensional. At roughly 60 integer dimensions the search space
is far too large for 30,000 evaluations to find anything a construction
heuristic has not already found. The warm start does not help the DE converge
faster -- it gives the DE something worth improving on.

`sel_de()` builds the warm start automatically from a greedy sweep, and
**removing it will silently degrade every result in this book.**

### Checking convergence against the bound

@sec-relaxation gives a valid upper bound. The gap between the DE and that bound
is the convergence diagnostic:

```{r}
relax <- ocs_relaxation(ctx, n_sel, lambdas = 10^seq(-1, 2.5, length.out = 10))
a_relax <- (null$gd_ref - relax[, "GD"]) / null$gd_ref
# the tightest bound at or below our diversity budget
ok <- a_relax <= alpha_max
bound <- if (any(ok)) max(relax[ok, "index"]) else NA_real_

c(DE_index = mean_index(de_idx, ctx),
  upper_bound = bound,
  gap_pct = 100 * (bound - mean_index(de_idx, ctx)) / bound)
```

A large gap means **the optimiser has not converged**, not that the problem is
intrinsically hard. A small gap means further optimiser effort is wasted.

::: {.callout-tip}
## If the DE ever looks unconvincing on real data

@debeukelaer2018, optimising core-collection subsets with local search rather
than DE, found that a basic stochastic hill-climber was already competitive,
that a considerably more complex genetic algorithm did *not* reliably beat it,
and that the method which won -- parallel tempering -- won by combining many
simple local searches rather than by being individually more sophisticated.

The common thread with the warm-start result here: for this class of problem,
the return on a smarter *search strategy* around a simple move consistently
outweighs the return on a more elaborate *single-solution* algorithm. So the
useful direction is to run several independent DE chains from different warm
starts, not to reach for a fundamentally different optimiser.
:::

## The full frontier

```{r}
#| fig-cap: "Everything together. The constrained DE, the greedy sweep, the continuous upper bound, and the fixed reference strategies."
alphas <- c(0, 0.0025, 0.005, 0.01, 0.02)
front <- do.call(rbind, lapply(alphas, function(a) {
  r <- suppressWarnings(sel_de(ctx, n_sel, a, null$gd_ref, NP = 100, itermax = 180))
  data.frame(alpha_max = a, alpha = alpha_loss(r, ctx, null$gd_ref),
             index = mean_index(r, ctx), Ne_par = ne_parents(r, ctx))
}))

plot(front$alpha * 100, front$index, type = "b", pch = 19, lwd = 2, col = "#b2182b",
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)",
     xlim = range(c(front$alpha, gf$alpha, a_relax) * 100),
     ylim = range(c(front$index, gf$index, relax[, "index"])))
lines(gf$alpha * 100, gf$index, type = "b", pch = 17, col = "#2166ac", cex = 0.7)
lines(sort(a_relax) * 100, relax[order(a_relax), "index"], type = "b", pch = 1,
      lty = 2, col = "grey55", cex = 0.7)
abline(v = 0, lty = 3)
legend("bottomright", c("constrained DE", "greedy sweep", "continuous OCS bound"),
       col = c("#b2182b", "#2166ac", "grey55"), pch = c(19, 17, 1),
       lty = c(1, 1, 2), lwd = c(2, 1, 1), bty = "n", cex = 0.8)
grid()
```

```{r}
fmt(front, 4)
```

Two things to read. The DE sits above the greedy sweep at every budget -- that
is what it is for. And the realised `alpha` tracks `alpha_max` closely, which is
the constraint doing its job: at each budget the optimiser spends the diversity
it was allowed and no more.

## Which strategy to use

```{r}
#| echo: false
knitr::kable(data.frame(
  Situation = c("The constraint is inactive (ceiling below your limit)",
                "A per-line cap already meets the constraint",
                "One-off analysis, small pool",
                "Production scale, several simultaneous restrictions"),
  Use = c("truncation -- nothing to optimise",
          "`sel_truncation_cap()` -- no matrix needed",
          "`sel_greedy()` swept over w",
          "`sel_de()` warm-started, on the **line** encoding of @sec-collapse"),
  check.names = FALSE))
```

The honest summary is that the optimiser earns its place only in the last row.
In the first three, something simpler is either exactly as good or good enough,
and @sec-choosing's cost table is the reason to check which row you are in
before writing any optimisation code.
