# What the plant literature establishes {#sec-plant-literature}

```{r}
#| include: false
source("_common.R")
bib <- read.csv(book_data("bibliography_plant.csv"), stringsAsFactors = FALSE)
ani <- read.csv(book_data("bibliography_animal.csv"), stringsAsFactors = FALSE)
mu  <- read.csv(book_data("plant_metric_usage.csv"))
```

The preceding chapters derived a recommendation. This one asks what the applied
literature actually does, where it agrees, and -- more usefully -- where it is
silent.

The corpus behind this chapter is `r nrow(bib)` screened plant-breeding records
(2017-2026), of which `r sum(bib$relevance == 3)` are directly relevant and
`r sum(grepl("maize", bib$crop, ignore.case = TRUE))` concern maize, plus
`r nrow(ani)` animal-breeding records over the same period. @sec-corpora
describes both and how to query them.

## The headline qualification

@sec-choosing recommended economising to a minimal non-redundant set: $\theta$
and $N_{e,\text{par}}$ as constraints, allelic richness at validation. That
recommendation stands on its own evidence. **But it is not what plant breeding
practice does, and the gap is worth stating plainly.**

```{r}
#| echo: false
m <- mu[order(-mu$n_records), c("metric", "n_records", "n_maize", "pct")]
names(m) <- c("metric", "records", "of which maize", "% of 687")
knitr::kable(m, row.names = FALSE,
  caption = "Diversity metrics reported across the 687 relevant plant-breeding records.")
```

Three observations follow directly.

**Applied plant work stacks metrics rather than economising.** A representative
sub-Saharan maize panel of 376 elite inbreds reports PIC = 0.39, gene diversity
= 0.37, MAF = 0.29, Shannon = 6.86, Simpson = 949.09, allelic richness = 787.70
and mean Rogers distance = 0.303 -- all in one analysis, presented as a reporting
package rather than as competing alternatives. Roughly 20% of relevant records
report three or more distinct metric families together.

**The metrics recommended here are among the least used.** Marker coancestry
appears in 6.7% of records, allelic richness in 6.7%, and effective size or
status number in 2.5%. Population structure analysis appears in 44% and $F_{ST}$
in 17%. The applied field is oriented toward *describing* structure, not toward
*constraining* a loss.

**Nothing in the retrieved set computes $\theta$ or $N_s$ on a maize
hybrid-usage vector.** The specific formalism of @sec-collapse is not
established in the retrieved set. That is a gap this work fills rather than a
contradiction -- but it means the approach has **no published plant-breeding
baseline to be validated against**, which is a real limitation and should be
read as one.

::: {.callout-note}
## These are different jobs, not a conflict

For **monitoring and description**, plant practice deliberately reports many
metrics because different audiences need different views. For **constraining an
optimiser**, redundancy is a cost rather than a virtue.

Report the wide package to collaborators and in publications; constrain the
optimiser on the narrow set.
:::

## Why plant practice differs from animal breeding

Three structural facts separate the two literatures, and they explain nearly
every divergence in metric choice.

**Lines are homozygous.** Observed heterozygosity in an inbred panel is a purity
control, not a diversity descriptor -- pearl millet parental lines report
$H_o = 0.031$ against $H_e = 0.334$, and the gap is diagnostic of complete
inbreeding rather than informative about the pool. Animal coancestry theory
assumes a randomly mating heterozygous population; the plant analogue computes
$\theta$ directly on fixed line genotypes, which is exactly the setting where
the line-level collapse applies.

**The product is the F1, not the line.** Diversity is tracked separately for
each parental pool and for cross-pool complementarity -- a structure with no
livestock analogue. GWAS models are now built to partition additive and dominance
effects by *heterotic group of origin*.

**Diversity is managed as two objects, not one.** Within-pool and between-pool
diversity trend in opposite directions under selection and require opposite
management, which no single pool-wide statistic can express -- the argument of
@sec-subdivided, arrived at from the applied side.

### Distance measures, and the absence of a standard

Modified Rogers distance dominates in maize (36 of 83 distance-reporting
records), used chiefly for heterotic assignment and tester choice rather than for
diversity monitoring. A recent methods paper works explicitly to unify similarity
and distance measures for inbred comparison -- the field still treats the choice
as open. Recall from @sec-stages that MRD and coancestry are the same statistic:
$f_{ij} = 1 - \text{MRD}_{ij}^2$.

One result is worth carrying into practice: in 298 African highland maize
inbreds, **91% of line pairs differ at 30-36% of scored alleles, yet only 32% of
pairs have kinship below 0.05.** The authors read the gap as allelic redundancy
from repeated use of narrow source germplasm.

> A pool can look diverse by marker counts while being genealogically shallow.
> That is a direct argument for a coancestry-based constraint over a
> distance-based one.

Other calibration benchmarks: North American ex-PVP dent inbreds resolve into 8
admixture subgroups aligned with company of origin; China summer maize (490
inbreds, 525,141 SNPs) shows 66.05% of pairwise kinship values near zero; sweet
corn shows systematically lower diversity, PIC and MAF than field corn across
514 versus 181 lines.

## Heterotic groups

### Assignment is not a solved problem

GCA-based, SCA-based, combined (HSGCA) and SNP-distance classification are
compared repeatedly with no consistent winner. Several studies report HSGCA as
most consistent with testcross performance; others find SNP-based and
testcross-based classification **disagree**. Notably, SSR markers linked to
yield QTL give **no advantage** over random SSRs for grouping -- against the
intuition that linked markers should be more predictive.

This matters for the pipeline: the line-level collapse assumes pool structure is
given. Group assignment is itself contested, so the pool-balance constraint
inherits whatever uncertainty the assignment method carries. Treat it as an
uncertain input, not a fixed given.

### The empirical trend that justifies the constraint

The single most load-bearing empirical result in this corpus: in a commercial
European dent maize programme, **differentiation between Stiff Stalk and
Non-Stiff Stalk groups increased significantly over time while genetic diversity
within both groups declined.** This is longitudinal commercial data, not
simulation.

It confirms the constraint is doing necessary work rather than solving a
hypothetical problem, and it confirms the two-object framing: what erodes is
within-pool diversity; what must be protected from erosion *and* from excessive
inflation is between-pool divergence. That is exactly the panel `s1_monitor()`
logs in @sec-stages.

Why divergence should be maintained rather than maximised now has a mechanistic
basis: additive and dominance QTL effects for hybrid performance are
**background-specific**, depending on allelic ancestry. Divergence pays through
complementarity at specific loci, not through generic distance -- so pushing it
further does not add hybrid value and risks removing shared favourable alleles.
That is Trap 3 of @sec-stages, with a mechanism attached.

The complication worth knowing: deliberate crossing *across* the heterotic
boundary for germplasm enhancement generated substantial usable variation, with
**10.8% of hybrids from one sub-population exceeding a high-yielding check.** The
boundary is maintained for commercial hybrid-parent pools while being
selectively porous for pre-breeding.

## Selecting within the pool

### The constraint alone is not sufficient

Optimal contribution and optimal cross selection are established across crops.
But a load-bearing qualifier appears repeatedly: diversity depletion in closed
elite programmes is described as **"ineluctable"** at realistic population sizes
even under active management, with bridging populations argued as the necessary
remedy.

> A $\theta$ constraint on the existing pool cannot indefinitely substitute for
> new germplasm. Plan periodic external injections rather than treating the line
> pool as closed.

### Cycle length, and what goes unreported

Rapid cycling delivers large gain-rate benefits: 12-88% relative gain advantage
for oligogenic architectures at 2-3 cycles per year versus one; cycle time cut
from 2-5 years to 1 year in intermediate wheatgrass with an explicit
diversity-preserving sampling step. A maize two-cycle genomewide selection scheme
reports larger gains at equal cost and time but **no diversity metric at all.**

The diversity cost per unit of cycle-time reduction is not established in the
retrieved set. Since shortening cycles is the single most common current
intervention in commercial programmes, that is a conspicuous hole.

Simulation work on hybrid rice reciprocal recurrent selection reports that larger
population sizes give higher gain *and* less depletion of genetic variance -- and
that diversity is managed at the parental-pool level that feeds hybrids, which is
exactly the line-usage level this book constrains.

### What is hybrid-specific, and what is not

This distinction decides which literature transfers. **Every usefulness-criterion
and progeny-variance study retrieved operates upstream, at the inbred-line
development stage** -- cross potential selection, UCPC for multi-parental
crosses, barley validation against realised populations, soybean family mean and
variance prediction, wheat progeny-variance formulas. None operates on F1 hybrids
of fixed inbreds.

That confirms rather than challenges the decision to drop the progeny-variance
term: the term is real and actively predicted *before* fixation, and vanishes
after it. It is why `usefulness_criterion()` belongs at S2 and nowhere else.

One caution that is genuinely hybrid-specific: hybrid **mean** prediction
benefits materially from additive-plus-dominance models, since dominance and SCA
change predicted performance versus additive-only GEBVs. That is a separate axis
from the diversity constraint -- it does not reopen the variance term, but it
means line-usage optimisation on $\theta$ should not be conflated with
mean-performance ranking of the same crosses.

## Genebanks, core collections and introgression

Core-subset construction converges on a shared objective -- maximise retained
diversity at minimum accession count -- but diverges on method: GenoCore
selecting 310 from 3,300+ sorghum accessions; distance-based optimisation on
cassava; rice cores of 247 and 30 from 2,242 accessions; a lettuce core of 268
from 811 capturing 99.4% of total variation.

The quantitatively strongest result: **marker-based core extraction captures
3-78-fold more haplotypic diversity than geography- or environment-based
surrogate sampling** across Arabidopsis, Populus and sorghum panels. If cores are
being assembled, assemble them on markers.

Maize-specific core-collection algorithm choice is **not established in the
retrieved set** -- maize genetic-resource work is oriented toward bridging and
targeted introgression instead:

- DNA-pool genotyping of 156 landraces (2,340 individuals) against 327 elite
  lines spanning the 20th century produced two indicators of a landrace's genomic
  contribution to elite germplasm, used to rank donors rather than introduce
  material blind.
- Haplotype mining in ~1,000 landrace-derived doubled haploids found landrace
  haplotypes **outperforming elite alternatives at the same loci**, stable across
  populations and environments.
- Rapid-cycling genomic selection *within* a landrace population (899 DH lines,
  600K array) improves donor material before it contacts the elite pool.
- For capturing landrace variation, the **doubled-haploid route beats gamete
  capture** -- higher prediction accuracy, and no masking of landrace variation
  behind a capture parent's background.

A $\theta$- or $N_s$-based accounting of introgression cost against a same-size
resampling baseline is not established in the retrieved set.

## The animal-breeding corpus: which matrix to manage

On the animal side there is broad convergence that some form of optimal
contribution or optimal cross selection on a **marker-based** coancestry matrix
is the right optimisation *framework*. There is **no consensus on which specific
genomic matrix** to place inside it, and multiple post-2018 papers report that
the choice changes allele-frequency trajectories and diversity outcomes:

- IBD versus homozygosity- versus drift-based matrices give substantially
  different results unless management is IBD-based (@sec-genomic-era).
- A deviation-based coancestry matrix retained more genetic variability, while
  VanRaden's kept allele frequencies closer to the base -- a genuine trade-off
  between "more heterozygosity" and "less frequency drift".
- A linkage-and-pedigree-informed matrix had the highest correlation with true
  IBD, the most accurate $\Delta F$, and was the only scheme keeping frequency
  drift down while meeting the inbreeding target.
- ROH-based genomic relationship outperformed SNP-based relationship for
  limiting inbreeding accumulation at equal gain, in pigs.

The consensus, such as it is: molecular or IBD-flavoured coancestry outperforms
simple VanRaden-style relationship for diversity management, but the field has
not converged on a single preferred matrix. As @sec-choosing notes, the
line-level collapse holds for **any** symmetric $\mathbf{f}^L$, so this open
question does not destabilise the computational results.

On allelic richness the post-2018 work sharpens rather than contradicts the
recommendation: allelic variation is far more sensitive to bottlenecks than
heterozygosity, there is a linear relationship between census-size reduction and
reduction in expected number of alleles at drift-mutation equilibrium, and
standard $N_e$ corrections do **not** predict allelic loss well. One paper
explicitly recommends expanding the $N_e > 500$ conservation guideline to include
an allelic-variation criterion.

And on effective size, two cautions beyond $N_s = 1/(2\theta)$: under *active*
optimal molecular management, rate-based $N_e$ is numerically pathological --
negative in early generations, with no long-term asymptote -- and $N_e$ however
estimated is informative about heterozygosity loss but not about allelic loss.
Both reinforce reporting only a static status-number descriptor under active
management.

## Optimisation machinery reported in plants

This is where the corpus is thinnest, and it confirms the computational gap.

- **Optimal contribution methods** are applied across crops, but no abstract
  states solver type, runtime, or complexity class.
- **Core-subset search** is greedy or distance-based heuristic; only input and
  output set sizes are reported, never runtime.
- **Evolutionary algorithms** appear mainly *outside* plants. Within plants the
  closest analogue is a black-box numerical optimiser for progeny allocation,
  with no named algorithm, problem size, or convergence statistic.
- Problem sizes are reported as candidate-cross counts (330,078 possible barley
  parent combinations) or collection sizes, never as optimiser performance.

**No retrieved plant study applies differential evolution -- or any metaheuristic
-- to an OCS-type hybrid selection problem, and none reports a runtime or
complexity comparison against quadratic-programming OCS solvers.**

On the animal side, AlphaMate remains the most cited applied tool. The most
directly relevant comparison benchmarked genetic algorithms, differential
evolution, particle swarm and simulated annealing for parent selection in a
583-line wheat dataset: GA gave the best convergence speed and fitness; **DE was
robust but required longer runtime and careful tuning**. That is consistent with
@sec-choosing's finding that the encoding, not the algorithm, is what decides
the outcome.

## What remains unestablished

Stated strictly, from the screened abstracts. These are the honest boundaries of
what this book can claim support for. The distinction that matters here is
between *unestablished in the literature* and *unaddressed in this book*: two of
these items now have a simulated answer in @sec-heterosis and @sec-cycles, which
narrows what the book leaves open without changing what the corpus establishes. A
simulation under stated assumptions is not evidence from an operating programme,
and the items below are still the honest boundary of the second.

- Application of DE or any metaheuristic to a $\theta$-constrained line-usage
  selection problem in maize.
- Runtime or complexity comparison between OCS quadratic solvers and
  metaheuristics for any plant breeding problem.
- Empirical proportional diversity loss measured against a same-size resampling
  baseline in an operating maize hybrid programme. (@sec-cycles simulates the
  multi-cycle consequence and finds a gain crossover; no published maize
  programme reports the measurement itself.)
- Longitudinal joint reporting of allelic richness alongside $\theta$ or $N_s$
  for the same hybrid maize programme.
- Hybrid-genome-level (rather than parental-pool-level) coancestry optimisation
  for F1 maize.
- Whether a dominance-informed term would change optimal-cross outcomes relative
  to the mean-only approach. (@sec-heterosis derives the term and shows expected
  heterosis is invariant to $F_{ST}$ under neutral divergence, so a
  dominance-informed ranking must differ from a frequency-based one. Whether it
  changes *realised* outcomes in a programme is untested either way.)
- Quantitative diversity cost per unit of cycle-time reduction under
  rapid-cycling hybrid maize.
- ROH-based diversity metrics applied to maize hybrid pools (22 records use ROH,
  only 5 in maize, none for hybrid diversity management).
- Per-evaluation computational cost of diversity metrics, as a function of
  population size or metric type, anywhere in either corpus.
- Diversity metrics defined on hybrids rather than on lines or parents.
- Combinatorial equal-contribution (0/1) selection compared directly against
  continuous-contribution designs.

## Consequences for the pipeline

1. Keep the narrow metric set inside the optimiser and report the wide package
   outside it.
2. Plan for periodic germplasm injection. A closed elite pool depletes regardless
   of how well the within-pool constraint is managed.
3. Prefer bridging over direct introduction for exotic material, and rank donors
   by their genomic contribution to the elite pool rather than by raw diversity.
4. Treat heterotic-group assignment as an uncertain input -- the pool-balance
   constraint is only as good as the assignment behind it.
5. Constrain between-pool divergence within a band rather than maximising it.
6. Do not conflate the diversity constraint with mean prediction. Hybrid means
   need additive-plus-dominance models; the diversity constraint needs neither.
7. Report the diversity cost per cycle explicitly when shortening cycles.
8. If assembling a core or donor set, do it on markers rather than passport data.
9. Watch for allelic redundancy: high marker-based distance can coexist with
   shallow genealogy, so a coancestry constraint is safer than a distance-based
   one.
10. Validate on real genotypes before treating the redundancy result as settled.
