# The limit, and what the metrics cannot see {#sec-limits}

```{r}
#| include: false
source("_common.R")
traj <- read.csv(book_data("cycles_trajectory.csv"))
vr   <- read.csv(book_data("limits_variance.csv"))
lk   <- read.csv(book_data("limits_linked.csv"))
```

@sec-cycles ran the programme forward and found the crossover: a constrained
programme trails an unconstrained one for about eight cycles and leads
thereafter. That is a measurement. This chapter asks the two questions a
measurement leaves open -- could it have been *predicted*, and is the quantity
being managed the quantity that actually matters.

The answers are yes-with-a-condition, and no-not-entirely.

## Robertson's limit

@sec-response wrote down the response to one cycle of selection. Robertson
[-@robertson1960] wrote down the sum of all of them. Selection fixes alleles; a
finite population fixes some of the favourable ones in the wrong direction; and
the total advance before everything is fixed is approximately

$$R_{\text{total}} \;\approx\; 2 N_e R_1$$

with a half-life of about $1.4 N_e$ generations. The result is the theoretical
ancestor of every optimal-contribution scheme in this book: it is *why* $N_e$ is
the thing to manage, and it says exactly what a unit of $N_e$ is worth in units
of cumulative gain.

Both inputs are already in @sec-cycles' trajectory. $R_1$ is the first cycle's
response, and $N_e$ comes from the rate of inbreeding, exactly as @sec-cycles
computed it:

```{r}
rob <- do.call(rbind, lapply(unique(traj$budget), function(b) {
  d  <- aggregate(cbind(index, F_drift) ~ cycle, traj[traj$budget == b, ], mean)
  f  <- lm(log(1 - F_drift) ~ cycle, d[-1, ])
  Ne <- 1 / (-2 * coef(f)[["cycle"]])
  R1 <- d$index[2] - d$index[1]
  data.frame(budget = b, R2 = summary(f)$r.squared, Ne = Ne, R1 = R1,
             limit = 2 * Ne * R1, half_life = 1.4 * Ne,
             observed_15 = d$index[16] - d$index[1],
             predicted_15 = 2 * Ne * R1 * (1 - (1 - 1 / (2 * Ne))^15))
}))
fmt(rob, 3)
```

Read the last two columns against each other, and then read both against the
first.

## The $R^2$ column decides whether the theory applies

For the tightest budget the prediction is
`r fmt(rob$predicted_15[rob$budget == "0.005"], 2)` and the observation is
`r fmt(rob$observed_15[rob$budget == "0.005"], 2)`. Robertson's theory, published
in 1960 and derived for a single locus under idealised drift, predicts fifteen
cycles of a genomic hybrid programme to better than one percent.

For the unconstrained programme it predicts
`r fmt(rob$predicted_15[rob$budget == "unconstrained"], 2)` against an observed
`r fmt(rob$observed_15[rob$budget == "unconstrained"], 2)` -- too high by about
half.

The theory did not fail arbitrarily, and @sec-cycles already said why. Robertson
assumes a *constant* rate of inbreeding, and the regression of $\log(1-F)$ on
cycle is where that assumption is testable. Its $R^2$ is
`r fmt(rob$R2[rob$budget == "0.005"], 3)` for the tightest budget and
`r fmt(rob$R2[rob$budget == "unconstrained"], 3)` for the unconstrained run,
whose drift decelerates because there is progressively less left to lose. That
low $R^2$ is not a nuisance. It is the theory declining to apply, in advance,
and in a column you can read before you trust the number.

::: {.callout-important}
## Do not quote a Robertson limit without its $R^2$

$N_e$ here is a *fitted* quantity, and @sec-cycles showed the fit is meaningless
for a saturating series -- its fitted $\Delta F$ comes out lowest for the run
that is actually losing diversity fastest, which is backwards. Feeding that
$N_e$ into $2 N_e R_1$ propagates the error into a number that looks like a
long-term plan.

Report the $R^2$ next to every limit, and treat a limit computed from a fit
below about 0.99 as an illustration rather than a projection.
:::

That yields an argument for the constraint that this book did not previously
have. @sec-cycles justified $\alpha$ by where the trajectories cross. The table
above adds: the constrained programme is the one whose future can be *computed*.
An unconstrained programme is not merely worse after cycle 8, it is
unforecastable, because the theory that would forecast it needs a constant
$\Delta F$ that it does not have.

## What the metrics cannot see

Now the second question. Every diversity metric in this book -- $\theta$, $GD$,
$N_s$, allelic richness, `F_hom`, `F_drift` -- is a function of allele
frequencies **one locus at a time**. @sec-response showed that additive variance
is not:

$$\operatorname{Var}(\mathbf{X}\boldsymbol\beta) \;=\; V_{\text{genic}}
\;+\; 2\sum_{k<l}\beta_k\beta_l\operatorname{Cov}(x_k, x_l)$$

Directional selection drives that covariance term negative (Bulmer
[-@bulmer1971]): it couples loci with like effects in repulsion, hiding genic
variance from the next generation without moving a single allele frequency. No
metric in the catalogue can detect it.

@sec-cycles assumed the effect away, in assumption **C1**: *loci are unlinked, so
no linkage disequilibrium builds up and there is no Bulmer effect on the genic
side.* That is worth testing rather than assuming, because Bulmer's 1971
treatment covers unlinked loci explicitly -- free recombination halves the
disequilibrium each generation, but selection regenerates it every generation,
and the equilibrium is not zero.

### The experiment

The gap $\operatorname{Var}(\mathbf{X}\boldsymbol\beta) - V_{\text{genic}}$ is
**not** on its own a measure of the Bulmer effect. A finite founder base leaves
disequilibrium in the very first cycle, and two divergent pools add more. So the
comparison is against a control: an arm identical in every respect -- same
founders, same seeds, same recycling, same population sizes -- in which the
hybrids are chosen at random instead of on the index.

```{r}
gapc <- aggregate(gap ~ regime + cycle, vr, mean)
w <- reshape(gapc, idvar = "cycle", timevar = "regime", direction = "wide")
names(w) <- sub("gap.", "", names(w))
w$selection_effect <- w$unconstrained - w$random
fmt(head(w[, c("cycle", "random", "unconstrained", "constrained",
               "selection_effect")], 7), 4)
```

Cycle 0 is identical across all three arms, to the digit. Nothing has been
selected yet, so the control is doing its job.

From cycle 1 the arms separate. Paired within seed and cycle:

```{r}
p  <- reshape(aggregate(gap ~ regime + seed + cycle, vr, mean),
              idvar = c("seed", "cycle"), timevar = "regime", direction = "wide")
p  <- p[p$cycle > 0, ]
dd <- p$gap.unconstrained - p$gap.random
c(mean_effect = mean(dd), se = sd(dd) / sqrt(length(dd)),
  t = mean(dd) / (sd(dd) / sqrt(length(dd))), n_pairs = length(dd))
```

::: {.callout-warning}
## Assumption C1 of @sec-cycles is too strong, and is restated here

Selection builds negative disequilibrium among **unlinked** loci, at
$t = `r fmt(mean(dd) / (sd(dd) / sqrt(length(dd))), 2)`$ over
`r length(dd)` paired observations. Unlinked loci do not prevent the Bulmer
effect; they make it decay fast.

What C1 should say is that with free recombination the disequilibrium is halved
every generation, so it reaches a low equilibrium and does not accumulate across
cycles the way frequency change does. That is what makes @sec-cycles' *frequency*
trajectories trustworthy, and it is the correct reason -- not the absence of the
effect.
:::

The effect is largest in the early cycles and decays toward zero late, which is
the expected shape: it is proportional to the genic variance still available to
be hidden, and by cycle 12 there is much less of it.

### With real linkage

The same question with real meiosis behind it. @sec-alphasimr's reciprocal
recurrent selection runs on a coalescent founder history with a genetic map, and
AlphaSimR reports `varA` (which carries the disequilibrium) and `genicVarA`
(which does not), so this arm needs no variance code of its own.

```{r}
lk$rel <- ((lk$varA_A - lk$genicVarA_A) + (lk$varA_B - lk$genicVarA_B)) /
          (lk$genicVarA_A + lk$genicVarA_B)
pl  <- reshape(lk[, c("regime", "seed", "cycle", "rel")],
               idvar = c("seed", "cycle"), timevar = "regime", direction = "wide")
pl  <- pl[pl$cycle > 0, ]
ddl <- pl$rel.selected - pl$rel.random
c(mean_effect = mean(ddl), se = sd(ddl) / sqrt(length(ddl)),
  t = mean(ddl) / (sd(ddl) / sqrt(length(ddl))), n_pairs = length(ddl))
```

Same sign, larger in magnitude, and measured with fewer replicates so the
evidence is weaker. Linkage does what linkage should: it slows the decay, so the
disequilibrium selection creates persists longer than it does under free
recombination.

::: {.callout-note}
## A reproducibility trap in the linked arm

`alphasimr_founder_pools()` is **not** reproducible from `cfg$seed`.
`AlphaSimR::runMacs2()` hands the coalescent to MaCS, which seeds itself outside
R's RNG, so two calls with the same configuration return different founder
haplotypes. Everything downstream of the founders is seeded normally.

The consequence for an experiment like this one is severe: founding the pools
separately for each arm would have compared two different base populations and
charged the difference to selection. `book/scripts/regenerate.R` founds once per
replicate and branches both arms off that single result, which is why cycle 0 of
the table above matches exactly.
:::

## The two costs of selection

The chapter's result stated once.

Selection pays for gain twice. It builds negative covariance between loci, which
is a real loss of usable variance, and **recombination repays it** -- halved
every generation under free recombination, more slowly under linkage. And it
changes allele frequencies, moving them toward fixation, and **nothing repays
that** short of mutation or migration.

The first cost is invisible to every metric in this book and is largely
self-correcting. The second is what every metric in this book measures, and is
permanent.

That is the justification for constraining the second and ignoring the first,
and it is a stronger statement than "the constraint is what we can compute". It
is worth being explicit that this ordering is a *consequence* of the two decay
rates, not an assumption: if the disequilibrium did not decay -- tight linkage,
few chromosomes, short cycle times -- the ranking would change, and a constraint
written on allele frequencies would be managing the smaller of the two bills.

## Where this connects

@sec-response supplies the variance decomposition and the breeder's equation
this chapter runs forward. @sec-cycles supplies the trajectories, and has one
assumption restated as a result of the test above. @sec-alphasimr supplies the
linked arm, which is the only place in the book where a genetic map, rather than
an assumption about one, decides how fast disequilibrium decays.

@sec-case-study's checklist gains one line from this chapter: report the $R^2$ of
the $\log(1-F)$ regression next to any projection built on $N_e$, and treat a
projection from a poor fit as an illustration.
