# End to end, and the operational checklist {#sec-case-study}

```{r}
#| include: false
source("_common.R")
```

Everything the book has built, run once, on one dataset, in the order a
programme would run it. Then the checklist.

## The whole pipeline in one script

```{r}
st1 <- stage1_simulate(book_cfg)
c(hybrids = nrow(st1$X), markers = ncol(st1$X),
  f_min = min(st1$f), f_max = max(st1$f))
```

```{r}
#| results: hide
n_sel <- round(0.10 * nrow(st1$X))
res <- suppressWarnings(stage2_select(
  X = st1$X, f = st1$f, hybrids = st1$hybrids,
  n_sel   = n_sel,
  alphas  = NULL,     # NULL = automatic grid, 0 to the attainable ceiling
  weights = NULL,     # NULL = equal weight across traits
  fL      = st1$fL,   # optional: enables theta_A / theta_B / theta_AB
  B_null = 500, B_full = 80, NP = 100, itermax = 200, verbose = FALSE))
```

::: {.callout-note}
## The same pipeline as a script

This chapter runs at the book's reduced scale so it renders in seconds. The
identical pipeline at the **production** scale of `sim_config`, writing its
tables to `report/`, is `inst/scripts/run_all.R`:

```sh
Rscript inst/scripts/run_all.R          # ~4 min
Rscript inst/scripts/run_benchmark.R    # ~6 min; the numbers quoted in README.md
```
:::

With real data only the first line changes:

```{r}
#| eval: false
st1 <- stage1_build(X = my_hybrid_matrix, ped = my_pedigree,
                    traits = my_predicted_traits, fL = my_line_coancestry)
```

### The reference values

```{r}
unlist(res$ref)
```

`alpha_max` is the attainable ceiling from @sec-baseline: the loss pure
truncation would incur. Any requested budget above it constrains nothing, and
`stage2_select()` says so rather than returning a solution that merely looks
constrained.

### Metrics per scenario

```{r}
fmt(res$metrics[, c("scenario", "alpha", "index", "Ns", "Ne_parents",
                    "Ne_lines_A", "Ne_lines_B", "max_line", "alleles_lost")], 4)
```

Read the `index` column against `alpha`. That is the price list: what each
percentage point of diversity costs in index units, on this dataset, at this
selection intensity.

Now read `Ne_parents` against `Ns` in the same rows. This is the diagnostic
@sec-genomic-era called for, and the reason `Ne_parents` is a second constraint
rather than a report line:

```{r}
d <- res$metrics
fmt(data.frame(scenario = d$scenario,
               Ns = d$Ns,
               Ne_parents = d$Ne_parents,
               Ns_relative = d$Ns / max(d$Ns),
               Ne_par_relative = d$Ne_parents / max(d$Ne_parents)), 3)
```

The status number barely moves across scenarios while the effective number of
parents collapses. A programme reporting only $N_s$ would see a healthy pool and
a shrinking parent base it never measured.

### Both lenses, and the pool partition

```{r}
fmt(res$metrics[, c("scenario", "alpha", "F_hom", "F_drift", "cov_diag",
                    "rare_retained", "GD_WI_hyb", "GD_BS", "F_ST")], 4)
```

`alpha` constrains the first lens only. `F_drift` is the unmonitored bill, and
`cov_diag` is exactly the gap between them -- verify it:

```{r}
c(max_deviation_from_Eq3 =
    max(abs(d$cov_diag - (d$F_hom - d$F_drift)), na.rm = TRUE))
```

`GD_BS` rising across scenarios is selection concentrating on the most
complementary line pairs, which pushes the pools further apart. That is not
automatically bad -- it is heterosis -- but @sec-stages is the reason to watch it
rather than celebrate it.

### Distance from the random null

```{r}
fmt(res$z_scores[, c("scenario", "z_GD", "z_Ne_parents", "z_F_drift",
                     "z_rare_retained", "z_GD_BS", "z_ANE")], 2)
```

Every metric in standard deviations from a random selection of the same size. A
metric near zero is not distinguishing this plan from chance, and cannot be
serving as a constraint.

### The selection table

```{r}
scn <- setdiff(names(res$selection), c(names(st1$hybrids), "n_scenarios"))
head(res$selection[order(-res$selection$n_scenarios),
                   c("hybrid", "line_A", "line_B", scn, "n_scenarios")], 8)
```

```{r}
de_cols <- grep("^DE_", scn, value = TRUE)
c(scenarios          = length(scn),
  chosen_in_all_DE   = sum(rowSums(res$selection[, de_cols, drop = FALSE]) ==
                             length(de_cols)),
  never_chosen       = sum(res$selection$n_scenarios == 0),
  total_candidates   = nrow(res$selection))
```

`n_scenarios` is the most directly useful column in the whole output. Hybrids
chosen under **every** diversity budget are robust to the choice of $\alpha$ --
they are the plan's backbone, and the argument for them does not depend on a
policy number nobody wants to defend. Hybrids that appear only at the loosest
budget are the ones being bought with diversity.

## Reading the frontier

```{r}
#| fig-cap: "The price list. Each point is one diversity budget; the curve is what index it buys."
plot(d$alpha * 100, d$index, type = "b", pch = 19, col = "#b2182b", lwd = 2,
     xlab = "gene diversity loss, alpha (%)", ylab = "mean index (s.d.)")
text(d$alpha * 100, d$index, d$scenario, pos = 4, cex = 0.65, col = "grey30")
abline(v = 0, lty = 3)
grid()
```

Choose the knee, not the endpoint. Past the knee, each additional unit of index
costs disproportionately more diversity, and the diversity is not recoverable
next cycle while the index gain is realised once.

## What this simulation cannot tell you

Stated plainly, because the numbers above are precise and precision invites more
confidence than the model supports.

**The GEBVs carry no prediction error.** The traits are built from *known* marker
effects (@sec-simulating-crosses). @sec-choosing Finding 5 measures what happens
when they do not: the reported index is biased upward, and error *correlated with
relatedness* costs several times the diversity that independent error of the same
size does. The plans below should be read as an optimistic ceiling.

**The model is additive.** An additive $\mathbf{G}$ misses heterosis and specific
combining ability, which is precisely what a hybrid programme sells. The
diversity constraint does not need a dominance model; the *mean-performance
ranking* does (@sec-heterosis). Do not conflate them, and do not use these `index`
values as performance predictions.

**The pools are symmetric by construction.** Both are simulated from the same
ancestral frequencies at the same $F_{ST}$. Real heterotic pools differ in size,
history and internal structure, and a real programme will find one pool eroding
faster than the other -- which is why `ne_lines_pool()` is reported per pool.

**$N_s$ is not a drift projection.** It is a static descriptor
(@sec-ns). Do not extrapolate it forward.

**One cycle, not a trajectory.** Every result here is a single round of hybrid
selection. @sec-stages showed that a constraint applied at one stage buys
stage-level diversity, not end-of-pipeline richness -- and the same logic applies
across cycles. A one-cycle constraint is not a multi-cycle policy: @sec-cycles
runs fifteen of them and finds the unconstrained programme ahead for eight cycles
and behind for the rest.

**$p_0$ must be frozen.** In this book it is set once in `build_ctx()`. In a real
multi-cycle programme it must be the founder cycle's frequencies, propagated
forward -- the single most common error in applying `F_hom` and `F_drift`
(@sec-genomic-era). @sec-cycles shows the failure directly: re-anchored, five
generations of real drift are reported as exactly none.

## The operational checklist

### Setting up

- The base $p_0$ is frozen at the founding cycle and documented.
- Diversity is computed on molecular coancestry `f`, never on a centered `G`.
  Check `sum(G) == 0` once, to see why.
- The marker panel's MAF spectrum is known -- it decides how severely the two
  lenses diverge.
- Heterotic-group assignment is recorded, with the method that produced it, and
  treated as uncertain.

### Choosing the budget

- The attainable $\alpha$ ceiling is computed *before* choosing $\alpha_{\max}$.
  A budget above the ceiling constrains nothing.
- Loss is measured against a **same-size resampling baseline**, never against the
  full pool.
- The constraint is applied at the target set size, not incrementally.
- The frontier is traced and the knee chosen, rather than a round number.

### Running the optimiser

- The search is encoded on **line weights**, not hybrid keys.
- The differential evolution is **warm-started** from greedy solutions.
- $N_{e,\text{par}}$ is a second constraint, not a report line.
- The per-line usage cap is enforced in the decoder, not by penalty.
- The result is checked against the continuous relaxation. A large gap means the
  optimiser has not converged; a slack constraint means the search never reached
  the boundary.

### Validating the plan

- `F_hom` and `F_drift` are computed **separately** and reported together, with
  `cov_diag` as the diagnostic.
- Allelic richness is computed on the returned plan -- it is the one metric
  carrying information $\theta$ does not.
- The within/between pool decomposition confirms diversity was not preserved by
  degrading pool separation.
- The number of selected parent lines is biologically plausible.
- $N_s$ is reported; a rate-based $N_e$ is **not**, under active management.
- Every metric is reported against the random null, in standard deviations.

### Across cycles

- The per-stage loss ledger is kept. DH induction, not hybrid mining, is usually
  where the diversity goes.
- The regression of $\log(1 - F)$ on cycle is checked for linearity. Curvature
  means a non-constant rate.
- Periodic germplasm injection is planned. A closed elite pool depletes however
  well the constraint is managed.
- The diversity cost per cycle is reported when cycle time is shortened.

## Where to go next

Three extensions the material here supports but does not cover, each needing new
work rather than new derivation:

**Real data ingestion.** `load_data()` is the single swap point and is currently
a stub. A worked path from a VCF to the `X`/`f`/`hybrids` contract is what most
readers will need first, and it is the one extension here that is pure
engineering rather than new derivation.

**Estimating the dominance deviations.** @sec-heterosis shows that expected
heterosis depends on the alignment between divergence and $d$, and that no
frequency-based statistic can see $d$. Everything downstream of that -- ranking
crosses on more than the mean, putting a number on the $\lambda$ of
@sec-catalogue -- waits on an estimate of $d$ the book does not attempt.

**Prediction error compounded across cycles.** @sec-choosing Finding 5 measures it
within one cycle and @sec-cycles assumes it away (C7). The two have not been run
together, and the interaction is not obviously benign: correlated error
concentrates the selection, and @sec-cycles shows concentration is what erodes the
programme's future.
