# Notation and glossary {#sec-notation}

```{r}
#| include: false
source("_common.R")
```

## Symbols

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$N$", "$n$, $n_{sel}$", "$L$", "$m$", "$x_{ik}$",
             "$\\mathbf{X}$", "$\\mathbf{M}$", "$\\mathbf{G}^L$",
             "$p_k$, $\\bar p_k$", "$p_{0,k}$"),
  Meaning = c("candidate hybrids in the pool",
              "hybrids selected",
              "parent lines",
              "markers",
              "within-individual reference-allele frequency of individual i at marker k, in 0/0.5/1",
              "hybrid marker matrix, N x m",
              "hybrid dosage matrix, 0/1/2",
              "line genotype matrix, L x m, coded 0/1",
              "allele frequency; the bar denotes a group mean",
              "**frozen** candidate-pool allele frequency: the base"),
  `In the code` = c("`ctx$N`", "`n_sel`", "`ctx$n_lines`", "`ctx$m`", "`ctx$X[i, k]`",
                    "`ctx$X`", "-", "`ctx$GL`", "-", "`ctx$p0`"),
  check.names = FALSE))
```

### Matrices

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$\\mathbf{f}$", "$\\mathbf{f}^L$", "$\\mathbf{G}$", "$\\mathbf{A}$",
             "$\\mathbf{G}_{0.5}$"),
  Meaning = c("molecular coancestry of the **hybrids**, N x N, in [0, 1]",
              "molecular coancestry **between lines**, L x L",
              "VanRaden genomic relationship, centered -- prediction only",
              "pedigree relationship",
              "VanRaden with the centering frequency fixed at 0.5"),
  Note = c("the diversity kernel; never centered",
           "the sufficient statistic in disguise -- see the collapse",
           "sum of all elements is exactly 0; unusable for diversity",
           "expectation of IBD, no markers",
           "**this is f**, arrived at from the VanRaden family"),
  check.names = FALSE))
```

### Coancestry and diversity

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$\\theta_S$", "$GD$, $H_e$", "$N_s$", "$N_{e,\\text{par}}$",
             "$f_{ii}$", "$s_i$", "$\\tilde x$", "$\\bar x$"),
  Definition = c("$n^{-2}\\sum_i\\sum_j f_{ij}$, diagonal **included**",
                 "$1 - \\theta$; these are the same quantity",
                 "$1/(2\\theta)$",
                 "$1/\\sum_i p_i^2$ over parent-line usage",
                 "coancestry within subpopulation i; $1 - f_{ii} = H_{e,i}$",
                 "self-coancestry; $F_i = 2s_i - 1$",
                 "weighted mean **within** subpopulations",
                 "mean over **all pairs**, between-subpopulation pairs included"),
  `Function` = c("`theta_group()`, `fast_theta()`", "`gene_diversity()`, `he_nei()`",
                 "`status_number()`", "`ne_parents()`", "`ct_partition()`",
                 "`ct_partition()`", "-", "-"),
  check.names = FALSE))
```

Always $\bar f \le \tilde f \le \tilde s$. The differences between those three
numbers *are* the partition of diversity.

### The partition

```{r}
#| echo: false
knitr::kable(data.frame(
  Component = c("$GD_T$", "$GD_{WI}$", "$GD_{BI}$", "$GD_{WS}$", "$GD_{BS}$"),
  Definition = c("$1 - \\bar f$", "$1 - \\tilde s$", "$\\tilde s - \\tilde f$",
                 "$1 - \\tilde f$", "$\\tilde f - \\bar f$"),
  Meaning = c("total", "within individuals", "between individuals, within subpopulation",
              "within subpopulations", "between subpopulations"),
  `In hybrid breeding` = c("the pool's diversity",
                           "0 for inbred lines; use `GD_WI_hyb` for the F1s",
                           "segregating variation available to selection",
                           "what erodes -- defend it",
                           "the heterosis engine -- **maintain**, do not maximise (@sec-heterosis)"),
  check.names = FALSE))
```

with $GD_T = GD_{WI} + GD_{BI} + GD_{BS}$ and $F_{ST} = GD_{BS}/GD_T$.

### Heterosis and cycles

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$d_k$", "$H_i$", "$\\bar p_k$", "$y_k$",
             "$g^A_i$, $g^B_j$", "$s_{ij}$", "$\\sigma^2_{SCA}$",
             "$\\lambda$", "$t$", "$p_0^{(0)}$"),
  Meaning = c("dominance deviation at locus $k$",
              "mid-parent heterosis of hybrid $i$; $\\sum_k d_k$ over its heterozygous loci",
              "$(p_{Ak} + p_{Bk})/2$, the merged-pool frequency",
              "$p_{Ak} - p_{Bk}$, the divergence at locus $k$",
              "general combining ability for heterosis; weighted by the **opposite** pool's frequencies",
              "specific combining ability; $-2\\sum_k d_k\\,\\delta u_{ik}\\,\\delta v_{jk}$",
              "$4\\sum_k d_k^2 p_{Ak}q_{Ak}p_{Bk}q_{Bk}$; falls as $(1 - F_{ST})^2$",
              "weight on within- against between-pool diversity",
              "cycle index",
              "base frequencies, frozen at the founding cycle"),
  `Function` = c("argument to `heterosis_value()`", "`heterosis_value()`",
                 "inside `heterosis_partition()`", "`pool_complementarity()$mean_sq_delta_p` is $\\overline{y^2}$",
                 "`heterosis_gca_sca()$gca_A`", "`heterosis_gca_sca()$sca`",
                 "closed form; see @sec-heterosis-relocation",
                 "not implemented; see @sec-catalogue", "`run_cycles()`",
                 "`ctx$p0`, set once"),
  check.names = FALSE))
```

Expected heterosis splits as $\sum_k d_k[\,2\bar p_k(1-\bar p_k) + y_k^2/2\,]$,
whose two halves are $GD_T$ and $GD_{BS}$ when $d_k \equiv 1$, and whose sum is
$1 - \theta_{AB}$ (@sec-heterosis). Over a complete factorial the realised values
split orthogonally as $H_{ij} = \mu + g^A_i + g^B_j + s_{ij}$.

### Inbreeding, the three lenses

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$F_{hom}$", "$F_{drift}$", "$F_{IBD}$", "$F_{ROH}$", "$\\text{cov}$"),
  Lens = c("homozygosity", "drift", "identity by descent", "hybrid", "-"),
  Definition = c("$1 - \\overline{H_t/H_0}$; **can be negative**",
                 "$\\overline{\\delta p^2 / [p_0(1-p_0)]}$; never negative",
                 "segments inherited from a defined base",
                 "genome fraction in runs of homozygosity",
                 "$2\\,\\text{cov}(\\delta p/s,\\,(p_0 - \\tfrac12)/s) = F_{hom} - F_{drift}$"),
  `Function` = c("`freq_metrics()`", "`freq_metrics()`", "requires pedigree + linkage",
                 "not implemented", "`freq_metrics()$cov_diag`"),
  check.names = FALSE))
```

### Selection and optimisation

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$u_i$", "$\\alpha$", "$\\alpha_{\\max}$", "$\\mathbf{c}$",
             "$\\mathbf{w}$", "$k_a$", "$\\lambda$", "$\\sigma^2_{DH}$", "UC"),
  Meaning = c("multi-trait index of hybrid i",
              "$(GD_{\\text{ref}} - GD_S)/GD_{\\text{ref}}$; negative means more diverse than chance",
              "the diversity budget -- a policy decision",
              "continuous contributions, summing to 1",
              "line-usage weights, $w_a = k_a/(2n)$",
              "times line a is used as a parent",
              "weight on the coancestry penalty in continuous OCS",
              "progeny variance of a DH family",
              "usefulness criterion, $\\mu + i h \\sigma$"),
  `Function` = c("`ctx$index`", "`alpha_loss()`", "argument to `stage2_select()`",
                 "`ocs_quadprog()`", "`line_weights()`, `line_usage_freq()`",
                 "`line_counts()`", "`ocs_relaxation()`", "`sigma2_dh()`",
                 "`usefulness_criterion()`"),
  check.names = FALSE))
```

### Genomic prediction

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$\\mathbf{u}$, $\\hat{\\mathbf{u}}$", "$\\boldsymbol\\beta$",
             "$\\sigma^2_u$, $\\sigma^2_e$", "$\\lambda$", "$h^2$",
             "$r$", "$r_i^2$", "$\\mathrm{PEV}_i$", "$\\pi$", "$\\nu$, $S$",
             "$M_e$"),
  Meaning = c("breeding values; the BLUP of them",
              "marker effects; $\\mathbf{u} = \\mathbf{Z}\\boldsymbol\\beta$",
              "genetic and residual variance",
              "$\\sigma^2_e/\\sigma^2_u$ -- **not** the OCS penalty above",
              "$\\sigma^2_u/(\\sigma^2_u + \\sigma^2_e)$",
              "accuracy: $\\mathrm{cor}(\\hat u, g)$, needs the truth",
              "reliability: $1 - \\mathrm{PEV}_i/(\\sigma^2_u \\mathbf{G}^*_{ii})$, does not",
              "prediction error variance, from the inverse of the MME",
              "prior probability a marker has no effect (BayesB)",
              "degrees of freedom and scale of the $\\chi^{-2}$ prior; `df`, `R2`",
              "effective number of independent chromosome segments"),
  Note = c("defined over **all** individuals, not just the phenotyped ones",
           "the same fit, two parametrisations -- exactly (@sec-genomic-prediction)",
           "supplied here, never estimated; REML in practice",
           "a variance ratio, not a tuning knob",
           "the one number the whole fit is conditioned on",
           "meaningless without its CV scheme (@sec-hybrid-cv)",
           "the early warning: it collapses before the accuracy does",
           "$\\sigma^2_e[\\mathbf{C}^{uu}]_{ii}$",
           "BayesC$\\pi$ estimates it; at book scale, badly",
           "hyperparameters, not fitted",
           "unlinked markers here, so $M_e$ is the marker count"),
  check.names = FALSE))
```

**CV0, CV1, CV2** name what the training set had already seen of a candidate
hybrid's *parents*: neither, one, or both. In a factorial between two pools these
are three different questions, and only CV2 is easy -- see @sec-hybrid-cv.

### Population sizing

```{r}
#| echo: false
knitr::kable(data.frame(
  Symbol = c("$H$", "$D$", "$N_{eq}$", "$L$ (Morgans)", "$A_t$", "$r$"),
  Meaning = c("panel heterozygosity; 0.125 in an F4",
              "diversity retained, $\\overline{2p(1-p)}/0.5$",
              "$(1-H)/(1-D)$ -- a communication device, **not** a Wright Ne",
              "total map length",
              "junctions per haplotype at generation t",
              "recombination fraction, Haldane"),
  `Function` = c("`panel_het()`", "`diversity_retained()`", "`n_equivalent()`",
                 "`map$total_cM / 100`", "`expected_junctions()`", "`build_map()`"),
  check.names = FALSE))
```

## Conventions

**Coancestry always includes the diagonal.** Group coancestry in the sense of
Cockerham. That is what makes $1 - \theta$ exactly Nei's $H_e$ and makes the
algebra close. Drop the diagonal and none of the identities hold.

**Everything is computed per locus and then averaged over loci.** Never average
frequencies first.

**Marker coding.** Lines are 0/1 (or 0/2) because they are homozygous. Hybrids
are 0/1/2 as dosage, rescaled to 0/0.5/1 as within-individual frequency.
`stage1_build()` auto-detects and rescales.

**`f` is never centered, `G` always is.** A negative coancestry is a symptom of
the wrong matrix.

**Loss is always relative to a same-size random baseline.** Never to the full
pool.

## Terms

**Balding-Nichols model.** A Beta distribution with mean $p$ and variance
$F_{ST}\,p(1-p)$, used to draw pool-specific allele frequencies from an
ancestral frequency. The parameter *is* the expected divergence.

**Group coancestry.** The probability that two alleles sampled at random from a
whole set are identical in state, including the case of sampling the same
individual twice.

**Heterotic group / pool.** A set of lines maintained as a breeding unit,
crossed to lines of a complementary group to produce hybrids. Divergence between
groups is where heterosis comes from -- but only the divergence that coincides
with dominance. Divergence produced by drift alone generates none, because it is
paid for exactly out of the within-pool term. See @sec-heterosis.

**Mid-parent heterosis.** The F1's value minus the mean of its two parents. With
inbred parents the additive contributions cancel exactly, leaving only the
dominance deviations at the F1's heterozygous loci.

**General and specific combining ability (GCA, SCA).** The two halves of a
hybrid's value that a two-way layout separates: what a line contributes to every
cross it enters, and what depends on which two lines met. For heterosis both are
exact functions of $d$ and the pool frequencies, with GCA weighted by the
*opposite* pool's frequencies -- so combining ability is a property of a line
against a tester pool, never of the line alone. See @sec-heterosis-gca-sca.

**Reciprocal recurrent selection.** Selecting on performance *in cross* and
recycling the winners within their own pool. Selection information crosses the
pool boundary; germplasm does not. See @sec-cycles.

**Identity by state (IBS) vs by descent (IBD).** IBS is what markers observe:
two alleles are the same allele. IBD adds the requirement of common ancestry
from a defined base. The gap between them is where @sec-genomic-era's whole
argument lives.

**Line-level collapse.** The exact identity $\theta_S = \mathbf{w}'
\mathbf{f}^L \mathbf{w}$ for F1 hybrids of homozygous lines, which removes the
need for the $N \times N$ hybrid matrix. See @sec-collapse.

**Status number.** $N_s = 1/(2\theta)$, the number of unrelated non-inbred
individuals with the same gene diversity. A static descriptor.

**Truncation selection.** Take the top $n$ by index, ignore everything else. The
ceiling on gain and the floor on diversity.

**Warm start.** Seeding an optimiser's initial population with heuristic
solutions. Not optional here: without it the differential evolution finishes
below a plain greedy.
