# The `hybdiv` package {#sec-package}

```{r}
#| include: false
source("_common.R")
```

Every function used in this book lives in one package, whose source is the `R/`
directory of this repository. This appendix is the index.

```{r}
#| echo: false
ns <- asNamespace("hybdiv")
exported <- sort(getNamespaceExports(ns))
c(exported_objects = length(exported))
```

## Installation

```{r}
#| eval: false
install.packages(c("DEoptim", "quadprog"))
devtools::install(".")        # from the repository root
library(hybdiv)
```

Only `DEoptim` is a hard dependency, used solely by `sel_de()`. `quadprog` is
suggested and needed only for `optimal_contributions()` and `ocs_quadprog()`.
Everything else, including every plot, is base R.

## File map

```{r}
#| echo: false
knitr::kable(data.frame(
  File = c("`R/00_data.R`", "`R/01_metrics.R`", "`R/02_benchmark.R`",
           "`R/03_de_select.R`", "`R/04_fast_metrics.R`", "`R/05_pipeline_metrics.R`",
           "`R/06_f2size.R`", "`R/07_caballero_toro.R`", "`R/08_mabc.R`",
           "`R/09_reference.R`",
           "`R/10_stage1.R`", "`R/11_stage2.R`", "`R/12_search.R`",
           "`R/13_dominance.R`", "`R/14_cycles.R`", "`R/15_response.R`"),
  Role = c("simulation, the two matrices, the `ctx` builder",
           "the diversity metrics, all `f(idx, ctx) -> scalar`",
           "null distribution, discriminatory power, cost, plots",
           "selection strategies: truncation, greedy, DE, OCS relaxation",
           "the line-level fast route -- no N x N matrix",
           "stage-specific metrics for S1 to S4",
           "F1 to F4 population sizing simulator",
           "Caballero & Toro subdivided-population analysis",
           "marker-assisted backcrossing: drag, the selection index, sizing",
           "slow literal oracles, for verification",
           "stage 1: lines to the X / f / hybrids contract",
           "stage 2: constrained selection across an alpha grid",
           "structure-aware search: local search, the bound, the exact oracle",
           "heterosis under a dominance model, and its link to the partition",
           "recurrent selection across cycles",
           "genic variance, and the Smith-Hazel and restricted indices"),
  Chapter = c("@sec-simulating-crosses", "@sec-catalogue", "@sec-catalogue",
              "@sec-combinatorial", "@sec-choosing", "@sec-stages",
              "@sec-sizing", "@sec-subdivided", "@sec-backcross",
              "@sec-verification",
              "@sec-simulating-crosses", "@sec-case-study", "@sec-advanced",
              "@sec-heterosis", "@sec-cycles", "@sec-response"),
  check.names = FALSE))
```

Files are numbered because the package has no `Collate` field and sourcing order
is alphabetical.

## The two-stage contract

Everything downstream of stage 1 talks only through three objects. This is the
seam for swapping in real data.

```{r}
#| eval: false
st1 <- stage1_simulate()                 # or stage1_build(X, ped, traits, fL)
#   $X        N x m hybrid markers, 0/0.5/1
#   $f        N x N molecular coancestry
#   $hybrids  N x (3 + n_traits): id, line_A, line_B, traits
#   $fL       L x L line coancestry (optional)

res <- stage2_select(st1$X, st1$f, st1$hybrids, n_sel = 100, fL = st1$fL)
#   $selection  0/1 column per scenario, plus n_scenarios
#   $metrics    one row per scenario
#   $z_scores   each metric in s.d. from the random null
#   $ref        gd_ref, its Monte Carlo s.e., alpha_max, n_sel
```

## Function index

### Simulation and context

```{r}
#| echo: false
idx <- function(...) {
  rows <- list(...)
  knitr::kable(do.call(rbind, lapply(rows, function(r)
    data.frame(Function = r[1], Does = r[2]))), row.names = FALSE)
}
idx(
  c("`sim_config`", "simulation scale: lines per pool, markers, traits, divergence"),
  c("`simulate_lines()`", "Balding-Nichols line genotypes in two divergent pools"),
  c("`simulate_data()`", "lines, all A x B crosses, both matrices, predicted traits"),
  c("`simulate_pools()`", "lighter simulator returning raw pieces, with `GL` in 0/1"),
  c("`build_ctx()`", "assemble the context; freezes `p0` and the index"),
  c("`build_ctx_from_stage1()`", "same, from the stage-1 contract"),
  c("`ctx_from_pools()`", "context from `simulate_pools()`, carrying `parents` and `GL`"),
  c("`load_data()`", "the single swap point for real data"),
  c("`stage1_simulate()` / `stage1_build()`", "produce the stage-1 contract"),
  c("`stage2_select()`", "the whole constrained selection pipeline")
)
```

### The two matrices

```{r}
#| echo: false
idx(
  c("`molecular_coancestry()`", "**the diversity kernel**; one `tcrossprod`, stays in [0, 1]"),
  c("`vanraden_G()`", "**prediction only**; `sum(G) == 0` by construction"),
  c("`mrd_matrix()` / `coancestry_from_mrd()`", "modified Rogers distance, the same statistic")
)
```

### Diversity metrics -- all `f(idx, ctx) -> scalar`

```{r}
#| echo: false
idx(
  c("`theta_group()`", "group coancestry, diagonal included"),
  c("`gene_diversity()` / `he_nei()`", "`1 - theta`; the same quantity, two routes"),
  c("`theta_from_freq()`", "theta via allele frequencies -- validation only, ~100x slower"),
  c("`status_number()`", "`Ns = 1/(2 theta)`; a static descriptor"),
  c("`ne_parents()`", "effective number of parent lines; independent of `f`"),
  c("`ne_lines_pool()`", "the same, per pool; needs only the pedigree"),
  c("`alleles_lost()`", "markers polymorphic in the pool, rare in the selection"),
  c("`freq_metrics()`", "**one marker sweep**: He, F_hom, F_drift, cov_diag, rare_retained"),
  c("`line_weights()` / `theta_pools()`", "per-pool usage weights and coancestries"),
  c("`gd_partition()`", "Caballero & Toro Eq. 8 over the heterotic pools"),
  c("`ene()` / `ane()`", "Core Hunter entry- and accession-to-nearest-entry"),
  c("`eff_dim()`", "effective dimensionality, on the **double-centered** submatrix"),
  c("`gst()`", "Nei's Gst, selected against non-selected"),
  c("`mean_offdiag()`", "the common production metric, kept for comparison"),
  c("`metrics_cheap()` / `metrics_full()`", "the inner-loop panel and the post-hoc panel")
)
```

### The fast line route

```{r}
#| echo: false
idx(
  c("`fast_ctx()`", "add `parents`, `GL` and the packed bitset to a context"),
  c("`line_counts()` / `line_usage_freq()`", "the sufficient statistic"),
  c("`fast_theta()` / `fast_gene_div()` / `fast_status_num()`", "`theta = w' fL w`; no N x N matrix"),
  c("`make_theta_fun()`", "precompiled closure for ~1e6 evaluations"),
  c("`theta_state()` / `theta_swap()`", "O(L) incremental update per swap"),
  c("`fast_ne_parents()` / `ne_parents_from_counts()`", "free once the counts exist"),
  c("`fast_pbar()` / `fast_he_nei()` / `fast_alleles_lost()`", "frequencies from the line matrix"),
  c("`pack_lines()` / `popcount32()` / `alleles_retained_fast()`", "bitset allelic richness"),
  c("`fast_theta_pools()`", "pool decomposition from the same counts"),
  c("`decode_rank()` / `decode_lines()`", "the two encodings; **use `decode_lines()`**"),
  c("`make_fitness_fast()`", "DE fitness with several simultaneous restrictions")
)
```

### Selection strategies

```{r}
#| echo: false
idx(
  c("`sel_random()`", "the null strategy"),
  c("`sel_truncation()`", "ceiling on gain, floor on diversity"),
  c("`sel_truncation_cap()`", "per-line cap; no matrix needed, usually most of the benefit"),
  c("`sel_greedy()`", "incremental greedy; `w = 0` is max diversity"),
  c("`decode()` / `make_fitness()` / `sel_de()`", "differential evolution, warm-started by default"),
  c("`project_capped_simplex()` / `ocs_relaxation()`", "the continuous upper bound"),
  c("`optimal_contributions()` / `ocs_quadprog()`", "quadratic-programming OCS")
)
```

### Benchmarking

```{r}
#| echo: false
idx(
  c("`null_distribution()`", "**the correct baseline**: random subsets of the same size"),
  c("`alpha_loss()`", "proportional diversity loss against that baseline"),
  c("`evaluate()` / `discriminatory_power()`", "metric panel and its z-scores"),
  c("`cost_per_eval()`", "microseconds per metric; decides what goes inside the loop"),
  c("`plot_null()` / `plot_correlation()` / `plot_frontier()`",
    "base-R diagnostics; run by `inst/scripts/run_benchmark.R`, not by the book")
)
```

## Runnable scripts

The book runs everything at a reduced scale so it renders quickly. Three scripts
ship with the package and run the same work at the production scale of
`sim_config`, writing their tables and plots to `report/`:

```sh
Rscript inst/scripts/run_all.R          # the two-stage pipeline, ~4 min
Rscript inst/scripts/run_benchmark.R    # the metrics benchmark, ~6 min
Rscript inst/scripts/run_mabc.R         # the MABC scans, ~3 min
```

`run_benchmark.R` is the source of the numeric results quoted in `README.md`,
and the only place the three `plot_*` helpers above are exercised. From an
installed copy of the package:

```{r}
#| eval: false
source(system.file("scripts", "run_all.R", package = "hybdiv"))
```

### Pipeline stages S1-S4

```{r}
#| echo: false
idx(
  c("`theta_decompose()` / `gst_nei()`", "**both** F_ST definitions, and Nei's Gst"),
  c("`pool_freqs()`", "allele frequencies per pool, one row each"),
  c("`pool_complementarity()` / `s1_monitor()`", "the S1 cycle panel; `mean_sq_delta_p` is the exact one"),
  c("`dh_sib_coancestry()` / `dh_cross_coancestry()` / `dh_pool_theta()`", "the three DH identities"),
  c("`sigma2_dh()` / `sigma_dh_crosses()`", "progeny variance from marker effects"),
  c("`usefulness_criterion()` / `sel_intensity()`", "UC -- applies at S2, not S4"),
  c("`s3_theta()` / `alleles_retained()`", "line-level metrics"),
  c("`loss_vs_resampling()` / `theta_ceiling()`", "calibrate against a same-size baseline"),
  c("`select_lines_constrained()`", "truncate-then-repair; the incremental greedy fails"),
  c("`hybrid_theta_from_usage()` / `line_usage()` / `pool_balance()`", "S4 on the line route"),
  c("`loss_ledger()`", "loss per stage, on one scale")
)
```

### Heterosis and cycles

```{r}
#| echo: false
idx(
  c("`heterosis_value()`", "mid-parent heterosis of each hybrid, from `d` alone"),
  c("`heterosis_expected()`", "the same, expected over a pool pair"),
  c("`heterosis_partition()`", "shared and divergence terms -- `GD_T` and `GD_BS`"),
  c("`run_cycles()`", "recurrent selection, one row per cycle"),
  c("`next_pool()`", "one generation of a pool: weighted intermating, DH derivation"),
  c("`sel_greedy_budget()`", "greedy sweep returning the best plan inside a budget")
)
```

### Caballero & Toro

```{r}
#| echo: false
idx(
  c("`ct_partition()`", "subpopulation coancestries, F statistics, the Eq. 7 partition"),
  c("`ct_loss_gain()`", "loss **or gain** from removing each subpopulation")
)
```

### Population sizing

```{r}
#| echo: false
idx(
  c("`maize_chr_len` / `f2_distorters` / `f2_opt`", "defaults; **replace the distorters** with your data"),
  c("`build_map()` / `marker_index()`", "Haldane map on a regular grid"),
  c("`meiosis()` / `meiosis_sel()` / `cross_indices()` / `make_F1()`", "the vectorised engine"),
  c("`make_distortion()` / `zyg_fitness()`", "gametophytic and zygotic distortion, by exact rejection"),
  c("`run_scheme()` / `reps_for_N()` / `bulk_weights()`", "SSD, bulk, Syn schemes"),
  c("`panel_freq()` / `panel_het()` / `diversity_retained()`", "panel metrics"),
  c("`min_minor_freq()` / `fixed_fraction()` / `skew_fraction()`", "worst-region metrics"),
  c("`n_equivalent()`", "equivalent N -- a communication device, **not** a Wright Ne"),
  c("`block_lengths()` / `junctions_per_hap()` / `expected_junctions()`", "map expansion"),
  c("`multilocus_recovery()` / `f1_sizing()`", "the objectives that keep paying for larger N"),
  c("`f2_run_experiment()`", "the full grid; minutes")
)
```

### Reference oracles

```{r}
#| echo: false
idx(
  c("`ref_theta()`, `ref_gene_div()`, `ref_theta_freq()`, `ref_he_nei()`", "literal transcriptions"),
  c("`ref_status_num()`, `ref_ne_parents()`, `ref_alleles_lost()`, `ref_alleles_fixed()`", "-"),
  c("`ref_ene()`, `ref_ane()`, `ref_eff_dim()`, `ref_theta_pools()`", "-")
)
```

## Full export list

```{r}
#| echo: false
knitr::kable(matrix(c(exported, rep("", (4 - length(exported) %% 4) %% 4)),
                    ncol = 4, byrow = FALSE),
             col.names = NULL)
```
