# Verification {#sec-verification}

```{r}
#| include: false
source("_common.R")
ic <- read.csv(book_data("theta_identity_check.csv"))
```

A derivation the reader cannot re-run against a naive implementation is a claim,
not a verification. This appendix says how each result in the book was checked,
and how to re-run the checks.

## The method: slow oracles

Every fast route in the package has a **deliberately slow, literal**
counterpart in `R/09_reference.R` that transcribes the definition with no
algebraic shortcut. The fast route must match it to machine precision.

```{r}
ref_theta
theta_group
fast_theta
```

Three implementations of one quantity: a literal mean over the submatrix, the
packaged version, and the line-route collapse that never forms the submatrix at
all. They must agree.

```{r}
sim  <- simulate_pools(n_A = 20, n_B = 20, m = 1500, seed = 7)
ctxp <- ctx_from_pools(sim)
set.seed(3); sel <- sample.int(ctxp$N, 50)

c(oracle     = ref_theta(sel, ctxp),
  packaged   = theta_group(sel, ctxp),
  line_route = fast_theta(sel, ctxp),
  max_dev    = max(abs(c(theta_group(sel, ctxp), fast_theta(sel, ctxp)) -
                       ref_theta(sel, ctxp))))
```

The oracles exist so that this comparison is possible for a reader who does not
trust the derivation -- which is the correct posture.

```{r}
fmt(rbind(
  he_nei       = c(oracle = ref_he_nei(sel, ctxp), packaged = he_nei(sel, ctxp)),
  theta_freq   = c(ref_theta_freq(sel, ctxp), theta_from_freq(sel, ctxp)),
  status_num   = c(ref_status_num(sel, ctxp), status_number(sel, ctxp)),
  ne_parents   = c(ref_ne_parents(sel, ctxp), ne_parents(sel, ctxp)),
  alleles_lost = c(ref_alleles_lost(sel, ctxp), alleles_lost(sel, ctxp)),
  ENE          = c(ref_ene(sel, ctxp), ene(sel, ctxp)),
  ANE          = c(ref_ane(sel, ctxp), ane(sel, ctxp))), 10)
```

## The identity sweep

The line-level collapse was checked across $F_{ST} \in \{0.05, 0.15, 0.35\}$,
two seeds and two selection sizes, thirty replicate selections each.

```{r}
#| echo: false
cols <- c(max_pairwise_f = "pairwise f: hybrid vs line formula",
          max_theta_sub_vs_line = "theta: line route vs f submatrix",
          max_theta_sub_vs_freq = "theta: line route vs allele-frequency route",
          max_pbar = "selection allele frequencies",
          max_He = "Nei He",
          max_Neparents = "effective number of parents",
          max_alleles_lost = "allele loss at MAF threshold",
          max_theta_pools = "pool decomposition",
          max_swap_update = "incremental swap vs full recompute",
          bitset_vs_literal = "bitset allelic richness vs literal count")
worst <- vapply(names(cols), function(k) max(abs(ic[[k]])), numeric(1))
knitr::kable(data.frame(
  Identity = unname(cols),
  `Worst deviation over the sweep` =
    ifelse(worst == 0, "0 (exact)", sprintf("%.1e", worst)),
  check.names = FALSE),
  caption = "Machine precision throughout. Anything above 1e-12 would mean an error, not a rounding artefact.")
```

## The test suite

The suite is `tests/testthat/`, run with `devtools::test()`. It takes about a
minute. What it asserts, and why each check earns its place:

```{r}
#| echo: false
knitr::kable(data.frame(
  Check = c("theta via submatrix == theta via allele frequencies",
            "1 - theta == Nei's He",
            "f is bounded to [0, 1]",
            "sum(VanRaden G) == 0",
            "theta and Ns analytical edge cases",
            "greedy < random < truncation on theta",
            "decode() is unique, in range, deterministic",
            "fast line route == hybrid-matrix route",
            "incremental swap == full recomputation",
            "popcount32 == naive bit count",
            "F_hom and F_drift are 0 when the selection is the base",
            "cov_diag == F_hom - F_drift",
            "max-diversity drives F_hom negative, F_drift positive",
            "freq_metrics reproduces he_nei and alleles_lost",
            "the Caballero & Toro partition closes",
            "GD_WI == 0 for inbred lines, GD_WI_hyb large for their F1s",
            "eff_dim in (0, n - 1]",
            "the two-stage output contract",
            "stage 2 runs without fL",
            "stage1_build accepts both marker codings",
            "MRD and coancestry are one statistic",
            "Gst < Fst always, and Gst matches the classical route",
            "the DH identities",
            "hybrid theta from usage == the hybrid matrix",
            "constrained line selection meets its ceiling",
            "the oracles agree with the packaged metrics",
            "map length and Haldane spacing",
            "heterozygosity halves per selfing",
            "D == 1 - (1 - H)/N under neutrality",
            "n_equivalent inverts diversity_retained",
            "a Syn cycle adds half a map length of junctions",
            "distortion shifts ancestry away from 0.5",
            "f1_sizing halves the loss probability per plant"),
  `If it fails` = c(
    "**f is wrong**: centered, rescaled, or the wrong marker coding",
    "the identity underpinning every 'loss of X%' statement is broken",
    "a centered matrix is being used as coancestry",
    "the demonstration in @sec-why-not-g no longer holds on this data",
    "the diagonal is being dropped somewhere",
    "there is no trade-off to optimise; the whole benchmark is pointless",
    "the DE fitness is noisy and cannot converge",
    "**the collapse is broken**; production scale becomes infeasible",
    "any local search built on the state vector is wrong",
    "allelic richness counts are wrong",
    "the base p0 is not frozen, or is being recomputed",
    "one of the three lens formulas is wrong",
    "the two-lens argument of @sec-genomic-era does not reproduce",
    "the consolidated marker sweep changed an existing number",
    "the Eq. 8 algebra is broken",
    "lines are not actually inbred, or the hybrid diagonal is wrong",
    "the double-centering was dropped and eff_dim re-measures theta",
    "downstream code reading `$selection` or `$metrics` will break",
    "the optional-fL path is broken",
    "real data in the other coding would give different answers",
    "Trap 1 of @sec-stages no longer holds",
    "Trap 2 is wrong, or the two definitions were conflated",
    "the S2 algebra is wrong and crossing plans cannot be pre-evaluated",
    "Trap 5's repair route is broken",
    "the constrained selection silently returns an infeasible plan",
    "a fast route diverged from its definition",
    "the sizing chapter's map is not the map it claims",
    "the simulator is not doing what selfing does",
    "the identity every sizing number rests on is broken",
    "N_eq is not the inverse it is described as",
    "the intercrossing budget in @sec-sizing is wrong by 2x",
    "the distortion mechanism is inactive",
    "the F1 sizing argument is wrong"),
  check.names = FALSE))
```

```{r}
#| eval: false
Rscript -e 'devtools::test()'
```

## Provenance of every figure and table

The book runs light examples live and reads heavy results from cache. This is
the complete accounting.

### Computed live at render

```{r}
#| echo: false
knitr::kable(data.frame(
  Chapter = c("@sec-coancestry", "@sec-subdivided", "@sec-simulating-crosses",
              "@sec-sizing", "@sec-genomic-era", "@sec-catalogue",
              "@sec-choosing", "@sec-stages", "@sec-ocs", "@sec-combinatorial",
              "@sec-case-study", "@sec-verification"),
  `Runs live` = c(
    "both matrices, the sum(G) demonstration, the f diagonal",
    "the whole worked example, reproduced against the published tables",
    "the simulation, the fst sweep, the contract, the sampling-bias histogram",
    "the theoretical identities, heterozygosity halving, map expansion, f1_sizing",
    "the Eq. 3 identity, the four-strategy lens panel",
    "every metric, the sampling-bias formula, the null distribution",
    "the collapse, the swap update, cost at book scale, the fixed-theta band",
    "all five traps, the S1 panel, the DH identities, the S4 collapse",
    "the OCS relaxation, the convergence check, the greedy frontier",
    "all baselines, the encodings, the DE, the warm-start comparison",
    "the full two-stage pipeline",
    "the oracle comparisons"),
  check.names = FALSE))
```

### Read from cache

```{r}
#| echo: false
knitr::kable(data.frame(
  File = c("`data/theta_identity_check.csv`", "`data/cost_per_eval.csv`",
           "`data/metric_redundancy.csv`", "`data/de_constraints.csv`",
           "`data/fst_definitions.csv`", "`data/dist_vs_variance.csv`",
           "`data/pipeline_loss_ledger.csv`", "`data/plant_metric_usage.csv`",
           "`data/bibliography_*.csv`", "`data/f2_results.csv`",
           "`data/f2_multilocus.csv`", "`data/mabc_*.csv`",
           "`figs/mabc_sizing.png`", "`figs/mabc_policies.png`",
           "`figs/cost_scaling.png`", "`figs/de_convergence.png`",
           "`figs/metric_redundancy.png`", "`figs/pipeline_loss.png`",
           "`figs/metric_pitfalls.png`"),
  Why = c("the full identity sweep, 12 configurations",
          "production scale: L = 300, N = 22500, m = 10000",
          "621 selections x 17 metrics",
          "9 DE runs at 60,000 evaluations each",
          "the divergence sweep",
          "600 candidate crosses with simulated DH",
          "8 founder pools, four stages, two strategies",
          "687 screened plant records",
          "the screened corpora themselves",
          "5 schemes x 20 sizes, adaptive replication",
          "multilocus recovery across the grid",
          "the MABC grid: 4 policies x 4 gene numbers x 11 sizes x 3 s, 300 reps",
          "analytical sizing by stage",
          "the four recombinant-selection policies",
          "cost scaling at production scale",
          "DE convergence under four restrictions",
          "the redundancy panels",
          "the loss ledger",
          "Traps 4 and 2 on the full simulation"),
  check.names = FALSE))
```

Nothing is read from `report/` or `data/` at the repository root: both are
gitignored, so a clean clone would fail to render. Everything cached lives in
`book/data` and `book/figs`, which are tracked.

## Regenerating the cached results

```{r}
#| eval: false
Rscript book/scripts/regenerate.R          # everything; hours
Rscript book/scripts/regenerate.R f2size   # one target
```

The script is the single place where the expensive computations live. It is
never invoked by `quarto render`.

## Reproducing the book

```{sh}
#| eval: false
Rscript -e 'devtools::document(); devtools::load_all()'
Rscript -e 'devtools::test()'
Rscript -e 'devtools::check()'
quarto render book
```

Quarto's `freeze: auto` means a chapter re-runs only when its source changes.
To force a full recomputation, delete `book/_freeze/`.

## Known limitations of the verification

Stated so they are not mistaken for oversights.

**The oracles share the author's understanding.** They check that the fast route
computes what the slow route computes. They cannot check that the slow route
implements the right definition. That is what the reproduction of the published
Caballero & Toro tables in @sec-subdivided is for -- an external check against
numbers this project did not produce.

**Monte Carlo checks have Monte Carlo error.** Where a test asserts agreement
with a simulated expectation rather than an algebraic identity, the tolerance is
loose by necessity, and a genuinely small bias would pass.

**Everything is verified on simulated data.** The identities are algebraic and
will hold on real genotypes. The *empirical* results -- which metrics are
redundant, how much index a constraint costs -- were derived at $F_{ST} = 0.15$
with simulated markers, and @sec-choosing says explicitly that they should be
re-derived on real genotypes before being treated as settled.
