# Preface {.unnumbered}

```{r}
#| include: false
source("_common.R")
```

A hybrid breeding programme has two objectives that pull against each other.
The first is short-term: pick the crosses with the highest predicted
performance. The second is long-term: do not spend the germplasm that will be
needed to make the next round of gain possible. Every programme knows this.
Very few can say, in a number, how much of the second they traded away to buy
the first.

This book is about producing that number, defending it, and putting it inside
an optimiser.

## The problem, stated once

Given a pool of $N$ candidate F1 hybrids -- each a cross between one line from
heterotic group A and one from group B, each with a predicted multi-trait index
-- choose $n_{sel}$ of them so that the mean index is as high as possible,
subject to losing no more than a fraction $\alpha$ of the molecular gene
diversity that a *random* set of the same size would have carried.

Three of those words carry the entire argument.

**Molecular.** Diversity is measured on the Caballero & Toro molecular
coancestry matrix $f$, never on a VanRaden genomic relationship matrix $G$.
This is not a stylistic preference. $G$ is centered on its own population by
construction, so the sum of all its elements is exactly zero; every diversity
quantity computed from it is meaningless. @sec-coancestry shows this happening.

**Random.** The reference for "no loss" is the distribution of random subsets of
size $n_{sel}$, not the full candidate pool. Comparing a selected set of 100
against a pool of 2500 confounds a selection effect with a sampling effect, and
the sampling effect alone can look like a loss of half a percent.
@sec-baseline separates them.

**Subject to.** $\alpha$ is a *constraint*, not a term in a weighted sum. When
you write $\alpha_{max} = 0.05$ you are saying "I accept losing five percent",
which is a statement a breeding committee can approve. A weighted sum says
"diversity is worth $\lambda$ index units", which is a statement nobody can
approve because nobody knows what $\lambda$ means. @sec-ocs makes the
distinction precise, and shows the standard optimal-contribution formulation it
specialises.

## What is unusual here

Most treatments of diversity management come from animal breeding, where the
population is randomly mating and the unit of selection is an individual. Plant
hybrid breeding breaks both assumptions:

- The parents are **inbred lines**, so their within-individual diversity is zero
  by construction, and the interesting heterozygosity lives in the F1 product,
  not in the parents.
- The population is **permanently subdivided** into heterotic groups, and the
  divergence between them is not a problem to be minimised -- it is the engine
  of heterosis, to be *maintained*. A single pool-wide statistic cannot express
  "conserve within, hold between". @sec-heterosis derives that claim as an exact
  identity, and then bounds it: divergence pays only where it coincides with
  dominance, and drifting the pools apart buys nothing at all.
- The candidates are **crosses of fixed lines**, so the progeny variance term
  that drives cross-selection criteria elsewhere collapses to zero, and the
  hybrid-by-hybrid coancestry matrix turns out never to be needed at all
  (@sec-collapse).

Each of these changes what should be measured. None of them changes the
underlying theory, which is why the book starts with the theory.

## How to read this

The book is meant to be run, not only read. Every derivation that could be
checked numerically is checked numerically, in a code block you can execute.
Where a result is an exact algebraic identity, the deviation from a simulated
oracle is printed next to it; where it is a Monte Carlo agreement, the residual
is printed.

The argument the parts assemble is a chain, and it is worth stating before the
table that lists them:

1. Diversity between the pools is what makes the hybrid worth selling, and
   diversity within them is what the next cycle has to select on
   (@sec-subdivided, @sec-heterosis).
2. Those two are in tension: every plan that raises one lowers the other, and the
   partition of @sec-catalogue is what makes the trade visible.
3. Within-pool diversity is worth protecting because it is the raw material of
   *response*: genic variance is the same sum over the same loci as gene
   diversity, weighted by squared effect instead of equally (@sec-response). That
   is what makes a diversity constraint a variance constraint in disguise, and
   what makes "restrict rather than price" a sixty-year-old theorem rather than a
   preference.
4. So the loss of within-pool diversity is a **budget**, set against what random
   selection of the same size would have cost (@sec-baseline), not a term traded
   off inside an objective (@sec-ocs).
5. How much diversity there is to spend depends on how the candidates were
   generated: a panel cannot be selected back up to variation it never sampled
   (@sec-sizing), and a pool of near-copies cannot be rescued by any optimiser
   (@sec-backcross). @sec-simulating-crosses and @sec-alphasimr build that panel
   two different ways, and everything downstream is blind to which.
6. Spending the budget optimally is a combinatorial problem with structure worth
   naming (@sec-combinatorial, @sec-advanced).
7. The budget is worth paying because of what it leaves for later cycles, which
   is a claim about trajectories and is checked as one (@sec-cycles) -- and then
   predicted, which turns out to work only where the constraint holds
   (@sec-limits).

The parts are ordered so that each depends only on what came before:

| Part | What it establishes |
|---|---|
| Foundations | The coancestry kernel, the partition of diversity in a subdivided population, and the equation that says what selection buys |
| Simulation | How the candidate pool is generated, and how large a segregating population has to be |
| Metrics | What can be measured, what it costs, and which measurements are redundant |
| Optimisation | How the constraint enters an optimiser, and which optimiser to use |
| Evidence and practice | What the literature does and does not establish, and a worked end-to-end run |

Readers who only want the operational answer can read @sec-choosing (which
metrics), @sec-stages (at which stage) and @sec-case-study (the whole pipeline
in one script), then come back for the derivations.

## Running the code

All code is in the `hybdiv` package that ships with this book.

```{r}
#| eval: false
# from the repository root
install.packages(c("DEoptim", "quadprog"))
devtools::install(".")

library(hybdiv)
st1 <- stage1_simulate()                                   # lines -> X, f, hybrids
res <- stage2_select(st1$X, st1$f, st1$hybrids, n_sel = 100, fL = st1$fL)
res$metrics
```

Every live example in the book runs at a deliberately small scale -- 25 lines
per pool and 2000 markers, giving 625 candidate hybrids -- so that a full render
takes minutes. The scale is set once, in `_common.R`:

```{r}
str(book_cfg)
```

Results that are expensive at any scale -- the differential-evolution sweeps,
the cost benchmark, the population-sizing grid -- are precomputed and shipped in
`book/data` and `book/figs`. `book/scripts/regenerate.R` regenerates all of
them, and @sec-verification says exactly which figures and tables come from
where.

A note on numbers: everything reported here is reproducible output from a
simulated dataset, not a fixed constant. Change the simulation configuration or
a random seed and the numbers move. The *relationships* between them -- which
metric dominates which, which identity holds exactly -- do not.
